### 2024-03-22 main.tex
Added the start of a paper.

\documentclass{article}

\usepackage{siunitx}

\title{Robotic Bicycle Balance Assist Reduces Probability of Falling When
Subjected to Mechanical Perturbations}

\begin{document}

\maketitle

\abstract{
  Bicycles are unstable at low speeds. We add a controlled steering motor to a
  consumer bicycle that stablizes the bicycle at low speeds and apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill. The probability of not recovering from a handlebar perturbation
  decreases when the balance assist is activated.
}

\section{Introduction}
%
Single actor bicycle crashes are associated with a large number of unreported
and reported injuries and even sometimes death. At low speeds (less than 15
km/h or so) bicycles are no longer self stable and become more difficult to
control for the rider. Low speed crashes may be reduced if the bicycle was self
stable. Since at least the 1980s it known that applying a steering torque
proportional to the roll rate can stabilize a single track vehicle at very low
speeds. If the robotic control of steering can stablize a bicycle, it may
reduce the control required by the rider to successfully manage balancing
tasks. We have developed such a balance assisting bicycle and desire to
evaluate whether it helps the rider in situations they are likley to fall.

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the head tube cable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded amperage to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the roll
rate of the bicycle with a MEMs rate gyroscope. The balance assist control
algorithm is implemented on microprocessor (Teensy).

\subsection{Balance Assist Control}
We use a forward speed gain scheduled proportional roll rate positive feedback
controller to stabilize the bicycle. Scaling the proprtional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar using an adapted
Bump'Em system~\cite{todo} making use of four motors working in tandem. The
four motors are programmed to apply a light force at all times to keep the
ropes taught and to track a commanded force profile using a PID controller
running on a microprocessor (Arduino XXX, Italy). We measure the force applied
by each motor at the handlebar via four inline load cells with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a 1
meter wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van Velde's MSc thesis~\cite{vanvelde2022}. Our modifications to that described
in the thesis include simplifying the controller with a inexpensive
microcontoller and the use of a simple safety harness.

\subsection{Protocol}
%
Subjects wore a helmet and fall safety harness attached to the ceiling. We
allowed them practive riding on the treadmill until they said they were
comfortable enough to apply perturbations. For most this was less than a
10~\si{\minute} warm up. We then ask the rider to ride for 90~\si{\second}
attempting to maintain the location of their front wheel on the center line of
the treadmill. We define a "fall" on the treadmill by two criteria: the rider
removes their foot from the pedal and places it on the ground and the bicycle
wheel exceeds the width of the treadmill. We then applied perturbations in
random directions (clockwise or counter clockwise) increasing the magnitude by
X~\si{\newton\meter} each time. We log the magnitude that causes the first fall
to characterize that subject's nominal fall threshold. Following this we apply
20 perturbations of random magnitudes to the cyclist while they ride at a
constant speed and record which perturbations cause a fall. We let the cyclist
rest and then perform another 20 perturbations. We randomize whether the
balance assist system is on or off during the first or second set of
perturbations.

\subsection{Subjects}
%
We recruited X healthy young adults to participate in the experiments. X
subjects performed the experiments at 6~\si{\kilo\meter\per\hour} and X
subjects performed the experiments at 10~\si{\kilo\meter\per\hour}. The
10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
6~\si{\kilo\meter\per\hour}.

\subsection{Measurements}

\section{Results}

\section{Discussion}

\section{Conclusion}

\end{document}

### 2024-03-22 main.tex
Added more in measurements and control algorithm sections, added references.

@@
 
-\usepackage{siunitx}
+\usepackage{amsmath}
+\usepackage[margin=25mm]{geometry}
+\usepackage{siunitx}  % use for all units
+\usepackage{todonotes}
 
-\title{Robotic Bicycle Balance Assist Reduces Probability of Falling When
+\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
 Subjected to Mechanical Perturbations}
+
+\author{Marten T. Haitjema \and Leila Alizadehsaravi \and Jason K. Moore}
 
@@
 \subsection{Balance Assist Control}
-We use a forward speed gain scheduled proportional roll rate positive feedback
-controller to stabilize the bicycle. Scaling the proprtional feedback gain
+We use a forward speed \(v\) gain scheduled proportional roll rate positive
+feedback controller to stabilize the bicycle. The commanded steer torque
+\(T_\delta\) from the steer motor follows
+\begin{align}
+  T_\delta = -k(4.7\si{\meter\per\second} - v)\dot{\phi}
+\end{align}
+where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
+open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
+and \(k=10\) as a high gain value. Scaling the proprtional feedback gain
 linearly with respect to speed stabilizes the normally unstable weave mode of
@@
 X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.
-%
+
+\todo[inline]{Show open loop bicycle rigid rider eigenvalue vs speed and then
+the closed loop versions with the 8 and 10 gain settings. Three plots.}
+
 \begin{figure}
@@
 We apply longitudinal forces to the ends of each handlebar using an adapted
-Bump'Em system~\cite{todo} making use of four motors working in tandem. The
+Bump'Em system~\cite{Tan2020} making use of four motors working in tandem. The
 four motors are programmed to apply a light force at all times to keep the
@@
 18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
-Van Velde's MSc thesis~\cite{vanvelde2022}. Our modifications to that described
-in the thesis include simplifying the controller with a inexpensive
+Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our modifications to that
+described in the thesis include simplifying the controller with a inexpensive
 microcontoller and the use of a simple safety harness.
@@
 %
-We recruited X healthy young adults to participate in the experiments. X
-subjects performed the experiments at 6~\si{\kilo\meter\per\hour} and X
-subjects performed the experiments at 10~\si{\kilo\meter\per\hour}. The
-10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
+We recruited 26 able-bodied young adults (20-36 years old) to participate in
+the experiments. The subjects all were confident in their cycling skills and
+had cycled at least once in the last month. Eleven subjects performed the
+experiments at 6~\si{\kilo\meter\per\hour} and fifteen subjects performed the
+experiments at 10~\si{\kilo\meter\per\hour}. All subjects consented to the
+experiment and could decline to continue at any time.  The study was approved
+by Delft University of Technology's Human Research Ethics Council (#XXX).
+
+\subsection{Two Experiments}
+%
+The 10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
 6~\si{\kilo\meter\per\hour}.
@@
 \subsection{Measurements}
+%
+We measure the time histories of the Bump'Em delivered perturbation forces and
+the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
+%
+\begin{table}
+  \caption{Raw measurements}
+  \begin{tabular}{llll}
+    Measure & Variable & Units & Sensor \\
+    Perturbation Order Number & \(j\) & integer & NA\\
+    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
+    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
+    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
+    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
+    Balance Assist Gain & \(k\) & TODO & NA \\
+    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
+  \end{tabular}
+\end{table}
+
+Based on the findings in \cite{Haitjema2023}, we calculate several variables.
+We use the angular impulse \(L\) due to the perturbation forces over a
+0.3~\si{\second} duration to characterize the magnitude of delivered
+pertrubation. The duration is selected based on the duration of the commanded
+force.
+
+\begin{align}
+  F_r = & F_{rf} - F_{rr} \\
+  F_l = & F_{lf} - F_{lr} \\
+  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
+\end{align}
+
+We characterize the skill of the rider by calculating the standard deviation of
+the steer angle \(\sigma_\delta\) over the 90~\si{\second} period they were
+asked to maintain their position at the center of the treadmill. At the
+intiation of each perturbation we log the steer and roll angles to characterize
+the configuration of the bicycle when perturbed. The gain setting on the
+balance assist controller indicates if the assistance is off \(k=0\) or on at
+two different levels: low \(k=8\) or high \(k\=10\). A recovery from the
+perturbation is successful if the person neither put their foot down onto the
+treadmill surface or either wheel of the bicycle exits the width of the
+treadmill belt. We record this as a binary variable \(f\).
+
+\begin{table}
+  \begin{tabular}{lll}
+    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
+    perturbation torque \\
+    \(j\) & integer & order number of perturbation \\
+    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
+    \(k\) & TODO & gain of balance assist control \\
+    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
+    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
+    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
+    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
+  \end{tabular}
+\end{table}
+
+\subsection{Statistics}
 
@@
 
+\bibliographystyle{plain}
+\bibliography{references}
+
 \end{document}

### 2024-03-22 main.tex
Reordered subjects section and started on statistics.

@@
   Bicycles are unstable at low speeds. We add a controlled steering motor to a
-  consumer bicycle that stablizes the bicycle at low speeds and apply varying
+  consumer bicycle that stabilizes the bicycle at low speeds and apply varying
   magnitude external handlebar perturbations to the bicycle while ridden on a
@@
 proportional to the roll rate can stabilize a single track vehicle at very low
-speeds. If the robotic control of steering can stablize a bicycle, it may
+speeds. If the robotic control of steering can stabilize a bicycle, it may
 reduce the control required by the rider to successfully manage balancing
 tasks. We have developed such a balance assisting bicycle and desire to
-evaluate whether it helps the rider in situations they are likley to fall.
+evaluate whether it helps the rider in situations they are likely to fall.
 
@@
 \begin{align}
-  T_\delta = -k(4.7\si{\meter\per\second} - v)\dot{\phi}
+  T_\delta = k(4.7\si{\meter\per\second} - v)\dot{\phi}
 \end{align}
@@
 open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
-and \(k=10\) as a high gain value. Scaling the proprtional feedback gain
+and \(k=10\) as a high gain value. Scaling the proportional feedback gain
 linearly with respect to speed stabilizes the normally unstable weave mode of
@@
 described in the thesis include simplifying the controller with a inexpensive
-microcontoller and the use of a simple safety harness.
+microcontroller and the use of a simple safety harness.
+
+\subsection{Subjects}
+%
+We recruited 26 able-bodied young adults (20-36 years old) to participate in
+the experiments. The subjects all were confident in their cycling skills and
+had cycled at least once in the last month. Eleven subjects performed the
+experiments at 1.7~\si{\meter\per\second} (6~\si{\kilo\meter\per\hour}) and
+fifteen subjects performed the experiments at 2.8~\si{\meter\per\second}
+(10~\si{\kilo\meter\per\hour}). All subjects consented to the experiment and
+could decline to continue at any time.  The study was approved by Delft
+University of Technology's Human Research Ethics Council (\#XXX).
+
+\subsection{Two Experiments}
+%
+The 10~\si{\kilo\meter\per\hour} experiments occurred X weeks before the
+6~\si{\kilo\meter\per\hour}.
 
@@
 Subjects wore a helmet and fall safety harness attached to the ceiling. We
-allowed them practive riding on the treadmill until they said they were
+allowed them practice riding on the treadmill until they indicated they were
 comfortable enough to apply perturbations. For most this was less than a
@@
 
-\subsection{Subjects}
-%
-We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments. The subjects all were confident in their cycling skills and
-had cycled at least once in the last month. Eleven subjects performed the
-experiments at 6~\si{\kilo\meter\per\hour} and fifteen subjects performed the
-experiments at 10~\si{\kilo\meter\per\hour}. All subjects consented to the
-experiment and could decline to continue at any time.  The study was approved
-by Delft University of Technology's Human Research Ethics Council (#XXX).
+Following the initial threshold determination, we choose perturbation forces
+according to a random and adaptive staircase procedure applying
+perturbations above and below the initial perturbation threshold, while
+allowing small progression of the perturbation threshold to accommodate
+learning effects. The goal of this adaptive staircase procedure is to have
+participants fall for approximately 50\% of the time. The procedure
+consists of twenty perturbations per condition. Participants undergo two
+conditions: twenty perturbations with the balance-assist turned on and twenty
+perturbations with the balance-assist turned off (in randomized order).
 
-\subsection{Two Experiments}
-%
-The 10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
-6~\si{\kilo\meter\per\hour}.
+Five possible perturbation forces are determined based on the initial
+perturbation threshold estimate: the initial estimate itself, two perturbations
+lower than the initial estimate and two perturbations higher than the initial
+estimate.  The five possible perturbations are separated by 10 N steps. For
+example, if the initial estimate of the perturbation threshold of a participant
+is 80 N, the five possible perturbations are 60, 70, 80, 90 and 100 N. Which
+one of these five perturbations is chosen is determined at random. If the
+perturbation results in a fall, the estimate of the perturbation threshold is
+decreased with 10 N, and vice versa if the perturbation did not result in a
+fall. Five new possible perturbation forces are determined around the updated
+perturbation threshold, and a new perturbation is selected at random. This
+process iterates until twenty perturbations are applied.
 
@@
 0.3~\si{\second} duration to characterize the magnitude of delivered
-pertrubation. The duration is selected based on the duration of the commanded
+perturbation. The duration is selected based on the duration of the commanded
 force.
@@
 asked to maintain their position at the center of the treadmill. At the
-intiation of each perturbation we log the steer and roll angles to characterize
+initiation of each perturbation we log the steer and roll angles to characterize
 the configuration of the bicycle when perturbed. The gain setting on the
 balance assist controller indicates if the assistance is off \(k=0\) or on at
-two different levels: low \(k=8\) or high \(k\=10\). A recovery from the
+two different levels: low \(k=8\) or high \(k=10\). A recovery from the
 perturbation is successful if the person neither put their foot down onto the
@@
 treadmill belt. We record this as a binary variable \(f\).
+
+\subsection{Statistics}
+%
+Our primary hypothesis is that the balance assistance controller will reduce
+the probability of falling when perturbed externally at the handlebar. We have
+a single binary outcome variable \(f\) that is dependent on several possible
+explanatory independent variables including the binary balance assistance on or
+off.
 
@@
     perturbation torque \\
-    \(j\) & integer & order number of perturbation \\
+    \(c\) & integer & order number of perturbation \\
     \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
@@
 
-\subsection{Statistics}
+We evaluate this hypothesis using a multivariate logistic regression model.
+
+\begin{align}
+  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij})
+\end{align}
+
+\(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
+which follows a Bernoulli distriction given the probability \(p_{ij}\) that a
+fall occurs. The log-odds of the probabiliy is then a linear function of our
+independent variables.
+
+\begin{align}
+  \log\left(\frac{p_{ij}}{1 - p_{ij}}\right) = \Beta + \sum^K_{k=0} \alpha_k x^k_{ij}
+\end{align}
+
+\todo[inline]{we already use k for gain}
+
+All of variables are scaled such they they have a mean of zero and a standard
+deviation of one. We use cluster-mean centering with the clusters being an
+individual subject.
+
+We divide the analysis into two separate model evalautions, one for the
+6~\si{\kilo\meter\per\hour} \(k=10\) trials and one for the
+10~\si{\kilo\meter\per\hour} \(k=8\) trials and we evaluate our hypothesis for
+each set of data.
+
+\todo[inline]{Marten writes "The skill variable is not included, because it was
+not collected correctly for all participants." This needs to be addressed.}
 
@@
 \bibliographystyle{plain}
+% NOTE : this file is automatically generated from Zotero, do not edit
+% manually!
 \bibliography{references}

### 2024-03-26 main.tex
Added stats results tables and some subcaption figures.

@@
 \usepackage{siunitx}  % use for all units
+\usepackage{subcaption}
 \usepackage{todonotes}
@@
 \abstract{
-  Bicycles are unstable at low speeds. We add a controlled steering motor to a
-  consumer bicycle that stabilizes the bicycle at low speeds and apply varying
-  magnitude external handlebar perturbations to the bicycle while ridden on a
-  treadmill. The probability of not recovering from a handlebar perturbation
-  decreases when the balance assist is activated.
+  Uncontrolled bicycles are unstable at low speeds. We add a controlled
+  steering motor to a consumer bicycle that stabilizes the bicycle at low
+  speeds and apply varying magnitude external handlebar perturbations to the
+  bicycle while ridden on a treadmill. The probability of not recovering from a
+  handlebar perturbation decreases when the balance assist is activated.
 }
@@
 We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
-with a custom motor in the head tube cable of applying up to
+with a custom motor in the head tube capable of applying up to
 7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
-motor controller converts the commanded amperage to applied torque. We measure
-the speed of the rear wheel with an encoder (Magura XXX) and measure the roll
-rate of the bicycle with a MEMs rate gyroscope. The balance assist control
-algorithm is implemented on microprocessor (Teensy).
+motor controller converts the commanded torque to applied torque. We measure
+the speed of the rear wheel with an encoder (Magura XXX) and measure the body
+fixed roll rate of the bicycle with a MEMs rate gyroscope. The balance assist
+control algorithm is implemented on microprocessor (Teensy) and data from all
+sensors is logged with a CAN bus.
 
@@
 feedback controller to stabilize the bicycle. The commanded steer torque
-\(T_\delta\) from the steer motor follows
+\(T_\delta\) from the steer motor follows the control law
 \begin{align}
@@
 \begin{figure}
+  \centering
+  \subcaptionbox{Uncontrolled root locus of eigenvalue components plotted versus speed.}{
+    \includegraphics[width=70mm]{example-image-a}
+  }
+  \subcaptionbox{Controlled root locus of eigenvalue components plotted versus speed.}{
+    \includegraphics[width=70mm]{example-image-b}
+  }
   \label{fig:eigenvalues}
@@
 %
-We apply longitudinal forces to the ends of each handlebar using an adapted
-Bump'Em system~\cite{Tan2020} making use of four motors working in tandem. The
-four motors are programmed to apply a light force at all times to keep the
-ropes taught and to track a commanded force profile using a PID controller
-running on a microprocessor (Arduino XXX, Italy). We measure the force applied
-by each motor at the handlebar via four inline load cells with a max load of
-XX~\si{\newton}. The commanded force profiles are designed to apply an external
-pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
-varying from X to X. The four motors are arranged at the four corners of a 1
-meter wide treadmill (make, Company, City) that can reach speeds of
-18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
-Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our modifications to that
-described in the thesis include simplifying the controller with a inexpensive
-microcontroller and the use of a simple safety harness.
+We apply external longitudinal forces to the ends of each handlebar using an
+adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
+working in tandem. The four motors are programmed to apply a light force at all
+times to keep the ropes taught and to track a commanded force profile using a
+PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
+force applied by each motor at the handlebar via four inline load cells with a
+max load of XX~\si{\newton}. The commanded force profiles are designed to apply
+an external pulsive torque to the front assembly (handlebars, forks, wheel) at
+magnitudes varying from X to X. The four motors are arranged at the four
+corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can reach
+speeds of 18~\si{\kilo\meter\per\hour}. The general design is described in
+detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
+modifications to that described in the thesis include simplifying the
+controller with a inexpensive microcontroller and the use of a simpler safety
+harness.
+
+\begin{figure}
+  \centering
+  % TODO : scale these pngs to a reasonable size for an A4 sheet
+  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
+  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
+}
+\end{figure}
 
@@
 \begin{align}
-  \log\left(\frac{p_{ij}}{1 - p_{ij}}\right) = \Beta + \sum^K_{k=0} \alpha_k x^k_{ij}
+  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
 \end{align}
@@
 
+The coefficient estimates for a single-level logistic regression at a speed of 6
+\unit{\kilo\meter\per\hour} and gain $k=-10$ are shown in table \ref{table:freq-coefs-6}. The
+associated multiplicative change in odds are shown in table \ref{table:freq-change-in-odds-6}.
+
+\begin{table}
+  \centering
+  \caption{Coefficient estimates at 6 \unit{\kilo\meter\per\hour} and gain
+  $k=-10$ based on frequentist single-level logistic regression.}
+  \label{table:freq-coefs-6}
+  \begin{tabular}{lllll}
+    \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
+    Intercept                                     & -0.29             & 0.17                    & 0.09             & \\
+    Angular impulse                               & 1.69              & 0.27                    & 0.00             & \\
+    Perturbation order                            & -0.77             & 0.22                    & 0.00             & \\
+    Balance-assist on                             & -0.64             & 0.27                    & 0.02             & \\
+    Roll angle                                    & -0.25             & 0.21                    & 0.24             & \\
+    Steer angle                                   & -0.14             & 0.21                    & 0.51             & \\
+    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             & \\
+    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             & \\
+    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             & \\
+    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             & \\
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \centering
+  \caption{Multiplicative change in odds of predictor variables at 6
+  \unit{\kilo\meter\per\hour} and gain $k=-10$ based on frequentist
+         single-level logistic regression.}
+  \label{table:freq-change-in-odds-6}
+  \begin{tabular}{lllll}
+      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
+      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
+      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
+      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
+      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
+      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
+      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
+      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
+      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
+      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
+      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
+  \end{tabular}
+\end{table}
+
+The coefficients estimates of a single-level logistic regression are shown in table
+\ref{table:freq-coefs}. The associated multiplicative changes in odds are shown in table
+\ref{table:freq-change-in-odds}.
+
+\begin{table}
+  \centering
+  \caption{Coefficient estimates at 10 \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
+      regression.} \label{table:freq-coefs}
+  \begin{tabular}{lllll}
+      \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
+      Intercept                                     & -0.24             & 0.16                    & 0.13             & \\
+      Angular impulse                               & 2.39              & 0.29                    & 0.00             & \\
+      Perturbation order                            & -1.16             & 0.21                    & 0.00             & \\
+      Balance-assist on                             & -0.44             & 0.24                    & 0.07             & \\
+      Roll angle                                    & 0.27              & 0.22                    & 0.22             & \\
+      Steer angle                                   & -0.37             & 0.24                    & 0.12             & \\
+      Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08             & \\
+      Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11             & \\
+      Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29             & \\
+      Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25             & \\
+  \end{tabular}
+\end{table}
+
+\begin{table}
+  \centering
+  \caption{Multiplicative change in odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
+      frequentist single-level logistic regression.} \label{table:freq-change-in-odds}
+  \begin{tabular}{lllll}
+    \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
+    Intercept                                     & 0.78                                   & 0.57           & 1.07            & 0.12             \\
+    Angular impulse                               & 10.92                                  & 6.23           & 19.13           & 0.00             \\
+    Perturbation order                            & 0.31                                   & 0.21           & 0.48            & 0.00             \\
+    Balance-assist on                             & 0.64                                   & 0.40           & 1.03            & 0.07             \\
+    Roll angle                                    & 1.31                                   & 0.85           & 2.04            & 0.22             \\
+    Steer angle                                   & 0.69                                   & 0.43           & 1.10            & 0.12             \\
+    Balance-assist on $\times$ roll angle         & 0.54                                   & 0.28           & 1.07            & 0.08             \\
+    Balance-assist on $\times$ steer angle        & 1.76                                   & 0.88           & 3.50            & 0.11             \\
+    Balance-assist on $\times$ angular impulse    & 1.59                                   & 0.68           & 3.74            & 0.29             \\
+    Balance-assist on $\times$ perturbation order & 0.69                                   & 0.37           & 1.30            & 0.25             \\
+  \end{tabular}
+\end{table}
+
 \section{Discussion}

### 2024-04-08 main.tex
Added more text to the results and added a first take on the discussion.

@@
 \usepackage{amsmath}
+\usepackage{booktabs}
 \usepackage[margin=25mm]{geometry}
@@
 
+\def\kph{\kilo\meter\per\hour}
+
 \title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
-Subjected to Mechanical Perturbations}
+Subjected to Mechanical Perturbations at Low Speeds}
 
@@
   \begin{tabular}{llll}
+    \toprule
     Measure & Variable & Units & Sensor \\
+    \midrule
     Perturbation Order Number & \(j\) & integer & NA\\
@@
     Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
+    \bottomrule
   \end{tabular}
@@
 %
-Our primary hypothesis is that the balance assistance controller will reduce
-the probability of falling when perturbed externally at the handlebar. We have
-a single binary outcome variable \(f\) that is dependent on several possible
-explanatory independent variables including the binary balance assistance on or
-off.
+We test the hypothesis that the balance assistance controller will reduce the
+probability of falling when perturbed externally at the handlebar. We have a
+single binary outcome variable \(f\) that is dependent on several possible
+explanatory independent variables including the binary balance assistance (on
+or off).
 
@@
 \(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
-which follows a Bernoulli distriction given the probability \(p_{ij}\) that a
+which follows a Bernoulli distribution given the probability \(p_{ij}\) that a
 fall occurs. The log-odds of the probabiliy is then a linear function of our
-independent variables.
+independent variables with \(\beta\) as the intercept and \(\alpha_k\) as the
+linear coefficients.
 
@@
 
-All of variables are scaled such they they have a mean of zero and a standard
-deviation of one. We use cluster-mean centering with the clusters being an
-individual subject.
+We scalle all independent variables such they they have a mean of zero and a
+standard deviation of one. We use cluster-mean centering, as recommended by
+\cite{Enders2007}, with the clusters being an individual subject because we are
+interested in the association between the state of the balance-assist system
+and the outcome of the perturbation.
+
+Cluster-mean centering shows there to be no variation between participants,
+i.e. that rider skill (the only participant differentiating factor) is not a
+predictor. We expected there to be a variation between participants in how well
+they are able to resist the perturbation. However, this was not true. This
+allowed us to use a simpler single-level logistic regression model, instead of
+a multilevel model. This left use with angular impulse, perturbation order,
+balance assist state, and roll & steer angles at the time of perturbation. We
+also include interaction effects between the balance assist state and the other
+four variables.
+
+\todo[inline]{Marten writes "The skill variable is not included, because it was
+not collected correctly for all participants." This needs to be addressed. What
+about skill being tied to participant clusters?}
 
@@
 
-\todo[inline]{Marten writes "The skill variable is not included, because it was
-not collected correctly for all participants." This needs to be addressed.}
-
 \section{Results}
 
-The coefficient estimates for a single-level logistic regression at a speed of 6
-\unit{\kilo\meter\per\hour} and gain $k=-10$ are shown in table \ref{table:freq-coefs-6}. The
-associated multiplicative change in odds are shown in table \ref{table:freq-change-in-odds-6}.
+The coefficient estimates for a single-level logistic regression at
+6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in Table
+\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and balance
+assist state are all statistically significant predictors. Larger angular
+impulse increases the probability to fall and enduring more perturbations or
+having the balance assist on, decrease the probability to fall.  The associated
+multiplicative change in odds are shown in
+Table~\ref{table:freq-change-in-odds-6}.
 
@@
   \centering
-  \caption{Coefficient estimates at 6 \unit{\kilo\meter\per\hour} and gain
-  $k=-10$ based on frequentist single-level logistic regression.}
+  \caption{Logisitic regression coefficient estimates at
+    6~\unit{\kilo\meter\per\hour} and gain $k=10$.}
   \label{table:freq-coefs-6}
-  \begin{tabular}{lllll}
-    \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
-    Intercept                                     & -0.29             & 0.17                    & 0.09             & \\
-    Angular impulse                               & 1.69              & 0.27                    & 0.00             & \\
-    Perturbation order                            & -0.77             & 0.22                    & 0.00             & \\
-    Balance-assist on                             & -0.64             & 0.27                    & 0.02             & \\
-    Roll angle                                    & -0.25             & 0.21                    & 0.24             & \\
-    Steer angle                                   & -0.14             & 0.21                    & 0.51             & \\
-    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             & \\
-    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             & \\
-    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             & \\
-    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             & \\
+  \begin{tabular}{llll}
+    \toprule
+    Variable                                      & Estimate          & Standard error          & p-value          \\
+    \midrule
+    Intercept                                     & -0.29             & 0.17                    & 0.09             \\
+    Angular impulse                               & 1.69              & 0.27                    & 0.00             \\
+    Perturbation order                            & -0.77             & 0.22                    & 0.00             \\
+    Balance assist state                          & -0.64             & 0.27                    & 0.02             \\
+    Roll angle                                    & -0.25             & 0.21                    & 0.24             \\
+    Steer angle                                   & -0.14             & 0.21                    & 0.51             \\
+    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             \\
+    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             \\
+    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             \\
+    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             \\
+    \bottomrule
   \end{tabular}
 \end{table}
+
+\todo[inline]{Combine the columns of the multiplicative change in odds into the
+coefficient table}
 
@@
   \caption{Multiplicative change in odds of predictor variables at 6
-  \unit{\kilo\meter\per\hour} and gain $k=-10$ based on frequentist
+  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
          single-level logistic regression.}
@@
 
-The coefficients estimates of a single-level logistic regression are shown in table
-\ref{table:freq-coefs}. The associated multiplicative changes in odds are shown in table
-\ref{table:freq-change-in-odds}.
+The coefficient estimates for a single-level logistic regression at
+10~\si{\kilo\meter\per\hour} with gain $k=8$ are shown in Table
+\ref{table:freq-coefs}. The angular impulse and perturbation order are
+statistically significant predictors. Larger angular impulse increases the
+probability to fall and enduring more perturbations decreases the probability
+to fall. The associated multiplicative change in odds are shown in
+Table~\ref{table:freq-change-in-odds}.
 
@@
   \centering
-  \caption{Coefficient estimates at 10 \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
-      regression.} \label{table:freq-coefs}
-  \begin{tabular}{lllll}
-      \textbf{Variable}                             & \textbf{Estimate} & \textbf{Standard error} & \textbf{p-value} & \\ \cline{1-4}
-      Intercept                                     & -0.24             & 0.16                    & 0.13             & \\
-      Angular impulse                               & 2.39              & 0.29                    & 0.00             & \\
-      Perturbation order                            & -1.16             & 0.21                    & 0.00             & \\
-      Balance-assist on                             & -0.44             & 0.24                    & 0.07             & \\
-      Roll angle                                    & 0.27              & 0.22                    & 0.22             & \\
-      Steer angle                                   & -0.37             & 0.24                    & 0.12             & \\
-      Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08             & \\
-      Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11             & \\
-      Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29             & \\
-      Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25             & \\
+  \caption{Logisitic regression coefficient estimates at
+    10~\unit{\kilo\meter\per\hour} and gain $k=8$.}
+  \label{table:freq-coefs}
+  \begin{tabular}{llll}
+    \toprule
+    Variable                                      & Estimate          & Standard error          & p-value \\
+    \midrule
+    Intercept                                     & -0.24             & 0.16                    & 0.13    \\
+    Angular impulse                               & 2.39              & 0.29                    & 0.00    \\
+    Perturbation order                            & -1.16             & 0.21                    & 0.00    \\
+    Balance-assist on                             & -0.44             & 0.24                    & 0.07    \\
+    Roll angle                                    & 0.27              & 0.22                    & 0.22    \\
+    Steer angle                                   & -0.37             & 0.24                    & 0.12    \\
+    Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08    \\
+    Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11    \\
+    Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29    \\
+    Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25    \\
+    \bottomrule
   \end{tabular}
@@
 
+\subsection{Balance-assist}
+
+Turning the balance-assist system on significantly \((p<0.05)\) reduces the
+odds that a fall occurs while cycling at a speed of 6~\si{\kph} but at
+10~\si{\kph} that is not the case \((p=0.07)\). If all other factors remain the
+same at 6~\si{\kph}, the balance-assist system halves the odds that a
+perturbation results in a fall. How much this changes the probability that a
+fall occurs depends on the values of the other predictor variables: if the odds
+that a perturbation results in a fall are a 1000:1, turning on the
+balance-assist system reduces the odds to 500:1. However, the probability that
+a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
+the odds that a fall will occur are smaller, halving the odds has a larger
+influence on the fall probability. For example, if the odds that a fall occurs
+is two, halving it to one reduces the probability from $0.66$ to $0.50$. Figure
+\ref{fig:probability} shows how the balance-assist system changes the fall
+probability for different magnitudes of angular impulse.
+
+\begin{figure}
+  \centering
+  \subcaptionbox{6~\si{\kph}}{
+  \includegraphics[width=70mm]{example-image-a}
+    %\includegraphics[width=\textwidth]{figures/predicted_fall_probability-6.png}
+  }
+  \subcaptionbox{10~\si{\kph}}{
+    \includegraphics[width=70mm]{example-image-b}
+  }
+  \caption{Comparison of predicted fall probability for balance-assist on or
+    off All other predictor variables are fixed at zero, except for the
+    interaction effect between the balance-assist and the angular impulse.}
+  \label{fig:probability}
+\end{figure}
+
+Angular impulse magnitude has the largest significant effect for predicting
+fall probability. An increase in angular impulse increases the fall probability
+both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
+is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
+more important predictor at higher speeds compared to lower speeds. The reason
+for this likely has to do with the width of the treadmill. As a bicycle travels
+at higher speeds, the same perturbation causes larger lateral deviations. At
+10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
+of the treadmill. If the same experiment was performed on an infinite plane,
+the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
+could often recover in the alloted treadmill width. Our results are very much
+dependent on the two modes of falling with use: exit the treadmill width or
+foot is placed on the belt. Cycle paths are a similar width as the treadmill,
+so ride's are often limited in width when recoverying from a fall.
+
+\todo[inline]{Could show an impulse response in lateral deviation for different
+speeds.}
+
+An increase in the number of perturbation that a participant already
+experienced, decreases the probability that a fall occurs. This is likely due
+to the participants learning how to better recover from the perturbation over
+the duration of the experiment. This effect is significant at both 6 and
+10~\si{\kph}, and in the same order of magnitude. This suggests that the
+learning effect that occurs during the experiment is not strongly dependent on
+the speed.
+
+Roll and steer angle are not a significant predictor of fall probability,
+neither at 6 or at 10~\si{\kph}. We expected this to have an effect. If you are
+in a rolled and steered state that is far from the upright equilibrium, then a
+perturbation that further pushes you from the equilibrium should have some
+additive or multiplicative effect on the resulting motion trajectory.
+
+None of the interaction effects are statistically significant. That means that
+whether the balance-assist system is on or off does not significantly change
+the effect that the roll angle, steer angle, angular impulse, and perturbation
+order have on the probability that a fall will occur.
+
 \section{Conclusion}

### 2024-04-09 main.tex
Right columns in two tables.

@@
   \centering
-  \caption{Logisitic regression coefficient estimates at
+  \caption{Logistic regression coefficient estimates at
     6~\unit{\kilo\meter\per\hour} and gain $k=10$.}
   \label{table:freq-coefs-6}
-  \begin{tabular}{llll}
+  \begin{tabular}{lrrr}
     \toprule
@@
   \centering
-  \caption{Logisitic regression coefficient estimates at
+  \caption{Logistic regression coefficient estimates at
     10~\unit{\kilo\meter\per\hour} and gain $k=8$.}
   \label{table:freq-coefs}
-  \begin{tabular}{llll}
+  \begin{tabular}{lrrr}
     \toprule

### 2024-04-09 main.tex
Added text to the intro based on my draft steer control paper.

@@
 \usepackage{amsmath}
-\usepackage{booktabs}
+\usepackage{booktabs}  % nice tables
 \usepackage[margin=25mm]{geometry}
+\usepackage{natbib}
 \usepackage{siunitx}  % use for all units
-\usepackage{subcaption}
+\usepackage{subcaption}  % for subfigures
 \usepackage{todonotes}
@@
 
-\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
-Subjected to Mechanical Perturbations at Low Speeds}
+\title{Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds
+When Subjected to Mechanical Perturbations}
 
@@
 Single actor bicycle crashes are associated with a large number of unreported
-and reported injuries and even sometimes death. At low speeds (less than 15
-km/h or so) bicycles are no longer self stable and become more difficult to
-control for the rider. Low speed crashes may be reduced if the bicycle was self
-stable. Since at least the 1980s it known that applying a steering torque
+and reported injuries. At low speeds (less than 15 km/h or so) bicycles
+themselves are not self-stable and become more difficult to control for the
+rider. Low speed crashes may be reduced if the bicycle was self stable at these
+speeds. Since at least the 1980s it known that applying a steering torque
 proportional to the roll rate can stabilize a single track vehicle at very low
@@
 evaluate whether it helps the rider in situations they are likely to fall.
+
+Automatic roll stabilization of single track vehicles began in earnest after
+predictive motorcycle models were developed and refined throughout the 1970s.
+\Citet{vanZytveld1975} seems to be the first to attempt to robotically
+stabilize a small motorbike with a controlled inverted pendulum that mimicked
+rider lean, but was not successful in demonstrating what his control model
+predicted. In his control model, he recognized that feedback of the vehicle
+roll angle and angular rate was essential to stabilize the vehicle. It was not
+until the mid 1980s that someone successfully demonstrated an automatically
+balanced motorcycle~\cite{Ruijs1986a} and they did so with a steer motor. Ruijs
+and Pacejka clearly showed that steer torque driven by roll angle feedback
+stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
+mode, and steer angular rate feedback stabilizes the wobble mode. They also
+show how the gains need to change with respect to vehicle speed. Roll motion
+feedback is the simplest controller that can stabilize a single track vehicle
+above a minimum speed when you are not concerned with wobble destablization.
+But Ruijs and Pacejka's work was not particularly concerned with low speed
+stability and their vehicle was fully automatic, no human rider was involved.
+Many more automatically balanced single track vehicles have been demonstrated
+over the last 40 years, but none have demonstrated that increasing low speed
+stability reduces falling when ridden.  Most of these robotic bicycles and
+motorcycles designers did not intend for a human rider to also control the
+stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
+controlled by a human rider if the motor controlled steer torque and the rider
+applied steer torque act on the steer in parallel. The effect of this automatic
+control gives the ability to change the dynamics, to a degree, of the human
+controlled plant.
+
+The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
+the simplest bicycle model that exhibits non-minimum phase behavior and
+self-stablity. It is ideal for showing the effects of a roll motion feedback
+driven steer motor on the dynamics. The linear model can be described by the
+fourth order state space equations:
+
+\begin{align}
+  \dot{\vec{x}} = \mathbf{A} \vec{x} +
+  \mathbf{B}
+  \vec{u}
+  \textrm{ where }
+  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
+  \textrm{ and }
+  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
+\end{align}
+
+The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
+their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
+\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
+populated with expressions that are functions of 27 geometric and inertial
+parameters of the nonholonomic mulitbody system made up of four rigid bodies:
+two wheels, front frame, and rear frame.
+
+If the steer torque is the sum of the (h)uman applied torque and the (m)otor
+applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
+\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
+\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
+the human controlled plant takes the form:
+
+\begin{align}
+  \dot{\vec{x}} = \left(\mathbf{A} - \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right] \mathbf{B}_\delta\right) \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
+\end{align}
+
+The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
+functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
+such that the eigenvalues of \(\left[\mathbf{A} - \left[0 \quad k_{\dot{\phi}}
+\quad 0 \quad 0\right] \mathbf{B}_\delta\right]\) have negative real parts for
+\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
+the speed range where the bicycle is stable can be maximized within the limits
+of any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
+elaborates on some of the possibilities in scheduling the gains for such a
+controller.
+
+In this paper, we test whether a gain scheduled steer motor controlled bicycle
+that is riderless stable at a low speed range is beneficial in preventing
+whether the rider will fall. To do this, we apply varying magintidue mechanical
+perturbations to the handlebars while the rider is balancing the bicycle on a
+treadmill and then assess the riders' probability of falling with the balance
+assist system on versus off.
 
@@
 a multilevel model. This left use with angular impulse, perturbation order,
-balance assist state, and roll & steer angles at the time of perturbation. We
+balance assist state, and roll \& steer angles at the time of perturbation. We
 also include interaction effects between the balance assist state and the other
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{apalike}
 % NOTE : this file is automatically generated from Zotero, do not edit

### 2024-04-12 main.tex
Editing of most of the sections.

--- 2024-04-09 main.tex
@@
 %
-Single actor bicycle crashes are associated with a large number of unreported
-and reported injuries and even sometimes death. At low speeds (less than 15
-km/h or so) bicycles are no longer self stable and become more difficult to
-control for the rider. Low speed crashes may be reduced if the bicycle was self
-stable. Since at least the 1980s it known that applying a steering torque
-proportional to the roll rate can stabilize a single track vehicle at very low
-speeds. If the robotic control of steering can stabilize a bicycle, it may
-reduce the control required by the rider to successfully manage balancing
-tasks. We have developed such a balance assisting bicycle and desire to
-evaluate whether it helps the rider in situations they are likely to fall.
+Single actor bicycle crashes are associated with a suprisingly large number of
+reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
+cruising speeds, bicycles themselves are not self-stable and are challenging to
+control for the rider. Low speed crashes may be reduced if the bicycle was
+self-stable at these speeds. Bicycles can be mechanically modified to lower the
+speeds at which it is self-stable~\citep{Astrom2005} but since at least the
+1980s it is known that applying a steering torque proportional to the roll rate
+can stabilize a single track vehicle at very low speeds. If the robotic control
+of steering can stabilize a bicycle, it may reduce the control required by the
+rider to successfully manage balancing tasks. We have developed such a balance
+assisting bicycle and hypothesize that it helps the rider in situations they
+are likely to fall.
+
+Automatic roll stabilization of single track vehicles began in earnest after
+predictive motorcycle models were developed and refined throughout the 1970s.
+\Citet{vanZytveld1975} seems to be the first
+\todo{Check if a Japanese researcher was first}
+to attempt to robotically stabilize a small motorbike with a controlled
+inverted pendulum that mimicked rider lean, but he was not successful in
+demonstrating what his control model predicted. In his model, he recognized
+that feedback of the vehicle roll angle and angular rate was essential to
+stabilize the vehicle. It was not until the mid 1980s that someone successfully
+demonstrated an automatically balanced motorcycle~\citep{Ruijs1986a} and they
+did so with a steering motor. Ruijs and Pacejka clearly showed that steer
+torque driven by roll angle feedback stabilizes the capsize mode, roll angular
+rate feedback stabilizes the weave mode, and steer angular rate feedback
+stabilizes the wobble mode. They also show how the gains need to change with
+respect to vehicle speed for favorable control at all speeds. This roll motion
+feedback is the simplest controller that can stabilize a single track vehicle
+above a minimum speed when not concerned with wobble instabilities,
+which are not prominent in bicycles. Ruijs and Pacejka's work was not
+particularly concerned with low speed stability and their vehicle was fully
+automatic, no human rider was involved.  Many more automatically balanced
+single track vehicles have been demonstrated over the last 40 years, but none
+have demonstrated that increasing low speed stability can assist in balancing
+and possibly reduce single actor falls. Most of these robotic bicycles and
+motorcycles designers did not intend for a human rider to also control the
+stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
+controlled by a human rider if the motor controlled steer torque and the rider
+applied steer torque act on the steer in parallel. The effect of this automatic
+control gives the ability to change the dynamics, up to some limits, of the
+human controlled plant.
+
+The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
+the simplest bicycle model that exhibits non-minimum phase behavior and
+self-stablity. Interestingly, Whipple showed that roll motion feedback
+stabilizes the system back in 1899 and the model is ideal for showing the
+effect of a roll motion feedback driven steer motor on the dynamics. The linear
+model can be described by the fourth order state space equations:
+
+\begin{align}
+  \dot{\vec{x}} = \mathbf{A} \vec{x} +
+  \mathbf{B}
+  \vec{u}
+  \textrm{ where }
+  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
+  \textrm{ and }
+  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
+\end{align}
+
+The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
+their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
+\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
+functions of the equilbirum forward speed \(v\) and  are otherwise populated
+with expressions that are functions of the geometric and inertial parameters of
+the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
+front frame, and rear frame.
+
+If the steer torque is the sum of the (h)uman applied torque and the (m)otor
+applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
+\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
+\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
+the human controlled plant takes the form:
+
+\begin{align}
+  \dot{\vec{x}} = \left(\mathbf{A} -
+  \mathbf{B}_\delta
+  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
+\right)
+  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
+\end{align}
+
+The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
+functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
+such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0 \quad k_{\dot{\phi}}
+\quad 0 \quad 0\right] \right)\) have negative real parts for
+\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
+the speed range where the bicycle is stable can be maximized within
+any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
+elaborates on some of the possibilities in scheduling the gains for such a
+controller.
+
+\todo[inline]{I'm not consistent with the arrow being a vector for $B_\delta$}
+
+In this paper, we test whether a gain scheduled steer motor controlled bicycle
+that is stable at a low speed range when riderless assists in balancing by
+investigating whether it is beneficial in preventing the rider from falling.
+To do this, we apply varying magnitude mechanical perturbations to the
+handlebars while the rider is balancing the bicycle on a treadmill and then
+assess the rider's probability of falling with the balance assist steer motor
+system on and off.
 
@@
 We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
-with a custom motor in the head tube capable of applying up to
+with a custom motor in the steering assembly capable of applying up to
 7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
@@
 the speed of the rear wheel with an encoder (Magura XXX) and measure the body
-fixed roll rate of the bicycle with a MEMs rate gyroscope. The balance assist
-control algorithm is implemented on microprocessor (Teensy) and data from all
-sensors is logged with a CAN bus.
+fixed roll rate of the bicycle with a MEMs rate gyroscope (Company, Location).
+The balance assist control algorithm is implemented on a microprocessor
+(Teensy, Company, Location) and data from all sensors is logged with a CAN bus
+(Company, Location) at approximately X Hz.
 
 \subsection{Balance Assist Control}
-We use a forward speed \(v\) gain scheduled proportional roll rate positive
-feedback controller to stabilize the bicycle. The commanded steer torque
-\(T_\delta\) from the steer motor follows the control law
+We use a forward speed \(v\) gain scheduled proportional roll rate feedback
+controller to stabilize the bicycle. The commanded steer torque
+\(T^\textrm{m}_\delta\) from the steer motor follows the control law
 \begin{align}
-  T_\delta = k(4.7\si{\meter\per\second} - v)\dot{\phi}
+  T^\textrm{m}_\delta = k(v_{weave} - v)\dot{\phi}
 \end{align}
 where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
-open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
-and \(k=10\) as a high gain value. Scaling the proportional feedback gain
-linearly with respect to speed stabilizes the normally unstable weave mode of
-the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
+open loop bicycle rigid-rider dynamics. We use \(k=8\) and \(k=10\) as gain
+values during the experiments. Scaling the proportional feedback gain linearly
+with respect to speed stabilizes the normally unstable weave mode of the
+bicycle from speeds of X~\si{\kilo\meter\per\hour} to
 X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.
@@
 PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
-force applied by each motor at the handlebar via four inline load cells with a
-max load of XX~\si{\newton}. The commanded force profiles are designed to apply
-an external pulsive torque to the front assembly (handlebars, forks, wheel) at
-magnitudes varying from X to X. The four motors are arranged at the four
-corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can reach
-speeds of 18~\si{\kilo\meter\per\hour}. The general design is described in
-detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
-modifications to that described in the thesis include simplifying the
+force applied by each motor at the handlebar via four inline load cells, each
+with a max load of XX~\si{\newton}. The commanded force profiles are designed
+to apply an external pulsive torque to the front assembly (handlebars, forks,
+wheel) at magnitudes varying from X to X. The four motors are arranged at the
+four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
+reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
+in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
+modifications relative to that described in the thesis include simplifying the
 controller with a inexpensive microcontroller and the use of a simpler safety
@@
 
-\subsection{Two Experiments}
-%
-The 10~\si{\kilo\meter\per\hour} experiments occurred X weeks before the
-6~\si{\kilo\meter\per\hour}.
-
 \subsection{Protocol}
 %
+The subjects were divided into two groups, where the first group performed the
+protocol at a belt speed of 10~\si{\kph} with the controller gain as $k=8$ and
+the second group performed the protocol at a belt speed of 6~\si{\kph}. The
+10~\si{\kph} group's experiments occured X weeks before the
+6~\si{\kilo\meter\per\hour} group.
+
 Subjects wore a helmet and fall safety harness attached to the ceiling. We
 allowed them practice riding on the treadmill until they indicated they were
-comfortable enough to apply perturbations. For most this was less than a
-10~\si{\minute} warm up. We then ask the rider to ride for 90~\si{\second}
+comfortable enough to have perturbations applied. For most, this was less than a
+10~\si{\minute} warm up. We then asked the rider to ride for 90~\si{\second},
 attempting to maintain the location of their front wheel on the center line of
 the treadmill. We define a "fall" on the treadmill by two criteria: the rider
-removes their foot from the pedal and places it on the ground and the bicycle
+removes their foot from the pedal and places it on the ground or the bicycle
 wheel exceeds the width of the treadmill. We then applied perturbations in

### 2024-04-21 main.tex
Wordsmithing the abstract and introduction.

@@
 \abstract{
-  Uncontrolled bicycles are unstable at low speeds. We add a controlled
-  steering motor to a consumer bicycle that stabilizes the bicycle at low
-  speeds and apply varying magnitude external handlebar perturbations to the
-  bicycle while ridden on a treadmill. The probability of not recovering from a
-  handlebar perturbation decreases when the balance assist is activated.
+  Uncontrolled bicycles are generally unstable at low speeds. We add a
+  controlled steering motor to a consumer bicycle that stabilizes the bicycle
+  at low speeds. To test the motor's assistance during falls, we apply varying
+  magnitude external handlebar perturbations to the bicycle while ridden on a
+  treadmill with the balance assist system activated and deactivated.  The
+  probability of not recovering from a handlebar perturbation decreases when
+  the balance assist is activated at a speed of 6~\si{\kph}.
 }
@@
 %
-Single actor bicycle crashes are associated with a suprisingly large number of
+Single-actor bicycle crashes are associated with a suprisingly large number of
 reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
-cruising speeds, bicycles themselves are not self-stable and are challenging to
-control for the rider. Low speed crashes may be reduced if the bicycle was
-self-stable at these speeds. Bicycles can be mechanically modified to lower the
-speeds at which it is self-stable~\citep{Astrom2005} but since at least the
-1980s it is known that applying a steering torque proportional to the roll rate
-can stabilize a single track vehicle at very low speeds. If the robotic control
-of steering can stabilize a bicycle, it may reduce the control required by the
-rider to successfully manage balancing tasks. We have developed such a balance
-assisting bicycle and hypothesize that it helps the rider in situations they
-are likely to fall.
+cruising speeds, bicycles themselves are not self-stable and can be challenging
+for the rider to balance. Low speed crashes may be reduced if the bicycle was
+self-stable at these speeds, possibly relieving the rider of a necessary
+control activity. Bicycles can be mechanically modified to lower the speeds at
+which they are self-stable~\citep{Astrom2005} but since at least the 1980s it
+is known that applying a steering torque proportional to the roll rate can
+stabilize a single track vehicle at very low speeds. If the robotic control of
+steering that implements this can stabilize a bicycle, it may reduce the
+control required by the rider to successfully manage balancing tasks. We have
+developed such a balance assisting bicycle and hypothesize that it helps the
+rider in situations they are likely to fall.
 
 Automatic roll stabilization of single track vehicles began in earnest after
-predictive motorcycle models were developed and refined throughout the 1970s.
-\Citet{vanZytveld1975} seems to be the first
-\todo{Check if a Japanese researcher was first}
-to attempt to robotically stabilize a small motorbike with a controlled
-inverted pendulum that mimicked rider lean, but he was not successful in
-demonstrating what his control model predicted. In his model, he recognized
-that feedback of the vehicle roll angle and angular rate was essential to
-stabilize the vehicle. It was not until the mid 1980s that someone successfully
-demonstrated an automatically balanced motorcycle~\citep{Ruijs1986a} and they
-did so with a steering motor. Ruijs and Pacejka clearly showed that steer
-torque driven by roll angle feedback stabilizes the capsize mode, roll angular
-rate feedback stabilizes the weave mode, and steer angular rate feedback
-stabilizes the wobble mode. They also show how the gains need to change with
-respect to vehicle speed for favorable control at all speeds. This roll motion
-feedback is the simplest controller that can stabilize a single track vehicle
-above a minimum speed when not concerned with wobble instabilities,
-which are not prominent in bicycles. Ruijs and Pacejka's work was not
-particularly concerned with low speed stability and their vehicle was fully
-automatic, no human rider was involved.  Many more automatically balanced
-single track vehicles have been demonstrated over the last 40 years, but none
-have demonstrated that increasing low speed stability can assist in balancing
-and possibly reduce single actor falls. Most of these robotic bicycles and
-motorcycles designers did not intend for a human rider to also control the
-stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
-controlled by a human rider if the motor controlled steer torque and the rider
-applied steer torque act on the steer in parallel. The effect of this automatic
-control gives the ability to change the dynamics, up to some limits, of the
-human controlled plant.
+predictive motorcycle models were developed and refined in throughout the
+1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
+researcher was first} to attempt to robotically stabilize a small motorbike
+with a controlled inverted pendulum that mimicked rider lean, but he was not
+successful in demonstrating what his theoretic control model predicted. In his
+model, he recognized that feedback of the vehicle roll angle and angular rate
+was essential to stabilize the vehicle. It was not until the mid 1980s that
+someone successfully demonstrated an automatically balanced
+motorcycle~\citep{Ruijs1986a} and they did so with a steering motor. Ruijs and
+Pacejka clearly showed that steer torque driven by roll angle feedback
+stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
+mode, and steer angular rate feedback stabilizes the wobble mode. They also
+propose how the gains need to change with respect to vehicle speed for
+favorable control at all speeds. This roll motion feedback is the simplest
+controller that can stabilize a single track vehicle above a minimum speed when
+not concerned with wobble instabilities, which are not prominent in bicycles at
+low speeds. Ruijs and Pacejka's work was not particularly concerned with low
+speed stability and their vehicle was fully automatic, no human rider was
+involved.  Many more automatically balanced single track vehicles have been
+demonstrated over the last 40 years, but none have demonstrated that increasing
+low speed stability can assist in balancing and possibly reduce single actor
+falls. Most of these robotic bicycles and motorcycles designers did not intend
+for a human rider to also control the stabilized vehicle. Nevertheless, an
+automatically stabilized bicycle can be controlled by a human rider if the
+motor controlled steer torque and the rider applied steer torque act on the
+steer in parallel. The effect of this automatic control gives the ability to
+change the dynamics, up to some limits, of the human controlled plant.
 
@@
 the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stablity. Interestingly, Whipple showed that roll motion feedback
-stabilizes the system back in 1899 and the model is ideal for showing the
-effect of a roll motion feedback driven steer motor on the dynamics. The linear
-model can be described by the fourth order state space equations:
+self-stablity. Whipple showed that roll motion feedback can stabilize the
+system. The model is ideal for showing the effect of a roll motion feedback
+driven steer motor on the dynamics. The linear model can be described by the
+fourth order state space equations:
 

### 2024-04-24 main.tex
Use example images.
Fix quotes.

@@
   % TODO : scale these pngs to a reasonable size for an A4 sheet
-  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
-  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
+  %\subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
+  \subcaptionbox{}{\includegraphics[width=70mm]{example-image-a}}
+  %\subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
+  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{example-image-b}
 }
@@
 attempting to maintain the location of their front wheel on the center line of
-the treadmill. We define a "fall" on the treadmill by two criteria: the rider
+the treadmill. We define a ``fall'' on the treadmill by two criteria: the rider
 removes their foot from the pedal and places it on the ground or the bicycle
@@
 
-\todo[inline]{Marten writes "The skill variable is not included, because it was
-not collected correctly for all participants." This needs to be addressed. What
-about skill being tied to participant clusters?}
+\todo[inline]{Marten writes ``The skill variable is not included, because it
+was not collected correctly for all participants.'' This needs to be addressed.
+What about skill being tied to participant clusters?}
 

### 2024-04-27 main.tex
Added source to generate figures for bicycle model analysis.

@@
   \centering
-  \subcaptionbox{Uncontrolled root locus of eigenvalue components plotted versus speed.}{
-    \includegraphics[width=70mm]{example-image-a}
-  }
-  \subcaptionbox{Controlled root locus of eigenvalue components plotted versus speed.}{
-    \includegraphics[width=70mm]{example-image-b}
-  }
+  \includegraphics{figures/balance-assist-controllers-eig-vs-speeds.png}
+  \caption{Uncontrolled and controlled root locus of eigenvalue components
+    plotted versus speed.}
   \label{fig:eigenvalues}

### 2024-04-27 src/control.py

@@ -1,0 +1,83 @@
+import os
+
+from bicycleparameters.parameter_sets import Meijaard2007ParameterSet
+from bicycleparameters.bicycle import sort_eigenmodes
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import numpy as np
+from scipy.constants import golden_ratio
+
+from data import bike_with_rider as bike_par
+#from data import bike_without_rider as bike_par
+#from data import rigid_bike_without_rider as bike_par
+from model import SteerControlModel
+
+SCRIPT_PATH = os.path.realpath(__file__)
+SRC_DIR = os.path.dirname(SCRIPT_PATH)
+ROOT_DIR = os.path.realpath(os.path.join(SRC_DIR, '..'))
+FIG_DIR = os.path.join(ROOT_DIR, 'figures')
+
+if not os.path.exists(FIG_DIR):
+    os.mkdir(FIG_DIR)
+
+parameter_set = Meijaard2007ParameterSet(bike_par, True)
+model = SteerControlModel(parameter_set)
+
+# FIGURE : Geometry and mass distribution
+fig, ax = plt.subplots()
+fig.set_size_inches((5.0, 5.0/golden_ratio))
+parameter_set.plot_all(ax=ax)
+fig.savefig(os.path.join(FIG_DIR, 'bicycle-geometry-mass.png'), dpi=300)
+
+# FIGURE : Compare eigenvalues vs speed for uncontrolled and two controllers.
+fig, ax = plt.subplots()
+fig.set_size_inches((6.0, 6.0/golden_ratio))
+speeds = np.linspace(0.0, 10.0, num=1001)
+ax = model.plot_eigenvalue_parts(ax=ax, v=speeds, colors=['k']*4)
+# TODO : This finds the uncontrolled weave speed for the controller design, but
+# the actual bike uses a specific value based on the benchmark bike values
+# (probably).
+evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
+idx = np.argmin(np.abs(evals_[:, -1].real))
+weave_speed = speeds[idx]
+print('Uncontrolled weave speed: {:1.2f} [m/s]'.format(weave_speed))
+kphidots = -8.0*(weave_speed - speeds)
+kphidots[idx:] = 0.0
+ax = model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
[…]

### 2024-04-27 src/data.py

@@ -1,0 +1,106 @@
+# NOTE : This is the Batavus Browser + rider Jason taken from HumanControl
+# repo. The parameters match those of the Meijaard2007 paper.
+bike_with_rider = {
+    'IBxx': 11.3557360401,
+    'IBxz': -1.96756380745,
+    'IByy': 12.2177848012,
+    'IBzz': 3.12354397008,
+    'IFxx': 0.0904106601579,
+    'IFyy': 0.149389340425,
+    'IHxx': 0.253379594731,
+    'IHxz': -0.0720452391817,
+    'IHyy': 0.246138810935,
+    'IHzz': 0.0955770796289,
+    'IRxx': 0.0883819364527,
+    'IRyy': 0.152467620286,
+    'c': 0.0685808540382,
+    'g': 9.81,
+    'lam': 0.399680398707,
+    'mB': 81.86,
+    'mF': 2.02,
+    'mH': 3.22,
+    'mR': 3.11,
+    'rF': 0.34352982332,
+    'rR': 0.340958858855,
+    'v': 1.0,
+    'w': 1.121,
+    'xB': 0.289099434117,
+    'xH': 0.866949640247,
+    'zB': -1.04029228321,
+    'zH': -0.748236400835,
+    'kdelta': 0.0,
+    'kdeltadot': 0.0,
+    'kphi': 0.0,
+    'kphidot': 0.0,
+}
+
+# Batavus Browser without rider
+bike_without_rider = {
+    'IBxx': 0.52962890621,
+    'IBxz': -0.116285607878,
+    'IByy': 1.3163960125,
+    'IBzz': 0.756786895402,
+    'IFxx': 0.0883826870796,
+    'IFyy': 0.149221207336,
+    'IHxx': 0.25335539588,
+    'IHxz': -0.0720217263293,
+    'IHyy': 0.245827908036,
+    'IHzz': 0.0955686343473,
+    'IRxx': 0.0904114316323,
+    'IRyy': 0.152391250767,
+    'c': 0.0685808540382,
+    'g': 9.81,
+    'lam': 0.399680398707,
+    'mB': 9.86,
+    'mF': 2.02,
+    'mH': 3.22,
+    'mR': 3.11,
+    'rF': 0.34352982332,
+    'rR': 0.340958858855,
+    'v': 1.0,
+    'w': 1.121,
+    'xB': 0.275951285677,
+    'xH': 0.866949640247,
[…]

### 2024-04-27 src/model.py

@@ -1,0 +1,101 @@
+import numpy as np
+from bicycleparameters.models import Meijaard2007Model
+from bicycleparameters.bicycle import ab_matrix
+
+
+class SteerControlModel(Meijaard2007Model):
+    """
+
+    Tdel_total = -kphidot*phidot - kphi*phi + Tdel
+
+    Tdel_total = -[kphi, kdelta, kphidot, kdeltadot] * x
+
+    x = [roll angle,
+         steer angle,
+         roll rate,
+         steer rate]
+
+        The inputs are [roll torque,
+                        steer torque]
+
+    """
+
+    def form_state_space_matrices(self, **parameter_overrides):
+        """Returns the A and B matrices for the Whipple-Carvallo model
+        linearized about the upright constant velocity configuration with a
+        full state feedback steer controller.
+
+        Returns
+        =======
+        A : ndarray, shape(4,4) or shape(n,4,4)
+            The state matrix.
+        B : ndarray, shape(4,2) or shape(n,4,2)
+            The input matrix.
+
+        Notes
+        =====
+        A, B, and K describe the model in state space form:
+
+            x' = (A - B*K)*x + B*u
+
+        where::
+
+        x = |phi     | = |roll angle |
+            |delta   |   |steer angle|
+            |phidot  |   |roll rate  |
+            |deltadot|   |steer rate |
+
+        K = |0    0      0       0        |
+            |kphi kdelta kphidot kdeltadot|
+
+        u = |Tphi  | = |roll torque |
+            |Tdelta|   |steer torque|
+
+        """
+        gain_names = ['kphi', 'kdelta', 'kphidot', 'kdeltadot']
+
+        par, array_keys, array_len = self._parse_parameter_overrides(
+            **parameter_overrides)
+
+        # g, v, and the contoller gains are not used in the computation of M,
+        # C1, K0, K2.
+
+        M, C1, K0, K2 = self.form_reduced_canonical_matrices(
+            **parameter_overrides)
+
+        # steer controller gains, 2x4, no roll control
+        if any(k in gain_names for k in array_keys):
+            # if one of the gains is an array, create a set of gain matrices
+            # where that single gain varies across the set
+            K = np.array([[0.0, 0.0, 0.0, 0.0],
+                          [par[p][0] if p in array_keys else par[p]
+                           for p in gain_names]])
+            # K is now shape(n, 2, 4)
[…]

### 2024-04-27 src/control.py
Separate the eigenvalue figures and show stable regions.

@@ -7,8 +7,8 @@ import matplotlib.pyplot as plt
 import numpy as np
 from scipy.constants import golden_ratio
 
-from data import bike_with_rider as bike_par
-#from data import bike_without_rider as bike_par
+#from data import bike_with_rider as bike_par
+from data import bike_without_rider as bike_par
 #from data import rigid_bike_without_rider as bike_par
 from model import SteerControlModel
 
@@ -29,26 +29,71 @@ fig.set_size_inches((5.0, 5.0/golden_ratio))
 parameter_set.plot_all(ax=ax)
 fig.savefig(os.path.join(FIG_DIR, 'bicycle-geometry-mass.png'), dpi=300)
 
-# FIGURE : Compare eigenvalues vs speed for uncontrolled and two controllers.
+
+def stable_ranges(speeds, evals):
+    where_stable = np.all(evals.real < 0.0, axis=1)
+    number_ranges = np.count_nonzero(np.diff(where_stable)) - 1
+    stable_speeds = speeds[where_stable]
+    if number_ranges == 1:
+        ranges = ((stable_speeds[0], stable_speeds[-1]),)
+    else:
+        # TODO
+        #idxs = np.argmax(np.abs(np.diff(where_stable)))
+        ranges = ((stable_speeds[0], stable_speeds[-1]),)
+    return ranges
+
+
+# FIGURE : Compare eigenvalues vs speed for uncontrolled.
 fig, ax = plt.subplots()
-fig.set_size_inches((6.0, 6.0/golden_ratio))
+fig.set_size_inches((3.0, 3.0/golden_ratio))
 speeds = np.linspace(0.0, 10.0, num=1001)
+evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
+weave_idx = np.argmin(np.abs(evals_[:, -1].real))
+#weave_speed = speeds[weave_idx]
+#capsize_idx = np.argmin(np.abs(evals_[:, 0].real))
+#capsize_speed = speeds[capsize_idx]
+weave_speed, capsize_speed = stable_ranges(speeds, evals_)[0]
+print('Uncontrolled weave speed: {:1.2f} [m/s]'.format(weave_speed))
+print('Uncontrolled capsize speed: {:1.2f} [m/s]'.format(capsize_speed))
+kph2mps = 1000.0/3600.0
+ax.axvline(6.0*kph2mps, ymin=-10.0, ymax=10.0)
[…]

### 2024-04-27 main.tex
Separate the eigenvalue figures and show stable regions.
Use the gain of -10 for the plot.

@@
 
-\todo[inline]{Show open loop bicycle rigid rider eigenvalue vs speed and then
-the closed loop versions with the 8 and 10 gain settings. Three plots.}
-
 \begin{figure}
   \centering
-  \includegraphics{figures/balance-assist-controllers-eig-vs-speeds.png}
-  \caption{Uncontrolled and controlled root locus of eigenvalue components
-    plotted versus speed.}
+  \subcaptionbox{}{
+    \includegraphics[width=3in]{figures/uncontrolled-eig-vs-speeds.png}
+  }
+  \subcaptionbox{}{
+    \includegraphics[width=3in]{figures/balance-assist-controllers-eig-vs-speeds.png}
+  }
+  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
+    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
+    blue lines indicate the two speeds we perform experiments at. Green shaded
+    region is the linear stable range.}
   \label{fig:eigenvalues}

### 2024-04-27 src/control.py

@@ -16,6 +16,7 @@ SCRIPT_PATH = os.path.realpath(__file__)
 SRC_DIR = os.path.dirname(SCRIPT_PATH)
 ROOT_DIR = os.path.realpath(os.path.join(SRC_DIR, '..'))
 FIG_DIR = os.path.join(ROOT_DIR, 'figures')
+KPH2MPS = 1000.0/3600.0
 
 if not os.path.exists(FIG_DIR):
     os.mkdir(FIG_DIR)
@@ -49,15 +50,11 @@ fig.set_size_inches((3.0, 3.0/golden_ratio))
 speeds = np.linspace(0.0, 10.0, num=1001)
 evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
 weave_idx = np.argmin(np.abs(evals_[:, -1].real))
-#weave_speed = speeds[weave_idx]
-#capsize_idx = np.argmin(np.abs(evals_[:, 0].real))
-#capsize_speed = speeds[capsize_idx]
 weave_speed, capsize_speed = stable_ranges(speeds, evals_)[0]
 print('Uncontrolled weave speed: {:1.2f} [m/s]'.format(weave_speed))
 print('Uncontrolled capsize speed: {:1.2f} [m/s]'.format(capsize_speed))
-kph2mps = 1000.0/3600.0
-ax.axvline(6.0*kph2mps, ymin=-10.0, ymax=10.0)
-ax.axvline(10.0*kph2mps, ymin=-10.0, ymax=10.0)
+ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0)
+ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0)
 ax = model.plot_eigenvalue_parts(ax=ax, v=speeds, colors=['k']*4)
 ax.fill_between(speeds, -10, 10,
                 where=(speeds > weave_speed) & (speeds < capsize_speed),
@@ -74,7 +71,7 @@ fig.set_size_inches((3.0, 3.0/golden_ratio))
 # TODO : This finds the uncontrolled weave speed for the controller design, but
 # the actual bike uses a specific value based on the benchmark bike values
 # (probably).
-kphidots = -8.0*(weave_speed - speeds)
+kphidots = -10.0*(weave_speed - speeds)
 kphidots[weave_idx:] = 0.0
[…]

### 2024-04-27 src/control.py
Improved the stable speed range function and plotting.

@@ -31,17 +31,13 @@ parameter_set.plot_all(ax=ax)
 fig.savefig(os.path.join(FIG_DIR, 'bicycle-geometry-mass.png'), dpi=300)
 
 
-def stable_ranges(speeds, evals):
+def stable_ranges(evals):
+    # TODO : This should return the indice after a rise and before a fall. It
+    # now always returns the indices after a rise or fall.
     where_stable = np.all(evals.real < 0.0, axis=1)
-    number_ranges = np.count_nonzero(np.diff(where_stable)) - 1
-    stable_speeds = speeds[where_stable]
-    if number_ranges == 1:
-        ranges = ((stable_speeds[0], stable_speeds[-1]),)
-    else:
-        # TODO
-        #idxs = np.argmax(np.abs(np.diff(where_stable)))
-        ranges = ((stable_speeds[0], stable_speeds[-1]),)
-    return ranges
+    padded = np.hstack(([False], where_stable, [False]))
+    start_stop_idxs = np.flatnonzero(np.diff(padded))
+    return start_stop_idxs.reshape(-1, 2)
 
 
 # FIGURE : Compare eigenvalues vs speed for uncontrolled.
@@ -49,15 +45,16 @@ fig, ax = plt.subplots()
 fig.set_size_inches((3.0, 3.0/golden_ratio))
 speeds = np.linspace(0.0, 10.0, num=1001)
 evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
-weave_idx = np.argmin(np.abs(evals_[:, -1].real))
-weave_speed, capsize_speed = stable_ranges(speeds, evals_)[0]
+weave_idx, capsize_idx = stable_ranges(evals_)[0]
+weave_speed = speeds[weave_idx]
+capsize_speed = speeds[capsize_idx]
 print('Uncontrolled weave speed: {:1.2f} [m/s]'.format(weave_speed))
 print('Uncontrolled capsize speed: {:1.2f} [m/s]'.format(capsize_speed))
 ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0)
 ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0)
 ax = model.plot_eigenvalue_parts(ax=ax, v=speeds, colors=['k']*4)
 ax.fill_between(speeds, -10, 10,
-                where=(speeds > weave_speed) & (speeds < capsize_speed),
[…]

### 2024-05-23 main.tex
Added placeholders for figures that Marten and I discussed.

@@
 change the dynamics, up to some limits, of the human controlled plant.
+
+\todo[inline]{We should cite Leila's first paper and also the Pforzhiem work
+(if there is something to cite).}
 
@@
 
+\begin{figure}
+  \includegraphics{example-image-a}
+  \caption{Force nd torque plts of a pterubation of 80N per rope in the
+  counterclockwise direction (from Marten's thesis, pg 26)}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{example-image-a}
+  \caption{Thesis Fig 3.13 on pg 27}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{example-image-a}
+  \caption{Thesis Fig 5.1 on pg 37}
+\end{figure}
+
 \subsection{Subjects}
@@
 %
-The subjects were divided into two groups, where the first group performed the
+The subjects were divided into two groups. Tthe first group performed the
 protocol at a belt speed of 10~\si{\kph} with the controller gain as $k=8$ and
@@
 
+\begin{figure}
+  \includegraphics{example-image-a}
+  \caption{Box plot or distrubtion plot showing graphical view of stats
+  tables.}
+\end{figure}
+
 \section{Discussion}

### 2024-05-23 main.tex
Added two papers to intro and improved wording.

@@
 control activity. Bicycles can be mechanically modified to lower the speeds at
-which they are self-stable~\citep{Astrom2005} but since at least the 1980s it
-is known that applying a steering torque proportional to the roll rate can
-stabilize a single track vehicle at very low speeds. If the robotic control of
-steering that implements this can stabilize a bicycle, it may reduce the
-control required by the rider to successfully manage balancing tasks. We have
-developed such a balance assisting bicycle and hypothesize that it helps the
-rider in situations they are likely to fall.
+which they are self-stable~\citep{Astrom2005} but since the 1980s it has been
+known that applying a motor acutated steering torque proportional to the roll
+rate can stabilize a single track vehicle at very low speeds. If this automatic
+control of steering can stabilize a bicycle, it may reduce the control required
+by the rider to successfully manage balancing tasks. We have developed such a
+balance assisting bicycle and hypothesize that it helps the rider in situations
+they are likely to fall.
 
@@
 with a controlled inverted pendulum that mimicked rider lean, but he was not
-successful in demonstrating what his theoretic control model predicted. In his
+successful in demonstrating what his theoretic control model correctly predicted. In his
 model, he recognized that feedback of the vehicle roll angle and angular rate
 was essential to stabilize the vehicle. It was not until the mid 1980s that
-someone successfully demonstrated an automatically balanced
-motorcycle~\citep{Ruijs1986a} and they did so with a steering motor. Ruijs and
+Ruijs and Pacjecka successfully demonstrated an automatically balanced
+motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
 Pacejka clearly showed that steer torque driven by roll angle feedback
@@
 mode, and steer angular rate feedback stabilizes the wobble mode. They also
-propose how the gains need to change with respect to vehicle speed for
+propose how the gains must change with respect to vehicle speed for
 favorable control at all speeds. This roll motion feedback is the simplest
@@
 steer in parallel. The effect of this automatic control gives the ability to
-change the dynamics, up to some limits, of the human controlled plant.
+change the dynamics, up to some limits, of the human controlled plant.In our
+prior study, we demonstrate reduced motion during
+distractions~\cite{Alizadehsaravi2023} due to the balance assist but
+researchers with a similar vehicle recently showed rider disatisfction with the
+stabilization~\cite{Hanakam2023}.
 
-\todo[inline]{We should cite Leila's first paper and also the Pforzhiem work
-(if there is something to cite).}
-
-The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
+The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
 the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stablity. Whipple showed that roll motion feedback can stabilize the
-system. The model is ideal for showing the effect of a roll motion feedback
-driven steer motor on the dynamics. The linear model can be described by the
-fourth order state space equations:
+self-stablity. Whipple even showed that roll motion feedback can stabilize the
+system, remarkably in the early years of developments automatic control. The
+bicycle model is ideal for showing the effect of a roll motion feedback driven
+steer motor on the dynamics. The linear model can be described by the fourth
+order state space equations:
 
@@
 elaborates on some of the possibilities in scheduling the gains for such a
-controller.
+controller on this model.
 
@@
 %
-We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
-with a custom motor in the steering assembly capable of applying up to
-7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
-motor controller converts the commanded torque to applied torque. We measure
-the speed of the rear wheel with an encoder (Magura XXX) and measure the body
-fixed roll rate of the bicycle with a MEMs rate gyroscope (Company, Location).
-The balance assist control algorithm is implemented on a microprocessor
-(Teensy, Company, Location) and data from all sensors is logged with a CAN bus
-(Company, Location) at approximately X Hz.
+We modified a standard electric bicycle (Gazelle Grenoble C7, Dieren, The
+Netherlands) with a custom motor in the steering assembly capable of applying
+up to 7~\si{\newton\meter} of torque between the head tube and steer tube. A
+custom motor controller converts the commanded torque to applied torque. We
+measure the speed of the rear wheel with an encoder (Magura XXX) and measure
+the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
+Location).  The balance assist control algorithm is implemented on a
+microprocessor (Teensy, Company, Location) and data from all sensors is logged
+with a CAN bus (Company, Location) at approximately X Hz.
 
@@
 
+\todo[inline]{Decide on what physical parameters to use.}
+
 \subsection{Perturbations}
@@
 reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
-in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
+in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022a}. Our
 modifications relative to that described in the thesis include simplifying the
@@
 
-Based on the findings in \cite{Haitjema2023}, we calculate several variables.
+Based on the findings in \cite{Haitjema2023a}, we calculate several variables.
 We use the angular impulse \(L\) due to the perturbation forces over a

### 2024-05-28 src/generate_time_series_imgs.py
Add script that reads example data and creates force and torque plots

@@ -1,0 +1,331 @@
+import os
+import pandas as pd
+import matplotlib.pyplot as plt
+from bisect import bisect_left
+
+EXAMPLE_DATA = os.path.join("data", "example_data.parquet")
+TRACKING_FORCE = 3
+DESIRED_FORCES = ["desforce13", "desforce24"]
+ALL_FORCES = ["force1", "force2", "force3", "force4"] + DESIRED_FORCES
+DURATION_BEFORE = 0.3
+DURATION_AFTER = 2.0
+DIRECTORY = "imgs"
+HANLDEBAR_LENGTH = 0.82
+
+
+def main():
+    """Load example time series data from ./data and generate figures."""
+    data = pd.read_parquet(EXAMPLE_DATA)
+    perturbation_dfs = get_perturbations(
+        data, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
+    )
+    generate_force_torque_plots(perturbation_dfs, DIRECTORY, ALL_FORCES)
+
+
+def get_perturbations(
+    data,
+    DESIRED_FORCES,
+    duration_before,
+    duration_after,
+):
+    """Returns each perturbation in a new dataframe. Flips the direction of the
+    roll angle and steer rate such that there is no difference between clockwise
+    and counterclockwise.
+
+    Parameters
+    ----------
+    data : pandas.DataFrame
+        Dataframe containing time series data of the experiment.
+    DESIRED_FORCES : List[str]
+        Names of the force column headers in the dataframe.
+    duration_before : float
+        Duration in seconds before the perturbation is applied that should be included.
+    duration_after : float
+        Duration in seconds after the perurbation has ended that should be included.
+
+    Returns
+    -------
+    df_list : List[pd.DataFrame]
+        List of dataframes where the direction is normalized for the perturbations.
+    """
+    start_indices, stop_indices = get_perturbation_indices(data, DESIRED_FORCES)
+    context_start_indices, context_stop_indices = get_context_around_perturbation(
+        data, start_indices, stop_indices, duration_before, duration_after
+    )
+
+    perturbation_dfs = []
+    for i, _ in enumerate(start_indices):
+        start = start_indices[i]
+        context_start = context_start_indices[i]
+        context_stop = context_stop_indices[i]
+
+        pert_data = pd.DataFrame()
+        start_time = data.loc[start, "seconds_since_start"]
+        pert_data["seconds_since_start"] = (
[…]

### 2024-05-28 src/generate_time_series_imgs.py
Create overlay plots, set time on x-axii relative to start of perturbation, add example data of balance-assist on and off

@@ -1,9 +1,11 @@
 import os
 import pandas as pd
 import matplotlib.pyplot as plt
+from matplotlib.lines import Line2D
 from bisect import bisect_left
 
-EXAMPLE_DATA = os.path.join("data", "example_data.parquet")
+EXAMPLE_DATA_OFF = os.path.join("data", "example_data_balance_assist_off.parquet")
+EXAMPLE_DATA_ON = os.path.join("data", "example_data_balance_assist_on.parquet")
 TRACKING_FORCE = 3
 DESIRED_FORCES = ["desforce13", "desforce24"]
 ALL_FORCES = ["force1", "force2", "force3", "force4"] + DESIRED_FORCES
@@ -15,11 +17,17 @@ HANLDEBAR_LENGTH = 0.82
 
 def main():
     """Load example time series data from ./data and generate figures."""
-    data = pd.read_parquet(EXAMPLE_DATA)
-    perturbation_dfs = get_perturbations(
-        data, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
+    data_on = pd.read_parquet(EXAMPLE_DATA_ON)
+    perturbation_dfs_on = get_perturbations(
+        data_on, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
     )
+    data_off = pd.read_parquet(EXAMPLE_DATA_OFF)
+    perturbation_dfs_off = get_perturbations(
+        data_off, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
+    )
+    perturbation_dfs = perturbation_dfs_on + perturbation_dfs_off
     generate_force_torque_plots(perturbation_dfs, DIRECTORY, ALL_FORCES)
+    generate_roll_steer_plots(perturbation_dfs, DIRECTORY)
 
 
 def get_perturbations(
@@ -66,7 +74,10 @@ def get_perturbations(
         )
 
         for var in data.columns.values:
-            if ("steer" or "roll" or "gyro") in var:
+            if var == "seconds_since_start":
+                continue
+
+            if "steer" in var or "roll" in var or "gyro" in var:
                 pert_data[var + "_orginal"] = data[var][context_start:context_stop]
                 if data["desforce24"][start] > TRACKING_FORCE:
                     pert_data[var] = -data[var][context_start:context_stop]
@@ -126,7 +137,7 @@ def generate_force_torque_plots(perturbation_dfs, directory, force_column_names)
[…]

### 2024-05-29 src/generate_time_series_imgs.py
Consolidate in the figures directory.

@@ -11,7 +11,7 @@ DESIRED_FORCES = ["desforce13", "desforce24"]
 ALL_FORCES = ["force1", "force2", "force3", "force4"] + DESIRED_FORCES
 DURATION_BEFORE = 0.3
 DURATION_AFTER = 2.0
-DIRECTORY = "imgs"
+DIRECTORY = "figures"
 HANLDEBAR_LENGTH = 0.82
 
 

### 2024-05-29 main.tex
Added some of the static figures.

@@
   % TODO : scale these pngs to a reasonable size for an A4 sheet
-  %\subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
-  \subcaptionbox{}{\includegraphics[width=70mm]{example-image-a}}
-  %\subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
-  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{example-image-b}
+  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
+  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
 }
@@
 \begin{figure}
-  \includegraphics{example-image-a}
+  \centering
+  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
+  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
+  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
+  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
   \caption{Thesis Fig 5.1 on pg 37}

### 2024-05-31 src/generate_time_series_imgs.py
Only plot the first perturbation.

@@ -26,7 +26,7 @@ def main():
         data_off, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
     )
     perturbation_dfs = perturbation_dfs_on + perturbation_dfs_off
-    generate_force_torque_plots(perturbation_dfs, DIRECTORY, ALL_FORCES)
+    generate_force_torque_plots(perturbation_dfs[:1], DIRECTORY, ALL_FORCES)
     generate_roll_steer_plots(perturbation_dfs, DIRECTORY)
 
 

### 2024-05-31 main.tex
Added in data processing figure.
Added perturbation figure.
Update for CI build corrections.

@@
 
-Automatic roll stabilization of single track vehicles began in earnest after
-predictive motorcycle models were developed and refined in throughout the
-1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
-researcher was first} to attempt to robotically stabilize a small motorbike
-with a controlled inverted pendulum that mimicked rider lean, but he was not
-successful in demonstrating what his theoretic control model correctly predicted. In his
-model, he recognized that feedback of the vehicle roll angle and angular rate
-was essential to stabilize the vehicle. It was not until the mid 1980s that
-Ruijs and Pacjecka successfully demonstrated an automatically balanced
+In the early years of developments automatic control\citep{Whipple1899}
+remarkably realized and showed that roll motion feedback can stabilize the a
+bicycle. Automatic roll stabilization of single track vehicles began in earnest
+after predictive motorcycle models were developed and refined in throughout the
+1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
+the first \todo{Check if a Japanese researcher was first} to attempt to
+robotically stabilize a small motorbike with a controlled inverted pendulum
+that mimicked rider lean, but he was not successful in demonstrating what his
+theoretic control model correctly predicted. In his model, he recognized that
+feedback of the vehicle roll angle and angular rate was essential to stabilize
+the vehicle. It was not until the mid 1980s that Ruijs and Pacjecka
+successfully demonstrated an automatically balanced
 motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
@@
 mode, and steer angular rate feedback stabilizes the wobble mode. They also
-propose how the gains must change with respect to vehicle speed for
-favorable control at all speeds. This roll motion feedback is the simplest
-controller that can stabilize a single track vehicle above a minimum speed when
-not concerned with wobble instabilities, which are not prominent in bicycles at
-low speeds. Ruijs and Pacejka's work was not particularly concerned with low
-speed stability and their vehicle was fully automatic, no human rider was
-involved. Many more automatically balanced single track vehicles have been
-demonstrated over the last 40 years, but none have demonstrated that increasing
-low speed stability can assist in balancing and possibly reduce single actor
-falls. Most of these robotic bicycles and motorcycles designers did not intend
-for a human rider to also control the stabilized vehicle. Nevertheless, an
-automatically stabilized bicycle can be controlled by a human rider if the
-motor controlled steer torque and the rider applied steer torque act on the
-steer in parallel. The effect of this automatic control gives the ability to
-change the dynamics, up to some limits, of the human controlled plant.In our
-prior study, we demonstrate reduced motion during
-distractions~\cite{Alizadehsaravi2023} due to the balance assist but
-researchers with a similar vehicle recently showed rider disatisfction with the
-stabilization~\cite{Hanakam2023}.
+propose how the gains must change with respect to vehicle speed for favorable
+control at all speeds. This roll motion feedback is the simplest controller
+that can stabilize a single track vehicle above a minimum speed when not
+concerned with wobble instabilities, which are not prominent in bicycles at low
+speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
+stability and their vehicle was fully automatic, no human rider was involved.
+Many more automatically balanced single track vehicles have been demonstrated
+over the last 40 years, but none have demonstrated that increasing low speed
+stability can assist in balancing and possibly reduce single actor falls. Most
+of these robotic bicycles and motorcycles designers did not intend for a human
+rider to also control the stabilized vehicle. Nevertheless, an automatically
+stabilized bicycle can be controlled by a human rider if the motor controlled
+steer torque and the rider applied steer torque act on the steer in parallel.
+The effect of this automatic control gives the ability to change the dynamics,
+up to some limits, of the human controlled plant.In our prior study, we
+demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
+the balance assist but researchers with a similar vehicle recently showed rider
+disatisfction with the stabilization~\cite{Hanakam2023}.
 
@@
 the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stablity. Whipple even showed that roll motion feedback can stabilize the
-system, remarkably in the early years of developments automatic control. The
-bicycle model is ideal for showing the effect of a roll motion feedback driven
-steer motor on the dynamics. The linear model can be described by the fourth
-order state space equations:
+self-stablity. The bicycle model is ideal for showing the effect of a roll
+motion feedback driven steer motor on the dynamics. The linear model can be
+described by the fourth order state space equations:
 
@@
 \begin{figure}
-  \includegraphics{example-image-a}
+  \includegraphics[width=\textwidth]{figures/perturbation_0.png}
   \caption{Force nd torque plts of a pterubation of 80N per rope in the
   counterclockwise direction (from Marten's thesis, pg 26)}
-\end{figure}
-
-\begin{figure}
-  \includegraphics{example-image-a}
-  \caption{Thesis Fig 3.13 on pg 27}
 \end{figure}
@@
 treadmill belt. We record this as a binary variable \(f\).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\columnwidth]{figures/data-processing.png}
+  \caption{Thesis Fig 3.13 on pg 27}
+\end{figure}
 

### 2024-05-31 main.tex
Removed bib url due to format issue.

@@
 for the rider to balance. Low speed crashes may be reduced if the bicycle was
-self-stable at these speeds, possibly relieving the rider of a necessary
-control activity. Bicycles can be mechanically modified to lower the speeds at
-which they are self-stable~\citep{Astrom2005} but since the 1980s it has been
-known that applying a motor acutated steering torque proportional to the roll
-rate can stabilize a single track vehicle at very low speeds. If this automatic
+self-stable at these speeds by  relieving the rider of a necessary control
+activity. Bicycles can be mechanically modified to lower the speeds at which
+they are self-stable~\citep{Astrom2005} but since the 1980s, it has been known
+that applying a motor acutated steering torque proportional to the roll rate
+can stabilize a single track vehicle at very low speeds. If this automatic
 control of steering can stabilize a bicycle, it may reduce the control required
@@
 
-In the early years of developments automatic control\citep{Whipple1899}
+In the early years of developments automatic control, \citep{Whipple1899}
 remarkably realized and showed that roll motion feedback can stabilize the a
-bicycle. Automatic roll stabilization of single track vehicles began in earnest
-after predictive motorcycle models were developed and refined in throughout the
-1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
-the first \todo{Check if a Japanese researcher was first} to attempt to
-robotically stabilize a small motorbike with a controlled inverted pendulum
-that mimicked rider lean, but he was not successful in demonstrating what his
-theoretic control model correctly predicted. In his model, he recognized that
-feedback of the vehicle roll angle and angular rate was essential to stabilize
-the vehicle. It was not until the mid 1980s that Ruijs and Pacjecka
-successfully demonstrated an automatically balanced
+bicycle. Much later, automatic roll stabilization of single track vehicles
+began in earnest after predictive motorcycle models were developed and refined
+in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
+work and seems to be the first \todo{Check if a Japanese researcher was first}
+to attempt to robotically stabilize a small motorbike with a controlled
+inverted pendulum that mimicked rider lean, but he was not successful in
+demonstrating what his theoretic control model correctly predicted. In his
+model, he recognized that feedback of the vehicle roll angle and angular rate
+was essential to stabilize the vehicle. It was not until the mid 1980s that
+Ruijs and Pacjecka successfully demonstrated an automatically balanced
 motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
@@
 control at all speeds. This roll motion feedback is the simplest controller
-that can stabilize a single track vehicle above a minimum speed when not
+that can stabilize a single track vehicle above a minimum speed when one is not
 concerned with wobble instabilities, which are not prominent in bicycles at low
@@
 stability can assist in balancing and possibly reduce single actor falls. Most
-of these robotic bicycles and motorcycles designers did not intend for a human
+of these robotic bicycle and motorcycle designers did not intend for a human
 rider to also control the stabilized vehicle. Nevertheless, an automatically
@@
 The effect of this automatic control gives the ability to change the dynamics,
-up to some limits, of the human controlled plant.In our prior study, we
+up to some limits, of the human controlled plant. In our prior study, we
 demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
-the balance assist but researchers with a similar vehicle recently showed rider
-disatisfction with the stabilization~\cite{Hanakam2023}.
+the balance assist but conversely researchers with a similar vehicle recently
+showed rider disatisfction with the stabilization~\cite{Hanakam2023}.
 

### 2024-05-31 src/generate_time_series_imgs.py
Improved size and fontsize of perturbation figure.

@@ -108,10 +108,11 @@ def generate_force_torque_plots(perturbation_dfs, directory, force_column_names)
         os.mkdir(directory)
 
     for i, df in enumerate(perturbation_dfs):
-        fig, axs = plt.subplots(2, 1, sharex=True)
-        fig.set_size_inches(10, 6)
+        fig, axs = plt.subplots(2, 1, sharex=True, layout='constrained')
+        fig.set_size_inches(150/25.4, 100/25.4)
 
-        df.plot(x="seconds_since_start", y=force_column_names, ax=axs[0], grid=True)
+        df.plot(x="seconds_since_start", y=force_column_names, ax=axs[0],
+                grid=True)
 
         actual_torque, desired_torque = calculate_torque_on_handlebars(df)
         axs[1].plot(
@@ -135,13 +136,13 @@ def generate_force_torque_plots(perturbation_dfs, directory, force_column_names)
                 "Motor 4",
                 "Desired motor 1 and 3",
                 "Desired motor 2 and 4",
-            ]
+            ],
+            fontsize='x-small'
         )
         axs[1].set_xlabel("Time relative to start of perturbation [s]")
         axs[1].set_ylabel("Torque [Nm]")
-        axs[1].legend()
+        axs[1].legend(fontsize='x-small')
         axs[1].grid()
-        axs[1].legend()
 
         filename = os.path.join(directory, "perturbation_" + str(i))
         fig.savefig(fname=filename, dpi=300, bbox_inches="tight")

### 2024-05-31 main.tex
Figure updates and text for figures.

@@
 in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
-work and seems to be the first \todo{Check if a Japanese researcher was first}
-to attempt to robotically stabilize a small motorbike with a controlled
-inverted pendulum that mimicked rider lean, but he was not successful in
-demonstrating what his theoretic control model correctly predicted. In his
-model, he recognized that feedback of the vehicle roll angle and angular rate
-was essential to stabilize the vehicle. It was not until the mid 1980s that
-Ruijs and Pacjecka successfully demonstrated an automatically balanced
+work and seems to be the first to attempt to robotically stabilize a small
+motorbike with a controlled inverted pendulum that mimicked rider lean, but he
+was not successful in demonstrating what his theoretic control model correctly
+predicted. In his model, he recognized that feedback of the vehicle roll angle
+and angular rate was essential to stabilize the vehicle. It was not until the
+early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
+bicycle on a treadmill with both steering control and a laterally moving mass.
+Ruijs and Pacjecka followed this by demonstrating an automatically balanced
 motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
@@
 Netherlands) with a custom motor in the steering assembly capable of applying
-up to 7~\si{\newton\meter} of torque between the head tube and steer tube. A
-custom motor controller converts the commanded torque to applied torque. We
-measure the speed of the rear wheel with an encoder (Magura XXX) and measure
-the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
-Location).  The balance assist control algorithm is implemented on a
-microprocessor (Teensy, Company, Location) and data from all sensors is logged
-with a CAN bus (Company, Location) at approximately X Hz.
+up to 7~\si{\newton\meter} of torque between the head tube and steer tube,
+Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
+commanded torque to applied torque. We measure the speed of the rear wheel with
+an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
+with a MEMs rate gyroscope (Company, Location). The balance assist control
+algorithm is implemented on a microprocessor (Teensy, Company, Location) and
+data from all sensors is logged with a CAN bus (Company, Location) at
+approximately X Hz.
+%
+\begin{figure}
+  \centering
+  \includegraphics[width=100mm]{figures/balance-assist-bicycle.jpg}
+  \caption{Balance assist bicycle prototype with steer motor and data
+    aquisition and control electronics mounted in the rear rack.}
+  \label{fig:balance-assist-bicycle}
+\end{figure}
 
@@
 adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
-working in tandem. The four motors are programmed to apply a light force at all
-times to keep the ropes taught and to track a commanded force profile using a
-PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
-force applied by each motor at the handlebar via four inline load cells, each
-with a max load of XX~\si{\newton}. The commanded force profiles are designed
-to apply an external pulsive torque to the front assembly (handlebars, forks,
-wheel) at magnitudes varying from X to X. The four motors are arranged at the
-four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
-reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
-in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022a}. Our
-modifications relative to that described in the thesis include simplifying the
-controller with a inexpensive microcontroller and the use of a simpler safety
-harness.
-
+working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
+programmed to apply a light force at all times to keep the ropes taught and to
+track a commanded force profile using a PID controller running on a
+microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
+at the handlebar via four inline load cells, each with a max load of
+XX~\si{\newton}. The commanded force profiles are designed to apply an external
+pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
+varying from X to X. The four motors are arranged at the four corners of a
+1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
+18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
+Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
+is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
+to that described in the thesis include simplifying the controller with a
+inexpensive microcontroller and the use of a simpler safety harness.
+%
 \begin{figure}
@@
   % TODO : scale these pngs to a reasonable size for an A4 sheet
-  \subcaptionbox{}{\includegraphics[width=70mm]{figures/participant-in-set-up.png}}
-  \subcaptionbox{Diagram of perturbation system.}{\includegraphics[width=70mm]{figures/setup-diagram.png}
+  \subcaptionbox{
+    A participant on the bicycle in the safety harness with the Bump'Em motor
+    lines attached to the ends of the handlebars.
+    \label{fig:participant-in-set-up}
+  }{
+    \includegraphics[width=70mm]{figures/participant-in-set-up.png}
+  }
+  \subcaptionbox{
+    Top view diagram of bicycle handlebar perturbation system.
+    \label{fig:setup-diagram}
+  }{
+    \includegraphics[width=70mm]{figures/setup-diagram.png
+  }
 }
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/perturbation_0.png}
-  \caption{Force nd torque plts of a pterubation of 80N per rope in the
-  counterclockwise direction (from Marten's thesis, pg 26)}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
-  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
-  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
-  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
-  \caption{Thesis Fig 5.1 on pg 37}
+  \includegraphics[width=150mm]{figures/perturbation_0.png}
+  \caption{Motor force and resulting  steer torque perturbation based on an 80N
+  counterclockwise applied force.}
 \end{figure}
@@
 
+\todo[inline]{Include the average age, mass, height and standard deviation if
+we recorded those.}
+
 \subsection{Protocol}
 %
-The subjects were divided into two groups. Tthe first group performed the
-protocol at a belt speed of 10~\si{\kph} with the controller gain as $k=8$ and
-the second group performed the protocol at a belt speed of 6~\si{\kph}. The
-10~\si{\kph} group's experiments occured X weeks before the
-6~\si{\kilo\meter\per\hour} group.
+The subjects were divided into two groups. The first group performed the
+protocol at a belt speed of 2.8~\si{\meter\per\second}
+(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
+group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
+(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occured X
+weeks before the 6~\si{\kilo\meter\per\hour} group.
 
-Subjects wore a helmet and fall safety harness attached to the ceiling. We
-allowed them practice riding on the treadmill until they indicated they were
-comfortable enough to have perturbations applied. For most, this was less than a
-10~\si{\minute} warm up. We then asked the rider to ride for 90~\si{\second},
-attempting to maintain the location of their front wheel on the center line of
-the treadmill. We define a ``fall'' on the treadmill by two criteria: the rider
-removes their foot from the pedal and places it on the ground or the bicycle
-wheel exceeds the width of the treadmill. We then applied perturbations in
-random directions (clockwise or counter clockwise) increasing the magnitude by
-X~\si{\newton\meter} each time. We log the magnitude that causes the first fall
-to characterize that subject's nominal fall threshold. Following this we apply
-20 perturbations of random magnitudes to the cyclist while they ride at a
-constant speed and record which perturbations cause a fall. We let the cyclist
-rest and then perform another 20 perturbations. We randomize whether the
-balance assist system is on or off during the first or second set of
-perturbations.
+Subjects wore a helmet and fall safety harness attached to the ceiling
+Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
+the treadmill until they indicated they were comfortable enough to have
+perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
+We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
+location of their front wheel on the center line of the treadmill. We define a
+``fall'' on the treadmill by two criteria: the rider removes their foot from
+the pedal and places it on the ground or the bicycle wheel exceeds the width of
+the treadmill. Figure~\ref{fig:fall} shows an example fall. We then applied
+perturbations in random directions (clockwise or counter clockwise) increasing
+the magnitude by X~\si{\newton\meter} each time.  We log the magnitude that
+causes the first fall to characterize that subject's nominal fall threshold.
+Following this we apply 20 perturbations of random magnitudes to the cyclist
+while they ride at a constant speed and record which perturbations cause a
+fall. We let the cyclist rest and then perform another 20 perturbations. We
+randomize whether the balance assist system is on or off during the first or
+second set of perturbations.
+%
+\begin{figure}
+  \centering
+  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
+  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
+  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
+  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
+  \caption{Thesis Fig 5.1 on pg 37}
+  \label{fig:fall}
+\end{figure}
 

### 2024-05-31 main.tex
Adjustment of figure location and size.

@@
   \subcaptionbox{}{
-    \includegraphics[width=3in]{figures/uncontrolled-eig-vs-speeds.png}
+    \includegraphics[width=70mm]{figures/uncontrolled-eig-vs-speeds.png}
   }
   \subcaptionbox{}{
-    \includegraphics[width=3in]{figures/balance-assist-controllers-eig-vs-speeds.png}
+    \includegraphics[width=70mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
   }
@@
 inexpensive microcontroller and the use of a simpler safety harness.
+Figure~\ref{fig:perturbation_0} shows an example perturbation.
 %
@@
   }{
-    \includegraphics[width=70mm]{figures/participant-in-set-up.png}
+    \includegraphics[width=75mm]{figures/participant-in-set-up.png}
   }
@@
   }{
-    \includegraphics[width=70mm]{figures/setup-diagram.png
+    \includegraphics[width=65mm]{figures/setup-diagram.png
   }
 }
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=150mm]{figures/perturbation_0.png}
-  \caption{Motor force and resulting  steer torque perturbation based on an 80N
-  counterclockwise applied force.}
 \end{figure}
@@
 the pedal and places it on the ground or the bicycle wheel exceeds the width of
-the treadmill. Figure~\ref{fig:fall} shows an example fall. We then applied
-perturbations in random directions (clockwise or counter clockwise) increasing
-the magnitude by X~\si{\newton\meter} each time.  We log the magnitude that
-causes the first fall to characterize that subject's nominal fall threshold.
-Following this we apply 20 perturbations of random magnitudes to the cyclist
-while they ride at a constant speed and record which perturbations cause a
-fall. We let the cyclist rest and then perform another 20 perturbations. We
-randomize whether the balance assist system is on or off during the first or
-second set of perturbations.
+the treadmill. We then applied perturbations in random directions (clockwise or
+counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
+Figure~\ref{fig:perturbation-sequence} shows an example reult of a
+perturbation.  We log the magnitude that causes the first fall to characterize
+that subject's nominal fall threshold.  Following this we apply 20
+perturbations of random magnitudes to the cyclist while they ride at a constant
+speed and record which perturbations cause a fall. We let the cyclist rest and
+then perform another 20 perturbations. We randomize whether the balance assist
+system is on or off during the first or second set of perturbations.
 %
@@
   \centering
-  \subcaptionbox{Start}{\includegraphics[width=60mm]{figures/start.png}}
-  \subcaptionbox{During}{\includegraphics[width=60mm]{figures/during.png}}
-  \subcaptionbox{After}{\includegraphics[width=60mm]{figures/after.png}}
-  \subcaptionbox{Recovery}{\includegraphics[width=60mm]{figures/recovery.png}}
-  \caption{Thesis Fig 5.1 on pg 37}
-  \label{fig:fall}
+  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
+  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
+  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
+  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
+  \caption{Video frames depicting the application of a perturbation and the
+  rider's response and recovery.}
+  \label{fig:perturbation-sequence}
 \end{figure}
@@
 Following the initial threshold determination, we choose perturbation forces
-according to a random and adaptive staircase procedure applying
-perturbations above and below the initial perturbation threshold, while
-allowing small progression of the perturbation threshold to accommodate
-learning effects. The goal of this adaptive staircase procedure is to have
-participants fall for approximately 50\% of the time. The procedure
-consists of twenty perturbations per condition. Participants undergo two
-conditions: twenty perturbations with the balance-assist turned on and twenty
-perturbations with the balance-assist turned off (in randomized order).
+according to a random and adaptive staircase procedure applying perturbations
+above and below the initial perturbation threshold, while allowing small
+progression of the perturbation threshold to accommodate learning effects. The
+goal of this adaptive staircase procedure is to have participants fall for
+approximately 50\% of the time. The procedure consists of twenty perturbations
+per condition. Participants undergo two conditions: twenty perturbations with
+the balance-assist turned on and twenty perturbations with the balance-assist
+turned off.
 
@@
 the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
+Figure~\ref{fig:perturbation} shows an example perturbation force measurement
+compared to our Bump'Em controller command.
+%
+\begin{figure}
+  \centering
+  \includegraphics[width=140mm]{figures/perturbation_0.png}
+  \caption{Motor force and resulting  steer torque perturbation based on an 80N
+  counterclockwise applied force.}
+  \label{fig:perturbation_0}
+\end{figure}
+
 %

### 2024-05-31 src/control.py

@@ -41,8 +41,8 @@ def stable_ranges(evals):
 
 
 # FIGURE : Compare eigenvalues vs speed for uncontrolled.
-fig, ax = plt.subplots()
-fig.set_size_inches((3.0, 3.0/golden_ratio))
+fig, ax = plt.subplots(layout='compressed')
+fig.set_size_inches((70/25.4, 70/25.4/golden_ratio))
 speeds = np.linspace(0.0, 10.0, num=1001)
 evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
 weave_idx, capsize_idx = stable_ranges(evals_)[0]
@@ -58,13 +58,12 @@ ax.fill_between(speeds, -10, 10,
                 color='green', alpha=0.5, transform=ax.get_xaxis_transform())
 ax.set_ylim((-10, 10))
 ax.set_ylabel('')
-fig.tight_layout()
 fig.savefig(os.path.join(FIG_DIR,
                          'uncontrolled-eig-vs-speeds.png'),
             dpi=300)
 
-fig, ax = plt.subplots()
-fig.set_size_inches((3.0, 3.0/golden_ratio))
+fig, ax = plt.subplots(layout='compressed')
+fig.set_size_inches((70/25.4, 70/25.4/golden_ratio))
 # TODO : This finds the uncontrolled weave speed for the controller design, but
 # the actual bike uses a specific value based on the benchmark bike values
 # (probably).
@@ -74,8 +73,10 @@ evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds, kphidot=kphidots))
 for ranges in stable_ranges(evals_):
     weave_speed = speeds[ranges[0]]
     capsize_speed = speeds[ranges[1]]
-    print('Controlled (gain=-10.0) lower speed: {:1.2f} [m/s]'.format(weave_speed))
-    print('Controlled (gain=-10.0) upped speed: {:1.2f} [m/s]'.format(capsize_speed))
+    msg = 'Controlled (gain=-10.0) lower speed: {:1.2f} [m/s]'
+    print(msg.format(weave_speed))
+    msg = 'Controlled (gain=-10.0) upped speed: {:1.2f} [m/s]'
+    print(msg.format(capsize_speed))
 ax = model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
                                  colors=['k']*4)
[…]

### 2024-05-31 src/generate_time_series_imgs.py

@@ -109,7 +109,7 @@ def generate_force_torque_plots(perturbation_dfs, directory, force_column_names)
 
     for i, df in enumerate(perturbation_dfs):
         fig, axs = plt.subplots(2, 1, sharex=True, layout='constrained')
-        fig.set_size_inches(150/25.4, 100/25.4)
+        fig.set_size_inches(140/25.4, 80/25.4)
 
         df.plot(x="seconds_since_start", y=force_column_names, ax=axs[0],
                 grid=True)

### 2024-05-31 main.tex
Some todo and image name fix.

@@
 activity. Bicycles can be mechanically modified to lower the speeds at which
-they are self-stable~\citep{Astrom2005} but since the 1980s, it has been known
+they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
 that applying a motor acutated steering torque proportional to the roll rate
@@
 functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
-such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0 \quad k_{\dot{\phi}}
-\quad 0 \quad 0\right] \right)\) have negative real parts for
-\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
-the speed range where the bicycle is stable can be maximized within
-any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
+such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0
+\quad k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts
+for \(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to
+\(v\), the speed range where the bicycle is stable can be maximized within any
+physical acutator magnitude and bandwidth limits.  \citet{Schwab2008}
 elaborates on some of the possibilities in scheduling the gains for such a
@@
 %
-We modified a standard electric bicycle (Gazelle Grenoble C7, Dieren, The
-Netherlands) with a custom motor in the steering assembly capable of applying
-up to 7~\si{\newton\meter} of torque between the head tube and steer tube,
-Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
-commanded torque to applied torque. We measure the speed of the rear wheel with
-an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
-with a MEMs rate gyroscope (Company, Location). The balance assist control
-algorithm is implemented on a microprocessor (Teensy, Company, Location) and
-data from all sensors is logged with a CAN bus (Company, Location) at
-approximately X Hz.
+We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
+this is the right model}, Dieren, The Netherlands) with a custom motor in the
+steering assembly capable of applying up to 7~\si{\newton\meter} of torque
+between the head tube and steer tube, Figure~\ref{fig:balance-assist-bicycle}.
+A custom motor controller converts the commanded torque to applied torque. We
+measure the speed of the rear wheel with an encoder (Magura XXX) and measure
+the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
+Location). The balance assist control algorithm is implemented on a
+microprocessor (Teensy, Company, Location) and data from all sensors is logged
+with a CAN bus (Company, Location) at approximately X Hz.
 %
@@
 the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
-Figure~\ref{fig:perturbation} shows an example perturbation force measurement
+Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
 compared to our Bump'Em controller command.

### 2024-06-04 main.tex
Improvements to measurements section.

@@
   \subcaptionbox{}{
-    \includegraphics[width=70mm]{figures/uncontrolled-eig-vs-speeds.png}
+    \includegraphics[width=100mm]{figures/uncontrolled-eig-vs-speeds.png}
   }
   \subcaptionbox{}{
-    \includegraphics[width=70mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
+    \includegraphics[width=100mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
   }
@@
     lines attached to the ends of the handlebars.
-    \label{fig:participant-in-set-up}
+    % TODO : This label causes a compile error. I think this is where it is
+    % supposed to go.
+    %\label{fig:participant-in-set-up}
   }{
@@
     Top view diagram of bicycle handlebar perturbation system.
-    \label{fig:setup-diagram}
+    %\label{fig:setup-diagram}
   }{
-    \includegraphics[width=65mm]{figures/setup-diagram.png
+    \includegraphics[width=65mm]{figures/setup-diagram.png}
   }
-}
+  \label{fig:experimental-setup}
 \end{figure}
@@
 
-%
-\begin{table}
-  \caption{Raw measurements}
-  \begin{tabular}{llll}
-    \toprule
-    Measure & Variable & Units & Sensor \\
-    \midrule
-    Perturbation Order Number & \(j\) & integer & NA\\
-    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
-    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
-    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
-    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
-    Balance Assist Gain & \(k\) & TODO & NA \\
-    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Based on the findings in \cite{Haitjema2023a}, we calculate several variables.
-We use the angular impulse \(L\) due to the perturbation forces over a
-0.3~\si{\second} duration to characterize the magnitude of delivered
-perturbation. The duration is selected based on the duration of the commanded
-force.
+Based on findings from measuring riders without balance
+assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
+may influence fall probability. We use the angular impulse \(L\) of the
+perturbation forces over a 0.3~\si{\second} duration to characterize the
+magnitude of delivered perturbation. The duration is selected based on the
+duration of the commanded force and is calculated as follows:
 
@@
   L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
+  \textrm{.}
 \end{align}
 
-We characterize the skill of the rider by calculating the standard deviation of
-the steer angle \(\sigma_\delta\) over the 90~\si{\second} period they were
-asked to maintain their position at the center of the treadmill. At the
-initiation of each perturbation we log the steer and roll angles to characterize
-the configuration of the bicycle when perturbed. The gain setting on the
-balance assist controller indicates if the assistance is off \(k=0\) or on at
-two different levels: low \(k=8\) or high \(k=10\). A recovery from the
+At the initiation of each perturbation we log the steer and roll angles to
+characterize the configuration of the bicycle when perturbed. The gain setting
+on the balance assist controller indicates if the assistance is off \(k=0\) or
+on at two different levels: low \(k=8\) or high \(k=10\). A recovery from the
 perturbation is successful if the person neither put their foot down onto the
 treadmill surface or either wheel of the bicycle exits the width of the
-treadmill belt. We record this as a binary variable \(f\).
-
-\begin{figure}
+treadmill belt. We record this as a binary variable \(f\). All measured
+variables are reported in Table~\ref{tab:raw}.
+%
+\begin{table}
   \centering
-  \includegraphics[width=\columnwidth]{figures/data-processing.png}
-  \caption{Thesis Fig 3.13 on pg 27}
-\end{figure}
+  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
+    ``Perturbation order number'' are recorded per perturbation instance.}
+  \begin{tabular}{llll}
+    \toprule
+    Measure & Variable & Units & Sensor \\
+    \midrule
+    Balance Assist Gain & \(k\) & TODO & NA \\
+    Bicycle Speed & \(v\) & \si{\meters\per\second} & wheel encoder \\
+    Fall outcome & \(f\) & boolean & NA \\
+    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
+    Perturbation Order Number & \(j\) & integer & NA\\
+    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\cite{Gabriel2023} \\
+    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
+    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
+    \bottomrule
+  \end{tabular}
+  \label{tab:raw}
+\end{table}
 
@@
 \begin{table}
+  \small
   \begin{tabular}{lll}
+    \toprule
+    \midrule
     \(L\) & \si{\newton\meter\second} & angular impulse of the applied
@@
     \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
+    \bottomrule
   \end{tabular}
@@
 \begin{figure}
+  \centering
   \includegraphics{example-image-a}

### 2024-06-04 src/control.py

@@ -42,7 +42,7 @@ def stable_ranges(evals):
 
 # FIGURE : Compare eigenvalues vs speed for uncontrolled.
 fig, ax = plt.subplots(layout='compressed')
-fig.set_size_inches((70/25.4, 70/25.4/golden_ratio))
+fig.set_size_inches((100/25.4, 100/25.4/golden_ratio))
 speeds = np.linspace(0.0, 10.0, num=1001)
 evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds))
 weave_idx, capsize_idx = stable_ranges(evals_)[0]
@@ -57,13 +57,14 @@ ax.fill_between(speeds, -10, 10,
                 where=np.all(evals_ < 0.0, axis=1),
                 color='green', alpha=0.5, transform=ax.get_xaxis_transform())
 ax.set_ylim((-10, 10))
-ax.set_ylabel('')
+ax.set_ylabel('Eigenvalue Components [1/s]')
+ax.set_xlabel('Speed [m/s]')
 fig.savefig(os.path.join(FIG_DIR,
                          'uncontrolled-eig-vs-speeds.png'),
             dpi=300)
 
 fig, ax = plt.subplots(layout='compressed')
-fig.set_size_inches((70/25.4, 70/25.4/golden_ratio))
+fig.set_size_inches((100/25.4, 100/25.4/golden_ratio))
 # TODO : This finds the uncontrolled weave speed for the controller design, but
 # the actual bike uses a specific value based on the benchmark bike values
 # (probably).
@@ -85,7 +86,8 @@ ax.fill_between(speeds, -10, 10,
                 where=np.all(evals_ < 0.0, axis=1),
                 color='green', alpha=0.5, transform=ax.get_xaxis_transform())
 ax.set_ylim((-10, 10))
-ax.set_ylabel('')
+ax.set_ylabel('Eigenvalue Components [1/s]')
+ax.set_xlabel('Speed [m/s]')
 fig.savefig(os.path.join(FIG_DIR,
                          'balance-assist-controllers-eig-vs-speeds.png'),
             dpi=300)

### 2024-06-04 main.tex
Fixed unit.

@@
     Balance Assist Gain & \(k\) & TODO & NA \\
-    Bicycle Speed & \(v\) & \si{\meters\per\second} & wheel encoder \\
+    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
     Fall outcome & \(f\) & boolean & NA \\

### 2024-06-04 main.tex
Combined the two 6 km/h stats tables.

@@
 \def\kph{\kilo\meter\per\hour}
+\def\mps{\meter\per\second}
 
@@
 \begin{table}
-  \small
+  \centering
+  \caption{Statisitcal model variables.}
   \begin{tabular}{lll}
     \toprule
+    Variable & Units & Description \\
     \midrule
-    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
-    perturbation torque \\
+    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
     \(c\) & integer & order number of perturbation \\
-    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
     \(k\) & TODO & gain of balance assist control \\
@@
     \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
-    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
+    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
     \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
@@
 
-We evaluate this hypothesis using a multivariate logistic regression model.
+We evaluate this hypothesis using a multivariate logistic regression model in
+the form
 
 \begin{align}
-  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij})
+  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
 \end{align}
 
-\(f_{ij}\) is the binary outcome of perturbation \(j\) of participant \(i\)
-which follows a Bernoulli distribution given the probability \(p_{ij}\) that a
-fall occurs. The log-odds of the probabiliy is then a linear function of our
-independent variables with \(\beta\) as the intercept and \(\alpha_k\) as the
-linear coefficients.
+Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) of
+participant \(i\) which follows a Bernoulli distribution given the probability
+\(p_{ij}\) that a fall occurs. The log-odds of the probabiliy is then a linear
+function of our independent variables with \(\beta\) as the intercept and
+\(\alpha_k\) as the linear coefficients.
 
@@
 
-We scalle all independent variables such they they have a mean of zero and a
+We scale all independent variables such they they have a mean of zero and a
 standard deviation of one. We use cluster-mean centering, as recommended by
@@
 
-Cluster-mean centering shows there to be no variation between participants,
-i.e. that rider skill (the only participant differentiating factor) is not a
-predictor. We expected there to be a variation between participants in how well
-they are able to resist the perturbation. However, this was not true. This
-allowed us to use a simpler single-level logistic regression model, instead of
-a multilevel model. This left use with angular impulse, perturbation order,
-balance assist state, and roll \& steer angles at the time of perturbation. We
-also include interaction effects between the balance assist state and the other
-four variables.
-
-\todo[inline]{Marten writes ``The skill variable is not included, because it
-was not collected correctly for all participants.'' This needs to be addressed.
-What about skill being tied to participant clusters?}
+Cluster-mean centering shows there to be no variation between participants. We
+expected there to be a variation between participants in how well they are able
+to resist the perturbation. However, this was not true. This allowed us to
+stick with a simple single-level logistic regression model, instead of a
+multilevel model. This left use with angular impulse, perturbation order,
+balance assist state, and roll \& steer angles at the time of perturbation as
+independen tvariables.. We also include interaction effects between the balance
+assist state and the other four variables.
 
@@
 The coefficient estimates for a single-level logistic regression at
-6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in Table
-\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and balance
-assist state are all statistically significant predictors. Larger angular
-impulse increases the probability to fall and enduring more perturbations or
-having the balance assist on, decrease the probability to fall.  The associated
-multiplicative change in odds are shown in
+6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in
+Table~\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and
+balance assist state are all statistically significant predictors. Larger
+angular impulse increases the probability to fall and enduring more
+perturbations or having the balance assist on, decrease the probability to
+fall.  The associated multiplicative change in odds are shown in
 Table~\ref{table:freq-change-in-odds-6}.
+
+\todo{Combine tables 3 and 4 into one}
 
@@
   \caption{Logistic regression coefficient estimates at
-    6~\unit{\kilo\meter\per\hour} and gain $k=10$.}
-  \label{table:freq-coefs-6}
-  \begin{tabular}{lrrr}
-    \toprule
-    Variable                                      & Estimate          & Standard error          & p-value          \\
-    \midrule
-    Intercept                                     & -0.29             & 0.17                    & 0.09             \\
-    Angular impulse                               & 1.69              & 0.27                    & 0.00             \\
-    Perturbation order                            & -0.77             & 0.22                    & 0.00             \\
-    Balance assist state                          & -0.64             & 0.27                    & 0.02             \\
-    Roll angle                                    & -0.25             & 0.21                    & 0.24             \\
-    Steer angle                                   & -0.14             & 0.21                    & 0.51             \\
-    Balance-assist on $\times$ roll angle         & 0.52              & 0.34                    & 0.12             \\
-    Balance-assist on $\times$ steer angle        & -0.41             & 0.34                    & 0.22             \\
-    Balance-assist on $\times$ angular impulse    & 0.41              & 0.41                    & 0.32             \\
-    Balance-assist on $\times$ perturbation order & -0.53             & 0.34                    & 0.12             \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\todo[inline]{Combine the columns of the multiplicative change in odds into the
-coefficient table}
-
-\begin{table}
-  \centering
-  \caption{Multiplicative change in odds of predictor variables at 6
+    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in odds of predictor variables at 6
   \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
          single-level logistic regression.}
-  \label{table:freq-change-in-odds-6}
-  \begin{tabular}{lllll}
-      \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
-      Intercept                                     & 0.75                                   & 0.53           & 1.05            & 0.09             \\
-      Angular impulse                               & 5.40                                   & 3.18           & 9.16            & 0.00             \\
-      Perturbation order                            & 0.46                                   & 0.30           & 0.72            & 0.00             \\
-      Balance-assist on                             & 0.53                                   & 0.31           & 0.89            & 0.02             \\
-      Roll angle                                    & 0.78                                   & 0.51           & 1.18            & 0.24             \\
-      Steer angle                                   & 0.87                                   & 0.58           & 1.32            & 0.51             \\
-      Balance-assist on $\times$ roll angle         & 1.68                                   & 0.86           & 3.29            & 0.12             \\
-      Balance-assist on $\times$ steer angle        & 0.66                                   & 0.34           & 1.28            & 0.22             \\
-      Balance-assist on $\times$ angular impulse    & 1.51                                   & 0.67           & 3.38            & 0.32             \\
-      Balance-assist on $\times$ perturbation order & 0.59                                   & 0.30           & 1.15            & 0.12             \\
+  \label{table:freq-coefs-6}
+  \begin{tabular}{lrrrrrr}
+    \toprule
+    Variable & Estimate & Standard error & p-value & MCIO & 2.5\% & 97.5\% \\
+    \midrule
+    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
+    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
+    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
+    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
+    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
+    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
+    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
+    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
+    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
+    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
+    \bottomrule
   \end{tabular}
@@
 Table~\ref{table:freq-change-in-odds}.
+
+\todo{Combine tables 4 and 5 into one}
 
@@
   \centering
-  \caption{Multiplicative change in odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
-      frequentist single-level logistic regression.} \label{table:freq-change-in-odds}
+  \caption{Multiplicative change in odds of predictor variables at 10
+    \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
+    regression.}
+  \label{table:freq-change-in-odds}
   \begin{tabular}{lllll}
-    \textbf{Variable}                             & \textbf{Multiplicative change in odds} & \textbf{2.5\%} & \textbf{97.5\%} & \textbf{p-value} \\ \hline
+    \toprule
+    Variable & Multiplicative change in odds & 2.5\% & 97.5\% & p-value \\ \hline
+    \midrule
     Intercept & 0.78 & 0.57 & 1.07 & 0.12 \\
@@
     Balance-assist on $\times$ perturbation order & 0.69 & 0.37 & 1.30 & 0.25 \\
+    \bottomrule
   \end{tabular}
@@
 Turning the balance-assist system on significantly \((p<0.05)\) reduces the
-odds that a fall occurs while cycling at a speed of 6~\si{\kph} but at
-10~\si{\kph} that is not the case \((p=0.07)\). If all other factors remain the
-same at 6~\si{\kph}, the balance-assist system halves the odds that a
+odds that a fall occurs while cycling at a speed of 1.7~\si{\mps} but at
+2.8~\si{\mps} that is not the case \((p=0.07)\). If all other factors remain
+the same at 1.7~\si{\mps}, the balance-assist system halves the odds that a
 perturbation results in a fall. How much this changes the probability that a
@@
   \centering
-  \subcaptionbox{6~\si{\kph}}{
+  \subcaptionbox{1.7~\si{\mps}}{
     \includegraphics[width=70mm]{example-image-a}
-    %\includegraphics[width=\textwidth]{figures/predicted_fall_probability-6.png}
   }
-  \subcaptionbox{10~\si{\kph}}{
+  \subcaptionbox{2.8~\si{\mps}}{
     \includegraphics[width=70mm]{example-image-b}

### 2024-06-04 main.tex
Combined second stats table.

@@
 
-In the early years of developments automatic control, \citep{Whipple1899}
+In the early years of developments in automatic control, \citep{Whipple1899}
 remarkably realized and showed that roll motion feedback can stabilize the a
@@
 mode, and steer angular rate feedback stabilizes the wobble mode. They also
-propose how the gains must change with respect to vehicle speed for favorable
+proposed how the gains must change with respect to vehicle speed for favorable
 control at all speeds. This roll motion feedback is the simplest controller
 that can stabilize a single track vehicle above a minimum speed when one is not
-concerned with wobble instabilities, which are not prominent in bicycles at low
+concerned with wobble instabilities, somethign not prominent in bicycles at low
 speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
@@
 The effect of this automatic control gives the ability to change the dynamics,
-up to some limits, of the human controlled plant. In our prior study, we
-demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
-the balance assist but conversely researchers with a similar vehicle recently
-showed rider disatisfction with the stabilization~\cite{Hanakam2023}.
+up to some limits, of the human controlled plant. In our prior
+study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
+distractions due to the balance assist but contrarily researchers with a
+similar vehicle recently showed rider disatisfction with the
+stabilization~\cite{Hanakam2023}.
 
@@
   \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
+  \textrm{.}
 \end{align}
-
+%
 The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
@@
 
-\todo{Combine tables 3 and 4 into one}
-
 \begin{table}
@@
   \caption{Logistic regression coefficient estimates at
-    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in odds of predictor variables at 6
-  \unit{\kilo\meter\per\hour} and gain $k=10$ based on frequentist
-         single-level logistic regression.}
+    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in
+    odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
+    $k=10$ based on frequentist single-level logistic regression.}
   \label{table:freq-coefs-6}
@@
     \toprule
-    Variable & Estimate & Standard error & p-value & MCIO & 2.5\% & 97.5\% \\
+    Variable & Estimate & Standard error & \(p\)-value & MCiO & 2.5\% & 97.5\% \\
     \midrule
@@
 
-\todo{Combine tables 4 and 5 into one}
-
 \begin{table}
@@
   \caption{Logistic regression coefficient estimates at
-    10~\unit{\kilo\meter\per\hour} and gain $k=8$.}
+    10~\unit{\kilo\meter\per\hour} and gain $k=8$.  Multiplicative change in
+    odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
+    frequentist single-level logistic regression.}
   \label{table:freq-coefs}
-  \begin{tabular}{lrrr}
+  \begin{tabular}{lrrrrrr}
     \toprule
-    Variable                                      & Estimate          & Standard error          & p-value \\
+    Variable & Estimate & Standard error & \(p\)-value & MCiO & 2.5\% & 97.5\% \\
     \midrule
-    Intercept                                     & -0.24             & 0.16                    & 0.13    \\
-    Angular impulse                               & 2.39              & 0.29                    & 0.00    \\
-    Perturbation order                            & -1.16             & 0.21                    & 0.00    \\
-    Balance-assist on                             & -0.44             & 0.24                    & 0.07    \\
-    Roll angle                                    & 0.27              & 0.22                    & 0.22    \\
-    Steer angle                                   & -0.37             & 0.24                    & 0.12    \\
-    Balance-assist on $\times$ roll angle         & -0.61             & 0.34                    & 0.08    \\
-    Balance-assist on $\times$ steer angle        & 0.56              & 0.35                    & 0.11    \\
-    Balance-assist on $\times$ angular impulse    & 0.46              & 0.44                    & 0.29    \\
-    Balance-assist on $\times$ perturbation order & -0.37             & 0.32                    & 0.25    \\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-\begin{table}
-  \centering
-  \caption{Multiplicative change in odds of predictor variables at 10
-    \unit{\kilo\meter\per\hour} based on frequentist single-level logistic
-    regression.}
-  \label{table:freq-change-in-odds}
-  \begin{tabular}{lllll}
-    \toprule
-    Variable & Multiplicative change in odds & 2.5\% & 97.5\% & p-value \\ \hline
-    \midrule
-    Intercept & 0.78 & 0.57 & 1.07 & 0.12 \\
-    Angular impulse & 10.92 & 6.23 & 19.13 & 0.00 \\
-    Perturbation order & 0.31 & 0.21 & 0.48 & 0.00 \\
-    Balance-assist on & 0.64 & 0.40 & 1.03 & 0.07 \\
-    Roll angle & 1.31 & 0.85 & 2.04 & 0.22 \\
-    Steer angle & 0.69 & 0.43 & 1.10 & 0.12 \\
-    Balance-assist on $\times$ roll angle & 0.54 & 0.28 & 1.07 & 0.08 \\
-    Balance-assist on $\times$ steer angle & 1.76 & 0.88 & 3.50 & 0.11 \\
-    Balance-assist on $\times$ angular impulse & 1.59 & 0.68 & 3.74 & 0.29 \\
-    Balance-assist on $\times$ perturbation order & 0.69 & 0.37 & 1.30 & 0.25 \\
+    Intercept                                     & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
+    Angular impulse                               & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
+    Perturbation order                            & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
+    Balance-assist on                             & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
+    Roll angle                                    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
+    Steer angle                                   & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
+    Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
+    Balance-assist on $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
+    Balance-assist on $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
+    Balance-assist on $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
     \bottomrule

### 2024-06-04 main.tex
Try with labels in subcaptions.

@@
   % TODO : scale these pngs to a reasonable size for an A4 sheet
-  \subcaptionbox{
-    A participant on the bicycle in the safety harness with the Bump'Em motor
-    lines attached to the ends of the handlebars.
-    % TODO : This label causes a compile error. I think this is where it is
-    % supposed to go.
-    %\label{fig:participant-in-set-up}
-  }{
-    \includegraphics[width=75mm]{figures/participant-in-set-up.png}
-  }
-  \subcaptionbox{
-    Top view diagram of bicycle handlebar perturbation system.
-    %\label{fig:setup-diagram}
-  }{
-    \includegraphics[width=65mm]{figures/setup-diagram.png}
-  }
+  \subcaptionbox{A participant on the bicycle in the safety harness with the
+    Bump'Em motor lines attached to the ends of the
+    handlebars.\label{fig:participant-in-set-up}}
+    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
+  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
+    system.\label{fig:setup-diagram}}
+    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
   \label{fig:experimental-setup}
@@
 fall.  The associated multiplicative change in odds are shown in
-Table~\ref{table:freq-change-in-odds-6}.
+Table~\ref{table:freq-coefs-6}.
 
@@
 to fall. The associated multiplicative change in odds are shown in
-Table~\ref{table:freq-change-in-odds}.
+Table~\ref{table:freq-coefs}.
 

### 2024-06-04 main.tex
Fixed compile error related to unprotected label.

@@
 %
+% NOTE : I get a compile error with the \protect calls here, not sure why.
 \begin{figure}
   \centering
-  % TODO : scale these pngs to a reasonable size for an A4 sheet
   \subcaptionbox{A participant on the bicycle in the safety harness with the
     Bump'Em motor lines attached to the ends of the
-    handlebars.\label{fig:participant-in-set-up}}
+    handlebars.\protect\label{fig:participant-in-set-up}}
     {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
   \subcaptionbox{Top view diagram of bicycle handlebar perturbation
-    system.\label{fig:setup-diagram}}
+    system.\protect\label{fig:setup-diagram}}
     {\includegraphics[width=65mm]{figures/setup-diagram.png}}
+  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
+    perturbations.}
   \label{fig:experimental-setup}

### 2024-06-05 main.tex
Move figure, spelling, results text.

@@
 %
-Single-actor bicycle crashes are associated with a suprisingly large number of
+Single-actor bicycle crashes are associated with a surprisingly large number of
 reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
@@
 they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
-that applying a motor acutated steering torque proportional to the roll rate
+that applying a motor actuated steering torque proportional to the roll rate
 can stabilize a single track vehicle at very low speeds. If this automatic
@@
 balance assisting bicycle and hypothesize that it helps the rider in situations
-they are likely to fall.
+they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
+%
+\begin{figure}
+  \centering
+  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
+  \caption{Balance assist bicycle prototype with steer motor and data
+    acquisition and control electronics mounted in the rear rack.}
+  \label{fig:balance-assist-bicycle}
+\end{figure}
 
 In the early years of developments in automatic control, \citep{Whipple1899}
-remarkably realized and showed that roll motion feedback can stabilize the a
+remarkably realized and showed that roll motion feedback can stabilize a
 bicycle. Much later, automatic roll stabilization of single track vehicles
@@
 and angular rate was essential to stabilize the vehicle. It was not until the
-early 1980s that \cite{Nagai1983} sucessfully demonstrated balancing a robotic
+early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
 bicycle on a treadmill with both steering control and a laterally moving mass.
@@
 that can stabilize a single track vehicle above a minimum speed when one is not
-concerned with wobble instabilities, somethign not prominent in bicycles at low
+concerned with wobble instabilities, something not prominent in bicycles at low
 speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
@@
 distractions due to the balance assist but contrarily researchers with a
-similar vehicle recently showed rider disatisfction with the
+similar vehicle recently showed rider dissatisfaction with the
 stabilization~\cite{Hanakam2023}.
@@
 the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stablity. The bicycle model is ideal for showing the effect of a roll
+self-stability. The bicycle model is ideal for showing the effect of a roll
 motion feedback driven steer motor on the dynamics. The linear model can be
@@
 \(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
-functions of the equilbirum forward speed \(v\) and  are otherwise populated
+functions of the equilibrium forward speed \(v\) and  are otherwise populated
 with expressions that are functions of the geometric and inertial parameters of
-the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
+the nonholonomic multibody system made up of four rigid bodies: two wheels,
 front frame, and rear frame.
@@
 \(v\), the speed range where the bicycle is stable can be maximized within any
-physical acutator magnitude and bandwidth limits.  \citet{Schwab2008}
+physical actuator magnitude and bandwidth limits.  \citet{Schwab2008}
 elaborates on some of the possibilities in scheduling the gains for such a
@@
 system on and off.
+
+\todo{Talk about the types of real life perturbations an how our perturbation
+mimics it.}
 
@@
 with a CAN bus (Company, Location) at approximately X Hz.
-%
-\begin{figure}
-  \centering
-  \includegraphics[width=100mm]{figures/balance-assist-bicycle.jpg}
-  \caption{Balance assist bicycle prototype with steer motor and data
-    aquisition and control electronics mounted in the rear rack.}
-  \label{fig:balance-assist-bicycle}
-\end{figure}
 
@@
 18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
-Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangment
+Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangement
 is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
@@
 group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
-(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occured X
+(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred X
 weeks before the 6~\si{\kilo\meter\per\hour} group.
@@
 counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
-Figure~\ref{fig:perturbation-sequence} shows an example reult of a
+Figure~\ref{fig:perturbation-sequence} shows an example result of a
 perturbation.  We log the magnitude that causes the first fall to characterize
@@
 \begin{align}
-  F_r = & F_{rf} - F_{rr} \\
-  F_l = & F_{lf} - F_{lr} \\
-  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
+  L =
+  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
+  =
+  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
+  - F_{lr})\right] dt
   \textrm{.}
@@
 probability of falling when perturbed externally at the handlebar. We have a
-single binary outcome variable \(f\) that is dependent on several possible
-explanatory independent variables including the binary balance assistance (on
-or off).
-
+single binary fall outcome variable \(f\) that is dependent on several possible
+explanatory independent variables, one of which is the binary balance
+assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
+variables.
+%
 \begin{table}
   \centering
-  \caption{Statisitcal model variables.}
+  \caption{Statistical model variables.}
+  \label{tab:stat-model-variables}
   \begin{tabular}{lll}
@@
 
-We evaluate this hypothesis using a multivariate logistic regression model in
-the form
-
+We evaluate this hypothesis using a multivariate logistic regression model that
+takes the general form
+%
 \begin{align}
@@
 \end{align}
-
-Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) of
+%
+Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
 participant \(i\) which follows a Bernoulli distribution given the probability
-\(p_{ij}\) that a fall occurs. The log-odds of the probabiliy is then a linear
+\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
 function of our independent variables with \(\beta\) as the intercept and
-\(\alpha_k\) as the linear coefficients.
-
+\(\alpha_k\) as the linear coefficients to the independent variables
+\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:statmodel-variables} except
+\(f\).
+%
 \begin{align}
@@
 
-We scale all independent variables such they they have a mean of zero and a
-standard deviation of one. We use cluster-mean centering, as recommended by
-\cite{Enders2007}, with the clusters being an individual subject because we are
+Before fitting the model, we scale each \(x_{ij}^k\) such they they have a mean
+of zero and a standard deviation of one by cluster-mean centering, as
+recommended by \cite{Enders2007}, with clusters being an individual subject.
+The clusters are all data from an individual subject because we are only
 interested in the association between the state of the balance-assist system
-and the outcome of the perturbation.
+and the outcome of the perturbation. We expected there to be a variation
+between participants in how well they are able to resist the perturbation.
+However, this was not true.  Cluster-mean centering shows there to be no
+variation between participants. This allowed us to stick with a simple
+single-level logistic regression model, instead of a multilevel model. This
+left us with angular impulse, perturbation order, balance assist state, and
+roll \& steer angles at the time of perturbation as independent variables. We
+also include interaction effects between the balance assist state and the other
+four variables.
 
-Cluster-mean centering shows there to be no variation between participants. We
-expected there to be a variation between participants in how well they are able
-to resist the perturbation. However, this was not true. This allowed us to
-stick with a simple single-level logistic regression model, instead of a
-multilevel model. This left use with angular impulse, perturbation order,
-balance assist state, and roll \& steer angles at the time of perturbation as
-independen tvariables.. We also include interaction effects between the balance
-assist state and the other four variables.
-
-We divide the analysis into two separate model evalautions, one for the
+We divide the analysis into two separate model evaluations, one for the
 6~\si{\kilo\meter\per\hour} \(k=10\) trials and one for the
@@
   \includegraphics{example-image-a}
-  \caption{Box plot or distrubtion plot showing graphical view of stats
+  \caption{Box plot or distribution plot showing graphical view of stats
   tables.}
@@
 the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
-could often recover in the alloted treadmill width. Our results are very much
+could often recover in the allotted treadmill width. Our results are very much
 dependent on the two modes of falling with use: exit the treadmill width or
 foot is placed on the belt. Cycle paths are a similar width as the treadmill,
-so ride's are often limited in width when recoverying from a fall.
+so ride's are often limited in width when recovering from a fall.
 

### 2024-06-05 main.tex
Edits to last parts of the intro for better wording.

@@
   \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
-  \caption{Balance assist bicycle prototype with steer motor and data
-    acquisition and control electronics mounted in the rear rack.}
+  \caption{Balance assist bicycle prototype with electric motor in the steering
+    column and data acquisition and control electronics mounted in the rear
+    rack.}
   \label{fig:balance-assist-bicycle}
@@
 applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
-\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
+\(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
 \end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
@@
   \dot{\vec{x}} = \left(\mathbf{A} -
-  \mathbf{B}_\delta
+  \vec{B}_\delta
   \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
@@
 functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
-such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0
-\quad k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts
-for \(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to
-\(v\), the speed range where the bicycle is stable can be maximized within any
-physical actuator magnitude and bandwidth limits.  \citet{Schwab2008}
-elaborates on some of the possibilities in scheduling the gains for such a
-controller on this model.
-
-\todo[inline]{I'm not consistent with the arrow being a vector for $B_\delta$}
+such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
+k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
+\(v_{min} < v < v_{capsize}\). With gain scheduling with respect to \(v\), the
+speed range where the bicycle is stable can be maximized within any physical
+actuator magnitude and bandwidth limits. \citet{Schwab2008} elaborates on some
+of the possibilities in scheduling the gains for such a controller and shows
+that a linear scheduling with respect to speed can give satisfactory stability.
 
 In this paper, we test whether a gain scheduled steer motor controlled bicycle
-that is stable at a low speed range when riderless assists in balancing by
-investigating whether it is beneficial in preventing the rider from falling.
-To do this, we apply varying magnitude mechanical perturbations to the
-handlebars while the rider is balancing the bicycle on a treadmill and then
-assess the rider's probability of falling with the balance assist steer motor
-system on and off.
+that is stable in a large low speed range (when riderless) assists the rider in
+balancing by investigating whether it is beneficial in preventing falling.  To
+do this, we apply varying magnitude mechanical perturbations to the handlebars
+while the rider is balancing the bicycle on a treadmill and then assess the
+rider's probability of falling with the balance assist steer motor system on
+and off.
 

### 2024-06-05 main.tex
Intro improvements and fixed some units.

@@
 they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
+\todo{Talk about the types of real life perturbations an how our perturbation
+mimics it.}
 %
@@
 In the early years of developments in automatic control, \citep{Whipple1899}
-remarkably realized and showed that roll motion feedback can stabilize a
-bicycle. Much later, automatic roll stabilization of single track vehicles
-began in earnest after predictive motorcycle models were developed and refined
-in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
-work and seems to be the first to attempt to robotically stabilize a small
-motorbike with a controlled inverted pendulum that mimicked rider lean, but he
-was not successful in demonstrating what his theoretic control model correctly
+not only derived the correct equations of motion of the bicycle but realized
+and showed that roll motion feedback can stabilize a bicycle. Much later,
+automatic roll stabilization of single track vehicles began in earnest after
+predictive motorcycle models were developed and refined in throughout the
+1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
+the first to attempt to robotically stabilize a small motorbike with a
+controlled inverted pendulum that mimicked rider lean, but he was not
+successful in demonstrating what his theoretic control model correctly
 predicted. In his model, he recognized that feedback of the vehicle roll angle
@@
 described by the fourth order state space equations:
-
+%
 \begin{align}
@@
 k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
-\(v_{min} < v < v_{capsize}\). With gain scheduling with respect to \(v\), the
-speed range where the bicycle is stable can be maximized within any physical
-actuator magnitude and bandwidth limits. \citet{Schwab2008} elaborates on some
-of the possibilities in scheduling the gains for such a controller and shows
-that a linear scheduling with respect to speed can give satisfactory stability.
+\(v_{min} < v < v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable
+speed given \(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which
+the uncontrolled bicycle's capsize mode goes unstable. With gain scheduling
+with respect to \(v\), the speed range where the bicycle is stable can be
+maximized within any physical actuator magnitude and bandwidth limits.
+\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
+gains for such a controller and shows that a linear scheduling with respect to
+speed can give satisfactory stability.
 
-In this paper, we test whether a gain scheduled steer motor controlled bicycle
-that is stable in a large low speed range (when riderless) assists the rider in
-balancing by investigating whether it is beneficial in preventing falling.  To
-do this, we apply varying magnitude mechanical perturbations to the handlebars
-while the rider is balancing the bicycle on a treadmill and then assess the
-rider's probability of falling with the balance assist steer motor system on
-and off.
-
-\todo{Talk about the types of real life perturbations an how our perturbation
-mimics it.}
+In this paper, we test whether a linearly gain scheduled steer motor controlled
+bicycle that is stable in a large low speed range (when riderless) assists the
+rider in balancing by investigating whether it is beneficial in preventing
+falling.  To do this, we apply varying magnitude mechanical perturbations to
+the handlebars while the rider is balancing the bicycle on a treadmill and then
+assess the rider's probability of falling with the balance assist steer motor
+system on and off.
 
@@
 steering assembly capable of applying up to 7~\si{\newton\meter} of torque
-between the head tube and steer tube, Figure~\ref{fig:balance-assist-bicycle}.
-A custom motor controller converts the commanded torque to applied torque. We
-measure the speed of the rear wheel with an encoder (Magura XXX) and measure
-the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
-Location). The balance assist control algorithm is implemented on a
-microprocessor (Teensy, Company, Location) and data from all sensors is logged
-with a CAN bus (Company, Location) at approximately X Hz.
+between the head tube and steer tube, see
+Figure~\ref{fig:balance-assist-bicycle}.  A custom motor controller converts
+the commanded torque to applied torque. We measure the speed of the rear wheel
+with an encoder (Magura XXX) and measure the body fixed roll rate of the
+bicycle with a MEMs rate gyroscope (Company, Location). The balance assist
+control algorithm is implemented on a microprocessor (Teensy, Company,
+Location) and data from all sensors is logged with a CAN bus (Company,
+Location) at approximately X Hz.
 
 \subsection{Balance Assist Control}
+%
 We use a forward speed \(v\) gain scheduled proportional roll rate feedback
@@
 \(T^\textrm{m}_\delta\) from the steer motor follows the control law
+%
 \begin{align}
@@
 \end{align}
+%
 where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
@@
 
-\subsection{Subjects}
+\subsection{Protocol}
 %
@@
 
-\subsection{Protocol}
-%
 The subjects were divided into two groups. The first group performed the
@@
 example, if the initial estimate of the perturbation threshold of a participant
-is 80 N, the five possible perturbations are 60, 70, 80, 90 and 100 N. Which
-one of these five perturbations is chosen is determined at random. If the
-perturbation results in a fall, the estimate of the perturbation threshold is
-decreased with 10 N, and vice versa if the perturbation did not result in a
-fall. Five new possible perturbation forces are determined around the updated
-perturbation threshold, and a new perturbation is selected at random. This
-process iterates until twenty perturbations are applied.
+is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
+100~\si{\newton}. Which one of these five perturbations is chosen is
+determined at random. If the perturbation results in a fall, the estimate of
+the perturbation threshold is decreased with 10~\si{\newton}, and vice versa
+if the perturbation did not result in a fall. Five new possible perturbation
+forces are determined around the updated perturbation threshold, and a new
+perturbation is selected at random. This process iterates until twenty
+perturbations are applied.
 
@@
     Perturbation Order Number & \(j\) & integer & NA\\
-    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\cite{Gabriel2023} \\
+    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\citep{Gabriel2023} \\
     Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
@@
 \(\alpha_k\) as the linear coefficients to the independent variables
-\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:statmodel-variables} except
+\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
 \(f\).

### 2024-06-06 main.tex
Define our k_phi.

@@
 \begin{align}
-  T^\textrm{m}_\delta = k(v_{weave} - v)\dot{\phi}
+  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = k(v_{weave} - v)\dot{\phi}
 \end{align}
@@
 the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
-Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
-compared to our Bump'Em controller command.
+Figure~\ref{fig:perturbation_0}\todo{Maybe remove the top plot of individual
+forces and add plots of roll angle, roll rate, steer rate.} shows an example
+perturbation force measurement compared to our Bump'Em controller command.
 %

### 2024-06-06 main.tex
Change gain from k to g.

@@
 and showed that roll motion feedback can stabilize a bicycle. Much later,
-automatic roll stabilization of single track vehicles began in earnest after
-predictive motorcycle models were developed and refined in throughout the
-1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
-the first to attempt to robotically stabilize a small motorbike with a
-controlled inverted pendulum that mimicked rider lean, but he was not
+attempts at automatic roll stabilization of single track vehicles began in
+earnest after predictive motorcycle models were developed and refined in
+throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
+and seems to be the first to attempt to robotically stabilize a small motorbike
+with a controlled inverted pendulum that mimicked rider lean, but he was not
 successful in demonstrating what his theoretic control model correctly
@@
 \begin{align}
-  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = k(v_{weave} - v)\dot{\phi}
+  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{weave} - v)\dot{\phi}
 \end{align}
@@
 where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
-open loop bicycle rigid-rider dynamics. We use \(k=8\) and \(k=10\) as gain
+open loop bicycle rigid-rider dynamics. We use \(g=8\) and \(g=10\) as gain
 values during the experiments. Scaling the proportional feedback gain linearly
@@
 protocol at a belt speed of 2.8~\si{\meter\per\second}
-(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
+(10~\si{\kilo\meter\per\hour}) with the controller gain at $g=8$ and the second
 group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
@@
 characterize the configuration of the bicycle when perturbed. The gain setting
-on the balance assist controller indicates if the assistance is off \(k=0\) or
-on at two different levels: low \(k=8\) or high \(k=10\). A recovery from the
+on the balance assist controller indicates if the assistance is off \(g=0\) or
+on at two different levels: low \(g=8\) or high \(g=10\). A recovery from the
 perturbation is successful if the person neither put their foot down onto the
@@
 \begin{align}
-  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
+  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) =
+  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
 \end{align}
-
-\todo[inline]{we already use k for gain}
 
@@
 We divide the analysis into two separate model evaluations, one for the
-6~\si{\kilo\meter\per\hour} \(k=10\) trials and one for the
-10~\si{\kilo\meter\per\hour} \(k=8\) trials and we evaluate our hypothesis for
+6~\si{\kph} \(g=10\) trials and one for the
+10~\si{\kph} \(g=8\) trials and we evaluate our hypothesis for
 each set of data.
@@
 \section{Results}
-
+%
 The coefficient estimates for a single-level logistic regression at
-6~\unit{\kilo\meter\per\hour} with gain $k=10$ are shown in
+6~\unit{\kph} with gain $g=10$ are shown in
 Table~\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and
@@
   \caption{Logistic regression coefficient estimates at
-    6~\unit{\kilo\meter\per\hour} and gain $k=10$. Multiplicative change in
+    6~\unit{\kilo\meter\per\hour} and gain $g=10$. Multiplicative change in
     odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
-    $k=10$ based on frequentist single-level logistic regression.}
+    $g=10$ based on frequentist single-level logistic regression.}
   \label{table:freq-coefs-6}
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kilo\meter\per\hour} with gain $k=8$ are shown in Table
+10~\si{\kilo\meter\per\hour} with gain $g=8$ are shown in Table
 \ref{table:freq-coefs}. The angular impulse and perturbation order are
@@
   \caption{Logistic regression coefficient estimates at
-    10~\unit{\kilo\meter\per\hour} and gain $k=8$.  Multiplicative change in
+    10~\unit{\kilo\meter\per\hour} and gain $g=8$.  Multiplicative change in
     odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on

### 2024-06-12 src/statistics.r
Add R script for statistics

@@ -1,0 +1,70 @@
+library(lme4)
+library(readxl)
+library(tidyverse)
+library(effects)
+library(jtools)
+library(ggeffects)
+library(ggplot2)
+
+data_path_6 <- "./data/all_perturbations_6kmh.csv"
+data_path_10 <- "./data/all_perturbations_10kmh.csv"
+
+for (i in 1:2) {
+  if (i == 1) {
+    data <- read.csv(data_path_6)
+  } else {
+    data <- read.csv(data_path_10)
+  }
+
+  data$balance_assist <- as.factor(data$balance_assist)
+
+  model_simple <- glm(
+      formula = fall ~ X + angular_impulse + balance_assist + roll_angle + steer_angle + balance_assist:roll_angle + balance_assist:steer_angle + balance_assist:X + balance_assist:angular_impulse,
+      data = data,
+      family = binomial,
+  )
+  print(summary(model_simple))
+  print(summ(model_simple, exp = TRUE))
+  
+  # Create dummy dataset with all variables set to zero, except for angular impulse
+  # and state of balance-assist. The model created above uses this dataset to predict
+  # fall probabilities and show the difference between balance-assist off and 
+  # balance-assist on.
+  dummy_data_on <- data.frame(
+    angular_impulse = seq(from = -2.5, to = 2.5, by = 0.01)
+  )
+  dummy_data_on$X <- 0
+  dummy_data_on$roll_angle <- 0
+  dummy_data_on$steer_angle <- 0
+  dummy_data_on$direction <- 0
+  dummy_data_off <- dummy_data_on
+
+  dummy_data_on$balance_assist <- as.factor(1)
+  dummy_data_on$prediction <- predict(
+    model_simple, dummy_data_on, type = "response"
+  )
+  dummy_data_off$balance_assist <- as.factor(0)
+  dummy_data_off$prediction <- predict(
+    model_simple, dummy_data_off, type = "response"
+  )
+
+  if (i == 1) {
+    fig_title <- "Predicted fall probability at 6 km/h with gain k=-10"
+    file_name <- "./figures/predicted_fall_probability_6kmh.png"
+  } else {
+    fig_title <- "Predicted fall probability at 10 km/h with gain k=-8"
+    file_name <- "./figures/predicted_fall_probability_10kmh.png"
+  }
+
+  plot <- ggplot() +
[…]

### 2024-06-13 src/generate_time_series_imgs.py
Create plots with torque and roll/steer angles

@@ -13,6 +13,7 @@ DURATION_BEFORE = 0.3
 DURATION_AFTER = 2.0
 DIRECTORY = "figures"
 HANLDEBAR_LENGTH = 0.82
+BALANCE_ASSIST_MOTOR_CONSTANT = 5
 
 
 def main():
@@ -26,6 +27,7 @@ def main():
         data_off, DESIRED_FORCES, DURATION_BEFORE, DURATION_AFTER
     )
     perturbation_dfs = perturbation_dfs_on + perturbation_dfs_off
+    generate_torque_angle_plots(perturbation_dfs, DIRECTORY)
     generate_force_torque_plots(perturbation_dfs[:1], DIRECTORY, ALL_FORCES)
     generate_roll_steer_plots(perturbation_dfs, DIRECTORY)
 
@@ -78,7 +80,7 @@ def get_perturbations(
                 continue
 
             if "steer" in var or "roll" in var or "gyro" in var:
-                pert_data[var + "_orginal"] = data[var][context_start:context_stop]
+                pert_data[var + "_original"] = data[var][context_start:context_stop]
                 if data["desforce24"][start] > TRACKING_FORCE:
                     pert_data[var] = -data[var][context_start:context_stop]
                 else:
@@ -89,6 +91,84 @@ def get_perturbations(
         perturbation_dfs.append(pert_data)
 
     return perturbation_dfs
+
+
+def generate_torque_angle_plots(perturbations_dfs, directory):
+    """Generates a plot showing the torque on the handlebars and the roll angle and rate,
+    steer angle and desired torque of the balance-assist controller.
+
+    Parameters
+    ----------
+    perturbation_dfs : List[pandas.DataFrame]
+        A list of dataframes that contain the data to be plotted.
+    directory : str
+        Name of the directory in which to store the generated images.
+    """
+    if not os.path.isdir(directory):
+        os.mkdir(directory)
+
+    for i, df in enumerate(perturbations_dfs):
+        # print(df.columns.values)
+        if "motor_current" in df.columns.values:
+            fig, axs = plt.subplots(5, 1, sharex=True, layout="constrained")
+            df["motor_torque"] = df["motor_current"] / BALANCE_ASSIST_MOTOR_CONSTANT
+        else:
+            fig, axs = plt.subplots(4, 1, sharex=True, layout="constrained")
+
+        fig.set_size_inches(140 / 25.4, 160 / 25.4)
+
+        actual_torque, desired_torque = calculate_torque_on_handlebars(df)
+        axs[0].plot(
[…]

### 2024-06-13 main.tex
Fill in some missing data

@@
 with an encoder (Magura XXX) and measure the body fixed roll rate of the
-bicycle with a MEMs rate gyroscope (Company, Location). The balance assist
-control algorithm is implemented on a microprocessor (Teensy, Company,
-Location) and data from all sensors is logged with a CAN bus (Company,
-Location) at approximately X Hz.
+bicycle with a MEMs rate gyroscope (CRS 03-02s, Silicon Sensing, United Kingdom).
+\todo{Do you mean the gyro inside the bicycle or the external? Currently filled in the 
+data of the external one. Not sure what make and model is inside the bicycle}
+The balance assist control algorithm is implemented on a microprocessor (Teensy, PJRC,
+USA) and data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
+Denmark) at approximately 100 Hz. \todo{Sample rate is not the same for each
+sensor: wheel speed, imu = 1ms, steer torque command is updated every 10ms, 
+can log is written every 10ms}
 
@@
 track a commanded force profile using a PID controller running on a
-microprocessor (Arduino XXX, Italy). We measure the force applied by each motor
-at the handlebar via four inline load cells, each with a max load of
-XX~\si{\newton}. The commanded force profiles are designed to apply an external
-pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
-varying from X to X. The four motors are arranged at the four corners of a
-1~\si{\meter} wide treadmill (make, Company, City) that can reach speeds of
-18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
-Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical arrangement
-is shown in Figure~\ref{fig:participant-in-set-up}. Our modifications relative
-to that described in the thesis include simplifying the controller with a
-inexpensive microcontroller and the use of a simpler safety harness.
+microprocessor (Arduino Mega 2560, Italy). We measure the force applied by each 
+motor at the handlebar via four inline load cells, each with a max load of
+250~\si{\newton}. One load cell was damaged during pilot experiments, and is
+replaced by a load cell with a max load of 500~\si{\newton}. The commanded 
+force profiles are designed to apply an external pulsive torque to the front 
+assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200. The four
+motors are arranged at the four corners of a 1~\si{\meter} wide treadmill 
+(EC-90 Flat, Maxon Group, Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.
+The general design is described in detail in the Van De Velde's MSc 
+hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in 
+Figure~\ref{fig:participant-in-set-up}. Our modifications relative to that 
+described in the thesis include simplifying the controller with a inexpensive 
+microcontroller and the use of a simpler safety harness. 
 Figure~\ref{fig:perturbation_0} shows an example perturbation.
@@
 group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
-(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred X
+(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred 2
 weeks before the 6~\si{\kilo\meter\per\hour} group.
@@
 We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
-location of their front wheel on the center line of the treadmill. We define a
+location of their front wheel on the center line of the treadmill. \todo{The results 
+from this are not included because the balance-assist was turned on for some participants}
+We define a
 ``fall'' on the treadmill by two criteria: the rider removes their foot from
@@
 the treadmill. We then applied perturbations in random directions (clockwise or
-counter clockwise) increasing the magnitude by X~\si{\newton\meter} each time.
+counter clockwise), starting at 20~\si{\newton} and increasing the magnitude by 
+20~\si{\newton\meter} until the participants falls.
 Figure~\ref{fig:perturbation-sequence} shows an example result of a

### 2024-06-13 src/generate_time_series_imgs.py
Set legend fontsize to x-small

@@ -129,7 +129,9 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
             label="Measured net torque on handlebars",
         )
         axs[0].set_ylabel("Bump'em torque" "\n" r"[Nm]")
-        axs[0].legend()
+        axs[0].legend(
+            fontsize="x-small",
+        )
 
         df.plot(
             x="seconds_since_start",

### 2024-06-13 src/statistics.r
Change sign for gain from k to g in fig title

@@ -49,10 +49,10 @@ for (i in 1:2) {
   )
 
   if (i == 1) {
-    fig_title <- "Predicted fall probability at 6 km/h with gain k=-10"
+    fig_title <- "Predicted fall probability at 6 km/h with gain g=-10"
     file_name <- "./figures/predicted_fall_probability_6kmh.png"
   } else {
-    fig_title <- "Predicted fall probability at 10 km/h with gain k=-8"
+    fig_title <- "Predicted fall probability at 10 km/h with gain g=-8"
     file_name <- "./figures/predicted_fall_probability_10kmh.png"
   }
 

### 2024-06-13 main.tex
Started improving discussion.

--- 2024-06-06 main.tex
@@
 Table~\ref{table:freq-coefs-6}.
-
+%
 \begin{table}
@@
 \section{Discussion}
+%
+Turning the balance-assist system on significantly \((p<0.05)\) reduces the
+odds that a fall occurs while cycling at a speed of 6~\si{\kph}.
+Figure~\ref{fig:probability-6kph} visualizes the probability of falling as a
+function of the angular impulse while keeping all other explantory factors at
+their mean values. This figure is derived from the result that the
+balance-assist system halves the odds that a perturbation results in a fall.
+This figure shows that for relatively large impulses the probability to fall is
+unity for both states of balance assist on and off. And for relatively small
+impulses the probability to fall is zero for both states. But for impulses in
+the magnitude region around the mean fall threshold, the probability of falling
+is significantly lowered.
 
-\subsection{Balance-assist}
-
-Turning the balance-assist system on significantly \((p<0.05)\) reduces the
-odds that a fall occurs while cycling at a speed of 1.7~\si{\mps} but at
-2.8~\si{\mps} that is not the case \((p=0.07)\). If all other factors remain
-the same at 1.7~\si{\mps}, the balance-assist system halves the odds that a
-perturbation results in a fall. How much this changes the probability that a
+but at
+10~\si{\kph} that is not the case \((p=0.07)\).  How much this changes the probability that a
 fall occurs depends on the values of the other predictor variables: if the odds
@@
   \centering
-  \subcaptionbox{1.7~\si{\mps}}{
-    \includegraphics[width=70mm]{example-image-a}
+  \subcaptionbox{1.7~\si{\mps}\label{fig:probability-6kph}}{
+    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
   }
   \subcaptionbox{2.8~\si{\mps}}{
-    \includegraphics[width=70mm]{example-image-b}
+    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
   }

### 2024-06-13 main.tex
Some text from co-writing with Marten and Jason.

@@
 with an encoder (Magura XXX) and measure the body fixed roll rate of the
-bicycle with a MEMs rate gyroscope (CRS 03-02s, Silicon Sensing, United Kingdom).
-\todo{Do you mean the gyro inside the bicycle or the external? Currently filled in the 
-data of the external one. Not sure what make and model is inside the bicycle}
-The balance assist control algorithm is implemented on a microprocessor (Teensy, PJRC,
-USA) and data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
-Denmark) at approximately 100 Hz. \todo{Sample rate is not the same for each
-sensor: wheel speed, imu = 1ms, steer torque command is updated every 10ms, 
-can log is written every 10ms}
+bicycle with a MEMs rate gyroscope (CRS 03-02s, Silicon Sensing, United
+Kingdom).  \todo{Do you mean the gyro inside the bicycle or the external?
+Currently filled in the data of the external one. Not sure what make and model
+is inside the bicycle} The balance assist control algorithm is implemented on a
+microprocessor (Teensy, PJRC, USA) and data from all sensors is logged with a
+CAN bus (CanEdge2, CSS Electronics, Denmark) at approximately 100 Hz.
+\todo{Sample rate is not the same for each sensor: wheel speed, imu = 1ms,
+steer torque command is updated every 10ms, can log is written every 10ms}
 
@@
 force profiles are designed to apply an external pulsive torque to the front
-assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200. The four
-motors are arranged at the four corners of a 1~\si{\meter} wide treadmill 
+assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200. The
+four motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
 (EC-90 Flat, Maxon Group, Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.
@@
 could decline to continue at any time. The study was approved by Delft
-University of Technology's Human Research Ethics Council (\#XXX).
+University of Technology's Human Research Ethics Council (\#3897).
 
@@
   \centering
-  \includegraphics[width=140mm]{figures/perturbation_0.png}
+  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
   \caption{Motor force and resulting  steer torque perturbation based on an 80N
@@
   \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
+  \label{eq:log-regress}
 \end{align}
@@
     odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
-    frequentist single-level logistic regression.}
+    frequentist single-level logistic regression. and 5\% confidence intervals
+  of MCiO.}
   \label{table:freq-coefs}
@@
 
-\begin{figure}
-  \centering
-  \includegraphics{example-image-a}
-  \caption{Box plot or distribution plot showing graphical view of stats
-  tables.}
-\end{figure}
+\todo{Revisit if we want box plots to visualize the tables.}
+Turning the balance-assist system on significantly \((p<0.05)\) reduces the
+odds that a pertubation results in a fall occurs while cycling at a speed of
+6~\si{\kph}.  Figure~\ref{fig:probability-6kph} visualizes the probability of
+falling as a function of the mean and centered angular impulse per participant
+while keeping all other explantory factors at their centered mean values. This
+figure is derived by setting all explanatory variables to there mean and
+calculating the probality from Eq~\ref{eq:log-regress} for only change in
+angular impulse given the estimates in Table~\ref{tab:freq-coeffs-6} which
+indicate that the balance-assist system halves (0.53) the odds that a
+perturbation results in a fall. This figure shows that for relatively large
+impulses the probability to fall is unity for both states of balance assist on
+and off. And for relatively small impulses the probability to fall is zero for
+both states. But for impulses in the magnitude region (-1 to 0.5 STD) around
+the mean-centered fall threshold, the probability of falling is significantly
+lowered. The skewedness of the probaility curves arrives from the interaction
+effects\todo[inline]{check this and think about it}.
 
-\section{Discussion}
-%
-Turning the balance-assist system on significantly \((p<0.05)\) reduces the
-odds that a fall occurs while cycling at a speed of 6~\si{\kph}.
-Figure~\ref{fig:probability-6kph} visualizes the probability of falling as a
-function of the angular impulse while keeping all other explantory factors at
-their mean values. This figure is derived from the result that the
-balance-assist system halves the odds that a perturbation results in a fall.
-This figure shows that for relatively large impulses the probability to fall is
-unity for both states of balance assist on and off. And for relatively small
-impulses the probability to fall is zero for both states. But for impulses in
-the magnitude region around the mean fall threshold, the probability of falling
-is significantly lowered.
+\todo{Discuss the 10 kph finding}
 
-but at
-10~\si{\kph} that is not the case \((p=0.07)\).  How much this changes the probability that a
-fall occurs depends on the values of the other predictor variables: if the odds
-that a perturbation results in a fall are a 1000:1, turning on the
-balance-assist system reduces the odds to 500:1. However, the probability that
-a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
-the odds that a fall will occur are smaller, halving the odds has a larger
-influence on the fall probability. For example, if the odds that a fall occurs
-is two, halving it to one reduces the probability from $0.66$ to $0.50$. Figure
-\ref{fig:probability} shows how the balance-assist system changes the fall
-probability for different magnitudes of angular impulse.
+but at 10~\si{\kph} that is not the case \((p=0.07)\).
 
@@
     off All other predictor variables are fixed at zero, except for the
-    interaction effect between the balance-assist and the angular impulse.}
+    interaction effect between the balance-assist and the angular impulse. The
+    x axis is standard deviation around the mean (all perturbations), centered
+    per participant.}
   \label{fig:probability}
 \end{figure}
+
+\section{Discussion}
+%
+
+It is important to understand the relationship between probability and odds to
+relate Figure~\ref{fig:probability} and Table~\ref{tab:freq-coeffs-6}.  How
+much the probability that a fall changes depends on the values of all predictor
+variables and we visualize only one view into the results here. If the odds
+that a perturbation results in a fall are a 1000:1, turning on the
+balance-assist system reduces the odds to 500:1. However, the probability that
+a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
+the odds that a fall will occur are smaller, halving the odds has a larger
+influence on the fall probability.For example, if the odds that a fall occurs
+is two, halving it to one reduces the probability from $0.66$ to $0.50$. This
+clearly shown in the behavior in Figure~\ref{fig:probablity} with the behavior
+at high standard deviations.
+
+\todo{Consider making a specific value of calcualting the probabilty for a set
+of inputs to the regression model.}
 
@@
 
+Difficult to layer this onto the fall statistics, because we don't have the
+details of how people fall. If such data existed we could make estimates on the
+number of falls reduced if everyone road such a bike.
+
 \section{Conclusion}

### 2024-06-14 main.tex
Rework some text in introduction

@@
 balance assisting bicycle and hypothesize that it helps the rider in situations
-they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
+in which they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
 \todo{Talk about the types of real life perturbations an how our perturbation
@@
 similar vehicle recently showed rider dissatisfaction with the
-stabilization~\cite{Hanakam2023}.
+stabilization~\cite{Hanakam2023}. \todo{Maybe this paragraph can be a bit 
+shorter? Of course important to mention previous research, but all this 
+info is not 100% necessary for understanding our paper.}
 
@@
 In this paper, we test whether a linearly gain scheduled steer motor controlled
-bicycle that is stable in a large low speed range (when riderless) assists the
-rider in balancing by investigating whether it is beneficial in preventing
-falling.  To do this, we apply varying magnitude mechanical perturbations to
-the handlebars while the rider is balancing the bicycle on a treadmill and then
-assess the rider's probability of falling with the balance assist steer motor
-system on and off.
+bicycle, that is stable in a large low speed range when riderless, is beneficial
+in helping to prevent the rider from falling. We test this by applying varying
+magnitude mechanical perturbations to the handlebars while the rider is cycling
+on a treadmill. We assess the difference in the rider's probability of falling 
+with the balance assist steer motor system on and off.
 

### 2024-06-22 main.tex
Some work on the discussion.

--- 2024-06-13 main.tex
@@
 6~\unit{\kph} with gain $g=10$ are shown in
-Table~\ref{table:freq-coefs-6}. The angular impulse, perturbation order, and
+Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
 balance assist state are all statistically significant predictors. Larger
@@
 fall.  The associated multiplicative change in odds are shown in
-Table~\ref{table:freq-coefs-6}.
+Table~\ref{tab:freq-coefs-6}.
 %
@@
     $g=10$ based on frequentist single-level logistic regression.}
-  \label{table:freq-coefs-6}
+  \label{tab:freq-coefs-6}
   \begin{tabular}{lrrrrrr}
@@
 10~\si{\kilo\meter\per\hour} with gain $g=8$ are shown in Table
-\ref{table:freq-coefs}. The angular impulse and perturbation order are
+\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
 statistically significant predictors. Larger angular impulse increases the
 probability to fall and enduring more perturbations decreases the probability
-to fall. The associated multiplicative change in odds are shown in
-Table~\ref{table:freq-coefs}.
+to fall.
 
@@
   of MCiO.}
-  \label{table:freq-coefs}
+  \label{tab:freq-coefs-10}
   \begin{tabular}{lrrrrrr}
@@
 %
+Starting with Figure~\ref{fig:probability} and Table~\ref{tab:freq-coeffs-6}.
 
-It is important to understand the relationship between probability and odds to
-relate Figure~\ref{fig:probability} and Table~\ref{tab:freq-coeffs-6}.  How
-much the probability that a fall changes depends on the values of all predictor
-variables and we visualize only one view into the results here. If the odds
-that a perturbation results in a fall are a 1000:1, turning on the
-balance-assist system reduces the odds to 500:1. However, the probability that
-a fall occurs is only reduced from $0.999$ to $0.998$. On the other hand, if
+To interpret the results in Tables~\ref{tab:freq-coeffs-6} and
+\ref{tab:freq-coeffs-10}
+it is important to understand the relationship between probability and odds.
+The probability that a fall occurs depends on the values of all the indepedent
+variables and but we can visualize one view into the results in
+Figure~\ref{fig:probability}. The estimate in Table~\ref{tab-freq-coeffs-6}
+shows that if the odds that the balance assist system halves the odds that a
+perturbation results in a fall. For example, if the odds are a 1000:1,
+turning on the balance-assist system reduces the odds to 500:1. However, the
+probability that
+a fall occurs is only reduced from $0.999$ to $0.998$. If
 the odds that a fall will occur are smaller, halving the odds has a larger

### 2024-06-24 main.tex
Small updates in text

@@
 motor at the handlebar via four inline load cells, each with a max load of
-250~\si{\newton}. One load cell was damaged during pilot experiments, and is
+250~\si{\newton}. The left rear load cell  was damaged during pilot experiments, and is
 replaced by a load cell with a max load of 500~\si{\newton}. The commanded
 force profiles are designed to apply an external pulsive torque to the front
-assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200. The
+assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200~\si{\Newton} per motor. 
+This results in a torque between 16 and 160~\si{\Newton\meter}. The
 four motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
@@
 counter clockwise), starting at 20~\si{\newton} and increasing the magnitude by 
-20~\si{\newton\meter} until the participants falls.
+30~\si{\newton\meter} until the participants falls.
 Figure~\ref{fig:perturbation-sequence} shows an example result of a
@@
 that subject's nominal fall threshold. Following this we apply 20
-perturbations of random magnitudes to the cyclist while they ride at a constant
+perturbations of varying magnitudes to the cyclist while they ride at a constant
 speed and record which perturbations cause a fall. We let the cyclist rest and
@@
 system is on or off during the first or second set of perturbations.
+
+The magnitude of the perturbation is determined using an adaptive staircase procedure. 
+\todo{Cite to correct paper from thesis}
+The adaptive staircase takes the nominal fall threshold as input and randomly chooses 
+one of five possible forces that are centred around the nominal fall threshold with 
+10~\si{\Newton} intervals. If the perturbation results in a fall, the nominal fall 
+threshold is decreased with 10~\si{\Newton}. If the perturbation does not results in a 
+fall, the nominal fall threshold is increased with 10~\si{\Newton}. This adaptive
+staircase procedure ensures that approximately 50\% of the perturbations applied to the
+rider result in a fall. 
 %

### 2024-06-27 main.tex
More improvements to the discussion section.

@@
 that can stabilize a single track vehicle above a minimum speed when one is not
-concerned with wobble instabilities, something not prominent in bicycles at low
-speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
-stability and their vehicle was fully automatic, no human rider was involved.
-Many more automatically balanced single track vehicles have been demonstrated
-over the last 40 years, but none have demonstrated that increasing low speed
-stability can assist in balancing and possibly reduce single actor falls. Most
-of these robotic bicycle and motorcycle designers did not intend for a human
-rider to also control the stabilized vehicle. Nevertheless, an automatically
-stabilized bicycle can be controlled by a human rider if the motor controlled
-steer torque and the rider applied steer torque act on the steer in parallel.
-The effect of this automatic control gives the ability to change the dynamics,
-up to some limits, of the human controlled plant. In our prior
-study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
-distractions due to the balance assist but contrarily researchers with a
-similar vehicle recently showed rider dissatisfaction with the
-stabilization~\cite{Hanakam2023}. \todo{Maybe this paragraph can be a bit 
-shorter? Of course important to mention previous research, but all this 
-info is not 100% necessary for understanding our paper.}
+concerned with wobble instabilities. Ruijs and Pacejka's work was not
+particularly concerned with low speed stability and their vehicle was fully
+automatic, no human rider was involved. Many more automatically balanced single
+track vehicles have been demonstrated over the last 40 years, but none have
+demonstrated that increasing low speed stability can assist in balancing and
+possibly reduce single actor falls. Most of these robotic bicycle and
+motorcycle designers did not intend for a human rider to also control the
+stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
+controlled by a human rider if the motor controlled steer torque and the rider
+applied steer torque act on the steer in parallel.  The effect of this
+automatic control gives the ability to change the dynamics, up to some limits,
+of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
+we demonstrate reduced motion during distractions due to the balance assist but
+contrarily researchers with a similar vehicle recently showed rider
+dissatisfaction with the stabilization~\cite{Hanakam2023}. \todo{Maybe this
+paragraph can be a bit shorter? Of course important to mention previous
+research, but all this info is not 100\% necessary for understanding our paper.
+(I've been working on this paragraph for over a year for a few different
+possible publications about the balance assist bike, I'd like to keep it in.
+This is an attempt to set the record straight on where this entire idea of
+doing "balance assist" derives from.-- JKM)}
 
@@
 self-stability. The bicycle model is ideal for showing the effect of a roll
-motion feedback driven steer motor on the dynamics. The linear model can be
-described by the fourth order state space equations:
+motion feedback driven steer motor on the dynamics. The linear version of this
+model can be described by the fourth order state space
+equations~\citep{Meijaard2007}:
 %
@@
 In this paper, we test whether a linearly gain scheduled steer motor controlled
-bicycle, that is stable in a large low speed range when riderless, is beneficial
-in helping to prevent the rider from falling. We test this by applying varying
-magnitude mechanical perturbations to the handlebars while the rider is cycling
-on a treadmill. We assess the difference in the rider's probability of falling 
-with the balance assist steer motor system on and off.
+bicycle, that is stable in a large low speed range when riderless, is
+beneficial in helping to prevent the rider from falling. We test this by
+applying varying magnitude mechanical perturbations to the handlebars while the
+rider is cycling on a treadmill. We assess the difference in the rider's
+probability of falling with the balance assist steer motor system on and off.
 
@@
 motor at the handlebar via four inline load cells, each with a max load of
-250~\si{\newton}. The left rear load cell  was damaged during pilot experiments, and is
-replaced by a load cell with a max load of 500~\si{\newton}. The commanded
-force profiles are designed to apply an external pulsive torque to the front
-assembly (handlebars, forks, wheel) at magnitudes varying from 20 to 200~\si{\Newton} per motor. 
-This results in a torque between 16 and 160~\si{\Newton\meter}. The
-four motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
-(EC-90 Flat, Maxon Group, Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.
-The general design is described in detail in the Van De Velde's MSc
+250~\si{\newton}. The left rear load cell  was damaged during pilot
+experiments, and is replaced by a load cell with a max load of
+500~\si{\newton}. The commanded force profiles are designed to apply an
+external pulsive torque to the front assembly (handlebars, forks, wheel) at
+magnitudes varying from 20 to 200~\si{\newton} per motor.  This results in a
+torque between 16 and 160~\si{\newton\meter}. The four motors are arranged at
+the four corners of a 1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group,
+Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.  The
+general design is described in detail in the Van De Velde's MSc
 hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in
@@
 We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
-location of their front wheel on the center line of the treadmill. \todo{The results 
-from this are not included because the balance-assist was turned on for some participants}
-We define a
-``fall'' on the treadmill by two criteria: the rider removes their foot from
-the pedal and places it on the ground or the bicycle wheel exceeds the width of
-the treadmill. We then applied perturbations in random directions (clockwise or
-counter clockwise), starting at 20~\si{\newton} and increasing the magnitude by 
-30~\si{\newton\meter} until the participants falls.
-Figure~\ref{fig:perturbation-sequence} shows an example result of a
-perturbation.  We log the magnitude that causes the first fall to characterize
-that subject's nominal fall threshold. Following this we apply 20
-perturbations of varying magnitudes to the cyclist while they ride at a constant
-speed and record which perturbations cause a fall. We let the cyclist rest and
-then perform another 20 perturbations. We randomize whether the balance assist
-system is on or off during the first or second set of perturbations.
+location of their front wheel on the center line of the treadmill. \todo{The
+  results from this are not included because the balance-assist was turned on
+for some participants} We define a ``fall'' on the treadmill by two criteria:
+the rider removes their foot from the pedal and places it on the ground or the
+bicycle wheel exceeds the width of the treadmill. We then applied perturbations
+in random directions (clockwise or counter clockwise), starting at
+20~\si{\newton} and increasing the magnitude by 30~\si{\newton\meter} until the
+participants falls.  Figure~\ref{fig:perturbation-sequence} shows an example
+result of a perturbation.  We log the magnitude that causes the first fall to
+characterize that subject's nominal fall threshold. Following this we apply 20
+perturbations of varying magnitudes to the cyclist while they ride at a
+constant speed and record which perturbations cause a fall. We let the cyclist
+rest and then perform another 20 perturbations. We randomize whether the
+balance assist system is on or off during the first or second set of
+perturbations.
 
-The magnitude of the perturbation is determined using an adaptive staircase procedure. 
-\todo{Cite to correct paper from thesis}
-The adaptive staircase takes the nominal fall threshold as input and randomly chooses 
-one of five possible forces that are centred around the nominal fall threshold with 
-10~\si{\Newton} intervals. If the perturbation results in a fall, the nominal fall 
-threshold is decreased with 10~\si{\Newton}. If the perturbation does not results in a 
-fall, the nominal fall threshold is increased with 10~\si{\Newton}. This adaptive
-staircase procedure ensures that approximately 50\% of the perturbations applied to the
-rider result in a fall. 
+The magnitude of the perturbation is determined using an adaptive staircase
+procedure.  \todo{Cite to correct paper from thesis} The adaptive staircase
+takes the nominal fall threshold as input and randomly chooses one of five
+possible forces that are centred around the nominal fall threshold with
+10~\si{\newton} intervals. If the perturbation results in a fall, the nominal
+fall threshold is decreased with 10~\si{\newton}. If the perturbation does not
+results in a fall, the nominal fall threshold is increased with
+10~\si{\newton}. This adaptive staircase procedure ensures that approximately
+50\% of the perturbations applied to the rider result in a fall.
 %
@@
   \centering
-  \subcaptionbox{1.7~\si{\mps}\label{fig:probability-6kph}}{
+  \subcaptionbox{6~\si{\kph}\label{fig:probability-6kph}}{
     \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
@@
 %
-Starting with Figure~\ref{fig:probability} and Table~\ref{tab:freq-coeffs-6}.
-
-To interpret the results in Tables~\ref{tab:freq-coeffs-6} and
-\ref{tab:freq-coeffs-10}
-it is important to understand the relationship between probability and odds.
-The probability that a fall occurs depends on the values of all the indepedent
-variables and but we can visualize one view into the results in
-Figure~\ref{fig:probability}. The estimate in Table~\ref{tab-freq-coeffs-6}
-shows that if the odds that the balance assist system halves the odds that a
-perturbation results in a fall. For example, if the odds are a 1000:1,
-turning on the balance-assist system reduces the odds to 500:1. However, the
-probability that
-a fall occurs is only reduced from $0.999$ to $0.998$. If
-the odds that a fall will occur are smaller, halving the odds has a larger
-influence on the fall probability.For example, if the odds that a fall occurs
-is two, halving it to one reduces the probability from $0.66$ to $0.50$. This
-clearly shown in the behavior in Figure~\ref{fig:probablity} with the behavior
-at high standard deviations.
+The probability that a fall occurs depends on the values of all the independent
+variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
+effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
+some insight.
+But to interpret the results in Tables~\ref{tab:freq-coeffs-6} and
+\ref{tab:freq-coeffs-10} it is important to understand the relationship between
+probability and odds.
+The estimate in Table~\ref{tab-freq-coeffs-6} shows that the balance assist
+system halves the odds that a perturbation results in a fall
+(MCiO=0.53)~\todo{Make a variable for MCiO}.
+This means if the odds are a 1000:1, turning on the balance-assist system
+reduces the odds to 500:1.
+However, the probability that a fall occurs is only reduced from $0.999$ to
+$0.998$ in that case.
+If the odds that a fall will occur are smaller, halving the odds has a larger
+influence on the fall probability.
+For example, if the odds that a fall occurs is two, halving it to one reduces
+the probability from $0.66$ to $0.50$.
+This can be seen in Figure~\ref{fig:probablity} which shows the fall
+probability as a function of the magnitude of the normalized angular impulse
+for both speeds.
+The difference in probability between the balance-assist on/off case becomes
+insignificant outside of approximately \(\pm1\) standard deviation of the
+average angular impulse.
+This means that the balance assist is most effective for perturbations close to
+the subject's personal threshold between falling or recovery and that large
+perturbations will make you fall regardless of the balance assist's help.
 
@@
 Angular impulse magnitude has the largest significant effect for predicting
-fall probability. An increase in angular impulse increases the fall probability
-both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds
-is approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a
-more important predictor at higher speeds compared to lower speeds. The reason
-for this likely has to do with the width of the treadmill. As a bicycle travels
-at higher speeds, the same perturbation causes larger lateral deviations. At
-10~\si{\kph} almost all falls were due to the bicycle exiting the maximum width
-of the treadmill. If the same experiment was performed on an infinite plane,
-the riders may have recovered from the perturbation. At 6~\si{\kph} the riders
-could often recover in the allotted treadmill width. Our results are very much
-dependent on the two modes of falling with use: exit the treadmill width or
-foot is placed on the belt. Cycle paths are a similar width as the treadmill,
-so ride's are often limited in width when recovering from a fall.
+fall probability, Tables~\ref{tab:freq-coeffs-6} and \ref{tab:freq-coeffs-10}.
+An increase in angular impulse increases the fall probability both at 6 and
+10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
+approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
+important predictor at higher speeds compared to lower speeds. The reason for
+this likely has to do with the width of the treadmill and may also be why
+balance assist system did not have a statistically significant effect at
+10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
+causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
+the bicycle exiting the maximum width of the treadmill. If the same experiment
+was performed on an infinite plane, the riders may have recovered from more
+perturbations. At 6~\si{\kph} the riders could often recover in the allotted
+treadmill width. Our results are very much dependent on the two modes of
+falling with use: exit the treadmill width or foot is placed on the belt. Cycle
+paths are a similar width as the treadmill, so rider's are often limited in
+width when recovering from a fall.
 
@@
 \section{Conclusion}
+%
+TODO
 

### 2024-06-27 main.tex
Remove duplicate paragraph

@@
 
-The magnitude of the perturbation is determined using an adaptive staircase
-procedure.  \todo{Cite to correct paper from thesis} The adaptive staircase
-takes the nominal fall threshold as input and randomly chooses one of five
-possible forces that are centred around the nominal fall threshold with
-10~\si{\newton} intervals. If the perturbation results in a fall, the nominal
-fall threshold is decreased with 10~\si{\newton}. If the perturbation does not
-results in a fall, the nominal fall threshold is increased with
-10~\si{\newton}. This adaptive staircase procedure ensures that approximately
-50\% of the perturbations applied to the rider result in a fall.
 %

### 2024-06-27 main.tex
b is not balance variable.

--- 2024-06-27 main.tex
@@
 CAN bus (CanEdge2, CSS Electronics, Denmark) at approximately 100 Hz.
-\todo{Sample rate is not the same for each sensor: wheel speed, imu = 1ms,
-steer torque command is updated every 10ms, can log is written every 10ms}
 
@@
     \(c\) & integer & order number of perturbation \\
-    \(k\) & TODO & gain of balance assist control \\
     \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
@@
     \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
+    \(b\) & boolean & balance assist control on or off \\
     \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
@@
 some insight.
-But to interpret the results in Tables~\ref{tab:freq-coeffs-6} and
+To interpret the results in Tables~\ref{tab:freq-coeffs-6} and
 \ref{tab:freq-coeffs-10} it is important to understand the relationship between

### 2024-06-28 main.tex
Add example with concrete numbers

--- 2024-06-27 main.tex
@@
 
+To illustrate the effect of the balance-assist system on fall probability, we will
+give an example of how the data collected during the experiments is used to predict
+fall probability. We use \ref{eq:log-regress} and the coefficients in 
+\ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are not included.
+Let's assume that the mean angular impulse $\bar{L}$ of all the perturbations applied to 
+a participant is 100~\si{\newton}, and the standard deviation $\sigma^{L}=15$. The centred
+and scaled angular impulse can be calculated by substracting $\bar{L}$ from the applied
+angular impulse $L$, and dividing this by $\sigma^{L}$. The same applies for the 
+perturbation order $c$, initial roll angle $\phi_0$ and initial steer angle $\delta_0$.
+If we take the coefficients estimated for cycling at 6~\si{\kilo\meter\per\hour}, the
+log-odds of falling can be calculated as follows.
+
+\begin{align}
+	\log \left( \frac{p_{ij}}{1-p_{ij}} \right) & = \beta + \sum_{k}^{k=0}\alpha_k 
+	\frac{x_{ij}^{k}-\bar{x_{ij}^{k}}} {\sigma^{x^{k}}} 
+	& = -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} 
+	-0.77\cdot\frac{10-20}{11.54} 
+	-0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}} 
+	-0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}}
+	-0.64\cdot s
+	& = 1.42 -0.64 \cdot s
+\end{align}
+
+The state of the balance-assist $s$ is a binary variable. If the balance-assist is turned on, 
+the log-odds that a fall occurs are decreased by 0.64. The odds and probability can be calculated:
+
+\begin{align}
+	\frac{p_{ij}}{1-p_{ij}} = e^{1.42-0.64s} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s
+\end{align}
+
+\begin{align}
+	p_{ij}^{s=0} = \frac{4.14}{1 + 4.14} = 0.81
+\end{align}
+
+\begin{align}
+	p_{ij}^{s=1} = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}]} = 0.69
+\end{align}
+
+Turning on the balance-assist system reduces the probability that the perturbation results in a 
+fall from 0.81 to 0.69.
+
 Angular impulse magnitude has the largest significant effect for predicting

### 2024-08-02 main.tex
Remove todos, reduce table headers.

--- 2024-06-27 main.tex
@@
 the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
-Figure~\ref{fig:perturbation_0}\todo{Maybe remove the top plot of individual
-forces and add plots of roll angle, roll rate, steer rate.} shows an example
-perturbation force measurement compared to our Bump'Em controller command.
+Figure~\ref{fig:perturbation_0} shows an example perturbation force
+measurement compared to our Bump'Em controller command.
 %
@@
     \toprule
-    Variable & Estimate & Standard error & \(p\)-value & MCiO & 2.5\% & 97.5\% \\
+    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
     \midrule
@@
 
-\todo{Revisit if we want box plots to visualize the tables.}
 Turning the balance-assist system on significantly \((p<0.05)\) reduces the

### 2024-08-08 src/data.py
Change to new measured parameters of the balance assist bicycle.

@@ -1,33 +1,36 @@
-# NOTE : This is the Batavus Browser + rider Jason taken from HumanControl
-# repo. The parameters match those of the Meijaard2007 paper.
+# NOTE : This is the Balanceassistv1 + rider Jason, see:
+# https://github.com/moorepants/BicycleParameters/pull/110
 bike_with_rider = {
-    'IBxx': 11.3557360401,
-    'IBxz': -1.96756380745,
-    'IByy': 12.2177848012,
-    'IBzz': 3.12354397008,
-    'IFxx': 0.0904106601579,
-    'IFyy': 0.149389340425,
-    'IHxx': 0.253379594731,
-    'IHxz': -0.0720452391817,
-    'IHyy': 0.246138810935,
-    'IHzz': 0.0955770796289,
-    'IRxx': 0.0883819364527,
-    'IRyy': 0.152467620286,
-    'c': 0.0685808540382,
-    'g': 9.81,
-    'lam': 0.399680398707,
-    'mB': 81.86,
-    'mF': 2.02,
-    'mH': 3.22,
-    'mR': 3.11,
-    'rF': 0.34352982332,
-    'rR': 0.340958858855,
+    'IBxx': 19.692864561933778,
+    'IBxz': -3.507187082718745,
+    'IByy': 22.616790645669592,
+    'IBzz': 4.971570470059646,
+    'IFxx': 0.09947076116547277,
+    'IFyy': 0.19230532606494322,
+    'IFzz': 0.09947076116547277,
+    'IHxx': 0.3113335653577265,
+    'IHxz': -0.051328838578187846,
+    'IHyy': 0.2566995843845019,
+    'IHzz': 0.06539535254969839,
+    'IRxx': 0.10407036275570299,
+    'IRyy': 0.18685868837700836,
+    'IRzz': 0.10407036275570299,
+    'c': 0.041754825960194364,
+    'g': 9.80665,
+    'lam': 0.25453392722979173,
+    'mB': 106.00000000000001,
+    'mF': 2.235,
+    'mH': 4.3,
+    'mR': 4.085,
+    'rF': 0.3523093609017968,
+    'rR': 0.3489472127289805,
     'v': 1.0,
-    'w': 1.121,
-    'xB': 0.289099434117,
-    'xH': 0.866949640247,
-    'zB': -1.04029228321,
-    'zH': -0.748236400835,
[…]

### 2024-08-08 main.tex
Update new balance assist data with the values calculated without truncating the oscillation measurements.

@@
 contrarily researchers with a similar vehicle recently showed rider
-dissatisfaction with the stabilization~\cite{Hanakam2023}. \todo{Maybe this
-paragraph can be a bit shorter? Of course important to mention previous
-research, but all this info is not 100\% necessary for understanding our paper.
-(I've been working on this paragraph for over a year for a few different
-possible publications about the balance assist bike, I'd like to keep it in.
-This is an attempt to set the record straight on where this entire idea of
-doing "balance assist" derives from.-- JKM)}
+dissatisfaction with the stabilization~\cite{Hanakam2023}.
 

### 2024-08-08 src/control.py

@@ -7,8 +7,8 @@ import matplotlib.pyplot as plt
 import numpy as np
 from scipy.constants import golden_ratio
 
-#from data import bike_with_rider as bike_par
-from data import bike_without_rider as bike_par
+from data import bike_with_rider as bike_par
+#from data import bike_without_rider as bike_par
 #from data import rigid_bike_without_rider as bike_par
 from model import SteerControlModel
 

### 2024-08-08 src/data.py

@@ -1,20 +1,20 @@
 # NOTE : This is the Balanceassistv1 + rider Jason, see:
 # https://github.com/moorepants/BicycleParameters/pull/110
 bike_with_rider = {
-    'IBxx': 19.692864561933778,
-    'IBxz': -3.507187082718745,
-    'IByy': 22.616790645669592,
-    'IBzz': 4.971570470059646,
-    'IFxx': 0.09947076116547277,
-    'IFyy': 0.19230532606494322,
-    'IFzz': 0.09947076116547277,
-    'IHxx': 0.3113335653577265,
-    'IHxz': -0.051328838578187846,
-    'IHyy': 0.2566995843845019,
-    'IHzz': 0.06539535254969839,
-    'IRxx': 0.10407036275570299,
-    'IRyy': 0.18685868837700836,
-    'IRzz': 0.10407036275570299,
+    'IBxx': 19.66343529689361,
+    'IBxz': -3.5176010009473533,
+    'IByy': 22.675459667916336,
+    'IBzz': 5.001009792291661,
+    'IFxx': 0.09953295608625136,
+    'IFyy': 0.19015619017713314,
+    'IFzz': 0.09953295608625136,
+    'IHxx': 0.29844713987805804,
+    'IHxz': -0.03826451175579229,
+    'IHyy': 0.25662501245035463,
+    'IHzz': 0.05656444304099169,
+    'IRxx': 0.10227397798804952,
+    'IRyy': 0.1886505218140362,
+    'IRzz': 0.10227397798804952,
     'c': 0.041754825960194364,
     'g': 9.80665,
     'lam': 0.25453392722979173,
@@ -39,20 +39,20 @@ bike_with_rider = {
 
 # Batavus Browser without rider
 bike_without_rider = {
-    'IBxx': 1.14925449488747,
-    'IBxz': 0.05719397280279859,
-    'IByy': 3.1019340904048214,
-    'IBzz': 2.0895833926673024,
-    'IFxx': 0.09947076116547277,
-    'IFyy': 0.19230532606494322,
-    'IFzz': 0.09947076116547277,
-    'IHxx': 0.3113335653577265,
-    'IHxz': -0.051328838578187846,
-    'IHyy': 0.2566995843845019,
-    'IHzz': 0.06539535254969839,
-    'IRxx': 0.10407036275570299,
-    'IRyy': 0.18685868837700836,
-    'IRzz': 0.10407036275570299,
[…]

### 2024-08-10 src/control.py
Use four panel figure for eigenvalues vs speed.
- Show uncontrolled/controlled & with/without rider for the 4 panels.
- Use the control law Marten included in his report to match what is on
  the bike.

@@ -2,14 +2,11 @@ import os
 
 from bicycleparameters.parameter_sets import Meijaard2007ParameterSet
 from bicycleparameters.bicycle import sort_eigenmodes
-import matplotlib as mpl
 import matplotlib.pyplot as plt
 import numpy as np
 from scipy.constants import golden_ratio
 
-from data import bike_with_rider as bike_par
-#from data import bike_without_rider as bike_par
-#from data import rigid_bike_without_rider as bike_par
+from data import bike_with_rider, bike_without_rider
 from model import SteerControlModel
 
 SCRIPT_PATH = os.path.realpath(__file__)
@@ -17,18 +14,41 @@ SRC_DIR = os.path.dirname(SCRIPT_PATH)
 ROOT_DIR = os.path.realpath(os.path.join(SRC_DIR, '..'))
 FIG_DIR = os.path.join(ROOT_DIR, 'figures')
 KPH2MPS = 1000.0/3600.0
+MPS2KPH = 1.0/KPH2MPS
 
 if not os.path.exists(FIG_DIR):
     os.mkdir(FIG_DIR)
 
-parameter_set = Meijaard2007ParameterSet(bike_par, True)
-model = SteerControlModel(parameter_set)
+par_set_with = Meijaard2007ParameterSet(bike_with_rider, True)
+model_with = SteerControlModel(par_set_with)
+
+par_set_without = Meijaard2007ParameterSet(bike_without_rider, False)
+model_without = SteerControlModel(par_set_without)
 
 # FIGURE : Geometry and mass distribution
-fig, ax = plt.subplots()
-fig.set_size_inches((5.0, 5.0/golden_ratio))
-parameter_set.plot_all(ax=ax)
-fig.savefig(os.path.join(FIG_DIR, 'bicycle-geometry-mass.png'), dpi=300)
+fig, axes = plt.subplots(1, 2, sharey=True, layout='compressed')
+fig.set_size_inches((160/25.4, 160/25.4/golden_ratio))
+par_set_without.plot_all(ax=axes[0])
+par_set_with.plot_all(ax=axes[1])
+fig.savefig(os.path.join(FIG_DIR, 'bicycle-with-geometry-mass.png'), dpi=300)
+
+speeds = np.linspace(0.0, 10.0, num=2001)
+
+# control law
+vmin, vmin_idx = 1.5, np.argmin(np.abs(speeds - 1.5))
[…]

### 2024-08-10 src/data.py

@@ -37,7 +37,8 @@ bike_with_rider = {
     'kphidot': 0.0,
 }
 
-# Batavus Browser without rider
+# NOTE : This is the Balanceassistv1, see:
+# https://github.com/moorepants/BicycleParameters/pull/110
 bike_without_rider = {
     'IBxx': 1.1198252298472995,
     'IBxz': 0.04678005457419021,
@@ -74,8 +75,7 @@ bike_without_rider = {
     'kphidot': 0.0,
 }
 
-# The "Rigid" bike (Davis Instrumented bicycle), weighs more than Batavus
-# browser.
+# The "Rigid" bike (Davis Instrumented bicycle):
 rigid_bike_without_rider = {
     'IBxx': 2.63751918806,
     'IBxz': 0.65355108991,

### 2024-08-11 src/control.py
Removed redundancy in eig plots.

@@ -19,14 +19,14 @@ MPS2KPH = 1.0/KPH2MPS
 if not os.path.exists(FIG_DIR):
     os.mkdir(FIG_DIR)
 
-par_set_with = Meijaard2007ParameterSet(bike_with_rider, True)
-model_with = SteerControlModel(par_set_with)
-
 par_set_without = Meijaard2007ParameterSet(bike_without_rider, False)
 model_without = SteerControlModel(par_set_without)
 
+par_set_with = Meijaard2007ParameterSet(bike_with_rider, True)
+model_with = SteerControlModel(par_set_with)
+
 # FIGURE : Geometry and mass distribution
-fig, axes = plt.subplots(1, 2, sharey=True, layout='compressed')
+fig, axes = plt.subplots(1, 2, sharey=True, layout='constrained')
 fig.set_size_inches((160/25.4, 160/25.4/golden_ratio))
 par_set_without.plot_all(ax=axes[0])
 par_set_with.plot_all(ax=axes[1])
@@ -37,12 +37,13 @@ speeds = np.linspace(0.0, 10.0, num=2001)
 # control law
 vmin, vmin_idx = 1.5, np.argmin(np.abs(speeds - 1.5))
 vmax, vmax_idx = 4.7, np.argmin(np.abs(speeds - 4.7))
-kphidots = -10.0*(vmax - speeds)
-kphidots[:vmin_idx] = -10.0*(vmax - vmin)/vmin*speeds[:vmin_idx]
+static_gain = 10.0
+kphidots = -static_gain*(vmax - speeds)
+kphidots[:vmin_idx] = -static_gain*(vmax - vmin)/vmin*speeds[:vmin_idx]
 kphidots[vmax_idx:] = 0.0
 
 # FIGURE : Plot the roll rate gains versus speed.
-fig, ax = plt.subplots(layout='compressed')
+fig, ax = plt.subplots(layout='constrained')
 fig.set_size_inches((80/25.4, 80/25.4/golden_ratio))
 ax.plot(speeds, kphidots)
 ax.set_ylabel(r'$k_\dot{\phi}$')
@@ -61,87 +62,45 @@ def stable_ranges(evals):
 
 
 fig_four, axes = plt.subplots(2, 2, sharex=True, sharey=True,
-                              layout='compressed')
+                              layout='constrained')
 fig.set_size_inches((160/25.4, 160/25.4/golden_ratio))
 
[…]

### 2024-08-11 src/control.py
Shade stable region grey, improve axes labels.

@@ -71,30 +71,32 @@ def plot_eig(ax, model, kphidots=0.0):
                                                        kphidot=kphidots))
     weave_idx, capsize_idx = stable_ranges(evals_)[0]
     weave_speed, capsize_speed = speeds[weave_idx], speeds[capsize_idx]
-    print('Weave speed: {:1.2f} [m/s]'.format(weave_speed))
-    print('Capsize speed: {:1.2f} [m/s]'.format(capsize_speed))
+    msg = 'Weave speed: {:1.2f} [m/s], {:1.1f} [km/h]'
+    print(msg.format(weave_speed, weave_speed*MPS2KPH))
+    msg = 'Capsize speed: {:1.2f} [m/s], {:1.1f} [km/h]'
+    print(msg.format(capsize_speed, capsize_speed*MPS2KPH))
     ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0)
     ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0)
     model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
                                 colors=['k']*4)
     ax.fill_between(speeds, -10, 10, where=np.all(evals_.real < 0.0, axis=1),
-                    color='green', alpha=0.4,
+                    color='gray', alpha=0.4,
                     transform=ax.get_xaxis_transform())
     ax.set_ylim((-10, 10))
     return ax
 
 
 ax = plot_eig(axes[0, 0], model_without)
-ax.set_title('Without Rider')
-ax.set_ylabel('Uncontrolled\nEigenvalue Components [1/s]')
+ax.set_title('Without Rigid Rider')
+ax.set_ylabel('Balance Assist Off\nEigenvalue Components [1/s]')
 ax.set_xlabel('')
 
 ax = plot_eig(axes[1, 0], model_without, kphidots=kphidots)
-ax.set_ylabel('Controlled\nEigenvalue Components [1/s]')
+ax.set_ylabel('Balance Assist On\nEigenvalue Components [1/s]')
 ax.set_xlabel('Speed [m/s]')
 
 ax = plot_eig(axes[0, 1], model_with)
-ax.set_title('With Rider')
+ax.set_title('With Rigid Rider')
 ax.set_ylabel('')
 ax.set_xlabel('')
 

### 2024-08-11 main.tex
Use four panel figure for eigenvalues vs speed.
- Show uncontrolled/controlled & with/without rider for the 4 panels.
- Use the control law Marten included in his report to match what is on
  the bike.
Text edits and improvements.

@@
   \centering
-  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
+  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
   \caption{Balance assist bicycle prototype with electric motor in the steering
@@
 the human controlled plant takes the form:
-
+%
 \begin{align}
@@
 %
-We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
-this is the right model}, Dieren, The Netherlands) with a custom motor in the
-steering assembly capable of applying up to 7~\si{\newton\meter} of torque
-between the head tube and steer tube, see
-Figure~\ref{fig:balance-assist-bicycle}.  A custom motor controller converts
-the commanded torque to applied torque. We measure the speed of the rear wheel
-with an encoder (Magura XXX) and measure the body fixed roll rate of the
-bicycle with a MEMs rate gyroscope (CRS 03-02s, Silicon Sensing, United
-Kingdom).  \todo{Do you mean the gyro inside the bicycle or the external?
-Currently filled in the data of the external one. Not sure what make and model
-is inside the bicycle} The balance assist control algorithm is implemented on a
-microprocessor (Teensy, PJRC, USA) and data from all sensors is logged with a
-CAN bus (CanEdge2, CSS Electronics, Denmark) at approximately 100 Hz.
-\todo{Sample rate is not the same for each sensor: wheel speed, imu = 1ms,
-steer torque command is updated every 10ms, can log is written every 10ms}
+We modified an electric bicycle (Grenoble Arroyo\todo{Double check this is the
+right model}, Royal Dutch Gazelle, Dieren, The Netherlands) with a custom motor
+in the steering assembly capable of applying up to 7~\si{\newton\meter} of
+torque between the head tube and steer tube, see
+Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
+commanded torque to applied torque. We measure the speed of the rear wheel with
+an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
+with a MEMs rate gyroscope (\todo{check bike for this}). The balance assist
+control algorithm is implemented on a microprocessor (Teensy, PJRC, USA) and
+data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
+Denmark) at at least 100 Hz. \todo{Sample rate is not the same for each sensor:
+wheel speed, imu = 1ms, steer torque command is updated every 10ms, can log is
+written every 10ms}
 
@@
 We use a forward speed \(v\) gain scheduled proportional roll rate feedback
-controller to stabilize the bicycle. The commanded steer torque
-\(T^\textrm{m}_\delta\) from the steer motor follows the control law
+controller to stabilize the bicycle. In the speed range tested, the commanded
+steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
+law
 %
 \begin{align}
-  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{weave} - v)\dot{\phi}
+  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
 \end{align}
 %
-where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
-open loop bicycle rigid-rider dynamics. We use \(g=8\) and \(g=10\) as gain
-values during the experiments. Scaling the proportional feedback gain linearly
-with respect to speed stabilizes the normally unstable weave mode of the
-bicycle from speeds of X~\si{\kilo\meter\per\hour} to
-X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.
-
+where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
+average stable speed predicted from the open loop bicycle rigid-rider dynamics.
+We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
+proportional feedback gain linearly with respect to speed stabilizes the
+normally unstable weave mode of the bicycle down to
+2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
+4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
+Figure~\ref{fig:eigenvalues}~\todo{Add the experimental weave mode
+identificatio to this figure and consider lowering the gain on the model to
+make the model match the experimental values.}.
+%
 \begin{figure}
   \centering
-  \subcaptionbox{}{
-    \includegraphics[width=100mm]{figures/uncontrolled-eig-vs-speeds.png}
-  }
-  \subcaptionbox{}{
-    \includegraphics[width=100mm]{figures/balance-assist-controllers-eig-vs-speeds.png}
-  }
-  \caption{Uncontrolled (a) and controlled (b) root locus of eigenvalue
-    components (real: solid, imaginary: dashed) plotted versus speed. Vertical
-    blue lines indicate the two speeds we perform experiments at. Green shaded
-    region is the linear stable range.}
+  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
+  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
+    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
+    for the bicycle without the inertia effect of a rigid rider (left) and with
+    a rigid rider (right). Vertical blue lines indicate the two speeds we
+    perform experiments at. Grey shaded region is the linear stable speed
+    range. Geometric and inertial parameters for these plots were estimated
+    with the methods presented in \cite{Moore2012} and software packages
+    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
   \label{fig:eigenvalues}
 \end{figure}
-
-\todo[inline]{Decide on what physical parameters to use.}
 
@@
 adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
-working in tandem, Figure~\ref{fig:setup-diagram}. The four motors are
+working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
 programmed to apply a light force at all times to keep the ropes taught and to
 track a commanded force profile using a PID controller running on a
-microprocessor (Arduino Mega 2560, Italy). We measure the force applied by each
-motor at the handlebar via four inline load cells, each with a max load of
-250~\si{\newton}. The left rear load cell  was damaged during pilot
-experiments, and is replaced by a load cell with a max load of
-500~\si{\newton}. The commanded force profiles are designed to apply an
-external pulsive torque to the front assembly (handlebars, forks, wheel) at
-magnitudes varying from 20 to 200~\si{\newton} per motor.  This results in a
-torque between 16 and 160~\si{\newton\meter}. The four motors are arranged at
-the four corners of a 1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group,
-Switzerland) that can reach speeds of 18~\si{\kilo\meter\per\hour}.  The
-general design is described in detail in the Van De Velde's MSc
-hesis~\citep{vandeVelde2022a} and the physical arrangement  is shown in
-Figure~\ref{fig:participant-in-set-up}. Our modifications relative to that
-described in the thesis include simplifying the controller with a inexpensive
-microcontroller and the use of a simpler safety harness.
-Figure~\ref{fig:perturbation_0} shows an example perturbation.
+microprocessor (Arduino Mega 2560, COMPANYXXX, Italy). We measure the force
+applied by each motor at the handlebar via four inline 250~\si{\newton} load
+cells. The left rear load cell was damaged during pilot experiments, and is
+replaced by a load cell with a max load of 500~\si{\newton}. The commanded
+force profiles are designed to apply an external pulsive torque to the front
+assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
+160~\si{\newton\meter}. The four motors are arranged at the four corners of a
+1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
+reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
+is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
+the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
+Our modifications relative to Van De Velde include simplifying the controller
+with a inexpensive microcontroller and the use of a simpler non-active safety
+harness.
 %
-% NOTE : I get a compile error with the \protect calls here, not sure why.
+% NOTE : I get a compile error without the \protect calls here, not sure why.
 \begin{figure}
@@
 group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
-(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred 2
-weeks before the 6~\si{\kilo\meter\per\hour} group.
+(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred
+two weeks before the 6~\si{\kilo\meter\per\hour} group.
 
@@
 perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
-We then asked the rider to ride for 90~\si{\second}, attempting to maintain the
-location of their front wheel on the center line of the treadmill. \todo{The
-  results from this are not included because the balance-assist was turned on
-for some participants} We define a ``fall'' on the treadmill by two criteria:
-the rider removes their foot from the pedal and places it on the ground or the
-bicycle wheel exceeds the width of the treadmill. We then applied perturbations
-in random directions (clockwise or counter clockwise), starting at
-20~\si{\newton} and increasing the magnitude by 30~\si{\newton\meter} until the
-participants falls.  Figure~\ref{fig:perturbation-sequence} shows an example
-result of a perturbation.  We log the magnitude that causes the first fall to
-characterize that subject's nominal fall threshold. Following this we apply 20
-perturbations of varying magnitudes to the cyclist while they ride at a
-constant speed and record which perturbations cause a fall. We let the cyclist
-rest and then perform another 20 perturbations. We randomize whether the
-balance assist system is on or off during the first or second set of
-perturbations.
-
+We then asked the rider to ride for 90~\si{\second}, while attempting to
+maintain the location of their front wheel on the center line of the treadmill.
+We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
+their foot from the pedal and places it on the ground or 2) the bicycle wheel
+exceeds the width of the treadmill belt. We then applied perturbations in
+random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
+and increasing the magnitude by 30~\si{\newton\meter} until the participants
+fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
+from a perturbation. We log the magnitude that causes the first fall to
+characterize that subject's nominal fall threshold.
 %
@@
   \centering
-  \subcaptionbox{Start}{\includegraphics[width=40mm]{figures/start.png}}
-  \subcaptionbox{During}{\includegraphics[width=40mm]{figures/during.png}}
-  \subcaptionbox{After}{\includegraphics[width=40mm]{figures/after.png}}
-  \subcaptionbox{Recovery}{\includegraphics[width=40mm]{figures/recovery.png}}
+  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
+  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
+  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
+  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
   \caption{Video frames depicting the application of a perturbation and the
@@
 Following the initial threshold determination, we choose perturbation forces
-according to a random and adaptive staircase procedure applying perturbations
-above and below the initial perturbation threshold, while allowing small
-progression of the perturbation threshold to accommodate learning effects. The
-goal of this adaptive staircase procedure is to have participants fall for
-approximately 50\% of the time. The procedure consists of twenty perturbations
-per condition. Participants undergo two conditions: twenty perturbations with
-the balance-assist turned on and twenty perturbations with the balance-assist
-turned off.
-
-Five possible perturbation forces are determined based on the initial
-perturbation threshold estimate: the initial estimate itself, two perturbations
-lower than the initial estimate and two perturbations higher than the initial
-estimate.  The five possible perturbations are separated by 10 N steps. For
-example, if the initial estimate of the perturbation threshold of a participant
-is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
-100~\si{\newton}. Which one of these five perturbations is chosen is
+according to a random and adaptive staircase procedure~\cite{TODO} applying
+perturbations above and below the initial perturbation threshold, while
+allowing small progression of the perturbation threshold to accommodate
+learning effects. Five possible perturbation forces are determined based on the
+initial perturbation threshold estimate: the initial estimate itself, two
+perturbations lower than the initial estimate and two perturbations higher than
+the initial estimate.  The five possible perturbations are separated by 10 N
+steps. For example, if the initial estimate of the perturbation threshold of a
+participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
+90 and 100~\si{\newton}. Which one of these five perturbations is chosen is
 determined at random. If the perturbation results in a fall, the estimate of
-the perturbation threshold is decreased with 10~\si{\newton}, and vice versa
-if the perturbation did not result in a fall. Five new possible perturbation
+the perturbation threshold is decreased with 10~\si{\newton}, and vice versa if
+the perturbation did not result in a fall. Five new possible perturbation
 forces are determined around the updated perturbation threshold, and a new
 perturbation is selected at random. This process iterates until twenty
-perturbations are applied.
+perturbations are applied. The goal of this adaptive staircase procedure is to
+have participants fall for approximately 50\% of the time. We apply 20
+perturbations in this manner while the subject rides at a constant speed and
+record which perturbations cause a fall. We let the cyclist rest and then
+perform another 20 perturbations. We randomize whether the balance assist
+system is on or off during the first or second set of perturbations.
 
@@
 the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
-Figure~\ref{fig:perturbation_0} shows an example perturbation force
-measurement compared to our Bump'Em controller command.
+Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
+compared to our Bump'Em controller command.
 %
@@
 duration of the commanded force and is calculated as follows:
-
+%
 \begin{align}

### 2024-08-11 src/generate_time_series_imgs.py
Use black and white for the perturbation time series.

@@ -115,34 +115,41 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
         else:
             fig, axs = plt.subplots(3, 1, sharex=True, layout="constrained")
 
-        fig.set_size_inches(140 / 25.4, 160 / 25.4)
+        fig.set_size_inches(140/25.4, 100/25.4)
 
         actual_torque, desired_torque = calculate_torque_on_handlebars(df)
         axs[0].plot(
             df["seconds_since_start"],
             desired_torque,
-            label="Desired net torque on handlebars",
+            label="Commanded",
+            color='black',
+            linestyle='-'
         )
         axs[0].plot(
             df["seconds_since_start"],
             actual_torque,
-            label="Measured net torque on handlebars",
+            label="Measured",
+            color='black',
+            linestyle='--',
         )
-        axs[0].set_ylabel("Bump'em torque" "\n" r"[Nm]")
+        axs[0].set_ylabel("Bump'Em\nHandlebar Torque\n[Nm]")
         axs[0].legend(fontsize="x-small")
 
         df.plot(
             x="seconds_since_start",
             y="roll_angle",
             ax=axs[1],
-            ylabel="Angle [deg]",
-            label="Roll angle",
+            ylabel="Angle\n[deg]",
+            label="Roll",
+            color='black',
         )
         df.plot(
             x="seconds_since_start",
             y="steer_angle",
             ax=axs[1],
-            label="Steer angle",
+            label="Steer",
+            color='black',
+            linestyle='--',
         )
         axs[1].legend(fontsize="x-small")
 
@@ -150,8 +157,9 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
             x="seconds_since_start",
             y="roll_rate",
             ax=axs[2],
-            ylabel=r"Roll rate" "\n" r"[deg/s]",
+            ylabel="Roll Rate\n[deg/s]",
             legend=False,
+            color='black',
         )
         if "motor_current" in df.columns.values:
             df.plot(
@@ -159,14 +167,16 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 y="motor_torque",
                 ax=axs[3],
                 xlabel="Time since start of perturbation [s]",
-                ylabel=r"Desired controller" "\n" r" torque [Nm]",
+                ylabel="Commanded\nMotor Torque\n[Nm]",
                 legend=False,
+                color='black',
             )
 
         for ax in axs:
             ax.grid()
 
[…]

### 2024-08-12 src/control.py
Add custom eig val plot function to hide zeros.

@@ -78,10 +78,7 @@ def plot_eig(ax, model, kphidots=0.0):
     ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0)
     ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0)
     model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
-                                colors=['k']*4)
-    ax.fill_between(speeds, -10, 10, where=np.all(evals_.real < 0.0, axis=1),
-                    color='gray', alpha=0.4,
-                    transform=ax.get_xaxis_transform())
+                                hide_zeros=True, colors=['k']*4)
     ax.set_ylim((-10, 10))
     return ax
 

### 2024-08-12 src/model.py

@@ -1,6 +1,7 @@
 import numpy as np
+import matplotlib.pyplot as plt
 from bicycleparameters.models import Meijaard2007Model
-from bicycleparameters.bicycle import ab_matrix
+from bicycleparameters.bicycle import ab_matrix, sort_eigenmodes
 
 
 class SteerControlModel(Meijaard2007Model):
@@ -99,3 +100,95 @@ class SteerControlModel(Meijaard2007Model):
             B = B
 
         return A, B
+
+    def plot_eigenvalue_parts(self, ax=None, colors=None,
+                              show_stable_regions=True, hide_zeros=False,
+                              **parameter_overrides):
+        """Returns a matplotlib axis of the real and imaginary parts of the
+        eigenvalues plotted against the provided parameter.
+
+        Parameters
+        ==========
+        ax : Axes
+            Matplotlib axes.
+        colors : sequence, len(4)
+            Matplotlib colors for the 4 modes.
+        show_stable_regions : boolean, optional
+            If true, a grey shaded background will indicate stable regions.
+        hide_zeros : boolean or float, optional
+            If true, real or imaginary parts that are smaller than 1e-12 will
+            not be plotted. Providing a float will set the tolerance.
+        **parameter_overrides : dictionary
+            Parameter keys that map to floats or array_like of floats
+            shape(n,). All keys that map to array_like must be of the same
+            length.
+
+        Examples
+        ========
+
+        .. plot::
+           :include-source: True
+           :context: reset
+
+           import numpy as np
+           from bicycleparameters.parameter_dicts import meijaard2007_browser_jason
+           from bicycleparameters.parameter_sets import Meijaard2007ParameterSet
+           from bicycleparameters.models import Meijaard2007Model
+           p = Meijaard2007ParameterSet(meijaard2007_browser_jason, True)
+           m = Meijaard2007Model(p)
+           m.plot_eigenvalue_parts(v=np.linspace(0.0, 10.0, num=101))
+
+        """
+
+        if ax is None:
+            fig, ax = plt.subplots()
+
+        evals, evecs = self.calc_eigen(**parameter_overrides)
+        if len(evals.shape) > 1:
+            evals, evecs = sort_eigenmodes(evals, evecs)
+        else:
+            evals, evecs = np.array([evals]), np.array([evecs])
+
[…]

### 2024-08-12 src/generate_time_series_imgs.py
Add torque limit lines.

@@ -171,6 +171,8 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 legend=False,
                 color='black',
             )
+            axs[3].axhline(7.0, color='black', linestyle='--')
+            axs[3].axhline(-7.0, color='black', linestyle='--')
 
         for ax in axs:
             ax.grid()

### 2024-09-12 main.tex
Made a background section separate from the more general introduction.

@@
 %
-Single-actor bicycle crashes are associated with a surprisingly large number of
-reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
-cruising speeds, bicycles themselves are not self-stable and can be challenging
-for the rider to balance. Low speed crashes may be reduced if the bicycle was
-self-stable at these speeds by relieving the rider of a necessary control
-activity. Bicycles can be mechanically modified to lower the speeds at which
-they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
-that applying a motor actuated steering torque proportional to the roll rate
-can stabilize a single track vehicle at very low speeds. If this automatic
-control of steering can stabilize a bicycle, it may reduce the control required
-by the rider to successfully manage balancing tasks. We have developed such a
-balance assisting bicycle and hypothesize that it helps the rider in situations
-in which they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
-\todo{Talk about the types of real life perturbations an how our perturbation
-mimics it.}
+Single-actor bicycle crashes are associated with a surprisingly large
+percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
+start up to typical cruising speeds, bicycles themselves are not self-stable
+and can be challenging for the rider to balance. Low speed crashes may be
+reduced if the bicycle was self-stable at these speeds by relieving the rider
+of some of the necessary control activity. Bicycles can be mechanically
+modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
+and since the 1980s it has been known that applying a motor actuated steering
+torque proportional to the roll rate can stabilize a single-track vehicle at
+lower speeds than without the control. If automatic control of steering can
+stabilize a bicycle, it may reduce the control required from the rider to
+successfully manage balancing tasks. We have developed such a balance assisting
+bicycle, Figure~\ref{fig:balance-assist-bicycle}, and hypothesize that it helps
+the rider in situations in which they are likely to fall.
 %
@@
 
+Riders fall when their control authority is not able to maintain a stable
+vehicle state. There are many real world scenarios that can put the vehicle
+into an uncontrollable state. External forces applied to the vehicle or rider
+are one such scenario type and examples include wind gusts, handlebars
+colliding with a neighbor's, a bag swinging from the handlebar, or simply
+hitting a bump in the road. To assess our balance assisting bicycle, we subject
+the rider to perturbations at the handledbar, which can easily cause a rider to
+fall, especially when the rider is not so attentive or has one or more hands
+removed from the handlebar.
+
+In this paper, we test whether a linearly gain scheduled steer motor controlled
+bicycle, that is stable in a large low speed range, is beneficial in helping to
+prevent the rider from falling. We test this by applying varying magnitude
+mechanical perturbations to the handlebars while the rider is cycling on a
+treadmill. We assess the difference in the rider's probability of falling with
+the balance assist steer motor system on and off.
+
+\subsection{Background}
+%
 In the early years of developments in automatic control, \citep{Whipple1899}
@@
 speed can give satisfactory stability.
-
-In this paper, we test whether a linearly gain scheduled steer motor controlled
-bicycle, that is stable in a large low speed range when riderless, is
-beneficial in helping to prevent the rider from falling. We test this by
-applying varying magnitude mechanical perturbations to the handlebars while the
-rider is cycling on a treadmill. We assess the difference in the rider's
-probability of falling with the balance assist steer motor system on and off.
 

### 2024-09-12 main.tex
Edits down to perturbations section.

@@
 and since the 1980s it has been known that applying a motor actuated steering
-torque proportional to the roll rate can stabilize a single-track vehicle at
-lower speeds than without the control. If automatic control of steering can
-stabilize a bicycle, it may reduce the control required from the rider to
-successfully manage balancing tasks. We have developed such a balance assisting
-bicycle, Figure~\ref{fig:balance-assist-bicycle}, and hypothesize that it helps
-the rider in situations in which they are likely to fall.
+torque proportional to the vehicle's roll rate can stabilize a single-track
+vehicle at lower speeds. If automatic control of steering can stabilize a
+bicycle, it may reduce the control required from the rider to successfully
+manage balancing tasks much like the high speed self-stability already does. We
+have developed a balance assisting bicycle,
+Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
+hypothesize that it helps the rider in situations in which they are likely to
+fall.
 %
@@
 into an uncontrollable state. External forces applied to the vehicle or rider
-are one such scenario type and examples include wind gusts, handlebars
+are one such scenario type and natural examples include wind gusts, handlebars
 colliding with a neighbor's, a bag swinging from the handlebar, or simply
 hitting a bump in the road. To assess our balance assisting bicycle, we subject
-the rider to perturbations at the handledbar, which can easily cause a rider to
-fall, especially when the rider is not so attentive or has one or more hands
-removed from the handlebar.
+the rider to perturbations at the handledbar, which can relatively easily cause
+a rider to fall.
 
-In this paper, we test whether a linearly gain scheduled steer motor controlled
-bicycle, that is stable in a large low speed range, is beneficial in helping to
-prevent the rider from falling. We test this by applying varying magnitude
-mechanical perturbations to the handlebars while the rider is cycling on a
-treadmill. We assess the difference in the rider's probability of falling with
-the balance assist steer motor system on and off.
+In this paper, we test whether a steer motor controlled bicycle, that is stable
+in a large low speed range, is beneficial in helping to prevent the rider from
+falling. We test this by applying varying magnitude mechanical perturbations to
+the handlebars while the rider is cycling on a treadmill. We assess the
+difference in the rider's probability of falling with the balance assist system
+on and off.
 
-\subsection{Background}
+\subsection{Technical Background}
 %
-In the early years of developments in automatic control, \citep{Whipple1899}
+During the early years of developments in automatic control, \citep{Whipple1899}
 not only derived the correct equations of motion of the bicycle but realized
 and showed that roll motion feedback can stabilize a bicycle. Much later,
-attempts at automatic roll stabilization of single track vehicles began in
+attempts at automatic roll stabilization of single-track vehicles began in
 earnest after predictive motorcycle models were developed and refined in
@@
 Ruijs and Pacjecka followed this by demonstrating an automatically balanced
-motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
+motorcycle~\citep{Ruijs1986} and they did so solely with a steering motor. Ruijs and
 Pacejka clearly showed that steer torque driven by roll angle feedback
@@
 proposed how the gains must change with respect to vehicle speed for favorable
-control at all speeds. This roll motion feedback is the simplest controller
-that can stabilize a single track vehicle above a minimum speed when one is not
+control at all speeds. This roll motion feedback enables the simplest controller
+that can stabilize a single-track vehicle above a minimum speed when one is not
 concerned with wobble instabilities. Ruijs and Pacejka's work was not
 particularly concerned with low speed stability and their vehicle was fully
-automatic, no human rider was involved. Many more automatically balanced single
-track vehicles have been demonstrated over the last 40 years, but none have
-demonstrated that increasing low speed stability can assist in balancing and
-possibly reduce single actor falls. Most of these robotic bicycle and
-motorcycle designers did not intend for a human rider to also control the
-stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
-controlled by a human rider if the motor controlled steer torque and the rider
-applied steer torque act on the steer in parallel.  The effect of this
-automatic control gives the ability to change the dynamics, up to some limits,
-of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
-we demonstrate reduced motion during distractions due to the balance assist but
-contrarily researchers with a similar vehicle recently showed rider
-dissatisfaction with the stabilization~\cite{Hanakam2023}.
+automatic, i.e no human rider was involved.
+
+Many more automatically balanced single track vehicles have been demonstrated
+over the last 40 years, but none have demonstrated that increasing low speed
+stability can assist in balancing and possibly reduce single-actor falls. Most
+of these robotic bicycle and motorcycle designers did not intend for a human
+rider to also control the stabilized vehicle. Nevertheless, an automatically
+stabilized bicycle can be controlled by a human rider if the motor controlled
+steer torque and the rider applied steer torque act on the steer in parallel.
+The effect of this automatic control gives the ability to effectively change
+the dynamics, up to some limits, of the human-controlled plant. In our prior
+study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
+distractions due to the balance assist system but contrarily researchers with a
+similar vehicle recently showed rider dissatisfaction with the
+stabilization~\cite{Hanakam2023}.
 
@@
 the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stability. The bicycle model is ideal for showing the effect of a roll
+self-stability. The bicycle model is suitable for showing the effect of a roll
 motion feedback driven steer motor on the dynamics. The linear version of this
-model can be described by the fourth order state space
-equations~\citep{Meijaard2007}:
+model~\citep{Meijaard2007} can be described by the fourth order state space
+equations:
 %
@@
 their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
-\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
-functions of the equilibrium forward speed \(v\) and  are otherwise populated
-with expressions that are functions of the geometric and inertial parameters of
-the nonholonomic multibody system made up of four rigid bodies: two wheels,
-front frame, and rear frame.
+\(T_\delta\). The state \(\mathbf{A}\) matrix is a function of the equilibrium
+forward speed \(v\). It and the input \(\mathbf{B}\) matrix are otherwise
+populated with expressions that are functions of the geometric and inertial
+parameters of the nonholonomic multibody system made up of four rigid bodies:
+two wheels, front frame, and rear frame.
 
@@
 
-The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
-functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
-such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
-k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
-\(v_{min} < v < v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable
-speed given \(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which
-the uncontrolled bicycle's capsize mode goes unstable. With gain scheduling
-with respect to \(v\), the speed range where the bicycle is stable can be
-maximized within any physical actuator magnitude and bandwidth limits.
+The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
+means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
+\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
+0\right] \right)\) have negative real parts for \(v_{min} < v <
+v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
+\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
+uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
+respect to \(v\), the speed range where the bicycle is stable can be maximized
+within any physical actuator magnitude and bandwidth limits.
 \citet{Schwab2008} elaborates on some of the possibilities in scheduling the
 gains for such a controller and shows that a linear scheduling with respect to
-speed can give satisfactory stability.
+speed can give satisfactory stability for a low speed range. We use this simple
+feedback principle as the basis for our balance assist controller.
 
@@
 %
-We modified an electric bicycle (Grenoble Arroyo\todo{Double check this is the
-right model}, Royal Dutch Gazelle, Dieren, The Netherlands) with a custom motor
-in the steering assembly capable of applying up to 7~\si{\newton\meter} of
-torque between the head tube and steer tube, see
-Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller converts the
-commanded torque to applied torque. We measure the speed of the rear wheel with
-an encoder (Magura XXX) and measure the body fixed roll rate of the bicycle
-with a MEMs rate gyroscope (\todo{check bike for this}). The balance assist
-control algorithm is implemented on a microprocessor (Teensy, PJRC, USA) and
-data from all sensors is logged with a CAN bus (CanEdge2, CSS Electronics,
-Denmark) at at least 100 Hz. \todo{Sample rate is not the same for each sensor:
-wheel speed, imu = 1ms, steer torque command is updated every 10ms, can log is
-written every 10ms}
+We modified an electric bicycle (Grenoble Arroyo, Royal Dutch Gazelle, Dieren,
+The Netherlands) with a custom motor in the steering assembly capable of
+applying up to 7~\si{\newton\meter} of torque between the head tube and steer
+tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
+converts the commanded torque to applied torque. We measure the speed of the
+rear wheel with an encoder (Magura XXX) and measure the body fixed roll rate of
+the bicycle with a MEMs rate gyroscope (BN0086, Sparkfun, Niwot, USA). The
+balance assist control algorithm is implemented on a microprocessor (Teensy,
+PJRC, USA) and data from all sensors is logged with a CAN bus (CanEdge2, CSS
+Electronics, Denmark) at at least 100 Hz.
 
@@
 4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
-Figure~\ref{fig:eigenvalues}~\todo{Add the experimental weave mode
-identificatio to this figure and consider lowering the gain on the model to
+Figure~\ref{fig:eigenvalues}~\todo{Could add the experimental weave mode
+identification to this figure and consider lowering the gain on the model to
 make the model match the experimental values.}.
@@
     eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
-    for the bicycle without the inertia effect of a rigid rider (left) and with
-    a rigid rider (right). Vertical blue lines indicate the two speeds we
+    for the bicycle without the inertial effect of a rigid rider (left) and
+    with a rigid rider (right). Vertical blue lines indicate the two speeds we
     perform experiments at. Grey shaded region is the linear stable speed
@@
 track a commanded force profile using a PID controller running on a
-microprocessor (Arduino Mega 2560, COMPANYXXX, Italy). We measure the force
+microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
 applied by each motor at the handlebar via four inline 250~\si{\newton} load

### 2024-09-13 src/control.py
Add com_torque to plot.

@@ -107,6 +107,7 @@ fig_four.savefig(os.path.join(FIG_DIR, 'balance-assist-eig-vs-speeds.png'),
 
 # FIGURE : Simulate an initial value problem at a low speed under control.
 idx = np.argmin(np.abs(speeds - 6.0*KPH2MPS))  # 6 km/h
+print('kphidot @ 6 km/h:', kphidots[idx])
 
 
 def controller(t, x):

### 2024-09-13 src/generate_time_series_imgs.py

@@ -1,4 +1,5 @@
 import os
+import numpy as np
 import pandas as pd
 import matplotlib.pyplot as plt
 from matplotlib.lines import Line2D
@@ -171,6 +172,8 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 legend=False,
                 color='black',
             )
+            com_torque = -30.35*np.deg2rad(df['roll_rate'])
+            axs[3].plot(df['seconds_since_start'], com_torque, color='blue')
             axs[3].axhline(7.0, color='black', linestyle='--')
             axs[3].axhline(-7.0, color='black', linestyle='--')
 

### 2024-09-14 src/generate_time_series_imgs.py
Only plot the single figure we will use.

@@ -109,82 +109,83 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
         os.mkdir(directory)
 
     for i, df in enumerate(perturbations_dfs):
-        # print(df.columns.values)
-        if "motor_current" in df.columns.values:
-            fig, axs = plt.subplots(4, 1, sharex=True, layout="constrained")
-            df["motor_torque"] = df["motor_current"] / BALANCE_ASSIST_MOTOR_CONSTANT
-        else:
-            fig, axs = plt.subplots(3, 1, sharex=True, layout="constrained")
+        if i == 10:  # only plot the figure we will use
+            # print(df.columns.values)
+            if "motor_current" in df.columns.values:
+                fig, axs = plt.subplots(4, 1, sharex=True, layout="constrained")
+                df["motor_torque"] = df["motor_current"] / BALANCE_ASSIST_MOTOR_CONSTANT
+            else:
+                fig, axs = plt.subplots(3, 1, sharex=True, layout="constrained")
 
-        fig.set_size_inches(140/25.4, 100/25.4)
+            fig.set_size_inches(140/25.4, 100/25.4)
 
-        actual_torque, desired_torque = calculate_torque_on_handlebars(df)
-        axs[0].plot(
-            df["seconds_since_start"],
-            desired_torque,
-            label="Commanded",
-            color='black',
-            linestyle='-'
-        )
-        axs[0].plot(
-            df["seconds_since_start"],
-            actual_torque,
-            label="Measured",
-            color='black',
-            linestyle='--',
-        )
-        axs[0].set_ylabel("Bump'Em\nHandlebar Torque\n[Nm]")
-        axs[0].legend(fontsize="x-small")
+            actual_torque, desired_torque = calculate_torque_on_handlebars(df)
+            axs[0].plot(
+                df["seconds_since_start"],
+                desired_torque,
+                label="Commanded",
+                color='black',
+                linestyle='-'
+            )
+            axs[0].plot(
+                df["seconds_since_start"],
+                actual_torque,
+                label="Measured",
+                color='black',
+                linestyle='--',
+            )
+            axs[0].set_ylabel("Bump'Em\nHandlebar Torque\n[Nm]")
+            axs[0].legend(fontsize="x-small")
 
-        df.plot(
-            x="seconds_since_start",
-            y="roll_angle",
-            ax=axs[1],
-            ylabel="Angle\n[deg]",
-            label="Roll",
-            color='black',
[…]

### 2024-09-14 main.tex
Fix force in caption.
Use biblatex.
Editing in methods.

@@
 \usepackage[margin=25mm]{geometry}
-\usepackage{natbib}
+\usepackage[natbib=true,style=authoryear]{biblatex}
+\addbibresource{references.bib}
 \usepackage{siunitx}  % use for all units
@@
 4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
-Figure~\ref{fig:eigenvalues}~\todo{Could add the experimental weave mode
-identification to this figure and consider lowering the gain on the model to
-make the model match the experimental values.}.
+Figure~\ref{fig:eigenvalues}.
+\todo[inline]{Could add the experimental weave mode identification to this
+figure and consider lowering the gain on the model to make the model match the
+experimental values.}
 %
@@
 %
-We apply external longitudinal forces to the ends of each handlebar using an
-adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
-working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
-programmed to apply a light force at all times to keep the ropes taught and to
-track a commanded force profile using a PID controller running on a
-microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
-applied by each motor at the handlebar via four inline 250~\si{\newton} load
-cells. The left rear load cell was damaged during pilot experiments, and is
-replaced by a load cell with a max load of 500~\si{\newton}. The commanded
-force profiles are designed to apply an external pulsive torque to the front
-assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
-160~\si{\newton\meter}. The four motors are arranged at the four corners of a
-1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
-reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
-is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
+We apply longitudinal forces to the ends of each handlebar from Earth fixed
+posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
+four motors working in cooperation, Figure~\ref{fig:setup-diagram}. The four
+motors are programmed to apply a light force at all times to keep the ropes
+taught and to track a commanded force profile using a PID controller running on
+a microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
+applied by each motor at the handlebar via four inline 250~\si{\newton} S-style
+load cells except the left rear load cell that has a max load of
+500~\si{\newton}. The commanded force profiles are designed to apply an
+external pulsive torque to the front assembly (handlebars, forks, wheel) at
+magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors
+(EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners of a
+1~\si{\meter} wide treadmill that can reach speeds of
+18~\si{\kilo\meter\per\hour}. The general base design of this system is
+described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
 the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
@@
 We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments. The subjects all were confident in their cycling skills and
+the experiments. The subjects were all confident in their cycling skills and
 had cycled at least once in the last month. Eleven subjects performed the
-experiments at 1.7~\si{\meter\per\second} (6~\si{\kilo\meter\per\hour}) and
-fifteen subjects performed the experiments at 2.8~\si{\meter\per\second}
-(10~\si{\kilo\meter\per\hour}). All subjects consented to the experiment and
+experiments at 6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and
+fifteen subjects performed the experiments at 10~\si{\kilo\meter\per\hour}
+(2.8~\si{\meter\per\second}). All subjects consented to the experiment and
 could decline to continue at any time. The study was approved by Delft
@@
 The subjects were divided into two groups. The first group performed the
-protocol at a belt speed of 2.8~\si{\meter\per\second}
-(10~\si{\kilo\meter\per\hour}) with the controller gain at $g=8$ and the second
-group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
-(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred
-two weeks before the 6~\si{\kilo\meter\per\hour} group.
+protocol at a belt speed of 10~\si{\kilo\meter\per\hour} with the controller
+gain set to $g=8$ and the second group performed the protocol at a belt speed
+of 6~\si{\kilo\meter\per\hour} with the controller gain set to $g=10$. The
+10~\si{\kph} group's experiments occurred two weeks before the
+6~\si{\kilo\meter\per\hour} group.
 
@@
 random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
-and increasing the magnitude by 30~\si{\newton\meter} until the participants
+and increasing the magnitude by 30~\si{\newton} until the participants
 fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
 from a perturbation. We log the magnitude that causes the first fall to
-characterize that subject's nominal fall threshold.
+characterize that subject's \emph{nominal fall threshold}.
 %
@@
 Following the initial threshold determination, we choose perturbation forces
-according to a random and adaptive staircase procedure~\cite{TODO} applying
-perturbations above and below the initial perturbation threshold, while
-allowing small progression of the perturbation threshold to accommodate
-learning effects. Five possible perturbation forces are determined based on the
-initial perturbation threshold estimate: the initial estimate itself, two
-perturbations lower than the initial estimate and two perturbations higher than
-the initial estimate.  The five possible perturbations are separated by 10 N
-steps. For example, if the initial estimate of the perturbation threshold of a
-participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
-90 and 100~\si{\newton}. Which one of these five perturbations is chosen is
-determined at random. If the perturbation results in a fall, the estimate of
-the perturbation threshold is decreased with 10~\si{\newton}, and vice versa if
-the perturbation did not result in a fall. Five new possible perturbation
-forces are determined around the updated perturbation threshold, and a new
+according to a random and adaptive staircase procedure applying perturbations
+above and below the initial perturbation threshold, while allowing small
+progression of the perturbation threshold to accommodate learning effects. Five
+possible perturbation forces are determined based on the initial perturbation
+threshold estimate: the initial estimate itself, two perturbations lower than
+the initial estimate, and two perturbations higher than the initial estimate.
+The five possible perturbations are separated by 10~\si{\newton} steps. For
+example, if the initial estimate of the perturbation threshold of a participant
+is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
+100~\si{\newton}. Which one of these five perturbations is chosen is determined
+at random. If the perturbation results in a fall, the estimate of the
+perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
+perturbation did not result in a fall. Five new possible perturbation forces
+are determined around the updated perturbation threshold, and a new
 perturbation is selected at random. This process iterates until twenty
 perturbations are applied. The goal of this adaptive staircase procedure is to
-have participants fall for approximately 50\% of the time. We apply 20
-perturbations in this manner while the subject rides at a constant speed and
-record which perturbations cause a fall. We let the cyclist rest and then
-perform another 20 perturbations. We randomize whether the balance assist
-system is on or off during the first or second set of perturbations.
+have participants fall for approximately 50\% of the time. After the first set
+of 20 perturbations, we let the cyclist rest and then perform another 20
+perturbations. We randomize whether the balance assist system is on or off
+during the first or second set of perturbations for each participant.
 
@@
   \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
-  \caption{Motor force and resulting  steer torque perturbation based on an 80N
+  \caption{Motor force and resulting  steer torque perturbation based on an 110N
     counterclockwise applied force.}
@@
   \textrm{.}
+  \label{eq:angular-impulse}
 \end{align}
 
-At the initiation of each perturbation we log the steer and roll angles to
-characterize the configuration of the bicycle when perturbed. The gain setting
-on the balance assist controller indicates if the assistance is off \(g=0\) or
-on at two different levels: low \(g=8\) or high \(g=10\). A recovery from the
-perturbation is successful if the person neither put their foot down onto the
-treadmill surface or either wheel of the bicycle exits the width of the
-treadmill belt. We record this as a binary variable \(f\). All measured
-variables are reported in Table~\ref{tab:raw}.
+where \(F_r\) and \(F_l\)  is the total force applied on the right and left
+handlebar ends, respectively which are the sum of the rear and front load cell
+readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
+as \(l\) in Equation \ref{eq:angular-impulse}.
+
+At the initiation of each perturbation we log the instananeous steer and roll
+angles to characterize the configuration of the bicycle when perturbed. The
+gain setting on the balance assist controller indicates if the assistance is
+off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
+recovery from the perturbation is successful if the person neither places their
+foot down onto the treadmill surface or their wheel of the bicycle exits the
+width of the treadmill belt. We record this as a binary variable \(f\) for
+"fall". All measured variables are reported in Table~\ref{tab:raw}.
 %
@@
     \midrule
-    Balance Assist Gain & \(k\) & TODO & NA \\
+    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
     Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
@@
     Perturbation Order Number & \(j\) & integer & NA\\
-    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\citep{Gabriel2023} \\
+    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
     Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & rate gyroscope \\
@@
     \(c\) & integer & order number of perturbation \\
-    \(k\) & TODO & gain of balance assist control \\
+    \(g\) & \si{\newton\second\squared} & gain of balance assist control \\
     \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
@@
 calculating the probality from Eq~\ref{eq:log-regress} for only change in
-angular impulse given the estimates in Table~\ref{tab:freq-coeffs-6} which
+angular impulse given the estimates in Table~\ref{tab:freq-coefs-6} which
 indicate that the balance-assist system halves (0.53) the odds that a
@@
 some insight.
-But to interpret the results in Tables~\ref{tab:freq-coeffs-6} and
-\ref{tab:freq-coeffs-10} it is important to understand the relationship between
+But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
+\ref{tab:freq-coefs-10} it is important to understand the relationship between
 probability and odds.
-The estimate in Table~\ref{tab-freq-coeffs-6} shows that the balance assist
+The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
 system halves the odds that a perturbation results in a fall
@@
 the probability from $0.66$ to $0.50$.
-This can be seen in Figure~\ref{fig:probablity} which shows the fall
+This can be seen in Figure~\ref{fig:probability} which shows the fall
 probability as a function of the magnitude of the normalized angular impulse
@@
 Angular impulse magnitude has the largest significant effect for predicting
-fall probability, Tables~\ref{tab:freq-coeffs-6} and \ref{tab:freq-coeffs-10}.
+fall probability, Tables~\ref{tab:freq-coefs-6} and \ref{tab:freq-coefs-10}.
 An increase in angular impulse increases the fall probability both at 6 and
@@
 
-\bibliographystyle{apalike}
 % NOTE : this file is automatically generated from Zotero, do not edit
 % manually!
-\bibliography{references}
+\printbibliography
 

### 2024-09-15 main.tex
Edits in statistics section.

@@
   \caption{Raw measurements taken during each trial. ``Fall outcome'' and
-    ``Perturbation order number'' are recorded per perturbation instance.}
+    ``Perturbation order number'' are recorded per perturbation instance. The
+    remaining measures are time varying.}
   \begin{tabular}{llll}
@@
     Fall outcome & \(f\) & boolean & NA \\
-    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
+    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
+    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
     Perturbation Order Number & \(j\) & integer & NA\\
     Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
-    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & rate gyroscope \\
+    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
     Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
@@
 %
-We test the hypothesis that the balance assistance controller will reduce the
+We test our hypothesis that the balance assistance controller will reduce the
 probability of falling when perturbed externally at the handlebar. We have a
@@
 explanatory independent variables, one of which is the binary balance
-assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
-variables.
+assistance (on or off). See Table~\ref{tab:stat-model-variables} for the
+statistical model variables.
 %
@@
   \centering
-  \caption{Statistical model variables.}
+  \caption{Independent and dependent variables used in the statistical model.}
   \label{tab:stat-model-variables}
-  \begin{tabular}{lll}
+  \begin{tabular}{llll}
     \toprule
-    Variable & Units & Description \\
+    Variable & Causality & Units & Description \\
     \midrule
-    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
-    \(c\) & integer & order number of perturbation \\
-    \(g\) & \si{\newton\second\squared} & gain of balance assist control \\
-    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
-    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
-    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
-    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
+    \(L\) & ind. & \si{\newton\meter\second} & angular impulse of perturbation torque \\
+    \(\delta_0\) & ind. & \si{\degree} & steer angle at start of perturbation \\
+    \(\phi_0\) & ind. & \si{\degree} & roll angle at start of perturbation \\
+    \(f\) & dep. & boolean & outcome: did not fall 0 or did fall 1 \\
+    \(c\) & ind. & boolean & balance assist: off 0 or on 1 \\
+    \(j\) & ind. & integer & order number of perturbation \\
     \bottomrule
@@
 \(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
-function of our independent variables with \(\beta\) as the intercept and
-\(\alpha_k\) as the linear coefficients to the independent variables
+function of our independent variables with \(\beta\) being the intercept and
+\(\alpha_k\) the linear coefficients to the \(K\) independent variables
 \(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
@@
 
-Before fitting the model, we scale each \(x_{ij}^k\) such they they have a mean
-of zero and a standard deviation of one by cluster-mean centering, as
-recommended by \cite{Enders2007}, with clusters being an individual subject.
-The clusters are all data from an individual subject because we are only
-interested in the association between the state of the balance-assist system
-and the outcome of the perturbation. We expected there to be a variation
-between participants in how well they are able to resist the perturbation.
-However, this was not true. Cluster-mean centering shows there to be no
-variation between participants. This allowed us to stick with a simple
-single-level logistic regression model, instead of a multilevel model. This
-left us with angular impulse, perturbation order, balance assist state, and
-roll \& steer angles at the time of perturbation as independent variables. We
-also include interaction effects between the balance assist state and the other
-four variables.
-
-We divide the analysis into two separate model evaluations, one for the
-6~\si{\kph} \(g=10\) trials and one for the
-10~\si{\kph} \(g=8\) trials and we evaluate our hypothesis for
-each set of data.
+Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
+that they have a mean of zero and a standard deviation of one by cluster-mean
+centering, as recommended by \cite{Enders2007}, with clusters being an
+individual subject. The clusters are chosend as all data from an individual
+subject because we are only interested in the association between the state of
+the balance-assist system and the outcome of the perturbation. We expected
+there to be a variation between participants in how well they are able to
+resist the perturbation. However, this was not true. Cluster-mean centering
+shows there to be no variation between participants. This fact allows us to
+utilize a simple single-level logistic regression model, instead of a
+multilevel model. This left us with angular impulse, perturbation order,
+balance assist state, and roll \& steer angles at the time of perturbation as
+independent variables. We also include interaction effects between the balance
+assist state and the other four variables. We divide the analysis into two
+separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
+for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
+set of data.
 
@@
     Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
-    Angular impulse & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
-    Perturbation order & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
-    Balance assist state & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
-    Roll angle & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
-    Steer angle & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
+    Angular impulse \(L\) & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
+    Perturbation order \(j\)  & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
+    Balance assist state \(c\) & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
+    Roll angle \(\phi_0\)& -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
+    Steer angle \(\delta_0\) & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
     Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
@@
   \centering
-  \caption{Logistic regression coefficient estimates at
-    10~\unit{\kilo\meter\per\hour} and gain $g=8$.  Multiplicative change in
-    odds of predictor variables at 10 \unit{\kilo\meter\per\hour} based on
+  \caption{Logistic regression coefficient estimates at 10~\kph and gain $g=8$.
+    Multiplicative change in odds of predictor variables at 10~\kph based on
     frequentist single-level logistic regression. and 5\% confidence intervals
@@
     Intercept                                     & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
-    Angular impulse                               & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
-    Perturbation order                            & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
-    Balance-assist on                             & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
-    Roll angle                                    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
-    Steer angle                                   & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
+    Angular impulse \(L\)                         & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
+    Perturbation order \(j\)                      & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
+    Balance-assist on \(c\)                       & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
+    Roll angle \(\phi_0\)                         & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
+    Steer angle \(\delta_0\)                      & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
     Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\

### 2024-09-15 main.tex
Edited through the results.

@@
 
+% NOTE : Use as \si{\kph} or \si{\mps}
 \def\kph{\kilo\meter\per\hour}
@@
 
-\title{Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds
-When Subjected to Mechanical Perturbations}
+\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling at Low
+Speeds When Subjected to Mechanical Perturbations}
 
@@
     ``Perturbation order number'' are recorded per perturbation instance. The
-    remaining measures are time varying.}
+    remaining measures are time varying during each trial.}
   \begin{tabular}{llll}
@@
     \midrule
-    \(L\) & ind. & \si{\newton\meter\second} & angular impulse of perturbation torque \\
-    \(\delta_0\) & ind. & \si{\degree} & steer angle at start of perturbation \\
-    \(\phi_0\) & ind. & \si{\degree} & roll angle at start of perturbation \\
-    \(f\) & dep. & boolean & outcome: did not fall 0 or did fall 1 \\
-    \(c\) & ind. & boolean & balance assist: off 0 or on 1 \\
-    \(j\) & ind. & integer & order number of perturbation \\
+    \(L\)        & independent & \si{\newton\meter\second} & angular impulse of perturbation torque \\
+    \(\delta_0\) & independent & \si{\degree} & steer angle at start of perturbation \\
+    \(\phi_0\)   & independent & \si{\degree} & roll angle at start of perturbation \\
+    \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
+    \(c\)        & independent & boolean & balance assist: off 0 or on 1 \\
+    \(j\)        & independent & integer & order number of perturbation \\
     \bottomrule
@@
 %
-The coefficient estimates for a single-level logistic regression at
-6~\unit{\kph} with gain $g=10$ are shown in
+The coefficient estimates for a single-level logistic regression for the data
+from the 6~\si{\kph}, gain $g=10$ trials are shown in
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
-balance assist state are all statistically significant predictors. Larger
-angular impulse increases the probability to fall and enduring more
-perturbations or having the balance assist on, decrease the probability to
-fall.  The associated multiplicative change in odds are shown in
-Table~\ref{tab:freq-coefs-6}.
+balance assist state are all statistically significant predictors with angular
+impulse having a dominant effect. Larger angular impulse increases the
+probability to fall and enduring more perturbations or having the balance
+assist on, decrease the probability to fall. The associated multiplicative
+change in odds are shown in Table~\ref{tab:freq-coefs-6}.
 %
@@
   \centering
-  \caption{Logistic regression coefficient estimates at
-    6~\unit{\kilo\meter\per\hour} and gain $g=10$. Multiplicative change in
-    odds of predictor variables at 6 \unit{\kilo\meter\per\hour} and gain
-    $g=10$ based on frequentist single-level logistic regression.}
+  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
+    6~\si{\kph} and gain $g=10$ along with the standard error SE, p-value
+    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
+    confidence interval bounds.}
   \label{tab:freq-coefs-6}
@@
     \toprule
-    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
+    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
     \midrule
-    Intercept & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
-    Angular impulse \(L\) & 1.69 & 0.27 & 0.00 & 5.40 & 3.18 & 9.16 \\
-    Perturbation order \(j\)  & -0.77 & 0.22 & 0.00 & 0.46 & 0.30 & 0.72 \\
-    Balance assist state \(c\) & -0.64 & 0.27 & 0.02 & 0.53 & 0.31 & 0.89 \\
-    Roll angle \(\phi_0\)& -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
+    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
+    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
+    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
+    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
+    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
     Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
-    Balance-assist on $\times$ roll angle & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
-    Balance-assist on $\times$ steer angle & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
-    Balance-assist on $\times$ angular impulse & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
-    Balance-assist on $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
+    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
+    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
+    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
+    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
     \bottomrule
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kilo\meter\per\hour} with gain $g=8$ are shown in Table
-\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
-statistically significant predictors. Larger angular impulse increases the
-probability to fall and enduring more perturbations decreases the probability
-to fall.
-
+10~\si{\kph} with gain $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
+angular impulse and perturbation order are statistically significant predictors
+with angular impulse being the dominant effect. Larger angular impulse
+increases the probability to fall and enduring more perturbations decreases the
+probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
+not a significant predictor. At both speeds the angular impulse has about twice
+the effect as the perturbation order.
+%
 \begin{table}
   \centering
-  \caption{Logistic regression coefficient estimates at 10~\kph and gain $g=8$.
-    Multiplicative change in odds of predictor variables at 10~\kph based on
-    frequentist single-level logistic regression. and 5\% confidence intervals
-    of MCiO.}
+  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
+    10~\si{\kph} and gain $g=8$ along with the standard error SE, p-value
+    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
+    confidence interval bounds.}
   \label{tab:freq-coefs-10}
@@
     \toprule
-    Variable & \(\alpha_k\) & SE & \(p\)-value & $e^{\alpha_k}$ & 2.5\% & 97.5\% \\
+    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
     \midrule
     Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
-    Angular impulse \(L\)                         & 2.39  & 0.29 & 0.00 & 10.92 & 6.23 & 19.13 \\
-    Perturbation order \(j\)                      & -1.16 & 0.21 & 0.00 & 0.31 & 0.21 & 0.48 \\
+    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
+    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
     Balance-assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
@@
     Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
-    Balance-assist on $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
-    Balance-assist on $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
-    Balance-assist on $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
-    Balance-assist on $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
+    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
+    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
+    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
+    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
     \bottomrule
@@
 Turning the balance-assist system on significantly \((p<0.05)\) reduces the
-odds that a pertubation results in a fall occurs while cycling at a speed of
+odds that a pertubation results in a fall while cycling at a speed of
 6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
 falling as a function of the mean and centered angular impulse per participant
-while keeping all other explantory factors at their centered mean values. This
-figure is derived by setting all explanatory variables to there mean and
-calculating the probality from Eq~\ref{eq:log-regress} for only change in
-angular impulse given the estimates in Table~\ref{tab:freq-coefs-6} which
-indicate that the balance-assist system halves (0.53) the odds that a
-perturbation results in a fall. This figure shows that for relatively large
-impulses the probability to fall is unity for both states of balance assist on
-and off. And for relatively small impulses the probability to fall is zero for
-both states. But for impulses in the magnitude region (-1 to 0.5 STD) around
-the mean-centered fall threshold, the probability of falling is significantly
-lowered. The skewedness of the probaility curves arrives from the interaction
-effects\todo[inline]{check this and think about it}.
-
-\todo{Discuss the 10 kph finding}
-
-but at 10~\si{\kph} that is not the case \((p=0.07)\).
-
+for the balance assist state on and off while keeping all other explanatory
+factors at their centered mean values. This figure is created by setting all
+explanatory variables to there mean and calculating the probality from
+Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
+in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
+system halves (0.53) the odds that a perturbation results in a fall. This
+figure shows that for relatively large impulses the probability to fall is
+unity for both states of balance assist on and off. And for relatively small
+impulses the probability to fall is null for both states. But for impulses in
+the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
+threshold, the probability of falling is significantly lowered with the balance
+assist system on. The skewedness of the probaility curves arrives from the
+interaction effects.\todo{check this statement about skewedness this and think
+about it} Figure~\ref{fig:probablity-10kph} shows the same result for the
+10~\kph trials which has a similar trend of reducing the probability to fall
+with the balance assist system turned on, but the effect is not significant.
+%
 \begin{figure}
   \centering
-  \subcaptionbox{6~\si{\kph}\label{fig:probability-6kph}}{
-    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
+  % NOTE : We could uncomment the longer subcaption if titles removed from
+  % figures. It is currently redundant.
+  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
+    \label{fig:probability-6kph}}{
+    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
   }
-  \subcaptionbox{2.8~\si{\mps}}{
-    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
+  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
+    \label{fig:probability-10kph}}{
+    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
   }
-  \caption{Comparison of predicted fall probability for balance-assist on or
-    off All other predictor variables are fixed at zero, except for the
-    interaction effect between the balance-assist and the angular impulse. The
-    x axis is standard deviation around the mean (all perturbations), centered
-    per participant.}
+  \caption{Comparison of predicted fall probability for balance-assist on
+    (blue) or off (red) when all other predictor variables are fixed at zero,
+    except for the interaction effect between the balance-assist and the
+    angular impulse. The ordinate axis is standard deviation around the mean
+    (all perturbations), centered per participant.}
   \label{fig:probability}

### 2024-09-16 main.tex
Started work on the discussion section.

@@
 \usepackage[natbib=true,style=authoryear]{biblatex}
+% NOTE : this file is automatically generated from Zotero, do not edit
+% manually!
 \addbibresource{references.bib}
@@
   T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
+  \label{eq:implemented-controller}
 \end{align}
@@
 interaction effects.\todo{check this statement about skewedness this and think
-about it} Figure~\ref{fig:probablity-10kph} shows the same result for the
-10~\kph trials which has a similar trend of reducing the probability to fall
-with the balance assist system turned on, but the effect is not significant.
+about it} Figure~\ref{fig:probability-10kph} shows the same result for the
+10~\si{\kph} trials which has a similar trend of reducing the probability to
+fall with the balance assist system turned on, but the effect is not
+significant.
 %
@@
 %
+We have shown that at 6~\si{\kph} the addition of balance assist control
+reduces the chance that a rider will fall when perturbed around the limits of their
+control authority. But this effect diminishes at the higher speed
+scenario of 10~\si{\kph}. We were only able to test these two speed-gain
+scenarios for mostly homogeneous sets of riders within the resources of this
+research project, but additional experimental work could help understand more
+completely the range and limits of the positive effect of the balance system.
+For example, it is possible that simply increasing the controller gain at
+10~\si{\kph} also results in a significant positive effect.
+
+\subsection{Interpretation of the Results}
+%
 The probability that a fall occurs depends on the values of all the independent
@@
 The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
-system halves the odds that a perturbation results in a fall
-(MCiO=0.53)~\todo{Make a variable for MCiO}.
+system halves the odds that a perturbation results in a fall:
+\(e^{\alpha_k}=0.53\).
 This means if the odds are a 1000:1, turning on the balance-assist system
 reduces the odds to 500:1.
-However, the probability that a fall occurs is only reduced from $0.999$ to
-$0.998$ in that case.
+However, in that case the probability that a fall occurs is only reduced from
+$0.999$ to $0.998$.
 If the odds that a fall will occur are smaller, halving the odds has a larger
@@
 
-\todo{Consider making a specific value of calcualting the probabilty for a set
-of inputs to the regression model.}
-
-To illustrate the effect of the balance-assist system on fall probability, we will
-give an example of how the data collected during the experiments is used to predict
-fall probability. We use \ref{eq:log-regress} and the coefficients in 
-\ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are not included.
-Let's assume that the mean angular impulse $\bar{L}$ of all the perturbations applied to 
-a participant is 100~\si{\newton}, and the standard deviation $\sigma^{L}=15$. The centred
-and scaled angular impulse can be calculated by substracting $\bar{L}$ from the applied
-angular impulse $L$, and dividing this by $\sigma^{L}$. The same applies for the 
-perturbation order $c$, initial roll angle $\phi_0$ and initial steer angle $\delta_0$.
-If we take the coefficients estimated for cycling at 6~\si{\kilo\meter\per\hour}, the
-log-odds of falling can be calculated as follows.
-
+To illustrate the effect of the balance-assist system on fall probability, we
+will give an example of how the data collected during the experiments is used
+to predict fall probability. We use \ref{eq:log-regress} and the coefficients
+in \ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are
+not included.  Let's assume that the mean angular impulse $\bar{L}$ of all the
+perturbations applied to a participant is 100~\si{\newton}, and the standard
+deviation $\sigma^{L}=15$. The centred and scaled angular impulse can be
+calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
+dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
+initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
+the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
+can be calculated as follows.
+%
 \begin{align}
-	\log \left( \frac{p_{ij}}{1-p_{ij}} \right) & = \beta + \sum_{k}^{k=0}\alpha_k 
-	\frac{x_{ij}^{k}-\bar{x_{ij}^{k}}} {\sigma^{x^{k}}} 
-	& = -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} 
-	-0.77\cdot\frac{10-20}{11.54} 
-	-0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}} 
-	-0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}}
-	-0.64\cdot s
-	& = 1.42 -0.64 \cdot s
+  \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
+  =
+  \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}}
+  =
+  & -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}}
+    -0.77\cdot\frac{10-20}{11.54}  \\
+  & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
+    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
+  =
+  & 1.42 -0.64 \cdot s
 \end{align}
 
-The state of the balance-assist $s$ is a binary variable. If the balance-assist is turned on, 
-the log-odds that a fall occurs are decreased by 0.64. The odds and probability can be calculated:
+The state of the balance-assist $s$ is a binary variable. If the balance-assist
+is turned on, the log-odds that a fall occurs are decreased by 0.64. The odds
+and probability can be calculated:
 
@@
 
-Turning on the balance-assist system reduces the probability that the perturbation results in a 
-fall from 0.81 to 0.69.
+Turning on the balance-assist system reduces the probability that the
+perturbation results in a fall from 0.81 to 0.69.
 
+\subsection{Stability and Human Controlled Plant Dynamics}
+%
+The linear Carvallo-Whipple model indicates that the steer controller
+stabilizes the bicycle-rider system, but this model assumes the rider's hands
+are not connected to the handlebars and that they clamp their body as rigidly
+as possible to the rear frame. In reality, the system's behavior is likely more
+akin to a marginally stable or an easily controllable unstable system due to
+the various un-modeled effects. Our system may not result in a definitely
+stable system, i.e. cannot fall, but having plant eigenvalues with very small
+unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}.
+
+The controller design we utilize,
+Equation~\ref{eq:implemented-controller}, also increases the weave mode
+frequency by a factor of about three up to about 1~\si{\hertz}. This bandwidth
+is still controllable by the human's neuromuscular system, but may feel
+unnatural as it is more akin to what the steering would feel like at in the
+\SIrange{30}{40}{\kph} speed range. \Citet{Hanakam2023} reported
+dissatisfaction in subjective rider feeling on their similar bicycle to ours
+and this effect to the human-controlled plant dynamics could be connected to
+this.
+
+\subsection{Treadmill Width}
+%
 Angular impulse magnitude has the largest significant effect for predicting
-fall probability, Tables~\ref{tab:freq-coefs-6} and \ref{tab:freq-coefs-10}.
-An increase in angular impulse increases the fall probability both at 6 and
-10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
-approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
-important predictor at higher speeds compared to lower speeds. The reason for
-this likely has to do with the width of the treadmill and may also be why
-balance assist system did not have a statistically significant effect at
-10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
-causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
-the bicycle exiting the maximum width of the treadmill. If the same experiment
-was performed on an infinite plane, the riders may have recovered from more
-perturbations. At 6~\si{\kph} the riders could often recover in the allotted
-treadmill width. Our results are very much dependent on the two modes of
+fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
+\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
+probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
+change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
+impulse is a more important predictor at higher speeds compared to lower
+speeds. We posit that this likely has to do with the width of the treadmill and
+that this could also be why balance assist system did not have a statistically
+significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
+same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
+almost all falls were due to the bicycle exiting the maximum width of the
+treadmill. If the same experiment was performed on an infinitely wide plane,
+the riders may have recovered from more perturbations. At 6~\si{\kph} the
+riders could often recover in the allotted treadmill width due to the smaller
+lateral deviations. Our results are very much dependent on the two modes of
 falling with use: exit the treadmill width or foot is placed on the belt. Cycle
@@
 
+\subsection{Learning Effect}
+%
 An increase in the number of perturbation that a participant already
@@
 the speed.
+
+\subsection{Non-significant Predictors}
 
@@
 
-Difficult to layer this onto the fall statistics, because we don't have the
-details of how people fall. If such data existed we could make estimates on the
-number of falls reduced if everyone road such a bike.
+\subsection{Extrapolation to Natural Falls}
+%
+The positive effect of the balance assist system is coupled to the assumptions
+and experimental scenarios we implemented and there is unfortunately no simple
+way to extrapolate our results to reductions of single-actor crashes we may see
+if such a system were deployed widely to bicyclists. Although, our results do
+indicate that we would see such a reduction, even if only in a class of
+single-actor crashes that most resemble our experimental design. If there were
+more comprehensive and detailed natural data of how people fall we could make
+estimates on the number of falls reduced if everyone road a balance assist
+bicycle.
 
@@
 
-% NOTE : this file is automatically generated from Zotero, do not edit
-% manually!
 \printbibliography

### 2024-09-16 main.tex
First full draft of the paper done, V1.

@@
 \usepackage{subcaption}  % for subfigures
-\usepackage{todonotes}
+\usepackage[disable]{todonotes}
 
@@
 
-\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling at Low
-Speeds When Subjected to Mechanical Perturbations}
+\title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
+Low Speeds When Subjected to Handlebar Perturbations\\Version 1}
 
@@
   Uncontrolled bicycles are generally unstable at low speeds. We add a
-  controlled steering motor to a consumer bicycle that stabilizes the bicycle
+  controlled steering motor to a consumer electric bicycle that stabilizes it
   at low speeds. To test the motor's assistance during falls, we apply varying
@@
   probability of not recovering from a handlebar perturbation decreases when
-  the balance assist is activated at a speed of 6~\si{\kph}.
+  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
+  balance assist system in real world bicycling will reduce the number of falls
+  that occur near riders' control authority limits.
 }
@@
 hitting a bump in the road. To assess our balance assisting bicycle, we subject
-the rider to perturbations at the handledbar, which can relatively easily cause
+the rider to perturbations at the handlebar, which can relatively easily cause
 a rider to fall.
@@
 
-At the initiation of each perturbation we log the instananeous steer and roll
+At the initiation of each perturbation we log the instantaneous steer and roll
 angles to characterize the configuration of the bicycle when perturbed. The
@@
 centering, as recommended by \cite{Enders2007}, with clusters being an
-individual subject. The clusters are chosend as all data from an individual
+individual subject. The clusters are chosen as all data from an individual
 subject because we are only interested in the association between the state of
@@
 Turning the balance-assist system on significantly \((p<0.05)\) reduces the
-odds that a pertubation results in a fall while cycling at a speed of
+odds that a perturbation results in a fall while cycling at a speed of
 6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
@@
 factors at their centered mean values. This figure is created by setting all
-explanatory variables to there mean and calculating the probality from
+explanatory variables to there mean and calculating the probability from
 Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
@@
 threshold, the probability of falling is significantly lowered with the balance
-assist system on. The skewedness of the probaility curves arrives from the
+assist system on. The skewness of the probability curves arrives from the
 interaction effects.\todo{check this statement about skewedness this and think
@@
     \label{fig:probability-6kph}}{
-    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
+    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
   }
@@
     \label{fig:probability-10kph}}{
-    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
+    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
   }
@@
 %
-We have shown that at 6~\si{\kph} the addition of balance assist control
-reduces the chance that a rider will fall when perturbed around the limits of their
-control authority. But this effect diminishes at the higher speed
-scenario of 10~\si{\kph}. We were only able to test these two speed-gain
-scenarios for mostly homogeneous sets of riders within the resources of this
-research project, but additional experimental work could help understand more
-completely the range and limits of the positive effect of the balance system.
-For example, it is possible that simply increasing the controller gain at
-10~\si{\kph} also results in a significant positive effect.
-
-\subsection{Interpretation of the Results}
-%
-The probability that a fall occurs depends on the values of all the independent
-variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
-effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
-some insight.
-But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
-\ref{tab:freq-coefs-10} it is important to understand the relationship between
-probability and odds.
-The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
-system halves the odds that a perturbation results in a fall:
-\(e^{\alpha_k}=0.53\).
-This means if the odds are a 1000:1, turning on the balance-assist system
-reduces the odds to 500:1.
-However, in that case the probability that a fall occurs is only reduced from
-$0.999$ to $0.998$.
-If the odds that a fall will occur are smaller, halving the odds has a larger
-influence on the fall probability.
-For example, if the odds that a fall occurs is two, halving it to one reduces
-the probability from $0.66$ to $0.50$.
-This can be seen in Figure~\ref{fig:probability} which shows the fall
-probability as a function of the magnitude of the normalized angular impulse
-for both speeds.
-The difference in probability between the balance-assist on/off case becomes
-insignificant outside of approximately \(\pm1\) standard deviation of the
-average angular impulse.
-This means that the balance assist is most effective for perturbations close to
-the subject's personal threshold between falling or recovery and that large
-perturbations will make you fall regardless of the balance assist's help.
-
-To illustrate the effect of the balance-assist system on fall probability, we
-will give an example of how the data collected during the experiments is used
-to predict fall probability. We use \ref{eq:log-regress} and the coefficients
-in \ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are
-not included.  Let's assume that the mean angular impulse $\bar{L}$ of all the
-perturbations applied to a participant is 100~\si{\newton}, and the standard
-deviation $\sigma^{L}=15$. The centred and scaled angular impulse can be
-calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
-dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
-initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
-the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
-can be calculated as follows.
-%
-\begin{align}
-  \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
-  =
-  \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}}
-  =
-  & -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}}
-    -0.77\cdot\frac{10-20}{11.54}  \\
-  & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
-    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
-  =
-  & 1.42 -0.64 \cdot s
-\end{align}
-
-The state of the balance-assist $s$ is a binary variable. If the balance-assist
-is turned on, the log-odds that a fall occurs are decreased by 0.64. The odds
-and probability can be calculated:
-
-\begin{align}
-  \frac{p_{ij}}{1-p_{ij}} = e^{1.42-0.64s} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s
-\end{align}
-
-\begin{align}
-  p_{ij}^{s=0} = \frac{4.14}{1 + 4.14} = 0.81
-\end{align}
-
-\begin{align}
-  p_{ij}^{s=1} = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}]} = 0.69
-\end{align}
-
-Turning on the balance-assist system reduces the probability that the
-perturbation results in a fall from 0.81 to 0.69.
+We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
+control reduces the chance that a rider will fall when perturbed around the
+limits of their control authority. But this effect diminishes just below
+significance at the higher speed scenario of 10~\si{\kph}. We were only able to
+test these two speed-gain scenarios for mostly homogeneous sets of riders
+within the resources of this research project, but additional experimental work
+could help understand more completely the range and limits of the positive
+effect of the balance system. For example, it is possible that simply
+increasing the controller gain at 10~\si{\kph} also results in a significant
+positive effect. A longitudinal study of normal use of the balance assist
+bicycle compared to a control group could provide the strongest evidence of any
+benefit we have seen in this more narrow scenario.
 
@@
 
+\subsection{Application of the Logistic Model}
+%
+The probability that a fall occurs depends on the values of all the independent
+variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
+effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
+insight.
+But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
+\ref{tab:freq-coefs-10} it is important to understand the relationship between
+probability and odds.
+The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
+system halves the odds that a perturbation results in a fall:
+\(e^{\alpha_k}=0.53\).
+This means if the odds are a 1000:1, turning on the balance-assist system
+reduces the odds to 500:1.
+However, that case the probability that a fall occurs is only reduced from
+$0.999$ to $0.998$.
+If the odds that a fall will occur are smaller, halving the odds has a larger
+influence on the fall probability.
+For example, if the odds that a fall occurs is two, halving it to one reduces
+the probability from $0.66$ to $0.50$.
+This can be seen in Figure~\ref{fig:probability} which shows the fall
+probability as a function of the magnitude of the normalized angular impulse
+for both speeds.
+The difference in probability between the balance-assist on/off case becomes
+insignificant outside of approximately \(\pm1\) standard deviation of the
+average angular impulse the rider was subjected to.
+This means that the balance assist is most effective for perturbations close to
+the subject's personal threshold between falling or recovery and that large
+perturbations will make you fall regardless of the balance assist's help.
+
+To illustrate the effect of the balance-assist system on fall probability, we
+will give an example of how the data collected during the experiments is used
+to predict fall probability. We use Equation~\ref{eq:log-regress} and the coefficients
+in \ref{tab:freq-coefs-6}. For simplicity's sake, the interaction effects are
+not included. For example, assuming that the mean angular impulse $\bar{L}$ of all the
+perturbations applied to a participant is 100~\si{\newton}, and the standard
+deviation $\sigma^{L}=15$. The centered and scaled angular impulse can be
+calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
+dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
+initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
+the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
+can be calculated as follows.
+%
+\begin{equation}
+  \begin{split}
+    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
+    = & \quad
+    \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}} \\
+    = &
+   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
+    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
+    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot c \\
+    = &
+    1.42 -0.64 \cdot c
+  \end{split}
+\end{equation}
+%
+The state of the balance-assist \(s\) is a binary variable. If the
+balance-assist is turned on, the log-odds that a fall occurs are decreased by
+0.64. The odds and probability can be calculated:
+%
+\begin{align}
+  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64c} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53c \\
+  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
+  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
+\end{align}
+%
+Turning on the balance-assist system reduces the probability that the
+perturbation results in a fall from 81\% to 69\%.
+
+This illustration alludes to the difficulty needed to apply the model in way
+that could predict how many falls may be averted in a natural setting if the
+balance assist system is used. Estimates of the predictor variables extracted
+from the limited data collected from natural cycling would be needed to
+populate the model.
+
 \subsection{Treadmill Width}
@@
 speeds. We posit that this likely has to do with the width of the treadmill and
-that this could also be why balance assist system did not have a statistically
+that this could also be why the balance assist system did not have a statistically
 significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
@@
 riders could often recover in the allotted treadmill width due to the smaller
-lateral deviations. Our results are very much dependent on the two modes of
-falling with use: exit the treadmill width or foot is placed on the belt. Cycle
-paths are a similar width as the treadmill, so rider's are often limited in
-width when recovering from a fall.
+lateral deviations. We believe our results are very much dependent on the two
+modes of falling with use: exit the treadmill width or foot is placed on the
+belt. Cycle paths are a similar width as the treadmill, so rider's are often
+limited in width when recovering from a fall.
 
@@
 
-Roll and steer angle are not a significant predictor of fall probability,
-neither at 6 or at 10~\si{\kph}. We expected this to have an effect. If you are
-in a rolled and steered state that is far from the upright equilibrium, then a
-perturbation that further pushes you from the equilibrium should have some
-additive or multiplicative effect on the resulting motion trajectory.
+Roll and steer angle are not a significant predictor of fall probability at 6
+or at 10~\si{\kph}. We expected these variables to have a significant effect
+based on the following reasoning. If you are in a rolled and steered state that
+is far from the upright equilibrium, then a perturbation that further pushes
+you from the equilibrium should have some additive or multiplicative effect on
+the resulting motion trajectory.
 
@@
 %
-TODO
+Automatically controlling a steering motor in a bicycle using roll rate
+feedback lowers the speed at which the bicycle is stable. This makes the
+bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
+which is easier for a rider to balance. Perturbation forces applied to the
+handlebar can cause a rider to fall and every rider has a threshold force at
+which they are more likely to fall than not. The probability of falling when
+mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
+when the balance assisting control is activated. This effect is present when
+traveling at 10~\si{\kph} but more investigation is needed to determine if the
+effect can be significant. The positive effect to balance is rider independent
+and most effective in the regime of perturbations near the rider's control
+authority threshold. Given that similar effects cause falls during bicycling,
+use of the balance assist system in real world use cases will reduce the number
+of falls at low bicycling speeds.
+
+\section*{Acknowledgements}
+%
+This study follows and draws from experimental and analysis methods originally
+developed in unpublished research by Marco Reijne. The authors acknowledge
+Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
+Reijne, and Arend Schwab for contributions to the development of the balance
+assist bicycle.
+
+\section*{Funding}
+%
+This study is funded by Dutch Research Council, Nederlandse Organisatie voor
+Wetenschappelijk Onderzoek (NWO), under the Citius Altius Sanius program and in
+collaboration with Bosch eBike Systems and Royal Dutch Gazelle. The funders had
+no role in the data collection and analysis or preparation of the manuscript.
 

### 2024-09-17 main.tex
Bump to v2.dev.
Show todo notes, move two to issues.

@@
 \usepackage{subcaption}  % for subfigures
-\usepackage[disable]{todonotes}
+\usepackage[]{todonotes}  % use [disable] to hide notes
 
@@
 \title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
-Low Speeds When Subjected to Handlebar Perturbations\\Version 1}
+Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev}
 
@@
 Figure~\ref{fig:eigenvalues}.
-\todo[inline]{Could add the experimental weave mode identification to this
-figure and consider lowering the gain on the model to make the model match the
-experimental values.}
 %
@@
 We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments. The subjects were all confident in their cycling skills and
-had cycled at least once in the last month. Eleven subjects performed the
-experiments at 6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and
-fifteen subjects performed the experiments at 10~\si{\kilo\meter\per\hour}
+the experiments.
+\todo[]{Include the average age, mass, height and standard deviation if we
+recorded those.}
+The subjects were all confident in their cycling skills and had cycled at least
+once in the last month. Eleven subjects performed the experiments at
+6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and fifteen subjects
+performed the experiments at 10~\si{\kilo\meter\per\hour}
 (2.8~\si{\meter\per\second}). All subjects consented to the experiment and
@@
 University of Technology's Human Research Ethics Council (\#3897).
-
-\todo[inline]{Include the average age, mass, height and standard deviation if
-we recorded those.}
 
@@
 
-\todo[inline]{Could show an impulse response in lateral deviation for different
-speeds.}
-
 \subsection{Learning Effect}

### 2024-09-17 main.tex
Add acknowledgement to the helpers in the experiments.
Corrected the bicycle name, fixes #7.

@@
 %
-We modified an electric bicycle (Grenoble Arroyo, Royal Dutch Gazelle, Dieren,
+We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
 The Netherlands) with a custom motor in the steering assembly capable of
@@
 Reijne, and Arend Schwab for contributions to the development of the balance
-assist bicycle.
+assist bicycle. We thank all of the assistants that helped run the experiments
+and the subjects for their time an enthusiasm.
 

### 2024-09-17 main.tex
Added equipment specs, fixes #10.

@@
 converts the commanded torque to applied torque. We measure the speed of the
-rear wheel with an encoder (Magura XXX) and measure the body fixed roll rate of
-the bicycle with a MEMs rate gyroscope (BN0086, Sparkfun, Niwot, USA). The
-balance assist control algorithm is implemented on a microprocessor (Teensy,
-PJRC, USA) and data from all sensors is logged with a CAN bus (CanEdge2, CSS
-Electronics, Denmark) at at least 100 Hz.
+rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
+and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
+(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
+implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
+is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at at
+least 100 Hz.
 
@@
 applied by each motor at the handlebar via four inline 250~\si{\newton} S-style
-load cells except the left rear load cell that has a max load of
-500~\si{\newton}. The commanded force profiles are designed to apply an
-external pulsive torque to the front assembly (handlebars, forks, wheel) at
-magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors
-(EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners of a
-1~\si{\meter} wide treadmill that can reach speeds of
-18~\si{\kilo\meter\per\hour}. The general base design of this system is
-described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
-the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
-Our modifications relative to Van De Velde include simplifying the controller
-with a inexpensive microcontroller and the use of a simpler non-active safety
-harness.
+load cells (Bosche GmbH, Damme, Germany) except the left rear load cell that
+has a max load of 500~\si{\newton}. The commanded force profiles are designed
+to apply an external pulsive torque to the front assembly (handlebars, forks,
+wheel) at magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
+motors (EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners
+of a 1~\si{\meter} wide treadmill (Bonte Technology B.V., Zwolle, The
+Netherlands) that can reach speeds of 18~\si{\kilo\meter\per\hour}. The general
+design of this system is described in detail in Van De Velde's MSc
+thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
+Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
+Velde include simplifying the Bump'Em motor controller with an inexpensive
+microcontroller and the use of a simpler non-active safety harness.
 %

### 2024-09-17 main.tex
Addressed Peter's feedback, fixes #17.

@@
 and since the 1980s it has been known that applying a motor actuated steering
-torque proportional to the vehicle's roll rate can stabilize a single-track
-vehicle at lower speeds. If automatic control of steering can stabilize a
-bicycle, it may reduce the control required from the rider to successfully
-manage balancing tasks much like the high speed self-stability already does. We
-have developed a balance assisting bicycle,
+torque proportional to the vehicle's roll rate~\footnote{Roll angular rate is
+defined herein as per the SAE J670 vehicle standards.} can stabilize a
+single-track vehicle at lower speeds. If automatic control of steering can
+stabilize a bicycle, it may reduce the control required from the rider to
+successfully manage balancing tasks much like the high speed self-stability
+already does. We have developed a balance assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
-mode, and steer angular rate feedback stabilizes the wobble mode. They also
-proposed how the gains must change with respect to vehicle speed for favorable
-control at all speeds. This roll motion feedback enables the simplest controller
-that can stabilize a single-track vehicle above a minimum speed when one is not
-concerned with wobble instabilities. Ruijs and Pacejka's work was not
-particularly concerned with low speed stability and their vehicle was fully
-automatic, i.e no human rider was involved.
+mode, and steer angular rate feedback stabilizes the wobble mode~\footnote{The
+motorcycle (and bicycle) eigenmodes are defined in \citep{Sharp1971}.}. They
+also proposed how the gains must change with respect to vehicle speed for
+favorable control at all speeds.  This roll motion feedback enables the
+simplest controller that can stabilize a single-track vehicle above a minimum
+speed when one is not concerned with wobble instabilities. Ruijs and Pacejka's
+work was not particularly concerned with low speed stability and their vehicle
+was fully automatic, i.e no human rider was involved.
 
@@
 self-stability. The bicycle model is suitable for showing the effect of a roll
-motion feedback driven steer motor on the dynamics. The linear version of this
-model~\citep{Meijaard2007} can be described by the fourth order state space
-equations:
+motion feedback driven steer motor on the dynamics. This model is equivalently
+valid for on-road or treadmill motion~\citep{Kooijman2009a}, which have the
+same dynamics. The linear version of this model~\citep{Meijaard2007} can be
+described by the fourth order state space equations:
 %
@@
     handlebars.\protect\label{fig:participant-in-set-up}}
-    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
+    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
   \subcaptionbox{Top view diagram of bicycle handlebar perturbation
     system.\protect\label{fig:setup-diagram}}
-    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
+    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
   \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
@@
 recovery from the perturbation is successful if the person neither places their
-foot down onto the treadmill surface or their wheel of the bicycle exits the
+foot down onto the treadmill surface nor their wheel of the bicycle exits the
 width of the treadmill belt. We record this as a binary variable \(f\) for

### 2024-09-17 src/generate_time_series_imgs.py
Some improvements to the time series figure, added speed.

@@ -112,12 +112,14 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
         if i == 10:  # only plot the figure we will use
             # print(df.columns.values)
             if "motor_current" in df.columns.values:
-                fig, axs = plt.subplots(4, 1, sharex=True, layout="constrained")
+                fig, axs = plt.subplots(5, 1, sharex=True,
+                                        layout="constrained",
+                                        figsize=(140.0/25.4, 140/25.4))
                 df["motor_torque"] = df["motor_current"] / BALANCE_ASSIST_MOTOR_CONSTANT
             else:
-                fig, axs = plt.subplots(3, 1, sharex=True, layout="constrained")
-
-            fig.set_size_inches(140/25.4, 100/25.4)
+                fig, axs = plt.subplots(5, 1, sharex=True,
+                                        layout="constrained",
+                                        figsize=(140.0/25.4, 140/25.4))
 
             actual_torque, desired_torque = calculate_torque_on_handlebars(df)
             axs[0].plot(
@@ -134,7 +136,7 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 color='black',
                 linestyle='--',
             )
-            axs[0].set_ylabel("Bump'Em\nHandlebar Torque\n[Nm]")
+            axs[0].set_ylabel("Bump'Em\nHandlebar\nTorque\n[Nm]")
             axs[0].legend(fontsize="x-small")
 
             df.plot(
@@ -164,26 +166,38 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 color='black',
             )
             if "motor_current" in df.columns.values:
+                com_torque = -30.35*np.deg2rad(df['roll_rate'])
+                axs[3].plot(df['seconds_since_start'], com_torque,
+                            color='black', alpha=0.25)
+                axs[3].axhline(7.0, color='grey', linestyle='--')
+                axs[3].axhline(-7.0, color='grey', linestyle='--')
                 df.plot(
                     x="seconds_since_start",
                     y="motor_torque",
                     ax=axs[3],
-                    xlabel="Time since start of perturbation [s]",
                     ylabel="Commanded\nMotor Torque\n[Nm]",
                     legend=False,
                     color='black',
                 )
-                com_torque = -30.35*np.deg2rad(df['roll_rate'])
-                axs[3].plot(df['seconds_since_start'], com_torque, color='blue')
-                axs[3].axhline(7.0, color='black', linestyle='--')
[…]

### 2024-09-17 src/generate_time_series_imgs.py
Added average speed, still needs more work.

@@ -182,6 +182,7 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
             axs[4].axhline(1.7, color='black', linestyle='--')
             avg_speed = df['speed'].mean()
             std_speed = df['speed'].std()
+            axs[4].axhline(avg_speed, color='black')
             df.plot(
                 x="seconds_since_start",
                 y="speed",

### 2024-09-17 src/generate_time_series_imgs.py
Made time series plot b&w with all labels. Fixes #14

@@ -166,32 +166,45 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 color='black',
             )
             if "motor_current" in df.columns.values:
-                com_torque = -30.35*np.deg2rad(df['roll_rate'])
-                axs[3].plot(df['seconds_since_start'], com_torque,
-                            color='black', alpha=0.25)
-                axs[3].axhline(7.0, color='grey', linestyle='--')
-                axs[3].axhline(-7.0, color='grey', linestyle='--')
+                #com_torque = -30.35*np.deg2rad(df['roll_rate'])
+                #axs[3].plot(df['seconds_since_start'], com_torque,
+                            #color='black', alpha=0.25)
+                axs[3].axhline(7.0, color='grey', linestyle='--',
+                               label='Maximum Torque')
+                axs[3].axhline(-7.0, color='grey', linestyle='--',
+                               label='__none__')
                 df.plot(
                     x="seconds_since_start",
                     y="motor_torque",
                     ax=axs[3],
                     ylabel="Commanded\nMotor Torque\n[Nm]",
+                    label='__none__',
                     legend=False,
                     color='black',
                 )
-            axs[4].axhline(1.7, color='black', linestyle='--')
+                axs[3].legend(fontsize="x-small")
+            # 1.7 m/s is 6 km/h
+            #axs[4].axhline(1.7, color='black', linestyle='--')
             avg_speed = df['speed'].mean()
             std_speed = df['speed'].std()
-            axs[4].axhline(avg_speed, color='black')
+            axs[4].axhline(avg_speed, color='black', linestyle='--',
+                           label='Mean Speed')
+            axs[4].axhspan(
+                avg_speed - std_speed,
+                avg_speed + std_speed,
+                color='black',
+                alpha=0.2, label='Standard Deviation')
             df.plot(
                 x="seconds_since_start",
                 y="speed",
                 ax=axs[4],
                 xlabel="Time since start of perturbation [s]",
                 ylabel="Speed\n[m/s]",
+                label='__none__',
                 legend=False,
                 color='black',
             )
+            axs[4].legend(fontsize="x-small")
 
             for ax in axs:
                 ax.grid()

### 2024-09-17 main.tex
Made time series plot b&w with all labels. Fixes #14
Improve eigenvalue plots, fixes #12.

@@
     for the bicycle without the inertial effect of a rigid rider (left) and
-    with a rigid rider (right). Vertical blue lines indicate the two speeds we
-    perform experiments at. Grey shaded region is the linear stable speed
-    range. Geometric and inertial parameters for these plots were estimated
-    with the methods presented in \cite{Moore2012} and software packages
-    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
+    with a rigid rider (right). Vertical dotted and dotted-dashed lines
+    indicate the two speeds we perform experiments at. Grey shaded region is
+    the linear stable speed range. Geometric and inertial parameters for these
+    plots were estimated with the methods presented in \cite{Moore2012} and
+    software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
+    1.5.0~\citep{Dembia2015}.}
   \label{fig:eigenvalues}
@@
   \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
-  \caption{Motor force and resulting  steer torque perturbation based on an 110N
-    counterclockwise applied force.}
+  \caption{Perturbation torque alongside the resulting motion and steer motor
+    torque based on a 110~\si{\newton} counterclockwise applied force at a a
+    6~\si{\kph} travel speed.}
   \label{fig:perturbation_0}

### 2024-09-17 src/control.py

@@ -75,8 +75,8 @@ def plot_eig(ax, model, kphidots=0.0):
     print(msg.format(weave_speed, weave_speed*MPS2KPH))
     msg = 'Capsize speed: {:1.2f} [m/s], {:1.1f} [km/h]'
     print(msg.format(capsize_speed, capsize_speed*MPS2KPH))
-    ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0)
-    ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0)
+    ax.axvline(6.0*KPH2MPS, ymin=-10.0, ymax=10.0, color='black', linestyle=':')
+    ax.axvline(10.0*KPH2MPS, ymin=-10.0, ymax=10.0, color='black', linestyle='-.')
     model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
                                 hide_zeros=True, colors=['k']*4)
     ax.set_ylim((-10, 10))
@@ -84,15 +84,21 @@ def plot_eig(ax, model, kphidots=0.0):
 
 
 ax = plot_eig(axes[0, 0], model_without)
+sax = ax.secondary_xaxis('top', functions=(lambda x: x*MPS2KPH,
+                                           lambda x: x*KPH2MPS))
+sax.set_xlabel('Speed [km/h]')
 ax.set_title('Without Rigid Rider')
-ax.set_ylabel('Balance Assist Off\nEigenvalue Components [1/s]')
+ax.set_ylabel('Balance Assist Off\nEigenvalue Components\n[1/s]')
 ax.set_xlabel('')
 
 ax = plot_eig(axes[1, 0], model_without, kphidots=kphidots)
-ax.set_ylabel('Balance Assist On\nEigenvalue Components [1/s]')
+ax.set_ylabel('Balance Assist On\nEigenvalue Components\n[1/s]')
 ax.set_xlabel('Speed [m/s]')
 
 ax = plot_eig(axes[0, 1], model_with)
+sax = ax.secondary_xaxis('top', functions=(lambda x: x*MPS2KPH,
+                                           lambda x: x*KPH2MPS))
+sax.set_xlabel('Speed [km/h]')
 ax.set_title('With Rigid Rider')
 ax.set_ylabel('')
 ax.set_xlabel('')

### 2024-09-18 src/statistics.r
Make r plots smaller.

@@ -49,10 +49,12 @@ for (i in 1:2) {
   )
 
   if (i == 1) {
-    fig_title <- "Predicted fall probability at 6 km/h with gain g=-10"
+    #fig_title <- "Predicted fall probability at 6 km/h with gain g=-10"
+    fig_title <- ""
     file_name <- "./figures/predicted_fall_probability_6kmh.png"
   } else {
-    fig_title <- "Predicted fall probability at 10 km/h with gain g=-8"
+    #fig_title <- "Predicted fall probability at 10 km/h with gain g=-8"
+    fig_title <- ""
     file_name <- "./figures/predicted_fall_probability_10kmh.png"
   }
 
@@ -66,5 +68,5 @@ for (i in 1:2) {
       theme_bw() +
       labs(title = fig_title, x = "Centred and scaled angular impulse", y = "Fall probability") +
       theme(plot.title = element_text(hjust = 0.5))
-  ggsave(file = file_name, width = 8, height = 5, dpi = 300)
+  ggsave(file = file_name, width = 100/25.4, height = 75/25.4, dpi = 300)
 }

### 2024-09-19 main.tex
Make R figures black and white and better size and legend.

@@
     \label{fig:probability-6kph}}{
-    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
+    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
   }
@@
     \label{fig:probability-10kph}}{
-    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
+    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
   }

### 2024-09-19 src/statistics.r

@@ -49,24 +49,43 @@ for (i in 1:2) {
   )
 
   if (i == 1) {
-    #fig_title <- "Predicted fall probability at 6 km/h with gain g=-10"
-    fig_title <- ""
     file_name <- "./figures/predicted_fall_probability_6kmh.png"
   } else {
-    #fig_title <- "Predicted fall probability at 10 km/h with gain g=-8"
-    fig_title <- ""
     file_name <- "./figures/predicted_fall_probability_10kmh.png"
   }
 
   plot <- ggplot() +
-      geom_smooth(data = dummy_data_on, aes(x = angular_impulse, y = prediction, colour = "Balance-assist on"), method = "glm", method.args = list(family = binomial), fill = "cornflowerblue") +
-      geom_smooth(data = dummy_data_off, aes(x = angular_impulse, y = prediction, colour = "Balance-assist off"), method = "glm", method.args = list(family = binomial), fill = "coral") +
+      geom_smooth(
+          data = dummy_data_on,
+          linetype = 'twodash',
+          aes(x = angular_impulse, y = prediction, colour = "Balance-assist on"),
+          method = "glm",
+          method.args = list(family = binomial),
+          fill = "grey"
+      ) +
+      geom_smooth(
+          data = dummy_data_off,
+          aes(x = angular_impulse, y = prediction, colour = "Balance-assist off"),
+          method = "glm",
+          method.args = list(family = binomial),
+          fill = "grey"
+      ) +
       scale_colour_manual("",
           breaks = c("Balance-assist on", "Balance-assist off"),
-          values = c("Balance-assist on" = "blue", "Balance-assist off" = "red")
-      ) +
+          values = c("Balance-assist on" = "black",
+                     "Balance-assist off" = "black")) +
+      # TODO : Legend shows a solid line for each line, instead of dashed for
+      # one.
+      scale_linetype_manual(
+          breaks = c("Balance-assist on", "Balance-assist off"),
+          values = c("Balance-assist on" = "twodash",
+                     "Balance-assist off" = "solid")) +
       theme_bw() +
-      labs(title = fig_title, x = "Centred and scaled angular impulse", y = "Fall probability") +
-      theme(plot.title = element_text(hjust = 0.5))
-  ggsave(file = file_name, width = 100/25.4, height = 75/25.4, dpi = 300)
+      theme(legend.position = "bottom") +
+      labs(x = "Centered and scaled angular impulse",
[…]

### 2024-09-19 src/generate_time_series_imgs.py
Less height on time series plot.

@@ -114,12 +114,12 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
             if "motor_current" in df.columns.values:
                 fig, axs = plt.subplots(5, 1, sharex=True,
                                         layout="constrained",
-                                        figsize=(140.0/25.4, 140/25.4))
+                                        figsize=(140.0/25.4, 120/25.4))
                 df["motor_torque"] = df["motor_current"] / BALANCE_ASSIST_MOTOR_CONSTANT
             else:
                 fig, axs = plt.subplots(5, 1, sharex=True,
                                         layout="constrained",
-                                        figsize=(140.0/25.4, 140/25.4))
+                                        figsize=(140.0/25.4, 120/25.4))
 
             actual_torque, desired_torque = calculate_torque_on_handlebars(df)
             axs[0].plot(

### 2024-09-19 src/statistics.r
Whitespace

@@ -25,11 +25,11 @@ for (i in 1:2) {
   )
   print(summary(model_simple))
   print(summ(model_simple, exp = TRUE))
-  
-  # Create dummy dataset with all variables set to zero, except for angular impulse
-  # and state of balance-assist. The model created above uses this dataset to predict
-  # fall probabilities and show the difference between balance-assist off and 
-  # balance-assist on.
+
+  # Create dummy dataset with all variables set to zero, except for angular
+  # impulse and state of balance-assist. The model created above uses this
+  # dataset to predict fall probabilities and show the difference between
+  # balance-assist off and balance-assist on.
   dummy_data_on <- data.frame(
     angular_impulse = seq(from = -2.5, to = 2.5, by = 0.01)
   )

### 2024-09-19 src/generate_time_series_imgs.py
Added speed in km/h on time series figure.

@@ -15,6 +15,8 @@ DURATION_AFTER = 2.0
 DIRECTORY = "figures"
 HANLDEBAR_LENGTH = 0.82
 BALANCE_ASSIST_MOTOR_CONSTANT = 5
+KPH2MPS = 1000.0/3600.0
+MPS2KPH = 1.0/KPH2MPS
 
 
 def main():
@@ -204,6 +206,11 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 legend=False,
                 color='black',
             )
+            sax = axs[4].secondary_yaxis('right',
+                                         functions=(
+                                             lambda x: x*MPS2KPH,
+                                             lambda x: x*KPH2MPS))
+            sax.set_ylabel('Speed\n[km/h]')
             axs[4].legend(fontsize="x-small")
 
             for ax in axs:

### 2024-09-19 main.tex
Added citations to lme4 and R.

@@
 for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
-set of data.
+set of data. Statistics were computed with the lme4 version
+1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.
 
@@
   % figures. It is currently redundant.
-  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
+  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
     \label{fig:probability-6kph}}{
@@
   }
-  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
+  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
     \label{fig:probability-10kph}}{
@@
   \caption{Comparison of predicted fall probability for balance-assist on
-    (blue) or off (red) when all other predictor variables are fixed at zero,
+    (solid) or off (dashed) when all other predictor variables are fixed at zero,
     except for the interaction effect between the balance-assist and the
-    angular impulse. The ordinate axis is standard deviation around the mean
-    (all perturbations), centered per participant.}
+    angular impulse. The abscissa is standard deviation around the mean (all
+    perturbations), centered per participant.}
   \label{fig:probability}

### 2024-09-19 src/generate_time_series_imgs.py
Multiple roll angle by 10 and added circles for phi0 and delta0.

@@ -141,14 +141,20 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
             axs[0].set_ylabel("Bump'Em\nHandlebar\nTorque\n[Nm]")
             axs[0].legend(fontsize="x-small")
 
+            df['roll_angle_times_10'] = df['roll_angle']*10.0
             df.plot(
                 x="seconds_since_start",
-                y="roll_angle",
+                y="roll_angle_times_10",
                 ax=axs[1],
                 ylabel="Angle\n[deg]",
-                label="Roll",
+                label="Roll X 10",
                 color='black',
             )
+            axs[1].plot(0.0, df['roll_angle_times_10'].iloc[0],
+                        'o',
+                        fillstyle='none',
+                        markeredgecolor='black',
+                        label="__none__")
             df.plot(
                 x="seconds_since_start",
                 y="steer_angle",
@@ -157,6 +163,11 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 color='black',
                 linestyle='--',
             )
+            axs[1].plot(0.0, df['steer_angle'].iloc[0],
+                        'o',
+                        fillstyle='none',
+                        markeredgecolor='black',
+                        label="__none__")
             axs[1].legend(fontsize="x-small")
 
             df.plot(
@@ -176,8 +187,7 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 axs[3].axhline(-7.0, color='grey', linestyle='--',
                                label='__none__')
                 df.plot(
-                    x="seconds_since_start",
-                    y="motor_torque",
+                    x="seconds_since_start", y="motor_torque",
                     ax=axs[3],
                     ylabel="Commanded\nMotor Torque\n[Nm]",
                     label='__none__',

### 2024-09-19 main.tex
updated the captions and added definition of perturbation order.

@@
   \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
-  \caption{Perturbation torque alongside the resulting motion and steer motor
-    torque based on a 110~\si{\newton} counterclockwise applied force at a a
-    6~\si{\kph} travel speed.}
+  \caption{Externally applied perturbation torque alongside the resulting
+    measured motion and steer motor induced torque based on a 110~\si{\newton}
+    counterclockwise applied force at a 6~\si{\kph} travel speed. The circles
+    on the roll and steer angle plot show the angles \(\phi_0\) and
+    \(\delta_0\) at the perturbation start.}
   \label{fig:perturbation_0}
@@
 
-At the initiation of each perturbation we log the instantaneous steer and roll
-angles to characterize the configuration of the bicycle when perturbed. The
-gain setting on the balance assist controller indicates if the assistance is
-off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
-recovery from the perturbation is successful if the person neither places their
-foot down onto the treadmill surface nor their wheel of the bicycle exits the
-width of the treadmill belt. We record this as a binary variable \(f\) for
-"fall". All measured variables are reported in Table~\ref{tab:raw}.
+We record the order number of each perturbation, i.e. first, second, third,
+\ldots, to measure how many perturbations a rider is subjected to before the
+current pertubration. At the initiation of each perturbation we log the
+instantaneous steer and roll angles to characterize the configuration of the
+bicycle when perturbed. The gain setting on the balance assist controller
+indicates if the assistance is off \(g=0\) or on at two different levels: low
+\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
+person neither places their foot down onto the treadmill surface nor their
+wheel of the bicycle exits the width of the treadmill belt. We record this as a
+binary variable \(f\) for "fall". All measured variables are reported in
+Table~\ref{tab:raw}.
 %
@@
 
-We evaluate this hypothesis using a multivariate logistic regression model that
-takes the general form
+Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
+participant \(i\) which follows a Bernoulli distribution given the probability
+\(p_{ij}\) that a fall occurs:
 %
 \begin{align}
-  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
+  f_{ij} | p_{ij} \sim \textrm{Bernoulli}(p_{ij}) \textrm{.}
 \end{align}
 %
-Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
-participant \(i\) which follows a Bernoulli distribution given the probability
-\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
-function of our independent variables with \(\beta\) being the intercept and
-\(\alpha_k\) the linear coefficients to the \(K\) independent variables
-\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
-\(f\).
+We evaluate our hypothesis using a multivariate logistic regression model that
+represents the probability of this binary outcome. The log-odds of the
+probability is then a linear function of our independent variables with
+\(\beta\) being the intercept and \(\alpha_k\) the linear coefficients to the
+\(K\) independent variables \(x_{ij}^k\), i.e. all variables in
+Table~\ref{tab:stat-model-variables} except \(f\).
 %
@@
 assist bicycle. We thank all of the assistants that helped run the experiments
-and the subjects for their time an enthusiasm.
+and the subjects for their time an enthusiasm. We thank Peter Stahlecker for
+valuable feedback in early revisions.
 

### 2024-09-20 main.tex
Added table of bicycle physical parameters as an appendix, fixes #13.

@@
 
+\section*{Appendix}
+%
+Physical parameters and estimated uncertainty (standard deviation) for the
+balance assist bicycle without and with the rider. These are derived from the
+measurement approach in \citet{Moore2012}. Variable names match the notation
+from the benchmark bicycle model presented in \citet{Meijaard2007}.
+
+\begin{tabular}{llrrrr}
+  \toprule
+  Variable & Unit & Without Rider & Uncertainty & With Rider & Uncertainty \\
+  \midrule
+  \(c\) & \si{\meter} & 0.04 & 0.01 & 0.04 & 0.01 \\
+  \(g\) & \si{\meter\per\second\squared} & 9.80665 & NA & 9.80665 & NA \\
+  \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 0.02 & 19.7 & 0.3 \\
+  \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & 0.03 & -3.5 & 0.2 \\
+  \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 0.08 & 22.7 & 0.4 \\
+  \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 0.03 & 5.0 & 0.1 \\
+  \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0005 & 0.0995 & 0.0005 \\
+  \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.0009 & 0.1902 & 0.0009 \\
+  \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.003 & 0.298 & 0.003 \\
+  \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & 0.007 & -0.038 & 0.007 \\
+  \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.006 & 0.257 & 0.006 \\
+  \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.003 & 0.057 & 0.003 \\
+  \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.0006 & 0.1023 & 0.0006 \\
+  \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.0007 & 0.1887 & 0.0007 \\
+  \(\lambda\) & \si{\radian} & 0.25 & 0.03 & 0.25 & 0.03 \\
+  \(m_B\) & \si{\kilogram} & 22.50 & 0.01 & 106.00 & 0.01 \\
+  \(m_F\) & \si{\kilogram} & 2.23 & 0.01 & 2.23 & 0.01 \\
+  \(m_H\) & \si{\kilogram} & 4.30 & 0.01 & 4.30 & 0.01 \\
+  \(m_R\) & \si{\kilogram} & 4.08 & 0.01 & 4.08 & 0.01 \\
+  \(r_F\) & \si{\meter} & 0.3523 & 0.0001 & 0.3523 & 0.0001 \\
+  \(r_R\) & \si{\meter} & 0.3489 & 0.0001 & 0.3489 & 0.0001 \\
+  \(w\) & \si{\meter} & 1.113 & 0.008 & 1.113 & 0.008 \\
+  \(x_B\) & \si{\meter} & 0.52 & 0.01 & 0.374 & 0.002 \\
+  \(x_H\) & \si{\meter} & 0.92 & 0.02 & 0.92 & 0.02 \\
+  \(z_B\) & \si{\meter} & -0.52 & 0.02 & -1.009 & 0.003 \\
+  \(z_H\) & \si{\meter} & -0.860 & 0.006 & -0.860 & 0.006 \\
+  \bottomrule
+\end{tabular}
+
 \end{document}

### 2024-09-20 main.tex
Noted which bicycleparameter rider was used.

@@
 balance assist bicycle without and with the rider. These are derived from the
-measurement approach in \citet{Moore2012}. Variable names match the notation
-from the benchmark bicycle model presented in \citet{Meijaard2007}.
+measurement approach in \citet{Moore2012}. The rider inertial parameters are
+derived from the rider ``Jason'' available in BicycleParameters
+1.1.1~\citep{Moore2024}. The variable names match the notation from the
+benchmark bicycle model presented in \citet{Meijaard2007}.
 

### 2024-09-23 main.tex
balance assist -> balance-assist

@@
   magnitude external handlebar perturbations to the bicycle while ridden on a
-  treadmill with the balance assist system activated and deactivated. The
+  treadmill with the balance-assist system activated and deactivated. The
   probability of not recovering from a handlebar perturbation decreases when
-  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
-  balance assist system in real world bicycling will reduce the number of falls
+  the balance-assist is activated at a travel speed of 6~\si{\kph}. Use of a
+  balance-assist system in real world bicycling will reduce the number of falls
   that occur near riders' control authority limits.
@@
 successfully manage balancing tasks much like the high speed self-stability
-already does. We have developed a balance assisting bicycle,
+already does. We have developed a balance-assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 colliding with a neighbor's, a bag swinging from the handlebar, or simply
-hitting a bump in the road. To assess our balance assisting bicycle, we subject
+hitting a bump in the road. To assess our balance-assisting bicycle, we subject
 the rider to perturbations at the handlebar, which can relatively easily cause
@@
 the handlebars while the rider is cycling on a treadmill. We assess the
-difference in the rider's probability of falling with the balance assist system
+difference in the rider's probability of falling with the balance-assist system
 on and off.
@@
 study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
-distractions due to the balance assist system but contrarily researchers with a
+distractions due to the balance-assist system but contrarily researchers with a
 similar vehicle recently showed rider dissatisfaction with the
@@
 speed can give satisfactory stability for a low speed range. We use this simple
-feedback principle as the basis for our balance assist controller.
+feedback principle as the basis for our balance-assist controller.
 
@@
 and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
-(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
+(VR IMU BN0086, Sparkfun, Niwot, USA). The balance-assist control algorithm is
 implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
@@
 of 20 perturbations, we let the cyclist rest and then perform another 20
-perturbations. We randomize whether the balance assist system is on or off
+perturbations. We randomize whether the balance-assist system is on or off
 during the first or second set of perturbations for each participant.
@@
 instantaneous steer and roll angles to characterize the configuration of the
-bicycle when perturbed. The gain setting on the balance assist controller
+bicycle when perturbed. The gain setting on the balance-assist controller
 indicates if the assistance is off \(g=0\) or on at two different levels: low
@@
 %
-We test our hypothesis that the balance assistance controller will reduce the
+We test our hypothesis that the balance-assistance controller will reduce the
 probability of falling when perturbed externally at the handlebar. We have a
@@
     \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
-    \(c\)        & independent & boolean & balance assist: off 0 or on 1 \\
+    \(c\)        & independent & boolean & balance-assist: off 0 or on 1 \\
     \(j\)        & independent & integer & order number of perturbation \\
@@
 multilevel model. This left us with angular impulse, perturbation order,
-balance assist state, and roll \& steer angles at the time of perturbation as
+balance-assist state, and roll \& steer angles at the time of perturbation as
 independent variables. We also include interaction effects between the balance
@@
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
-balance assist state are all statistically significant predictors with angular
+balance-assist state are all statistically significant predictors with angular
 impulse having a dominant effect. Larger angular impulse increases the
@@
 increases the probability to fall and enduring more perturbations decreases the
-probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
+probability to fall. Unlike the 6~\si{\kph} trials, the balance-assist state is
 not a significant predictor. At both speeds the angular impulse has about twice
@@
 falling as a function of the mean and centered angular impulse per participant
-for the balance assist state on and off while keeping all other explanatory
+for the balance-assist state on and off while keeping all other explanatory
 factors at their centered mean values. This figure is created by setting all
@@
 figure shows that for relatively large impulses the probability to fall is
-unity for both states of balance assist on and off. And for relatively small
+unity for both states of balance-assist on and off. And for relatively small
 impulses the probability to fall is null for both states. But for impulses in
@@
 10~\si{\kph} trials which has a similar trend of reducing the probability to
-fall with the balance assist system turned on, but the effect is not
+fall with the balance-assist system turned on, but the effect is not
 significant.
@@
 %
-We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
+We have shown that at a 6~\si{\kph} riding speed the addition of balance-assist
 control reduces the chance that a rider will fall when perturbed around the
@@
 increasing the controller gain at 10~\si{\kph} also results in a significant
-positive effect. A longitudinal study of normal use of the balance assist
+positive effect. A longitudinal study of normal use of the balance-assist
 bicycle compared to a control group could provide the strongest evidence of any
@@
 probability and odds.
-The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
+The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance-assist
 system halves the odds that a perturbation results in a fall:
@@
 average angular impulse the rider was subjected to.
-This means that the balance assist is most effective for perturbations close to
+This means that the balance-assist is most effective for perturbations close to
 the subject's personal threshold between falling or recovery and that large
-perturbations will make you fall regardless of the balance assist's help.
+perturbations will make you fall regardless of the balance-assist's help.
 
@@
 that could predict how many falls may be averted in a natural setting if the
-balance assist system is used. Estimates of the predictor variables extracted
+balance-assist system is used. Estimates of the predictor variables extracted
 from the limited data collected from natural cycling would be needed to
@@
 speeds. We posit that this likely has to do with the width of the treadmill and
-that this could also be why the balance assist system did not have a statistically
+that this could also be why the balance-assist system did not have a statistically
 significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
@@
 %
-The positive effect of the balance assist system is coupled to the assumptions
+The positive effect of the balance-assist system is coupled to the assumptions
 and experimental scenarios we implemented and there is unfortunately no simple
@@
 more comprehensive and detailed natural data of how people fall we could make
-estimates on the number of falls reduced if everyone road a balance assist
+estimates on the number of falls reduced if everyone road a balance-assist
 bicycle.
@@
 mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
-when the balance assisting control is activated. This effect is present when
+when the balance-assisting control is activated. This effect is present when
 traveling at 10~\si{\kph} but more investigation is needed to determine if the
@@
 authority threshold. Given that similar effects cause falls during bicycling,
-use of the balance assist system in real world use cases will reduce the number
+use of the balance-assist system in real world use cases will reduce the number
 of falls at low bicycling speeds.
@@
 Physical parameters and estimated uncertainty (standard deviation) for the
-balance assist bicycle without and with the rider. These are derived from the
+balance-assist bicycle without and with the rider. These are derived from the
 measurement approach in \citet{Moore2012}. The rider inertial parameters are

### 2024-09-23 main.tex
Use builtin appendix feature and reference the appendix.

@@
     software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
-    1.5.0~\citep{Dembia2015}.}
+    1.5.0~\citep{Dembia2015} and are shown in
+    Appendix~\ref{sec:physical-parameters}.}
   \label{fig:eigenvalues}
@@
 
-\section*{Appendix}
+\appendix
 %
-Physical parameters and estimated uncertainty (standard deviation) for the
-balance-assist bicycle without and with the rider. These are derived from the
-measurement approach in \citet{Moore2012}. The rider inertial parameters are
-derived from the rider ``Jason'' available in BicycleParameters
-1.1.1~\citep{Moore2024}. The variable names match the notation from the
-benchmark bicycle model presented in \citet{Meijaard2007}.
-
+\section{Physical Parameters}
+\label{sec:physical-parameters}
+%
+Physical parameters and estimated uncertainties (standard deviation) for the
+balance-assist bicycle with and without a rider. These are found using the
+measurement approach for geometric and inertial parameters for a bicycle
+explained in \citet{Moore2012}. The rider inertial parameters are derived from
+the 83.5~\si{\kilogram} rider ``Jason'' available in BicycleParameters
+1.1.1~\citep{Moore2024}. The variable names match the notation and definitions
+from the benchmark bicycle model presented in
+\citet{Meijaard2007}.\todo{Uncertainties are based on best guess for
+uncertainty in the raw measurements, may be better to remove them as they
+aren't so meaningful.}
+%
+\begin{table}[h]  % have to use "h" so that it stays within the appendix
+  \centering
+  \caption{Physical parameters of the balance-assist bicycle with and without
+    the inertia effects of a rigid rider.}
   \begin{tabular}{llrrrr}
@@
   \end{tabular}
+\end{table}
 

### 2024-09-23 main.tex
Fixed a few more balance-assist spots.

@@
   \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
-  \caption{Balance assist bicycle prototype with electric motor in the steering
+  \caption{Balance-assist bicycle prototype with electric motor in the steering
     column and data acquisition and control electronics mounted in the rear
@@
     Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
-    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
+    Balance-assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
     Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
     Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
-    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
-    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
-    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
-    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
+    Balance-assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
+    Balance-assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
+    Balance-assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
+    Balance-assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
     \bottomrule
@@
     Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
-    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
-    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
-    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
-    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
+    Balance-assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
+    Balance-assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
+    Balance-assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
+    Balance-assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
     \bottomrule
@@
 Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
-Reijne, and Arend Schwab for contributions to the development of the balance
-assist bicycle. We thank all of the assistants that helped run the experiments
-and the subjects for their time an enthusiasm. We thank Peter Stahlecker for
-valuable feedback in early revisions.
+Reijne, and Arend Schwab for contributions to the development of the
+balance-assist bicycle. We thank all of the assistants that helped run the
+experiments and the subjects for their time an enthusiasm. We thank Peter
+Stahlecker for valuable feedback in early revisions.
 

### 2024-09-24 main.tex
Do not use a hyphen in balance-assist.

--- 2024-09-23 main.tex
@@
   magnitude external handlebar perturbations to the bicycle while ridden on a
-  treadmill with the balance-assist system activated and deactivated. The
+  treadmill with the balance assist system activated and deactivated. The
   probability of not recovering from a handlebar perturbation decreases when
-  the balance-assist is activated at a travel speed of 6~\si{\kph}. Use of a
-  balance-assist system in real world bicycling will reduce the number of falls
+  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
+  balance assist system in real world bicycling will reduce the number of falls
   that occur near riders' control authority limits.
@@
 successfully manage balancing tasks much like the high speed self-stability
-already does. We have developed a balance-assisting bicycle,
+already does. We have developed a balance assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 colliding with a neighbor's, a bag swinging from the handlebar, or simply
-hitting a bump in the road. To assess our balance-assisting bicycle, we subject
+hitting a bump in the road. To assess our balance assisting bicycle, we subject
 the rider to perturbations at the handlebar, which can relatively easily cause
@@
 the handlebars while the rider is cycling on a treadmill. We assess the
-difference in the rider's probability of falling with the balance-assist system
+difference in the rider's probability of falling with the balance assist system
 on and off.
@@
 study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
-distractions due to the balance-assist system but contrarily researchers with a
+distractions due to the balance assist system but contrarily researchers with a
 similar vehicle recently showed rider dissatisfaction with the
@@
 speed can give satisfactory stability for a low speed range. We use this simple
-feedback principle as the basis for our balance-assist controller.
+feedback principle as the basis for our balance assist controller.
 
@@
 and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
-(VR IMU BN0086, Sparkfun, Niwot, USA). The balance-assist control algorithm is
+(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
 implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
@@
 of 20 perturbations, we let the cyclist rest and then perform another 20
-perturbations. We randomize whether the balance-assist system is on or off
+perturbations. We randomize whether the balance assist system is on or off
 during the first or second set of perturbations for each participant.
@@
 instantaneous steer and roll angles to characterize the configuration of the
-bicycle when perturbed. The gain setting on the balance-assist controller
+bicycle when perturbed. The gain setting on the balance assist controller
 indicates if the assistance is off \(g=0\) or on at two different levels: low
@@
 %
-We test our hypothesis that the balance-assistance controller will reduce the
+We test our hypothesis that the balance assistance controller will reduce the
 probability of falling when perturbed externally at the handlebar. We have a
@@
     \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
-    \(c\)        & independent & boolean & balance-assist: off 0 or on 1 \\
+    \(c\)        & independent & boolean & balance assist: off 0 or on 1 \\
     \(j\)        & independent & integer & order number of perturbation \\
@@
 subject because we are only interested in the association between the state of
-the balance-assist system and the outcome of the perturbation. We expected
+the balance assist system and the outcome of the perturbation. We expected
 there to be a variation between participants in how well they are able to
@@
 multilevel model. This left us with angular impulse, perturbation order,
-balance-assist state, and roll \& steer angles at the time of perturbation as
+balance assist state, and roll \& steer angles at the time of perturbation as
 independent variables. We also include interaction effects between the balance
@@
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
-balance-assist state are all statistically significant predictors with angular
+balance assist state are all statistically significant predictors with angular
 impulse having a dominant effect. Larger angular impulse increases the
@@
 increases the probability to fall and enduring more perturbations decreases the
-probability to fall. Unlike the 6~\si{\kph} trials, the balance-assist state is
+probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
 not a significant predictor. At both speeds the angular impulse has about twice
@@
     Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
-    Balance-assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
+    Balance assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
     Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
@@
 
-Turning the balance-assist system on significantly \((p<0.05)\) reduces the
+Turning the balance assist system on significantly \((p<0.05)\) reduces the
 odds that a perturbation results in a fall while cycling at a speed of
@@
 falling as a function of the mean and centered angular impulse per participant
-for the balance-assist state on and off while keeping all other explanatory
+for the balance assist state on and off while keeping all other explanatory
 factors at their centered mean values. This figure is created by setting all
@@
 Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
-in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
+in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance assist
 system halves (0.53) the odds that a perturbation results in a fall. This
 figure shows that for relatively large impulses the probability to fall is
-unity for both states of balance-assist on and off. And for relatively small
+unity for both states of balance assist on and off. And for relatively small
 impulses the probability to fall is null for both states. But for impulses in
@@
 10~\si{\kph} trials which has a similar trend of reducing the probability to
-fall with the balance-assist system turned on, but the effect is not
+fall with the balance assist system turned on, but the effect is not
 significant.
@@
   }
-  \caption{Comparison of predicted fall probability for balance-assist on
+  \caption{Comparison of predicted fall probability for balance assist on
     (solid) or off (dashed) when all other predictor variables are fixed at zero,
-    except for the interaction effect between the balance-assist and the
+    except for the interaction effect between the balance assist and the
     angular impulse. The abscissa is standard deviation around the mean (all
@@
 %
-We have shown that at a 6~\si{\kph} riding speed the addition of balance-assist
+We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
 control reduces the chance that a rider will fall when perturbed around the
@@
 increasing the controller gain at 10~\si{\kph} also results in a significant
-positive effect. A longitudinal study of normal use of the balance-assist
+positive effect. A longitudinal study of normal use of the balance assist
 bicycle compared to a control group could provide the strongest evidence of any
@@
 probability and odds.
-The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance-assist
+The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
 system halves the odds that a perturbation results in a fall:
 \(e^{\alpha_k}=0.53\).
-This means if the odds are a 1000:1, turning on the balance-assist system
+This means if the odds are a 1000:1, turning on the balance assist system
 reduces the odds to 500:1.
@@
 for both speeds.
-The difference in probability between the balance-assist on/off case becomes
+The difference in probability between the balance assist on/off case becomes
 insignificant outside of approximately \(\pm1\) standard deviation of the
 average angular impulse the rider was subjected to.
-This means that the balance-assist is most effective for perturbations close to
+This means that the balance assist is most effective for perturbations close to
 the subject's personal threshold between falling or recovery and that large
-perturbations will make you fall regardless of the balance-assist's help.
+perturbations will make you fall regardless of the balance assist's help.
 
-To illustrate the effect of the balance-assist system on fall probability, we
+To illustrate the effect of the balance assist system on fall probability, we
 will give an example of how the data collected during the experiments is used
@@
 %
-The state of the balance-assist \(s\) is a binary variable. If the
-balance-assist is turned on, the log-odds that a fall occurs are decreased by
+The state of the balance assist \(s\) is a binary variable. If the
+balance assist is turned on, the log-odds that a fall occurs are decreased by
 0.64. The odds and probability can be calculated:
@@
 %
-Turning on the balance-assist system reduces the probability that the
+Turning on the balance assist system reduces the probability that the
 perturbation results in a fall from 81\% to 69\%.
@@
 that could predict how many falls may be averted in a natural setting if the
-balance-assist system is used. Estimates of the predictor variables extracted
+balance assist system is used. Estimates of the predictor variables extracted
 from the limited data collected from natural cycling would be needed to
@@
 speeds. We posit that this likely has to do with the width of the treadmill and
-that this could also be why the balance-assist system did not have a statistically
+that this could also be why the balance assist system did not have a statistically
 significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
@@
 None of the interaction effects are statistically significant. That means that
-whether the balance-assist system is on or off does not significantly change
+whether the balance assist system is on or off does not significantly change
 the effect that the roll angle, steer angle, angular impulse, and perturbation
@@
 %
-The positive effect of the balance-assist system is coupled to the assumptions
+The positive effect of the balance assist system is coupled to the assumptions
 and experimental scenarios we implemented and there is unfortunately no simple
@@
 more comprehensive and detailed natural data of how people fall we could make
-estimates on the number of falls reduced if everyone road a balance-assist
+estimates on the number of falls reduced if everyone road a balance assist
 bicycle.
@@
 mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
-when the balance-assisting control is activated. This effect is present when
+when the balance assisting control is activated. This effect is present when
 traveling at 10~\si{\kph} but more investigation is needed to determine if the
@@
 authority threshold. Given that similar effects cause falls during bicycling,
-use of the balance-assist system in real world use cases will reduce the number
+use of the balance assist system in real world use cases will reduce the number
 of falls at low bicycling speeds.
@@
 Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
-Reijne, and Arend Schwab for contributions to the development of the balance
-assist bicycle. We thank all of the assistants that helped run the experiments
-and the subjects for their time an enthusiasm. We thank Peter Stahlecker for
-valuable feedback in early revisions.
+Reijne, and Arend Schwab for contributions to the development of the
+balance assist bicycle. We thank all of the assistants that helped run the
+experiments and the subjects for their time an enthusiasm. We thank Peter
+Stahlecker for valuable feedback in early revisions.
 
@@
 Physical parameters and estimated uncertainties (standard deviation) for the
-balance-assist bicycle with and without a rider. These are found using the
+balance assist bicycle with and without a rider. These are found using the
 measurement approach for geometric and inertial parameters for a bicycle
@@
   \centering
-  \caption{Physical parameters of the balance-assist bicycle with and without
+  \caption{Physical parameters of the balance assist bicycle with and without
     the inertia effects of a rigid rider.}

### 2024-09-24 main.tex
Improved the abstract to be clearer about testing on subjects and also other better wording.

@@
 \abstract{
-  Uncontrolled bicycles are generally unstable at low speeds. We add a
-  controlled steering motor to a consumer electric bicycle that stabilizes it
-  at low speeds. To test the motor's assistance during falls, we apply varying
-  magnitude external handlebar perturbations to the bicycle while ridden on a
-  treadmill with the balance assist system activated and deactivated. The
-  probability of not recovering from a handlebar perturbation decreases when
-  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
-  balance assist system in real world bicycling will reduce the number of falls
-  that occur near riders' control authority limits.
+  Uncontrolled bicycles are generally unstable at low speeds. We add an
+  automatically controlled steering motor to a consumer electric bicycle that
+  has a stabilizing effect down to about 1~\si{\kph}. We hypothesize that a
+  stabilized bicycle will reduce the probability of falling. To test the
+  motor's possible assistance during falls, we apply varying magnitude external
+  handlebar perturbations to twenty-six subjects who rode on a treadmill with
+  the balance assist system activated and deactivated. The probability of not
+  recovering from a handlebar perturbation significantly decreases when the
+  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
+  effect is most prominent at and around the individual riders' perturbation
+  resistance threshold. We conclude that use of a balance assist system in real
+  world bicycling would reduce the number of falls that occur near riders'
+  control authority limits.
 }

### 2024-09-24 main.tex
Edits to the introduction.

@@
 reduced if the bicycle was self-stable at these speeds by relieving the rider
-of some of the necessary control activity. Bicycles can be mechanically
+of some of their necessary control activity. Bicycles can be mechanically
 modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
 and since the 1980s it has been known that applying a motor actuated steering
-torque proportional to the vehicle's roll rate~\footnote{Roll angular rate is
-defined herein as per the SAE J670 vehicle standards.} can stabilize a
-single-track vehicle at lower speeds. If automatic control of steering can
-stabilize a bicycle, it may reduce the control required from the rider to
-successfully manage balancing tasks much like the high speed self-stability
-already does. We have developed a balance assisting bicycle,
+torque proportional to the vehicle's roll angular rate~\footnote{Roll angular
+rate is defined herein as per the SAE J670 vehicle standards.} can stabilize a
+single-track vehicle down to very low speeds. If automatic control of steering
+can stabilize a bicycle, it may reduce the control required from the rider to
+successfully manage balancing tasks much like the natural high speed
+self-stability already does. We have developed a balance-assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 %
-During the early years of developments in automatic control, \citep{Whipple1899}
-not only derived the correct equations of motion of the bicycle but realized
-and showed that roll motion feedback can stabilize a bicycle. Much later,
-attempts at automatic roll stabilization of single-track vehicles began in
-earnest after predictive motorcycle models were developed and refined in
-throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
-and seems to be the first to attempt to robotically stabilize a small motorbike
-with a controlled inverted pendulum that mimicked rider lean, but he was not
-successful in demonstrating what his theoretic control model correctly
-predicted. In his model, he recognized that feedback of the vehicle roll angle
-and angular rate was essential to stabilize the vehicle. It was not until the
-early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
-bicycle on a treadmill with both steering control and a laterally moving mass.
-Ruijs and Pacjecka followed this by demonstrating an automatically balanced
-motorcycle~\citep{Ruijs1986} and they did so solely with a steering motor. Ruijs and
-Pacejka clearly showed that steer torque driven by roll angle feedback
-stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
-mode, and steer angular rate feedback stabilizes the wobble mode~\footnote{The
-motorcycle (and bicycle) eigenmodes are defined in \citep{Sharp1971}.}. They
-also proposed how the gains must change with respect to vehicle speed for
-favorable control at all speeds.  This roll motion feedback enables the
-simplest controller that can stabilize a single-track vehicle above a minimum
-speed when one is not concerned with wobble instabilities. Ruijs and Pacejka's
-work was not particularly concerned with low speed stability and their vehicle
-was fully automatic, i.e no human rider was involved.
+During the early years of developments in automatic control,
+\citep{Whipple1899} not only derived the correct equations of motion of the
+bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
+can stabilize a bicycle. Much later, attempts at automatic roll stabilization
+of single-track vehicles began in earnest after predictive motorcycle models
+were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
+influenced by Whipple's work and seems to be the first to attempt to
+robotically stabilize a small motorbike with a controlled inverted pendulum
+that mimicked rider lean, but he was not successful in demonstrating what his
+theoretic control model correctly predicted. In his model, he recognized that
+feedback of the vehicle roll angle and angular rate was essential to stabilize
+the vehicle. It was not until the early 1980s that \cite{Nagai1983}
+successfully demonstrated balancing a robotic bicycle on a treadmill with both
+steering control and a laterally moving mass. \citet{Ruijs1986} followed this
+breakthrough by demonstrating an automatically balanced motorcycle and they did
+so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
+torque driven by roll angle feedback stabilizes the capsize mode, by roll
+angular rate feedback stabilizes the weave mode, and by steer angular rate
+feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
+eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
+gains must change with respect to vehicle speed for favorable control across
+all speeds. This roll motion feedback enables the simplest controller that can
+stabilize a single-track vehicle above a minimum speed when one is not
+concerned with wobble instabilities. But Ruijs and Pacejka's work was not
+particularly concerned with low speed stability and their vehicle was fully
+automatic, i.e no human rider was involved.
 
-Many more automatically balanced single track vehicles have been demonstrated
+Many more automatically balanced single-track vehicles have been demonstrated
 over the last 40 years, but none have demonstrated that increasing low speed
-stability can assist in balancing and possibly reduce single-actor falls. Most
-of these robotic bicycle and motorcycle designers did not intend for a human
-rider to also control the stabilized vehicle. Nevertheless, an automatically
-stabilized bicycle can be controlled by a human rider if the motor controlled
-steer torque and the rider applied steer torque act on the steer in parallel.
-The effect of this automatic control gives the ability to effectively change
-the dynamics, up to some limits, of the human-controlled plant. In our prior
-study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
-distractions due to the balance assist system but contrarily researchers with a
-similar vehicle recently showed rider dissatisfaction with the
-stabilization~\cite{Hanakam2023}.
+stability can assist in human balancing and what effect it may have towards
+single-actor falls. Most of these robotic bicycle and motorcycle designers did
+not intend for a human rider to also control the stabilized vehicle.
+Nevertheless, an automatically stabilized bicycle can be controlled by a human
+rider if the motor controlled steer torque and the rider applied steer torque
+act on the steer in parallel. The effect of this automatic control gives the
+ability to effectively change the human-controlled plant dynamics, up to some
+limits. In our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced
+motion during distractions due to the balance assist system but
+\citet{Hanakam2023} recently showed rider dissatisfaction with the
+stabilization of a similar vehicle so the possible benefits are not
+definitively established.
 
-The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
-the simplest bicycle model that exhibits non-minimum phase behavior and
-self-stability. The bicycle model is suitable for showing the effect of a roll
-motion feedback driven steer motor on the dynamics. This model is equivalently
-valid for on-road or treadmill motion~\citep{Kooijman2009a}, which have the
-same dynamics. The linear version of this model~\citep{Meijaard2007} can be
-described by the fourth order state space equations:
+The design of a balance assist system relies on the linear Carvallo-Whipple
+bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
+model that exhibits non-minimum phase behavior and self-stability. The bicycle
+model is well suited for showing the effect of a roll motion feedback driven
+steer motor on the dynamics. This model is equivalently valid for on-road or
+treadmill riding~\citep{Kooijman2009a}, which have the same fundamental
+dynamics. The linear version of this model can be described by the fourth order
+state space equations:
 %
@@
 %
-The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
-their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
-\(T_\delta\). The state \(\mathbf{A}\) matrix is a function of the equilibrium
-forward speed \(v\). It and the input \(\mathbf{B}\) matrix are otherwise
-populated with expressions that are functions of the geometric and inertial
-parameters of the nonholonomic multibody system made up of four rigid bodies:
-two wheels, front frame, and rear frame.
+The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
+along with their time derivatives and the inputs \(\vec{u}\) are roll torque
+\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
+function of the equilibrium forward speed \(v\). It and the input matrix
+\(\mathbf{B}\) are otherwise populated with expressions that are functions of
+the geometric and inertial parameters of the nonholonomic multibody system made
+up of four rigid bodies: two wheels, front frame, and rear frame.
 
@@
 \(\mathbf{B} = \begin{bmatrix} \vec{B}_\phi \quad \vec{B}_\delta
-\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
-the human controlled plant takes the form:
+\end{bmatrix}\), and the motor torque is a proportional feedback controller
+\(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then the human-controlled
+plant takes the form:
 %
@@
 The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
-means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
-\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
-0\right] \right)\) have negative real parts for \(v_{min} < v <
+means the control gain \(k_{\dot{\phi}}\) can be selected such that the
+eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}}
+\quad 0 \quad 0\right] \right)\) have negative real parts for \(v_{min} < v <
 v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
 \(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
-uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
+uncontrolled bicycle's capsize mode goes unstable. By gain scheduling with
 respect to \(v\), the speed range where the bicycle is stable can be maximized
 within any physical actuator magnitude and bandwidth limits.
-\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
-gains for such a controller and shows that a linear scheduling with respect to
-speed can give satisfactory stability for a low speed range. We use this simple
-feedback principle as the basis for our balance assist controller.
+\citet{Schwab2008} elaborate on some of the possibilities in scheduling the
+gains for such a controller for a bicycle and shows that a linear scheduling
+with respect to speed can give satisfactory stability for a low speed range. We
+use this simple feedback principle as the basis for our balance assist
+controller.
 

### 2024-09-24 main.tex
Edits in the methods section.

@@
   automatically controlled steering motor to a consumer electric bicycle that
-  has a stabilizing effect down to about 1~\si{\kph}. We hypothesize that a
+  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
   stabilized bicycle will reduce the probability of falling. To test the
   motor's possible assistance during falls, we apply varying magnitude external
-  handlebar perturbations to twenty-six subjects who rode on a treadmill with
+  handlebar perturbations to twenty-six participants who rode on a treadmill with
   the balance assist system activated and deactivated. The probability of not
@@
 %
-We use a forward speed \(v\) gain scheduled proportional roll rate feedback
-controller to stabilize the bicycle. In the speed range tested, the commanded
-steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
-law
+We use a forward speed \(v\) gain scheduled proportional roll angular rate
+feedback controller to stabilize the bicycle. In the speed range tested, the
+commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
+control law
 %
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
-proportional feedback gain linearly with respect to speed stabilizes the
-normally unstable weave mode of the bicycle down to
+We use \(g=8\) and \(g=10\) to tune the gain magnitude during the experiments.
+Scaling the proportional feedback gain linearly with respect to speed
+stabilizes the normally unstable weave mode of the bicycle down to
 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
@@
     the linear stable speed range. Geometric and inertial parameters for these
-    plots were estimated with the methods presented in \cite{Moore2012} and
+    plots were estimated with the methods presented in \citet{Moore2012} and
     software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
@@
 posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
-four motors working in cooperation, Figure~\ref{fig:setup-diagram}. The four
-motors are programmed to apply a light force at all times to keep the ropes
-taught and to track a commanded force profile using a PID controller running on
-a microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
-applied by each motor at the handlebar via four inline 250~\si{\newton} S-style
-load cells (Bosche GmbH, Damme, Germany) except the left rear load cell that
-has a max load of 500~\si{\newton}. The commanded force profiles are designed
-to apply an external pulsive torque to the front assembly (handlebars, forks,
-wheel) at magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
-motors (EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners
-of a 1~\si{\meter} wide treadmill (Bonte Technology B.V., Zwolle, The
-Netherlands) that can reach speeds of 18~\si{\kilo\meter\per\hour}. The general
-design of this system is described in detail in Van De Velde's MSc
-thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
-Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
-Velde include simplifying the Bump'Em motor controller with an inexpensive
-microcontroller and the use of a simpler non-active safety harness.
+four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
+Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
+force at all times to keep the ropes taught and to track a commanded force
+profile using a PID controller running on a microprocessor (Arduino Mega 2560,
+Arduino, Italy). We measure the force applied by each motor at the handlebar
+via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
+rated for 250~\si{\newton} except the left rear load cell that was rated for
+500~\si{\newton}. The commanded force profiles are designed to apply an
+external pulsive torque to the front assembly (handlebars, forks, wheel) at
+magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
+motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
+(Bonte Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
+18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
+described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
+the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
+Our modifications relative to Van De Velde's design include simplifying the
+Bump'Em motor controller with an inexpensive microcontroller and the use of a
+simpler non-active safety harness.
 %
@@
   \centering
+  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
+    system.\protect\label{fig:setup-diagram}}
+    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
   \subcaptionbox{A participant on the bicycle in the safety harness with the
@@
     {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
-  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
-    system.\protect\label{fig:setup-diagram}}
-    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
   \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
@@
 We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments.
-\todo[]{Include the average age, mass, height and standard deviation if we
-recorded those.}
-The subjects were all confident in their cycling skills and had cycled at least
-once in the last month. Eleven subjects performed the experiments at
-6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and fifteen subjects
-performed the experiments at 10~\si{\kilo\meter\per\hour}
-(2.8~\si{\meter\per\second}). All subjects consented to the experiment and
-could decline to continue at any time. The study was approved by Delft
-University of Technology's Human Research Ethics Council (\#3897).
+the experiments. The participants were all confident in their cycling skills and
+had cycled at least once in the last month. All participants consented to the
+experiment and could decline to continue at any time. The study was approved by
+Delft University of Technology's Human Research Ethics Council (\#3897).
 
-The subjects were divided into two groups. The first group performed the
-protocol at a belt speed of 10~\si{\kilo\meter\per\hour} with the controller
-gain set to $g=8$ and the second group performed the protocol at a belt speed
-of 6~\si{\kilo\meter\per\hour} with the controller gain set to $g=10$. The
+The participants were divided into two groups. The first group of fifteen
+participants
+performed the protocol at a belt speed of
+10~\si{\kilo\meter\per\hour}(2.8~\si{\meter\per\second}) with the controller
+gain set to $g=8$ and the second group of eleven particpants performed the
+protocol at a belt speed of 6~\si{\kilo\meter\per\hour}
+(1.7~\si{\meter\per\second}) with the controller gain set to $g=10$. The
 10~\si{\kph} group's experiments occurred two weeks before the
@@
 
-Subjects wore a helmet and fall safety harness attached to the ceiling
-Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
-the treadmill until they indicated they were comfortable enough to have
-perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
-We then asked the rider to ride for 90~\si{\second}, while attempting to
-maintain the location of their front wheel on the center line of the treadmill.
-We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
-their foot from the pedal and places it on the ground or 2) the bicycle wheel
-exceeds the width of the treadmill belt. We then applied perturbations in
-random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
-and increasing the magnitude by 30~\si{\newton} until the participants
-fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
-from a perturbation. We log the magnitude that causes the first fall to
-characterize that subject's \emph{nominal fall threshold}.
+Subjects wore a helmet and they were attached to the ceiling via a fall safety
+harness, Figure~\ref{fig:participant-in-set-up}. We allowed them to practice
+riding on the treadmill until they indicated they were comfortable enough to
+have perturbations applied. For most, this was less than a 10~\si{\minute} warm
+up.  We then asked the rider to ride for 90~\si{\second}, while attempting to
+maintain the location of their front wheel on the center line of the treadmill
+as a final preparation before the perturbations.  We then applied perturbations
+in random directions (clockwise or counter-clockwise), starting at
+20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
+participants fell. We define a ``fall'' on the treadmill by two criteria: 1)
+the rider removes their foot from the pedal and places it on the ground or 2)
+the bicycle wheel exceeds the width of the treadmill belt.
+Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
+perturbation. We log the magnitude that causes the first fall to characterize
+that participant's \emph{perturbation resistance threshold}.
 %
@@
 \ldots, to measure how many perturbations a rider is subjected to before the
-current pertubration. At the initiation of each perturbation we log the
+current perturbation. At the initiation of each perturbation we log the
 instantaneous steer and roll angles to characterize the configuration of the
@@
 centering, as recommended by \cite{Enders2007}, with clusters being an
-individual subject. The clusters are chosen as all data from an individual
-subject because we are only interested in the association between the state of
+individual participant. The clusters are chosen as all data from an individual
+participant because we are only interested in the association between the state of
 the balance assist system and the outcome of the perturbation. We expected
@@
 resist the perturbation. However, this was not true. Cluster-mean centering
-shows there to be no variation between participants. This fact allows us to
+showed there to be no variation between participants. This fact allows us to
 utilize a simple single-level logistic regression model, instead of a
@@
 This means that the balance assist is most effective for perturbations close to
-the subject's personal threshold between falling or recovery and that large
+the participant's personal threshold between falling or recovery and that large
 perturbations will make you fall regardless of the balance assist's help.
@@
 Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
-Reijne, and Arend Schwab for contributions to the development of the
-balance assist bicycle. We thank all of the assistants that helped run the
-experiments and the subjects for their time an enthusiasm. We thank Peter
-Stahlecker for valuable feedback in early revisions.
+Reijne, and Arend Schwab for contributions to the development of the balance
+assist bicycle. We acknowledge Shannon van de Velde's and Marco Reijne's
+development of the bicycle perturbation system. We thank all of the assistants
+that helped run the experiments and the participants for their time an enthusiasm.
+We thank Peter Stahlecker for valuable feedback in early revisions.
 

### 2024-09-25 main.tex
Edits to methods down to statistics.

@@
 via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
-rated for 250~\si{\newton} except the left rear load cell that was rated for
+rated for 250~\si{\newton} except the left rear load cell which was rated for
 500~\si{\newton}. The commanded force profiles are designed to apply an
 external pulsive torque to the front assembly (handlebars, forks, wheel) at
-magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four
-motors are arranged at the four corners of a 1~\si{\meter} wide treadmill
-(Bonte Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
-18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
-described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
-the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
-Our modifications relative to Van De Velde's design include simplifying the
-Bump'Em motor controller with an inexpensive microcontroller and the use of a
-simpler non-active safety harness.
+magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors are
+arranged at the four corners of a 1~\si{\meter} wide by 2~\si{\meter} long
+treadmill (Bonte Technology B.V., Zwolle, The Netherlands) that can reach
+speeds up to 18~\si{\kilo\meter\per\hour}. The general design of this
+perturbation system is described in detail in Van De Velde's MSc
+thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
+Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
+Velde's design include simplifying the Bump'Em motor controller with an
+inexpensive microcontroller and the use of a simpler non-active safety harness.
 %
@@
   \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
-    perturbations.}
+    perturbations and the various system components.}
   \label{fig:experimental-setup}
@@
 We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments. The participants were all confident in their cycling skills and
-had cycled at least once in the last month. All participants consented to the
-experiment and could decline to continue at any time. The study was approved by
-Delft University of Technology's Human Research Ethics Council (\#3897).
+the experiments. The participants were all confident in their cycling skills
+and had cycled at least once in the last month. All participants consented to
+the experiment and could decline to continue at any time. The study was
+approved by Delft University of Technology's Human Research Ethics Council
+(record \#3897).
 
 The participants were divided into two groups. The first group of fifteen
-participants
-performed the protocol at a belt speed of
-10~\si{\kilo\meter\per\hour}(2.8~\si{\meter\per\second}) with the controller
-gain set to $g=8$ and the second group of eleven particpants performed the
-protocol at a belt speed of 6~\si{\kilo\meter\per\hour}
-(1.7~\si{\meter\per\second}) with the controller gain set to $g=10$. The
-10~\si{\kph} group's experiments occurred two weeks before the
-6~\si{\kilo\meter\per\hour} group.
+participants performed the protocol at a belt speed of
+10~\si{\kph}(2.8~\si{\mps}) with the controller gain value set to $g=8$ and the
+second group of eleven particpants performed the protocol at a belt speed of
+6~\si{\kph} (1.7~\si{\mps}) with the controller gain set to $g=10$. The
+10~\si{\kph} group's experiments occurred two weeks before the 6~\si{\kph}
+group.
 
-Subjects wore a helmet and they were attached to the ceiling via a fall safety
-harness, Figure~\ref{fig:participant-in-set-up}. We allowed them to practice
-riding on the treadmill until they indicated they were comfortable enough to
-have perturbations applied. For most, this was less than a 10~\si{\minute} warm
-up.  We then asked the rider to ride for 90~\si{\second}, while attempting to
-maintain the location of their front wheel on the center line of the treadmill
-as a final preparation before the perturbations.  We then applied perturbations
-in random directions (clockwise or counter-clockwise), starting at
-20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
-participants fell. We define a ``fall'' on the treadmill by two criteria: 1)
-the rider removes their foot from the pedal and places it on the ground or 2)
-the bicycle wheel exceeds the width of the treadmill belt.
-Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
-perturbation. We log the magnitude that causes the first fall to characterize
-that participant's \emph{perturbation resistance threshold}.
+Participants wore a helmet and they were attached to the ceiling via a fall
+safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
+free movement pre-fall. The participants practiced riding on the treadmill
+until they indicated they were comfortable enough to have perturbations
+applied. For most, this was less than a 10~\si{\minute} warm up. We then asked
+the rider to ride for 90~\si{\second}, while attempting to maintain the
+location of their front wheel on the center line of the treadmill as a baseline
+measure before the perturbations. We then applied perturbations in random
+directions (clockwise or counter-clockwise), starting at 20~\si{\newton} and
+increasing the magnitude by 30~\si{\newton} until the participants fell. We
+defined a ``fall'' on the treadmill by two criteria: 1) the rider removes their
+foot from the pedal and places it on the ground or 2) the bicycle wheel exceeds
+the width of the treadmill belt. Figure~\ref{fig:perturbation-sequence} shows
+an example resulting motion from a perturbation. We logged the force magnitude
+that caused the first fall to characterize that participant's
+\emph{perturbation resistance threshold}.
 %
@@
 is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
-100~\si{\newton}. Which one of these five perturbations is chosen is determined
-at random. If the perturbation results in a fall, the estimate of the
-perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
-perturbation did not result in a fall. Five new possible perturbation forces
-are determined around the updated perturbation threshold, and a new
+100~\si{\newton}. Which one of these five perturbations is chosen first is
+determined at random. If the perturbation results in a fall, the estimate of
+the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
+the perturbation did not result in a fall. Five new possible perturbation
+forces are determined around the updated perturbation threshold, and a new
 perturbation is selected at random. This process iterates until twenty
@@
 perturbations. We randomize whether the balance assist system is on or off
-during the first or second set of perturbations for each participant.
+during the first or second set of 20 perturbations for each participant.
 
@@
 We measure the time histories of the Bump'Em delivered perturbation forces and
-the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
-Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
-compared to our Bump'Em controller command.
+the bicycle's steer angle \(\delta\), roll angle \(\phi\), roll angular rate
+\(\dot{\phi}\), and forward speed \(v\). Figure~\ref{fig:perturbation_0} shows
+an example perturbation force measurement compared to our Bump'Em controller
+command.
 %
@@
 
-Based on findings from measuring riders without balance
-assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
-may influence fall probability. We use the angular impulse \(L\) of the
+Based on findings from measuring riders without balance assist in
+\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
+influence fall probability. We use the angular impulse \(L\) of the
 perturbation forces over a 0.3~\si{\second} duration to characterize the
@@
 readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
-as \(l\) in Equation \ref{eq:angular-impulse}.
+as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
+of peak torque to normalize for the duration of the applied perturbation.
 
-We record the order number of each perturbation, i.e. first, second, third,
-\ldots, to measure how many perturbations a rider is subjected to before the
-current perturbation. At the initiation of each perturbation we log the
+We record the order number of each perturbation \(j\), i.e. first, second,
+third, \ldots, to measure how many perturbations a rider is subjected to before
+the current perturbation. At the initiation of each perturbation we log the
 instantaneous steer and roll angles to characterize the configuration of the
@@
   \centering
-  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
-    ``Perturbation order number'' are recorded per perturbation instance. The
+  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
+    ``Perturbation Order Number'' are recorded per perturbation instance. The
     remaining measures are time varying during each trial.}
@@
     Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
-    Fall outcome & \(f\) & boolean & NA \\
+    Fall Outcome & \(f\) & boolean & NA \\
     Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &

### 2024-09-25 main.tex
Finished editing methods.

@@
 %
-We test our hypothesis that the balance assistance controller will reduce the
+We test our hypothesis that the balance assist controller will reduce the
 probability of falling when perturbed externally at the handlebar. We have a
 single binary fall outcome variable \(f\) that is dependent on several possible
-explanatory independent variables, one of which is the binary balance
-assistance (on or off). See Table~\ref{tab:stat-model-variables} for the
-statistical model variables.
+explanatory independent variables, one of which is the binary balance assist
+state (on or off). See Table~\ref{tab:stat-model-variables} for the statistical
+model variables.
 %
@@
   \centering
-  \caption{Independent and dependent variables used in the statistical model.}
+  \caption{Independent and dependent variables used in the statistical
+    regression model.}
   \label{tab:stat-model-variables}
@@
 that they have a mean of zero and a standard deviation of one by cluster-mean
-centering, as recommended by \cite{Enders2007}, with clusters being an
+centering, as recommended by \citet{Enders2007}, with clusters being an
 individual participant. The clusters are chosen as all data from an individual
-participant because we are only interested in the association between the state of
-the balance assist system and the outcome of the perturbation. We expected
+participant because we are only interested in the association between the state
+of the balance assist system and the outcome of the perturbation. We expected
 there to be a variation between participants in how well they are able to

### 2024-09-25 main.tex
Added gain 10 to eigenvalue plot.

@@
   \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
-    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
-    for the bicycle without the inertial effect of a rigid rider (left) and
-    with a rigid rider (right). Vertical dotted and dotted-dashed lines
-    indicate the two speeds we perform experiments at. Grey shaded region is
-    the linear stable speed range. Geometric and inertial parameters for these
-    plots were estimated with the methods presented in \citet{Moore2012} and
-    software packages BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon
-    1.5.0~\citep{Dembia2015} and are shown in
-    Appendix~\ref{sec:physical-parameters}.}
+  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
+    root locus of eigenvalue components (real: solid, imaginary: dashed)
+    plotted versus speed for the bicycle without the inertial effect of a rigid
+    rider (left) and with a rigid rider (right). Vertical dotted and
+    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
+    shaded region is the linear stable speed range. Geometric and inertial
+    parameters for these plots were estimated with the methods presented in
+    \citet{Moore2012} and software packages BicycleParameters
+    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
+    in Appendix~\ref{sec:physical-parameters}.}
   \label{fig:eigenvalues}

### 2024-09-25 main.tex
Clarify a couple of sentences in results.

@@
 impulse having a dominant effect. Larger angular impulse increases the
-probability to fall and enduring more perturbations or having the balance
+probability to fall and both enduring more perturbations or having the balance
 assist on, decrease the probability to fall. The associated multiplicative
@@
 probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
-not a significant predictor. At both speeds the angular impulse has about twice
-the effect as the perturbation order.
+not a significant predictor but similarly shows that the balance assist state
+reduces the probability to fall. At both speeds, the angular impulse has about
+twice the magnitude in effect as the perturbation order.
 %

### 2024-09-25 main.tex
Finished editing the results.

@@
 probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
-not a significant predictor but similarly shows that the balance assist state
-reduces the probability to fall. At both speeds, the angular impulse has about
+not a significant predictor but the effect would be a reduction in the
+probability to fall if it were. At both speeds, the angular impulse has about
 twice the magnitude in effect as the perturbation order.
@@
 factors at their centered mean values. This figure is created by setting all
-explanatory variables to there mean and calculating the probability from
-Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
-in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance assist
-system halves (0.53) the odds that a perturbation results in a fall. This
-figure shows that for relatively large impulses the probability to fall is
+explanatory variables to their mean and calculating the probability from
+Equation~\ref{eq:log-regress} for only change in angular impulse given the
+estimates in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance
+assist system halves (0.53) the odds that a perturbation results in a fall.
+This figure shows that for relatively large impulses the probability to fall is
 unity for both states of balance assist on and off. And for relatively small
 impulses the probability to fall is null for both states. But for impulses in
-the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
-threshold, the probability of falling is significantly lowered with the balance
-assist system on. The skewness of the probability curves arrives from the
-interaction effects.\todo{check this statement about skewedness this and think
-about it} Figure~\ref{fig:probability-10kph} shows the same result for the
-10~\si{\kph} trials which has a similar trend of reducing the probability to
-fall with the balance assist system turned on, but the effect is not
-significant.
+the magnitude region (about -1 to 0.5 STD), i.e. around the mean-centered
+perturbation resistance threshold, the probability of falling is significantly
+lowered with the balance assist system on. The skewness of the probability
+curves arrives from the interaction effects. Figure~\ref{fig:probability-10kph}
+shows the same result for the 10~\si{\kph} trials which has a similar trend of
+reducing the probability to fall with the balance assist system turned on, but
+the effect is not significant.
 %
@@
   \centering
-  % NOTE : We could uncomment the longer subcaption if titles removed from
-  % figures. It is currently redundant.
   \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
@@
     except for the interaction effect between the balance assist and the
-    angular impulse. The abscissa is standard deviation around the mean (all
-    perturbations), centered per participant.}
+    angular impulse. The abscissa is the standard deviation around the mean
+    (all perturbations), centered per participant.}
   \label{fig:probability}

### 2024-09-26 main.tex
Finished an editing pass of the whole paper.

@@
 bicycle compared to a control group could provide the strongest evidence of any
-benefit we have seen in this more narrow scenario.
+benefit we have seen in this specific scenario.
 
@@
 
-The controller design we utilize,
-Equation~\ref{eq:implemented-controller}, also increases the weave mode
-frequency by a factor of about three up to about 1~\si{\hertz}. This bandwidth
-is still controllable by the human's neuromuscular system, but may feel
-unnatural as it is more akin to what the steering would feel like at in the
-\SIrange{30}{40}{\kph} speed range. \Citet{Hanakam2023} reported
-dissatisfaction in subjective rider feeling on their similar bicycle to ours
-and this effect to the human-controlled plant dynamics could be connected to
-this.
+The controller design we utilize, Equation~\ref{eq:implemented-controller},
+also increases the weave mode frequency by a factor of about three up in a
+bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
+human's neuromuscular system, but may feel unusual as it is more akin to what
+the steering would feel like at in the \SIrange{30}{40}{\kph} speed range.
+\Citet{Hanakam2023} reported dissatisfaction in subjective rider feeling on
+their similar bicycle to ours and this effect to the human-controlled plant
+dynamics could be an explanation to their findings.
 
-\subsection{Application of the Logistic Model}
+\subsection{Utility of the Logistic Model and Extrapolation to Natural Falls}
 %
@@
 insight.
-But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
+To interpret the results in Tables~\ref{tab:freq-coefs-6} and
 \ref{tab:freq-coefs-10} it is important to understand the relationship between
@@
 reduces the odds to 500:1.
-However, that case the probability that a fall occurs is only reduced from
+However, in that case the probability that a fall occurs is only reduced from
 $0.999$ to $0.998$.
-If the odds that a fall will occur are smaller, halving the odds has a larger
+If the odds that a fall occurs are smaller, halving the odds has a larger
 influence on the fall probability.
@@
 This means that the balance assist is most effective for perturbations close to
-the participant's personal threshold between falling or recovery and that large
+the participant's perturbation resistance threshold and that large
 perturbations will make you fall regardless of the balance assist's help.
@@
 will give an example of how the data collected during the experiments is used
-to predict fall probability. We use Equation~\ref{eq:log-regress} and the coefficients
-in \ref{tab:freq-coefs-6}. For simplicity's sake, the interaction effects are
-not included. For example, assuming that the mean angular impulse $\bar{L}$ of all the
-perturbations applied to a participant is 100~\si{\newton}, and the standard
-deviation $\sigma^{L}=15$. The centered and scaled angular impulse can be
-calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
-dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
-initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
-the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
-can be calculated as follows.
+to predict fall probability. We use Equation~\ref{eq:log-regress} and the
+coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
+interaction effects are not included. For example, assuming that the mean
+angular impulse $\bar{L}$ of all the perturbations applied to a participant is
+100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
+scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
+applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
+applies for the perturbation order $j$, initial roll angle $\phi_0$, and
+initial steer angle $\delta_0$. If we take the coefficients estimated for
+cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows.
+\todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
 %
@@
     = & \quad
-    \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}} \\
+    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
     = &
@@
 %
-The state of the balance assist \(s\) is a binary variable. If the
+The state of the balance assist \(c\) is a binary variable. If the
 balance assist is turned on, the log-odds that a fall occurs are decreased by
@@
 Turning on the balance assist system reduces the probability that the
-perturbation results in a fall from 81\% to 69\%.
+perturbation results in a fall from 81\% to 69\% in this case.
 
@@
 populate the model.
+
+The positive effect of the balance assist system is coupled to the assumptions
+and experimental scenarios we implemented and there is unfortunately no simple
+way to extrapolate our results to reductions of single-actor crashes we may see
+if such a system were deployed widely to bicyclists. Although, our results do
+indicate that we would see such a reduction, even if only in a class of
+single-actor crashes that most resemble our experimental design. If there were
+more comprehensive and detailed natural data of how people fall we could make
+estimates on the number of falls reduced if everyone road a balance assist
+bicycle.
 
@@
 speeds. We posit that this likely has to do with the width of the treadmill and
-that this could also be why the balance assist system did not have a statistically
-significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
-same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
-almost all falls were due to the bicycle exiting the maximum width of the
-treadmill. If the same experiment was performed on an infinitely wide plane,
-the riders may have recovered from more perturbations. At 6~\si{\kph} the
-riders could often recover in the allotted treadmill width due to the smaller
-lateral deviations. We believe our results are very much dependent on the two
-modes of falling with use: exit the treadmill width or foot is placed on the
-belt. Cycle paths are a similar width as the treadmill, so rider's are often
-limited in width when recovering from a fall.
+that this could also be why the balance assist system did not have a
+statistically significant effect at 10~\si{\kph}. As a bicycle travels at
+higher speeds, the same perturbation magnitude causes larger lateral
+deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
+the maximum width of the treadmill. If the same experiment was performed on an
+infinitely wide plane, the riders may have recovered from more perturbations.
+At 6~\si{\kph} the riders could often recover in the allotted treadmill width
+due to the smaller lateral deviations. We believe our results are very much
+dependent on the two modes of falling, i.e. exit the treadmill width or foot is
+placed on the belt. On the other hand, Cycle paths are a similar width as the
+treadmill, so rider's are often limited in width when recovering from a fall
+thus exiting the treadmill width may be an appropriate measure for indicating a
+fall.
 
@@
 %
-An increase in the number of perturbation that a participant already
+An increase in the number of perturbations that a participant already
 experienced, decreases the probability that a fall occurs. This is likely due
@@
 
-Roll and steer angle are not a significant predictor of fall probability at 6
-or at 10~\si{\kph}. We expected these variables to have a significant effect
-based on the following reasoning. If you are in a rolled and steered state that
-is far from the upright equilibrium, then a perturbation that further pushes
-you from the equilibrium should have some additive or multiplicative effect on
-the resulting motion trajectory.
+Roll and steer angle at the initiation of a perturbation are not a significant
+predictor of fall probability at 6 or at 10~\si{\kph}. We expected these
+variables to have a significant effect based on the following reasoning. If you
+are in a rolled and steered state that is far from the upright equilibrium,
+then a perturbation that further pushes you from the equilibrium should have
+some additive or multiplicative effect on the resulting motion trajectory and
+make it harder to recover from a fall.
 
@@
 
-\subsection{Extrapolation to Natural Falls}
-%
-The positive effect of the balance assist system is coupled to the assumptions
-and experimental scenarios we implemented and there is unfortunately no simple
-way to extrapolate our results to reductions of single-actor crashes we may see
-if such a system were deployed widely to bicyclists. Although, our results do
-indicate that we would see such a reduction, even if only in a class of
-single-actor crashes that most resemble our experimental design. If there were
-more comprehensive and detailed natural data of how people fall we could make
-estimates on the number of falls reduced if everyone road a balance assist
-bicycle.
-
 \section{Conclusion}
@@
 bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
-which is easier for a rider to balance. Perturbation forces applied to the
-handlebar can cause a rider to fall and every rider has a threshold force at
-which they are more likely to fall than not. The probability of falling when
-mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
-when the balance assisting control is activated. This effect is present when
-traveling at 10~\si{\kph} but more investigation is needed to determine if the
-effect can be significant. The positive effect to balance is rider independent
-and most effective in the regime of perturbations near the rider's control
-authority threshold. Given that similar effects cause falls during bicycling,
-use of the balance assist system in real world use cases will reduce the number
-of falls at low bicycling speeds.
+which is easier for a rider to control in balance. Perturbation forces applied
+to the handlebar can cause a rider to fall and every rider has a perturbation
+magnitude threshold at which they are more likely to fall than not. The
+probability of falling when mechanically perturbed is significantly reduced
+when traveling at 6~\si{\kph} when the balance assisting control is activated.
+This effect is present when traveling at 10~\si{\kph} but more investigation is
+needed to determine if the effect can be significant. The positive effect to
+balance is rider independent and most effective in the regime of perturbations
+near the limits of the rider's control authority. Given that similar effects
+cause falls during bicycling, use of the balance assist system in real world
+use cases will reduce the number of falls at low bicycling speeds.
 
@@
 Reijne, and Arend Schwab for contributions to the development of the balance
-assist bicycle. We acknowledge Shannon van de Velde's and Marco Reijne's
+assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
 development of the bicycle perturbation system. We thank all of the assistants
-that helped run the experiments and the participants for their time an enthusiasm.
-We thank Peter Stahlecker for valuable feedback in early revisions.
+that helped run the experiments and the participants for their time an
+enthusiasm. We thank Peter Stahlecker for valuable feedback in early revisions.
 

### 2024-09-26 src/control.py
Improved stdout.

@@ -83,6 +83,9 @@ def plot_eig(ax, model, kphidots=0.0):
     return ax
 
 
+msg = 'Without rigid rider and balance assist off:'
+print(msg)
+print("-"*len(msg))
 ax = plot_eig(axes[0, 0], model_without)
 sax = ax.secondary_xaxis('top', functions=(lambda x: x*MPS2KPH,
                                            lambda x: x*KPH2MPS))
@@ -91,10 +94,16 @@ ax.set_title('Without Rigid Rider')
 ax.set_ylabel('Balance Assist Off\nEigenvalue Components\n[1/s]')
 ax.set_xlabel('')
 
+msg = '\nWithout rigid rider and balance assist on:'
+print(msg)
+print("-"*len(msg))
 ax = plot_eig(axes[1, 0], model_without, kphidots=kphidots)
 ax.set_ylabel('Balance Assist On\nEigenvalue Components\n[1/s]')
 ax.set_xlabel('Speed [m/s]')
 
+msg = '\nWith rigid rider and balance assist off:'
+print(msg)
+print("-"*len(msg))
 ax = plot_eig(axes[0, 1], model_with)
 sax = ax.secondary_xaxis('top', functions=(lambda x: x*MPS2KPH,
                                            lambda x: x*KPH2MPS))
@@ -103,6 +112,9 @@ ax.set_title('With Rigid Rider')
 ax.set_ylabel('')
 ax.set_xlabel('')
 
+msg = '\nWith rigid rider and balance assist on:'
+print(msg)
+print("-"*len(msg))
 ax = plot_eig(axes[1, 1], model_with, kphidots=kphidots)
 ax.set_ylabel('')
 ax.set_xlabel('Speed [m/s]')

### 2024-09-26 main.tex
Added acknowledgement to Allen Downey.
Added coda and data availability.
Address Marten's feedback.

@@
   motor's possible assistance during falls, we apply varying magnitude external
-  handlebar perturbations to twenty-six participants who rode on a treadmill with
-  the balance assist system activated and deactivated. The probability of not
-  recovering from a handlebar perturbation significantly decreases when the
+  handlebar perturbations to twenty-six participants who rode on a treadmill
+  with the balance assist system activated and deactivated. The probability of
+  recovering from a handlebar perturbation significantly increases when the
   balance assist is activated at a travel speed of 6~\si{\kph}. This positive
@@
 percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
-start up to typical cruising speeds, bicycles themselves are not self-stable
-and can be challenging for the rider to balance. Low speed crashes may be
-reduced if the bicycle was self-stable at these speeds by relieving the rider
-of some of their necessary control activity. Bicycles can be mechanically
-modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
-and since the 1980s it has been known that applying a motor actuated steering
-torque proportional to the vehicle's roll angular rate~\footnote{Roll angular
-rate is defined herein as per the SAE J670 vehicle standards.} can stabilize a
-single-track vehicle down to very low speeds. If automatic control of steering
-can stabilize a bicycle, it may reduce the control required from the rider to
-successfully manage balancing tasks much like the natural high speed
-self-stability already does. We have developed a balance-assisting bicycle,
+start up to typical cruising speeds, bicycles are not self-stable and can be
+challenging for the rider to balance. Low speed crashes may be reduced if the
+bicycle was self-stable at these speeds by relieving the rider of some of their
+necessary control activity. Bicycles can be mechanically modified to lower the
+speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
+has been known that applying a motor actuated steering torque proportional to
+the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
+as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
+down to very low speeds. If automatic control of steering can stabilize a
+bicycle, it may reduce the control required from the rider to successfully
+manage balancing tasks much like the natural high speed self-stability already
+does. We have developed a balance-assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use \(g=8\) and \(g=10\) to tune the gain magnitude during the experiments.
-Scaling the proportional feedback gain linearly with respect to speed
-stabilizes the normally unstable weave mode of the bicycle down to
+We use \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude during the
+experiments.  Scaling the proportional feedback gain linearly with respect to
+speed stabilizes the normally unstable weave mode of the bicycle down to
 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
@@
 assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
-development of the bicycle perturbation system. We thank all of the assistants
-that helped run the experiments and the participants for their time an
-enthusiasm. We thank Peter Stahlecker for valuable feedback in early revisions.
+development of the bicycle perturbation system. Allen Downey provided useful
+insight in the statistical condsiderations. We thank all of the assistants that
+helped run the experiments and the participants for their time an enthusiasm.
+We thank Peter Stahlecker for valuable feedback in early revisions.
 
@@
 no role in the data collection and analysis or preparation of the manuscript.
+
+\section*{Code and Data Availability}
+%
+The data and code to reproduce the paper are available at:
+
+\noindent
+\url{https://github.com/mechmotum/fall-probability-paper}
 

### 2024-09-26 src/statistics.R
Don't use hyphen in balance assist.

@@ -58,29 +58,30 @@ for (i in 1:2) {
       geom_smooth(
           data = dummy_data_on,
           linetype = 'twodash',
-          aes(x = angular_impulse, y = prediction, colour = "Balance-assist on"),
+          aes(x = angular_impulse, y = prediction, color = "Balance assist on"),
           method = "glm",
           method.args = list(family = binomial),
-          fill = "grey"
+          fill = "gray41"
       ) +
       geom_smooth(
           data = dummy_data_off,
-          aes(x = angular_impulse, y = prediction, colour = "Balance-assist off"),
+          linetype = 'solid',
+          aes(x = angular_impulse, y = prediction, color = "Balance assist off"),
           method = "glm",
           method.args = list(family = binomial),
-          fill = "grey"
+          fill = "gray41"
       ) +
-      scale_colour_manual("",
-          breaks = c("Balance-assist on", "Balance-assist off"),
-          values = c("Balance-assist on" = "black",
-                     "Balance-assist off" = "black")) +
+      scale_colour_manual(name = "",  # don't display the word "colour"
+          breaks = c("Balance assist on", "Balance assist off"),
+          values = c("Balance assist on" = "black",
+                     "Balance assist off" = "black")) +
       # TODO : Legend shows a solid line for each line, instead of dashed for
       # one.
       scale_linetype_manual(
-          breaks = c("Balance-assist on", "Balance-assist off"),
-          values = c("Balance-assist on" = "twodash",
-                     "Balance-assist off" = "solid")) +
-      theme_bw() +
+          breaks = c("Balance assist on", "Balance assist off"),
+          values = c("Balance assist on" = "twodash",
+                     "Balance assist off" = "solid")) +
+      theme_bw() +  # makes a white background instead of grey
       theme(legend.position = "bottom") +
       labs(x = "Centered and scaled angular impulse",
            y = "Fall Probability")

### 2024-09-26 src/statistics.R
Fixed the R plots to have correct line types and legend.

@@ -48,6 +48,8 @@ for (i in 1:2) {
     model_simple, dummy_data_off, type = "response"
   )
 
+  all_data <- rbind(dummy_data_on, dummy_data_off)
+
   if (i == 1) {
     file_name <- "./figures/predicted_fall_probability_6kmh.png"
   } else {
@@ -56,31 +58,14 @@ for (i in 1:2) {
 
   plot <- ggplot() +
       geom_smooth(
-          data = dummy_data_on,
-          linetype = 'twodash',
-          aes(x = angular_impulse, y = prediction, color = "Balance assist on"),
+          data = all_data,
+          colour = 'black',
+          aes(x = angular_impulse, y = prediction, linetype = balance_assist),
           method = "glm",
           method.args = list(family = binomial),
           fill = "gray41"
       ) +
-      geom_smooth(
-          data = dummy_data_off,
-          linetype = 'solid',
-          aes(x = angular_impulse, y = prediction, color = "Balance assist off"),
-          method = "glm",
-          method.args = list(family = binomial),
-          fill = "gray41"
-      ) +
-      scale_colour_manual(name = "",  # don't display the word "colour"
-          breaks = c("Balance assist on", "Balance assist off"),
-          values = c("Balance assist on" = "black",
-                     "Balance assist off" = "black")) +
-      # TODO : Legend shows a solid line for each line, instead of dashed for
-      # one.
-      scale_linetype_manual(
-          breaks = c("Balance assist on", "Balance assist off"),
-          values = c("Balance assist on" = "twodash",
-                     "Balance assist off" = "solid")) +
+      scale_linetype_discrete(name = "Balance Assist:", labels = c("On", "Off")) +
       theme_bw() +  # makes a white background instead of grey
       theme(legend.position = "bottom") +
       labs(x = "Centered and scaled angular impulse",

### 2024-09-26 main.tex
Minor wording.

@@
   \caption{Comparison of predicted fall probability for balance assist on
-    (solid) or off (dashed) when all other predictor variables are fixed at zero,
+    (solid) or off (dashed) when all other predictor variables are set to zero,
     except for the interaction effect between the balance assist and the

### 2024-09-26 main.tex
Switch to numbered citations.

@@
 \usepackage[margin=25mm]{geometry}
-\usepackage[natbib=true,style=authoryear]{biblatex}
+\usepackage[natbib=true,style=numeric]{biblatex}
 % NOTE : this file is automatically generated from Zotero, do not edit
@@
 as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
-down to very low speeds. If automatic control of steering can stabilize a
-bicycle, it may reduce the control required from the rider to successfully
-manage balancing tasks much like the natural high speed self-stability already
-does. We have developed a balance-assisting bicycle,
+down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
+stabilize a bicycle, it may reduce the control required from the rider to
+successfully manage balancing tasks much like the natural high speed
+self-stability already does. We have developed a balance-assisting bicycle,
 Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
@@
 During the early years of developments in automatic control,
-\citep{Whipple1899} not only derived the correct equations of motion of the
+\citet{Whipple1899} not only derived the correct equations of motion of the
 bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
@@
 feedback of the vehicle roll angle and angular rate was essential to stabilize
-the vehicle. It was not until the early 1980s that \cite{Nagai1983}
+the vehicle. It was not until the early 1980s that \citet{Nagai1983}
 successfully demonstrated balancing a robotic bicycle on a treadmill with both
@@
 from the benchmark bicycle model presented in
-\citet{Meijaard2007}.\todo{Uncertainties are based on best guess for
+\citet{Meijaard2007}.\todo[inline]{Uncertainties are based on best guess for
 uncertainty in the raw measurements, may be better to remove them as they

### 2024-09-26 main.tex
Finished formatting for the journal, fixes #9.

@@
 }
+
+\section*{Affliation}
+Department of BioMechanical Engineering \\
+Delft University of Technology \\
+Delft, The Netherlands \\
+Correspondence: j.k.moore@tudelft.nl
+
+\section*{Keywords}
+bicycle, fall, prevention, automatic control, stability
+
+\section*{Highlights}
+%
+\begin{itemize}
+  \item A bicycle steering motor with automatic roll rate feedback control can
+    stabilize the bicycle-rider system at speeds as low as 4.6~\si{\kph}.
+  \item Every bicycle rider has their own perturbation resistance threshold
+    when subjected to mechanical handlebar perturbations.
+  \item Automatic balance assistance reduces the probability of falling while
+    riding at 6~\si{\kph} when mechanically perturbed.
+  \item Balance assist technology can reduce natural single-actor falls around
+    the limits of the rider's control authority.
+\end{itemize}
 
@@
 \appendix
+\setcounter{table}{0}
+\renewcommand{\thetable}{A.\arabic{table}}
 %

### 2024-09-26 main.tex
Sort citations by location in text.
Removed uncertainty estimation from parameter table.

@@
 \usepackage[margin=25mm]{geometry}
-\usepackage[natbib=true,style=numeric]{biblatex}
+\usepackage[natbib=true,style=numeric,sorting=none]{biblatex}
 % NOTE : this file is automatically generated from Zotero, do not edit
@@
 %
-Physical parameters and estimated uncertainties (standard deviation) for the
-balance assist bicycle with and without a rider. These are found using the
-measurement approach for geometric and inertial parameters for a bicycle
-explained in \citet{Moore2012}. The rider inertial parameters are derived from
-the 83.5~\si{\kilogram} rider ``Jason'' available in BicycleParameters
-1.1.1~\citep{Moore2024}. The variable names match the notation and definitions
-from the benchmark bicycle model presented in
-\citet{Meijaard2007}.\todo[inline]{Uncertainties are based on best guess for
-uncertainty in the raw measurements, may be better to remove them as they
-aren't so meaningful.}
+Physical parameters for the balance assist bicycle with and without a rider.
+These are found using the measurement approach for geometric and inertial
+parameters for a bicycle explained in \citet{Moore2012}. The rider inertial
+parameters are derived from the 83.5~\si{\kilogram} rider ``Jason'' available
+in BicycleParameters 1.1.1~\citep{Moore2024}. The variable names match the
+notation and definitions from the benchmark bicycle model presented in
+\citet{Meijaard2007}.
 %
-\begin{table}[h]  % have to use "h" so that it stays within the appendix
+\begin{table}[ht]  % have to use "h" so that it stays within the appendix
   \centering
@@
     the inertia effects of a rigid rider.}
-  \begin{tabular}{llrrrr}
+  \begin{tabular}{llrr}
     \toprule
-    Variable & Unit & Without Rider & Uncertainty & With Rider & Uncertainty \\
+    Variable & Unit & Without Rider & With Rider \\
     \midrule
-    \(c\) & \si{\meter} & 0.04 & 0.01 & 0.04 & 0.01 \\
-    \(g\) & \si{\meter\per\second\squared} & 9.80665 & NA & 9.80665 & NA \\
-    \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 0.02 & 19.7 & 0.3 \\
-    \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & 0.03 & -3.5 & 0.2 \\
-    \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 0.08 & 22.7 & 0.4 \\
-    \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 0.03 & 5.0 & 0.1 \\
-    \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0005 & 0.0995 & 0.0005 \\
-    \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.0009 & 0.1902 & 0.0009 \\
-    \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.003 & 0.298 & 0.003 \\
-    \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & 0.007 & -0.038 & 0.007 \\
-    \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.006 & 0.257 & 0.006 \\
-    \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.003 & 0.057 & 0.003 \\
-    \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.0006 & 0.1023 & 0.0006 \\
-    \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.0007 & 0.1887 & 0.0007 \\
-    \(\lambda\) & \si{\radian} & 0.25 & 0.03 & 0.25 & 0.03 \\
-    \(m_B\) & \si{\kilogram} & 22.50 & 0.01 & 106.00 & 0.01 \\
-    \(m_F\) & \si{\kilogram} & 2.23 & 0.01 & 2.23 & 0.01 \\
-    \(m_H\) & \si{\kilogram} & 4.30 & 0.01 & 4.30 & 0.01 \\
-    \(m_R\) & \si{\kilogram} & 4.08 & 0.01 & 4.08 & 0.01 \\
-    \(r_F\) & \si{\meter} & 0.3523 & 0.0001 & 0.3523 & 0.0001 \\
-    \(r_R\) & \si{\meter} & 0.3489 & 0.0001 & 0.3489 & 0.0001 \\
-    \(w\) & \si{\meter} & 1.113 & 0.008 & 1.113 & 0.008 \\
-    \(x_B\) & \si{\meter} & 0.52 & 0.01 & 0.374 & 0.002 \\
-    \(x_H\) & \si{\meter} & 0.92 & 0.02 & 0.92 & 0.02 \\
-    \(z_B\) & \si{\meter} & -0.52 & 0.02 & -1.009 & 0.003 \\
-    \(z_H\) & \si{\meter} & -0.860 & 0.006 & -0.860 & 0.006 \\
+    \(c\) & \si{\meter} & 0.04 & 0.04 \\
+    \(g\) & \si{\meter\per\second\squared} & 9.80665 & 9.80665 \\
+    \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 19.7 \\
+    \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & -3.5 \\
+    \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 22.7 \\
+    \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 5.0 \\
+    \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0995 \\
+    \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.1902 \\
+    \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.298 \\
+    \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & -0.038 \\
+    \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.257 \\
+    \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.057 \\
+    \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.1023 \\
+    \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.1887 \\
+    \(\lambda\) & \si{\radian} & 0.25 & 0.25 \\
+    \(m_B\) & \si{\kilogram} & 22.50 & 106.00 \\
+    \(m_F\) & \si{\kilogram} & 2.23 & 2.23 \\
+    \(m_H\) & \si{\kilogram} & 4.30 & 4.30 \\
+    \(m_R\) & \si{\kilogram} & 4.08 & 4.08 \\
+    \(r_F\) & \si{\meter} & 0.3523 & 0.3523 \\
+    \(r_R\) & \si{\meter} & 0.3489 & 0.3489 \\
+    \(w\) & \si{\meter} & 1.113 & 1.113 \\
+    \(x_B\) & \si{\meter} & 0.52 & 0.374 \\
+    \(x_H\) & \si{\meter} & 0.92 & 0.92 \\
+    \(z_B\) & \si{\meter} & -0.52 & -1.009 \\
+    \(z_H\) & \si{\meter} & -0.860 & -0.860 \\
     \bottomrule

### 2024-09-28 main.tex
Addressed many of Benjamin's feedback comments.

@@
   automatically controlled steering motor to a consumer electric bicycle that
-  has a stabilizing effect down to about 3~\si{\kph}. We hypothesize that a
+  has a stabilizing effect down to about 4~\si{\kph}. We hypothesize that a
   stabilized bicycle will reduce the probability of falling. To test the
@@
     column and data acquisition and control electronics mounted in the rear
-    rack.}
+    rack first introduced in \citet{Alizadehsaravi2023}.}
   \label{fig:balance-assist-bicycle}
@@
 
-In this paper, we test whether a steer motor controlled bicycle, that is stable
-in a large low speed range, is beneficial in helping to prevent the rider from
-falling. We test this by applying varying magnitude mechanical perturbations to
-the handlebars while the rider is cycling on a treadmill. We assess the
-difference in the rider's probability of falling with the balance assist system
-on and off.
+In this paper, we investigate whether an automatically controlled bicycle, that
+is stable in a large low speed range for balance assist, is beneficial in
+helping to prevent the rider from falling. We test this possible benefit by
+applying varying magnitude mechanical perturbations to the handlebars while the
+rider is cycling on a treadmill. We assess the difference in the rider's
+probability of falling with the balance assist system on and off.
 
@@
   \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
+  \textrm{.}
 \end{align}
@@
 implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
-is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at at
+is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at
 least 100 Hz.
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude during the
-experiments.  Scaling the proportional feedback gain linearly with respect to
-speed stabilizes the normally unstable weave mode of the bicycle down to
-2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
+We use the factor \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude
+during the experiments. Scaling the proportional feedback gain linearly with
+respect to speed stabilizes the normally unstable weave mode of the bicycle
+down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
 4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
@@
 %
-We apply longitudinal forces to the ends of each handlebar from Earth fixed
+We apply longitudinal forces to the ends of each handlebar from earth-fixed
 posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
@@
 500~\si{\newton}. The commanded force profiles are designed to apply an
-external pulsive torque to the front assembly (handlebars, forks, wheel) at
-magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors are
-arranged at the four corners of a 1~\si{\meter} wide by 2~\si{\meter} long
-treadmill (Bonte Technology B.V., Zwolle, The Netherlands) that can reach
-speeds up to 18~\si{\kilo\meter\per\hour}. The general design of this
-perturbation system is described in detail in Van De Velde's MSc
-thesis~\citep{vandeVelde2022a} and the physical arrangement is shown in
-Figure~\ref{fig:participant-in-set-up}. Our modifications relative to Van De
-Velde's design include simplifying the Bump'Em motor controller with an
-inexpensive microcontroller and the use of a simpler non-active safety harness.
+external pulsive torque to the front frame at magnitudes varying from
+\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
+corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
+Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
+18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
+described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
+the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
+Our modifications relative to Van De Velde's design include simplifying the
+Bump'Em motor controller with an inexpensive microcontroller and the use of a
+simpler non-active safety harness.
 %
@@
 participants performed the protocol at a belt speed of
-10~\si{\kph}(2.8~\si{\mps}) with the controller gain value set to $g=8$ and the
-second group of eleven particpants performed the protocol at a belt speed of
-6~\si{\kph} (1.7~\si{\mps}) with the controller gain set to $g=10$. The
-10~\si{\kph} group's experiments occurred two weeks before the 6~\si{\kph}
-group.
+10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to $g=8$ and two
+weeks later the second group of eleven particpants performed the protocol at a
+belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the controller factor set to
+$g=10$.
 
@@
 perturbations. We randomize whether the balance assist system is on or off
-during the first or second set of 20 perturbations for each participant.
+during the first or second set of 20 perturbations for each participant and the
+participants were not told the order.
 
@@
   - F_{lr})\right] dt
-  \textrm{.}
   \label{eq:angular-impulse}
 \end{align}
-
+%
 where \(F_r\) and \(F_l\)  is the total force applied on the right and left
@@
 instantaneous steer and roll angles to characterize the configuration of the
-bicycle when perturbed. The gain setting on the balance assist controller
+bicycle when perturbed. The gain factor setting on the balance assist controller
 indicates if the assistance is off \(g=0\) or on at two different levels: low
@@
 wheel of the bicycle exits the width of the treadmill belt. We record this as a
-binary variable \(f\) for "fall". All measured variables are reported in
+binary variable \(f\) for ``fall''. All measured variables are reported in
 Table~\ref{tab:raw}.
@@
 there to be a variation between participants in how well they are able to
-resist the perturbation. However, this was not true. Cluster-mean centering
+resist the perturbation; however, this was not true. Cluster-mean centering
 showed there to be no variation between participants. This fact allows us to
@@
 The coefficient estimates for a single-level logistic regression for the data
-from the 6~\si{\kph}, gain $g=10$ trials are shown in
+from the 6~\si{\kph}, gain factor $g=10$ trials are shown in
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
@@
   \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    6~\si{\kph} and gain $g=10$ along with the standard error SE, p-value
+    6~\si{\kph} and gain factor $g=10$ along with the standard error SE, p-value
     \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kph} with gain $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
+10~\si{\kph} with gain factor $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
 angular impulse and perturbation order are statistically significant predictors
@@
   \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    10~\si{\kph} and gain $g=8$ along with the standard error SE, p-value
+    10~\si{\kph} and gain factor $g=8$ along with the standard error SE, p-value
     \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
@@
 bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
-human's neuromuscular system, but may feel unusual as it is more akin to what
-the steering would feel like at in the \SIrange{30}{40}{\kph} speed range.
-\Citet{Hanakam2023} reported dissatisfaction in subjective rider feeling on
-their similar bicycle to ours and this effect to the human-controlled plant
-dynamics could be an explanation to their findings.
+human's neuromuscular system~\citep{Magdaleno1971}, but may feel unusual as it
+is more akin to what the steering would feel like at in the
+\SIrange{30}{40}{\kph} speed range.  \Citet{Hanakam2023} reported
+dissatisfaction in subjective rider feeling on their similar bicycle to ours
+and this effect to the human-controlled plant dynamics could be an explanation
+to their findings.
 
@@
 helped run the experiments and the participants for their time an enthusiasm.
-We thank Peter Stahlecker for valuable feedback in early revisions.
+We thank Peter Stahlecker and Benjamin Gonzalez for valuable feedback in early
+revisions.
 

### 2024-09-28 main.tex
Corrected speed sensor name, as per Felix's feedback.

@@
 %
-We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
+We modified a Grenoble C8 HMB electric bicycle (Royal Dutch Gazelle, Dieren,
 The Netherlands) with a custom motor in the steering assembly capable of
@@
 converts the commanded torque to applied torque. We measure the speed of the
-rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
-and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
-(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
-implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
-is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at
-least 100 Hz.
+rear wheel with a DF30 Speed Sensor (Bosch eBike Systems, Reutlingen, Germany)
+and measure the body fixed roll rate of the bicycle with a VR IMU BN0086 MEMs
+rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
+implemented on a Teensy microprocessor (PJRC, USA) and data from all sensors is
+logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhoej, Denmark) at least
+100 Hz.
 
@@
 posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
-four motors (EC-90 Flat, Maxon Group, Switzerland) working in cooperation,
+four EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
 Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
 force at all times to keep the ropes taught and to track a commanded force
-profile using a PID controller running on a microprocessor (Arduino Mega 2560,
-Arduino, Italy). We measure the force applied by each motor at the handlebar
+profile using a PID controller running on an Arduino Mega 2560 microprocessor
+(Arduino, Italy). We measure the force applied by each motor at the handlebar
 via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
@@
     \midrule
-    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
+    Balance Assist Gain Factor & \(g\) & \si{\newton\second\squared} & NA \\
     Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\

### 2024-09-28 main.tex
Addressed some of Riender's feedback.

@@
     angular impulse. The abscissa is the standard deviation around the mean
-    (all perturbations), centered per participant.}
+    of all perturbations, centered per participant.}
   \label{fig:probability}
@@
 
+\subsection{On Rider Bicycling Skill and Experience}
+%
+Each individual participant had their own perturbation resistance threshold,
+i.e. the perturbation magnitude defining whether they would fall 50\% of the
+time.
+The value of this threshold is likely correlated with the skill and experience
+of the participant at balancing a bicycle.
+With this study, we cannot answer whether the balance assist system helps low
+skilled participants more than highly skilled participants because that is
+simply a function of how often a perturbation in a natural setting exceeds an
+individual's threshold.
+But, we did find that the system can help regardless of a participant's
+individual balance performance.
+On average, the balance assist system pushes the perturbation resistance
+threshold up for all participants.
+The vast majority of perturbations experienced in natural bicycling are below
+an individual's perturbation resistance threshold, otherwise we would observe
+bicyclists falling much more but low skilled riders may experience more
+perturbations that exceed their threshold.
+Determining how often certain perturbation types and magnitudes are experienced
+in natural bicycling is an open research need and would help understand whether
+a balance assist intervention would have a strong effect when used in the wild.
+
 \subsection{Learning Effect}
@@
 magnitude threshold at which they are more likely to fall than not. The
-probability of falling when mechanically perturbed is significantly reduced
-when traveling at 6~\si{\kph} when the balance assisting control is activated.
-This effect is present when traveling at 10~\si{\kph} but more investigation is
-needed to determine if the effect can be significant. The positive effect to
-balance is rider independent and most effective in the regime of perturbations
-near the limits of the rider's control authority. Given that similar effects
-cause falls during bicycling, use of the balance assist system in real world
-use cases will reduce the number of falls at low bicycling speeds.
+probability of falling when mechanically perturbed is significantly
+(\(p=0.02\)) reduced when traveling at 6~\si{\kph} when the balance assisting
+control is activated. This effect is present when traveling at 10~\si{\kph}
+and is very close to significance (\(p=0.07\)) but more investigation is needed
+to determine if the effect can be significant. The positive effect to balance
+is rider independent and most effective in the regime of perturbations near the
+limits of the rider's control authority. Given that similar effects cause falls
+during bicycling, use of the balance assist system in real world use cases will
+reduce the number of falls at low bicycling speeds.
 
@@
 helped run the experiments and the participants for their time an enthusiasm.
-We thank Peter Stahlecker and Benjamin Gonzalez for valuable feedback in early
-revisions.
+We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable
+feedback in early revisions.
 

### 2024-09-28 main.tex
Changed g to eta and c to s, fixes #20.

@@
 \begin{align}
-  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
+  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \eta(v_{stable} - v)\dot{\phi}
   \label{eq:implemented-controller}
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use the factor \(g=8\) (low) and \(g=10\) (high) to tune the gain magnitude
-during the experiments. Scaling the proportional feedback gain linearly with
-respect to speed stabilizes the normally unstable weave mode of the bicycle
-down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
+We use the gain factor \(\eta=8\) (low) and \(\eta=10\) (high) to tune the gain
+magnitude during the experiments. Scaling the proportional feedback gain
+linearly with respect to speed stabilizes the normally unstable weave mode of
+the bicycle down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
 4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
@@
   \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled with \(g=10\) (lower row)
+  \caption{Uncontrolled (upper row) and controlled with \(\eta=10\) (lower row)
     root locus of eigenvalue components (real: solid, imaginary: dashed)
@@
 bicycle when perturbed. The gain factor setting on the balance assist controller
-indicates if the assistance is off \(g=0\) or on at two different levels: low
-\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
+indicates if the assistance is off \(\eta=0\) or on at two different levels: low
+\(\eta=8\) or high \(\eta=10\). A recovery from the perturbation is successful if the
 person neither places their foot down onto the treadmill surface nor their
@@
     \midrule
-    Balance Assist Gain Factor & \(g\) & \si{\newton\second\squared} & NA \\
+    Balance Assist Gain Factor & \(\eta\) & \si{\newton\second\squared} & NA \\
     Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
@@
     \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
-    \(c\)        & independent & boolean & balance assist: off 0 or on 1 \\
+    \(s\)        & independent & boolean & balance assist: off 0 or on 1 \\
     \(j\)        & independent & integer & order number of perturbation \\
@@
 assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
-for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
+separate model evaluations, one for the 6~\si{\kph}, \(\eta=10\) trials and one
+for the 10~\si{\kph}, \(\eta=8\) trials and we evaluate our hypothesis for each
 set of data. Statistics were computed with the lme4 version
@@
     Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
-    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
+    Balance assist state \(s\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
     Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
@@
     Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
-    Balance assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
+    Balance assist on \(s\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
     Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
@@
 initial steer angle $\delta_0$. If we take the coefficients estimated for
-cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows.
+cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows:
 \todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
@@
     & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
-    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot c \\
+    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
     = &
-    1.42 -0.64 \cdot c
+    1.42 -0.64 \cdot s
   \end{split}
+  \textrm{.}
 \end{equation}
 %
-The state of the balance assist \(c\) is a binary variable. If the
+The state of the balance assist \(s\) is a binary variable. If the
 balance assist is turned on, the log-odds that a fall occurs are decreased by
@@
 \begin{align}
-  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64c} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53c \\
+  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64s} =
+  e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s \\
   p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\

### 2024-10-03 src/model.py
Docstring edit.

@@ -16,8 +16,8 @@ class SteerControlModel(Meijaard2007Model):
          roll rate,
          steer rate]
 
-        The inputs are [roll torque,
-                        steer torque]
+    The inputs are [roll torque,
+                    steer torque]
 
     """
 

### 2024-10-03 main.tex
Fix probability example
Addressed Leila's comments.

@@
 \usepackage[margin=25mm]{geometry}
-\usepackage[natbib=true,style=numeric,sorting=none]{biblatex}
+\usepackage[natbib=true,style=numeric,sorting=none,giveninits=true]{biblatex}
 % NOTE : this file is automatically generated from Zotero, do not edit
@@
   automatically controlled steering motor to a consumer electric bicycle that
-  has a stabilizing effect down to about 4~\si{\kph}. We hypothesize that a
-  stabilized bicycle will reduce the probability of falling. To test the
+  stabilizes the riderless bicycle down to about 3~\si{\kph}. We hypothesize
+  that a stabilized bicycle will reduce the probability of falling. To test the
   motor's possible assistance during falls, we apply varying magnitude external
   handlebar perturbations to twenty-six participants who rode on a treadmill
-  with the balance assist system activated and deactivated. The probability of
-  recovering from a handlebar perturbation significantly increases when the
-  balance assist is activated at a travel speed of 6~\si{\kph}. This positive
-  effect is most prominent at and around the individual riders' perturbation
-  resistance threshold. We conclude that use of a balance assist system in real
-  world bicycling would reduce the number of falls that occur near riders'
-  control authority limits.
+  with the balance assist system both activated and deactivated. The
+  probability of recovering from a handlebar perturbation significantly
+  increases when the balance assist is activated at a travel speed of
+  6~\si{\kph}. This positive effect is most prominent at and around the
+  individual riders' perturbation resistance threshold. We conclude that use of
+  a balance assist system in real world bicycling would reduce the number of
+  falls that occur near riders' control authority limits.
 }
@@
 \section*{Keywords}
-bicycle, fall, prevention, automatic control, stability
+bicycle, fall prevention, automatic control, stability
 
@@
 breakthrough by demonstrating an automatically balanced motorcycle and they did
-so solely with a steering motor. Ruijs and Pacejka clearly showed that steer
-torque driven by roll angle feedback stabilizes the capsize mode, by roll
-angular rate feedback stabilizes the weave mode, and by steer angular rate
-feedback stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle)
-eigenmodes are defined in \citep{Sharp1971}.} They also showed how the control
-gains must change with respect to vehicle speed for favorable control across
-all speeds. This roll motion feedback enables the simplest controller that can
-stabilize a single-track vehicle above a minimum speed when one is not
-concerned with wobble instabilities. But Ruijs and Pacejka's work was not
-particularly concerned with low speed stability and their vehicle was fully
-automatic, i.e no human rider was involved.
+so solely with a steering motor. Ruijs and Pacejka showed that steer torque
+driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
+feedback stabilizes the weave mode, and by steer angular rate feedback
+stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
+are defined in \citep{Sharp1971}.} They also showed how the control gains must
+change with respect to vehicle speed for favorable control across all speeds.
+This roll motion feedback enables the simplest controller that can stabilize a
+single-track vehicle above a minimum speed when one is not concerned with
+wobble instabilities. But Ruijs and Pacejka's work was not particularly
+concerned with low speed stability and their vehicle was fully automatic, i.e
+no human rider was involved.
 
@@
 safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
-free movement pre-fall. The participants practiced riding on the treadmill
-until they indicated they were comfortable enough to have perturbations
-applied. For most, this was less than a 10~\si{\minute} warm up. We then asked
-the rider to ride for 90~\si{\second}, while attempting to maintain the
-location of their front wheel on the center line of the treadmill as a baseline
-measure before the perturbations. We then applied perturbations in random
-directions (clockwise or counter-clockwise), starting at 20~\si{\newton} and
-increasing the magnitude by 30~\si{\newton} until the participants fell. We
-defined a ``fall'' on the treadmill by two criteria: 1) the rider removes their
-foot from the pedal and places it on the ground or 2) the bicycle wheel exceeds
-the width of the treadmill belt. Figure~\ref{fig:perturbation-sequence} shows
-an example resulting motion from a perturbation. We logged the force magnitude
-that caused the first fall to characterize that participant's
-\emph{perturbation resistance threshold}.
+natural free movement pre-fall. The participants practiced riding on the
+treadmill until they indicated they were comfortable enough to have
+perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
+We then asked the rider to ride for 90~\si{\second}, while attempting to
+maintain the location of their front wheel on the center line of the treadmill
+as a baseline measure before the perturbations. We then applied perturbations
+in random directions (clockwise or counter-clockwise), starting at
+20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
+participants fell. We defined a ``fall'' on the treadmill by two criteria: 1)
+the rider removes their foot from the pedal and places it on the ground or 2)
+the bicycle wheel exceeds the width of the treadmill belt.
+Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
+perturbation. We logged the force magnitude that caused the first fall to
+characterize that participant's \emph{perturbation resistance threshold}.
 %
@@
     = &
-   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
+   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 100\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
     & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
@@
     = &
-    1.42 -0.64 \cdot s
+    1.59 -0.64 \cdot s
   \end{split}
@@
 \begin{align}
-  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64s} =
-  e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s \\
-  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
-  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
+  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64c} =
+  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot 0.53c \\
+  p_{ij}|_{c=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
+  p_{ij}|_{c=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
 \end{align}
@@
 Turning on the balance assist system reduces the probability that the
-perturbation results in a fall from 81\% to 69\% in this case.
+perturbation results in a fall from 83\% to 72\% in this case.
 
@@
 dependent on the two modes of falling, i.e. exit the treadmill width or foot is
-placed on the belt. On the other hand, Cycle paths are a similar width as the
+placed on the belt. On the other hand, cycle paths are a similar width as the
 treadmill, so rider's are often limited in width when recovering from a fall
 thus exiting the treadmill width may be an appropriate measure for indicating a
-fall.
+fall. While the treadmill simulates narrow cycle paths, real-world paths may
+offer more lateral recovery space, providing riders with additional
+opportunities to regain balance after a perturbation. However, testing in such
+narrow conditions is still highly relevant, as system design and validation
+should focus on extreme conditions like narrow paths, because even in wider
+paths, obstacles such as parked cars or barriers can limit lateral space.
 

### 2024-10-03 main.tex
Use kappa for the gain factor.

@@
 \begin{align}
-  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \eta(v_{stable} - v)\dot{\phi}
+  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
   \label{eq:implemented-controller}
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use the gain factor \(\eta=8\) (low) and \(\eta=10\) (high) to tune the gain
+We use the gain factor \(\kappa=8\) (low) and \(\kappa=10\) (high) to tune the gain
 magnitude during the experiments. Scaling the proportional feedback gain
@@
   \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled with \(\eta=10\) (lower row)
+  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower row)
     root locus of eigenvalue components (real: solid, imaginary: dashed)
@@
     parameters for these plots were estimated with the methods presented in
-    \citet{Moore2012} and software packages BicycleParameters
-    1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015} and are shown
-    in Appendix~\ref{sec:physical-parameters}.}
+    \citet{Moore2012} and software packages
+    BicycleParameters~1.1.1~\citep{Moore2024} and
+    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
+    Appendix~\ref{sec:physical-parameters}.}
   \label{fig:eigenvalues}
@@
 participants performed the protocol at a belt speed of
-10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to $g=8$ and two
+10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to \(\kappa=8\) and two
 weeks later the second group of eleven particpants performed the protocol at a
 belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the controller factor set to
-$g=10$.
+\(\kappa=10\).
 
@@
 bicycle when perturbed. The gain factor setting on the balance assist controller
-indicates if the assistance is off \(\eta=0\) or on at two different levels: low
-\(\eta=8\) or high \(\eta=10\). A recovery from the perturbation is successful if the
+indicates if the assistance is off \(\kappa=0\) or on at two different levels: low
+\(\kappa=8\) or high \(\kappa=10\). A recovery from the perturbation is successful if the
 person neither places their foot down onto the treadmill surface nor their
@@
     \midrule
-    Balance Assist Gain Factor & \(\eta\) & \si{\newton\second\squared} & NA \\
+    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
     Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
@@
 assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(\eta=10\) trials and one
-for the 10~\si{\kph}, \(\eta=8\) trials and we evaluate our hypothesis for each
+separate model evaluations, one for the 6~\si{\kph}, \(\kappa=10\) trials and one
+for the 10~\si{\kph}, \(\kappa=8\) trials and we evaluate our hypothesis for each
 set of data. Statistics were computed with the lme4 version
@@
 The coefficient estimates for a single-level logistic regression for the data
-from the 6~\si{\kph}, gain factor $g=10$ trials are shown in
+from the 6~\si{\kph}, gain factor \(\kappa=10\) trials are shown in
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
@@
   \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    6~\si{\kph} and gain factor $g=10$ along with the standard error SE, p-value
+    6~\si{\kph} and gain factor \(\kappa=10\) along with the standard error SE, p-value
     \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kph} with gain factor $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
+10~\si{\kph} with gain factor \(\kappa=8\) are shown in Table \ref{tab:freq-coefs-10}. The
 angular impulse and perturbation order are statistically significant predictors
@@
   \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    10~\si{\kph} and gain factor $g=8$ along with the standard error SE, p-value
+    10~\si{\kph} and gain factor \(\kappa=8\) along with the standard error SE, p-value
     \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
@@
 parameters are derived from the 83.5~\si{\kilogram} rider ``Jason'' available
-in BicycleParameters 1.1.1~\citep{Moore2024}. The variable names match the
+in BicycleParameters~1.1.1~\citep{Moore2024}. The variable names match the
 notation and definitions from the benchmark bicycle model presented in

### 2024-10-03 main.tex
Slight correction to Marten's equations.

@@
 initial steer angle $\delta_0$. If we take the coefficients estimated for
-cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows:
-\todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
+cycling at 6~\si{\kph} and example means, the log-odds of falling can be
+calculated as follows:
 %
@@
     & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
-    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
+    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64 s \\
     = &
-    1.59 -0.64 \cdot s
+    1.59 -0.64 s
+  \textrm{.}
   \end{split}
-  \textrm{.}
 \end{equation}
@@
 \begin{align}
-  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64c} =
-  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot 0.53c \\
-  p_{ij}|_{c=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
-  p_{ij}|_{c=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
+  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64s} =
+  e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot e^{-0.64s} \\
+  p_{ij}|_{s=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
+  p_{ij}|_{s=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
 \end{align}
@@
 This study is funded by Dutch Research Council, Nederlandse Organisatie voor
-Wetenschappelijk Onderzoek (NWO), under the Citius Altius Sanius program and in
-collaboration with Bosch eBike Systems and Royal Dutch Gazelle. The funders had
-no role in the data collection and analysis or preparation of the manuscript.
+Wetenschappelijk Onderzoek (NWO), under the Citius Altius Sanius Perspectief
+program and in collaboration with Bosch eBike Systems and Royal Dutch Gazelle.
+The funders had no role in the data collection and analysis or preparation of
+the manuscript.
 

### 2024-10-04 main.tex
Added balance assist in second line of abstract. Thanks Frans.

@@
   automatically controlled steering motor to a consumer electric bicycle that
-  stabilizes the riderless bicycle down to about 3~\si{\kph}. We hypothesize
-  that a stabilized bicycle will reduce the probability of falling. To test the
-  motor's possible assistance during falls, we apply varying magnitude external
-  handlebar perturbations to twenty-six participants who rode on a treadmill
-  with the balance assist system both activated and deactivated. The
-  probability of recovering from a handlebar perturbation significantly
-  increases when the balance assist is activated at a travel speed of
-  6~\si{\kph}. This positive effect is most prominent at and around the
-  individual riders' perturbation resistance threshold. We conclude that use of
-  a balance assist system in real world bicycling would reduce the number of
-  falls that occur near riders' control authority limits.
+  stabilizes the riderless bicycle down to about 3~\si{\kph} to assist the
+  rider in balancing the vehicle. We hypothesize that a stabilized bicycle will
+  reduce the probability of falling. To test the motor's possible assistance
+  during falls, we apply varying magnitude external handlebar perturbations to
+  twenty-six participants who rode on a treadmill with the balance assist
+  system both activated and deactivated. The probability of recovering from a
+  handlebar perturbation significantly increases when the balance assist is
+  activated at a travel speed of 6~\si{\kph}. This positive effect is most
+  prominent at and around the individual riders' perturbation resistance
+  threshold. We conclude that use of a balance assist system in real world
+  bicycling would reduce the number of falls that occur near riders' control
+  authority limits.
 }
@@
 We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable
-feedback in early revisions.
+feedback in early revisions. Lastly, we thank Frans van der Helm for the
+consistent encouragement to execute this experiment.
 

### 2024-10-09 main.tex
Shortened header and mentioned Frans role in funding.
Final pass of edits just before submitting.

--- 2024-10-03 main.tex
@@
   automatically controlled steering motor to a consumer electric bicycle that
-  stabilizes the riderless bicycle down to about 3~\si{\kph}. We hypothesize
-  that a stabilized bicycle will reduce the probability of falling. To test the
-  motor's possible assistance during falls, we apply varying magnitude external
-  handlebar perturbations to twenty-six participants who rode on a treadmill
-  with the balance assist system both activated and deactivated. The
+  stabilizes the riderless bicycle down to about 3~\si{\kph} to assist a rider
+  in balancing the vehicle. We hypothesize that a such a stabilized bicycle
+  will reduce the probability of falling. To test the system's possible
+  assistance during falls, we applied varying magnitude external handlebar
+  perturbations to twenty-six participants who rode on a treadmill with the
+  balance assist system both activated and deactivated. We show that the
   probability of recovering from a handlebar perturbation significantly
@@
   individual riders' perturbation resistance threshold. We conclude that use of
-  a balance assist system in real world bicycling would reduce the number of
+  a balance assist system in real world bicycling can reduce the number of
   falls that occur near riders' control authority limits.
@@
 stabilize a bicycle, it may reduce the control required from the rider to
-successfully manage balancing tasks much like the natural high speed
+successfully manage balancing tasks much like natural high speed
 self-stability already does. We have developed a balance-assisting bicycle,
@@
 hitting a bump in the road. To assess our balance assisting bicycle, we subject
-the rider to perturbations at the handlebar, which can relatively easily cause
-a rider to fall.
+the rider to perturbations at the handlebar, which can easily cause a rider to
+fall.
 
@@
 change with respect to vehicle speed for favorable control across all speeds.
-This roll motion feedback enables the simplest controller that can stabilize a
+Thus roll motion feedback enables the simplest controller that can stabilize a
 single-track vehicle above a minimum speed when one is not concerned with
@@
 rider if the motor controlled steer torque and the rider applied steer torque
-act on the steer in parallel. The effect of this automatic control gives the
-ability to effectively change the human-controlled plant dynamics, up to some
-limits. In our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced
-motion during distractions due to the balance assist system but
-\citet{Hanakam2023} recently showed rider dissatisfaction with the
-stabilization of a similar vehicle so the possible benefits are not
-definitively established.
+act in parallel. The effect of this automatic control gives the ability to
+effectively change the human-controlled plant dynamics, up to some limits. In
+our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion
+during distractions and light perturbations due to the balance assist system
+but \citet{Hanakam2023} recently showed rider dissatisfaction with the
+stabilization of a similar vehicle so the overall possible benefits are not
+yet definitively established.
 
-The design of a balance assist system relies on the linear Carvallo-Whipple
+The design of our balance assist system relies on the linear Carvallo-Whipple
 bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
-model that exhibits non-minimum phase behavior and self-stability. The bicycle
-model is well suited for showing the effect of a roll motion feedback driven
-steer motor on the dynamics. This model is equivalently valid for on-road or
-treadmill riding~\citep{Kooijman2009a}, which have the same fundamental
-dynamics. The linear version of this model can be described by the fourth order
-state space equations:
+model that exhibits both the non-minimum phase behavior ``countersteering'' and
+self-stability. The bicycle model is well suited for showing the effect of a
+roll motion feedback driven steer motor on the dynamics. This model is
+equivalently valid for on-road or treadmill riding~\citep{Kooijman2009a}, which
+have the same fundamental dynamics. The linear version of this model can be
+described by the fourth order state space equations:
 %
@@
 rate gyroscope (Sparkfun, Niwot, USA). The balance assist control algorithm is
-implemented on a Teensy microprocessor (PJRC, USA) and data from all sensors is
-logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhoej, Denmark) at least
-100 Hz.
+implemented on a Teensy microprocessor (PJRC, Sherwood, USA) and data from all
+sensors is logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhøj, Denmark)
+at least 100 Hz.
 
@@
   \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower row)
-    root locus of eigenvalue components (real: solid, imaginary: dashed)
-    plotted versus speed for the bicycle without the inertial effect of a rigid
-    rider (left) and with a rigid rider (right). Vertical dotted and
-    dotted-dashed lines indicate the two speeds we perform experiments at. Grey
-    shaded region is the linear stable speed range. Geometric and inertial
-    parameters for these plots were estimated with the methods presented in
+  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower
+    row) root locus of the eigenvalue components (real: solid, imaginary:
+    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
+    without the inertial effect of a rigid rider (left) and with a rigid rider
+    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
+    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
+    region is the linear stable speed range. Geometric and inertial parameters
+    for these plots were estimated with the methods presented in
     \citet{Moore2012} and software packages
@@
 participants performed the protocol at a belt speed of
-10~\si{\kph}(2.8~\si{\mps}) with the controller factor set to \(\kappa=8\) and two
-weeks later the second group of eleven particpants performed the protocol at a
-belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the controller factor set to
+10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=8\) and two
+weeks later the second group of eleven participants performed the protocol at a
+belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
 \(\kappa=10\).
@@
   \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
-  \caption{Externally applied perturbation torque alongside the resulting
-    measured motion and steer motor induced torque based on a 110~\si{\newton}
-    counterclockwise applied force at a 6~\si{\kph} travel speed. The circles
-    on the roll and steer angle plot show the angles \(\phi_0\) and
-    \(\delta_0\) at the perturbation start.}
+  \caption{Example of the externally applied perturbation torque alongside the
+    resulting measured motion and steer motor induced torque based on a
+    110~\si{\newton} counterclockwise applied force at a 6~\si{\kph} travel
+    speed. The circles on the roll and steer angle plot show the angles
+    \(\phi_0\) and \(\delta_0\) at the perturbation start.}
   \label{fig:perturbation_0}
@@
 Based on findings from measuring riders without balance assist in
-\citet{Haitjema2023a}, we calculate several variables that we hypothesize may
+\citep{Haitjema2023a}, we calculate several variables that we hypothesize may
 influence fall probability. We use the angular impulse \(L\) of the
@@
 magnitude of delivered perturbation. The duration is selected based on the
-duration of the commanded force and is calculated as follows:
+duration of the commanded perturbation force and is calculated as follows:
 %
@@
   L =
-  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
+  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
   =
-  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
-  - F_{lr})\right] dt
+  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
+    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
   \label{eq:angular-impulse}
@@
 as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
-of peak torque to normalize for the duration of the applied perturbation.
+of peak torque to normalize for the duration of the applied perturbation to
+capture the total effect of the perturbation pulse.
 
@@
 the current perturbation. At the initiation of each perturbation we log the
-instantaneous steer and roll angles to characterize the configuration of the
-bicycle when perturbed. The gain factor setting on the balance assist controller
-indicates if the assistance is off \(\kappa=0\) or on at two different levels: low
-\(\kappa=8\) or high \(\kappa=10\). A recovery from the perturbation is successful if the
-person neither places their foot down onto the treadmill surface nor their
-wheel of the bicycle exits the width of the treadmill belt. We record this as a
-binary variable \(f\) for ``fall''. All measured variables are reported in
-Table~\ref{tab:raw}.
+instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
+characterize the configuration of the bicycle when perturbed. The gain factor
+setting on the balance assist controller indicates if the assistance is off
+\(\kappa=0\) or on at two different levels: low \(\kappa=8\) or high
+\(\kappa=10\). As mentioned earlier, a recovery from the perturbation is
+successful if the person neither places their foot down onto the treadmill
+surface nor their wheel of the bicycle exits the width of the treadmill belt.
+We record this as a binary variable \(f\) for ``fall''.  All measured variables
+are reported in Table~\ref{tab:raw}.
 %
@@
     Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
-    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
+    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
     Fall Outcome & \(f\) & boolean & NA \\
@@
 %
-We test our hypothesis that the balance assist controller will reduce the
+We test our hypothesis that the balance assist control will reduce the
 probability of falling when perturbed externally at the handlebar. We have a
@@
 assist on, decrease the probability to fall. The associated multiplicative
-change in odds are shown in Table~\ref{tab:freq-coefs-6}.
+change in odds are also shown in Table~\ref{tab:freq-coefs-6} and can be used
+to calculate the probability of falling.
 %
@@
   \centering
-  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    6~\si{\kph} and gain factor \(\kappa=10\) along with the standard error SE, p-value
-    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
-    confidence interval bounds.}
+  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
+    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=10\) along with the
+    standard error SE, p-value \(p\), multiplicative change in odds
+    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
   \label{tab:freq-coefs-6}
@@
     \toprule
-    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
+    Variable & \(\beta\) / \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) / \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
     \midrule
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kph} with gain factor \(\kappa=8\) are shown in Table \ref{tab:freq-coefs-10}. The
-angular impulse and perturbation order are statistically significant predictors
-with angular impulse being the dominant effect. Larger angular impulse
-increases the probability to fall and enduring more perturbations decreases the
-probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
-not a significant predictor but the effect would be a reduction in the
-probability to fall if it were. At both speeds, the angular impulse has about
-twice the magnitude in effect as the perturbation order.
+10~\si{\kph} with gain factor \(\kappa=8\) are shown in Table
+\ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
+statistically significant predictors with angular impulse being the dominant
+effect. Larger angular impulse increases the probability to fall and enduring
+more perturbations decreases the probability to fall. Unlike the 6~\si{\kph}
+trials, the balance assist state is not a significant predictor (\(p=0.07\))
+but the effect would be a reduction in the probability to fall if it were. At
+both speeds, the angular impulse has about twice the magnitude in effect as the
+perturbation order.
 %
@@
   \centering
-  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
-    10~\si{\kph} and gain factor \(\kappa=8\) along with the standard error SE, p-value
-    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
-    confidence interval bounds.}
+  \caption{Logistic regression intercept \(\beta\) and coefficient estimates
+    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=8\) along with the
+    standard error SE, p-value \(p\), multiplicative change in odds
+    \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
   \label{tab:freq-coefs-10}
@@
 
-Turning the balance assist system on significantly \((p<0.05)\) reduces the
+Turning the balance assist system on significantly \((p=0.02<0.05)\) reduces the
 odds that a perturbation results in a fall while cycling at a speed of
@@
 falling as a function of the mean and centered angular impulse per participant
-for the balance assist state on and off while keeping all other explanatory
-factors at their centered mean values. This figure is created by setting all
-explanatory variables to their mean and calculating the probability from
-Equation~\ref{eq:log-regress} for only change in angular impulse given the
-estimates in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance
-assist system halves (0.53) the odds that a perturbation results in a fall.
-This figure shows that for relatively large impulses the probability to fall is
-unity for both states of balance assist on and off. And for relatively small
-impulses the probability to fall is null for both states. But for impulses in
-the magnitude region (about -1 to 0.5 STD), i.e. around the mean-centered
-perturbation resistance threshold, the probability of falling is significantly
-lowered with the balance assist system on. The skewness of the probability
-curves arrives from the interaction effects. Figure~\ref{fig:probability-10kph}
-shows the same result for the 10~\si{\kph} trials which has a similar trend of
-reducing the probability to fall with the balance assist system turned on, but
-the effect is not significant.
+for the balance assist state on and off. This figure is created by setting all
+explanatory variables to their centered mean values and calculating the
+probability from Equation~\ref{eq:log-regress} for only change in angular
+impulse given the estimates in Table~\ref{tab:freq-coefs-6}. The table
+indicates that the balance assist system halves (0.53) the odds that a
+perturbation results in a fall.  This figure shows that for relatively large
+impulses the probability to fall is unity for both states of balance assist on
+and off. And for relatively small impulses the probability to fall is null for
+both states. But for impulses in the magnitude region (about -1 to 0.5 STD),
+i.e. around the mean-centered perturbation resistance threshold, the
+probability of falling is significantly lowered with the balance assist system
+on. The skewness of the probability curves arrives from the interaction
+effects. Figure~\ref{fig:probability-10kph} shows the same result for the
+10~\si{\kph} trials which has a similar trend of reducing the probability to
+fall with the balance assist system turned on, but the effect is not
+significant.
 %
@@
 could help understand more completely the range and limits of the positive
-effect of the balance system. For example, it is possible that simply
-increasing the controller gain at 10~\si{\kph} also results in a significant
-positive effect. A longitudinal study of normal use of the balance assist
-bicycle compared to a control group could provide the strongest evidence of any
-benefit we have seen in this specific scenario.
+effect of the balance assist system. For example, it is possible that simply
+increasing the controller gain factor at 10~\si{\kph} also results in a
+significant positive effect. A longitudinal study of normal use of the balance
+assist bicycle compared to a control group could provide the strongest evidence
+of any benefit we have seen in this specific scenario.
 
@@
 stable system, i.e. cannot fall, but having plant eigenvalues with very small
-unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}.
+unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}
+which can be beneficial for avoiding falls.
 
 The controller design we utilize, Equation~\ref{eq:implemented-controller},
-also increases the weave mode frequency by a factor of about three up in a
+also increases the weave mode frequency by a factor of about three in a
 bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
@@
 is more akin to what the steering would feel like at in the
-\SIrange{30}{40}{\kph} speed range.  \Citet{Hanakam2023} reported
-dissatisfaction in subjective rider feeling on their similar bicycle to ours
-and this effect to the human-controlled plant dynamics could be an explanation
-to their findings.
+\SIrange{30}{40}{\kph} speed range for a normal bicycle. \Citet{Hanakam2023}
+reported dissatisfaction in subjective rider feeling on their similar bicycle
+to ours and this high frequency effect to the human-controlled plant dynamics
+could be an explanation to their findings.
 
@@
 the participant's perturbation resistance threshold and that large
-perturbations will make you fall regardless of the balance assist's help.
+perturbations will make you fall regardless of the balance assist system's
+help.
 
@@
 more comprehensive and detailed natural data of how people fall we could make
-estimates on the number of falls reduced if everyone road a balance assist
+estimates on the number of falls reduced if everyone rode a balance assist
 bicycle.
@@
 due to the smaller lateral deviations. We believe our results are very much
-dependent on the two modes of falling, i.e. exit the treadmill width or foot is
-placed on the belt. On the other hand, cycle paths are a similar width as the
-treadmill, so rider's are often limited in width when recovering from a fall
-thus exiting the treadmill width may be an appropriate measure for indicating a
-fall. While the treadmill simulates narrow cycle paths, real-world paths may
-offer more lateral recovery space, providing riders with additional
+dependent on the two modes of falling, i.e. exiting the treadmill width or
+placing a foot on the belt. On the other hand, cycle paths are a similar width
+as the treadmill, so rider's are often limited in width when recovering from a
+fall thus exiting the treadmill width may be an appropriate measure for
+indicating a fall. While the treadmill simulates narrow cycle paths, real-world
+paths may offer more lateral recovery space, providing riders with additional
 opportunities to regain balance after a perturbation. However, testing in such
@@
 
-\subsection{On Rider Bicycling Skill and Experience}
+\subsection{Rider Skill and Experience}
 %
@@
 learning effect that occurs during the experiment is not strongly dependent on
-the speed.
+the speed but that similar experiments should consider the learning effect.
 
@@
 some additive or multiplicative effect on the resulting motion trajectory and
-make it harder to recover from a fall.
+make it harder to recover from a fall, but our results did not confirm this
+reasoning and we do not yet understand why.
 
@@
 to the handlebar can cause a rider to fall and every rider has a perturbation
-magnitude threshold at which they are more likely to fall than not. The
+resistance threshold at which they are more likely to fall than not. The
 probability of falling when mechanically perturbed is significantly
 (\(p=0.02\)) reduced when traveling at 6~\si{\kph} when the balance assisting
-control is activated. This effect is present when traveling at 10~\si{\kph}
-and is very close to significance (\(p=0.07\)) but more investigation is needed
-to determine if the effect can be significant. The positive effect to balance
-is rider independent and most effective in the regime of perturbations near the
+control is activated. This effect is present when traveling at 10~\si{\kph} and
+is very close to significance (\(p=0.07\)) but more investigation is needed to
+determine if the effect can be significant. The positive effect to balance is
+rider independent and most effective in the regime of perturbations near the
 limits of the rider's control authority. Given that similar effects cause falls
@@
 We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable
-feedback in early revisions.
+feedback in early revisions. Lastly, we thank Frans van der Helm for the
+consistent encouragement to execute this experiment and his early vision for
+obtaining the funding.
 

### 2024-10-09 main.tex
Mint Version 2.
Bump to version 3.dev.
Add acknowledgement to Christoph.

@@
 \title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
-Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev}
+Low Speeds When Subjected to Handlebar Perturbations\\Version 3.dev}
 
@@
 helped run the experiments and the participants for their time an enthusiasm.
+Thanks to Christoph Schmidt for motivation to do yet another round of trials.
 We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable

### 2024-10-22 main.tex
an -> and
Added thanks to DEMO guys.

@@
 insight in the statistical condsiderations. We thank all of the assistants that
-helped run the experiments and the participants for their time an enthusiasm.
-Thanks to Christoph Schmidt for motivation to do yet another round of trials.
-We thank Peter Stahlecker, Benjamin Gonzalez, and Riender Happee for valuable
-feedback in early revisions. Lastly, we thank Frans van der Helm for the
-consistent encouragement to execute this experiment and his early vision for
-obtaining the funding.
+helped run the experiments and the participants for their time and enthusiasm.
+Thanks to Wouter Gregoor, Hans van der Does, and Cor Meijneke for reviving the
+bicycles at a critical moment. Thanks to Christoph Schmidt for motivation to
+do yet another round of trials. We thank Peter Stahlecker, Benjamin Gonzalez,
+and Riender Happee for valuable feedback in early revisions. Lastly, we thank
+Frans van der Helm for the consistent encouragement to execute this experiment
+and his early vision for obtaining the funding.
 

### 2025-02-26 main.tex
Change probability variable from p to P.
Version 3
Bump to 4.dev
Fixed possessive to plural.

@@
 \title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
-Low Speeds When Subjected to Handlebar Perturbations\\Version 3.dev}
+Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev}
 
@@
 participant \(i\) which follows a Bernoulli distribution given the probability
-\(p_{ij}\) that a fall occurs:
+\(P_{ij}\) that a fall occurs:
 %
 \begin{align}
-  f_{ij} | p_{ij} \sim \textrm{Bernoulli}(p_{ij}) \textrm{.}
+  f_{ij} | P_{ij} \sim \textrm{Bernoulli}(P_{ij}) \textrm{.}
 \end{align}
@@
 \begin{align}
-  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) =
+  \log \left(\frac{P_{ij}}{1-P_{ij}} \right) =
   \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
@@
   \begin{split}
-    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
+    \log \left( \frac{P_{ij}}{1-P_{ij}} \right)
     = & \quad
@@
 \begin{align}
-  \frac{p_{ij}}{1-p_{ij}} & = e^{1.59-0.64s} =
+  \frac{P_{ij}}{1-P_{ij}} & = e^{1.59-0.64s} =
   e^{1.59}\cdot e^{-0.64s} = 4.90 \cdot e^{-0.64s} \\
-  p_{ij}|_{s=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
-  p_{ij}|_{s=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
+  P_{ij}|_{s=0} & = \frac{4.90}{1 + 4.90} = 0.83 \\
+  P_{ij}|_{s=1} & = \frac{4.90\cdot0.53}{1 + 4.90\cdot{0.53}} = 0.72
 \end{align}
@@
 placing a foot on the belt. On the other hand, cycle paths are a similar width
-as the treadmill, so rider's are often limited in width when recovering from a
+as the treadmill, so riders are often limited in width when recovering from a
 fall thus exiting the treadmill width may be an appropriate measure for

### 2025-04-15 main.tex
Added citation to the preprint and fixed Wegman reference.

@@
 Single-actor bicycle crashes are associated with a surprisingly large
-percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
-start up to typical cruising speeds, bicycles are not self-stable and can be
-challenging for the rider to balance. Low speed crashes may be reduced if the
-bicycle was self-stable at these speeds by relieving the rider of some of their
-necessary control activity. Bicycles can be mechanically modified to lower the
-speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
-has been known that applying a motor actuated steering torque proportional to
-the vehicle's roll angular rate~\footnote{Roll angular rate is defined herein
-as per the SAE J670 vehicle standards.} can stabilize a single-track vehicle
-down to very low speeds~\citep{Ruijs1986}. If automatic control of steering can
-stabilize a bicycle, it may reduce the control required from the rider to
-successfully manage balancing tasks much like natural high speed
-self-stability already does. We have developed a balance-assisting bicycle,
-Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
+percentage of reported serious injuries~\citep{Wegman2024a}. At low speeds,
+from start up to typical cruising speeds, bicycles are not self-stable and can
+be challenging for the rider to balance. Low speed crashes may be reduced if
+the bicycle was self-stable at these speeds by relieving the rider of some of
+their necessary control activity. Bicycles can be mechanically modified to
+lower the speeds at which they are self-stable~\citep{Astrom2005} and since the
+1980s it has been known that applying a motor actuated steering torque
+proportional to the vehicle's roll angular rate~\footnote{Roll angular rate is
+defined herein as per the SAE J670 vehicle standards.} can stabilize a
+single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
+control of steering can stabilize a bicycle, it may reduce the control required
+from the rider to successfully manage balancing tasks much like natural high
+speed self-stability already does. We have developed a balance-assisting
+bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
 hypothesize that it helps the rider in situations in which they are likely to
@@
 
-In this paper, we investigate whether an automatically controlled bicycle, that
-is stable in a large low speed range for balance assist, is beneficial in
-helping to prevent the rider from falling. We test this possible benefit by
-applying varying magnitude mechanical perturbations to the handlebars while the
-rider is cycling on a treadmill. We assess the difference in the rider's
-probability of falling with the balance assist system on and off.
+In this paper~\footnote{Also available as a preprint \cite{Haitjema2025}.}, we
+investigate whether an automatically controlled bicycle, that is stable in a
+large low speed range for balance assist, is beneficial in helping to prevent
+the rider from falling. We test this possible benefit by applying varying
+magnitude mechanical perturbations to the handlebars while the rider is cycling
+on a treadmill. We assess the difference in the rider's probability of falling
+with the balance assist system on and off.
 

### 2025-04-15 main.tex
Fixed Wegman citation key.
Use verbatim for the github url because the url command is not available when biblatex is not used.

@@
 Single-actor bicycle crashes are associated with a surprisingly large
-percentage of reported serious injuries~\citep{Wegman2024a}. At low speeds,
+percentage of reported serious injuries~\citep{Wegman2024}. At low speeds,
 from start up to typical cruising speeds, bicycles are not self-stable and can
@@
 \noindent
-\url{https://github.com/mechmotum/fall-probability-paper}
+\verb|https://github.com/mechmotum/fall-probability-paper|
 

### 2025-10-21 src/control.py
Move gain calc to function, add experimental weave id results.

@@ -12,6 +12,7 @@ from model import SteerControlModel
 SCRIPT_PATH = os.path.realpath(__file__)
 SRC_DIR = os.path.dirname(SCRIPT_PATH)
 ROOT_DIR = os.path.realpath(os.path.join(SRC_DIR, '..'))
+DAT_DIR = os.path.join(ROOT_DIR, 'data')
 FIG_DIR = os.path.join(ROOT_DIR, 'figures')
 KPH2MPS = 1000.0/3600.0
 MPS2KPH = 1.0/KPH2MPS
@@ -34,13 +35,19 @@ fig.savefig(os.path.join(FIG_DIR, 'bicycle-with-geometry-mass.png'), dpi=300)
 
 speeds = np.linspace(0.0, 10.0, num=2001)
 
+
 # control law
-vmin, vmin_idx = 1.5, np.argmin(np.abs(speeds - 1.5))
-vmax, vmax_idx = 4.7, np.argmin(np.abs(speeds - 4.7))
-static_gain = 10.0
-kphidots = -static_gain*(vmax - speeds)
-kphidots[:vmin_idx] = -static_gain*(vmax - vmin)/vmin*speeds[:vmin_idx]
-kphidots[vmax_idx:] = 0.0
+def generate_gains(static_gain):
+    vmin, vmin_idx = 1.5, np.argmin(np.abs(speeds - 1.5))
+    vmax, vmax_idx = 4.7, np.argmin(np.abs(speeds - 4.7))
+    kphidots = -static_gain*(vmax - speeds)
+    kphidots[:vmin_idx] = -static_gain*(vmax - vmin)/vmin*speeds[:vmin_idx]
+    kphidots[vmax_idx:] = 0.0
+    return kphidots
+
+
+static_gain = 3.8
+kphidots = generate_gains(static_gain)
 
 # FIGURE : Plot the roll rate gains versus speed.
 fig, ax = plt.subplots(layout='constrained')
@@ -49,9 +56,8 @@ ax.plot(speeds, kphidots)
 ax.set_ylabel(r'$k_\dot{\phi}$')
 fig.savefig(os.path.join(FIG_DIR, 'gains-vs-speed.png'), dpi=300)
 
+
 # FIGURE : Compare eigenvalues vs speed for uncontrolled.
-
-
 def stable_ranges(evals):
     # TODO : This should return the indice after a rise and before a fall. It
     # now always returns the indices after a rise or fall.
[…]

### 2025-10-21 src/control.py
Create an eigenvalue plot for both bicycle teensy gain settings, which correspond to the two experiments at 6 and 10 km/h.

@@ -1,10 +1,10 @@
 import os
 
-from bicycleparameters.parameter_sets import Meijaard2007ParameterSet
-from bicycleparameters.bicycle import sort_eigenmodes
-import matplotlib.pyplot as plt
 import numpy as np
 from scipy.constants import golden_ratio
+import matplotlib.pyplot as plt
+from bicycleparameters.parameter_sets import Meijaard2007ParameterSet
+from bicycleparameters.bicycle import sort_eigenmodes
 
 from data import bike_with_rider, bike_without_rider
 from model import SteerControlModel
@@ -16,6 +16,9 @@ DAT_DIR = os.path.join(ROOT_DIR, 'data')
 FIG_DIR = os.path.join(ROOT_DIR, 'figures')
 KPH2MPS = 1000.0/3600.0
 MPS2KPH = 1.0/KPH2MPS
+# NOTE : The theorectical gains (values) are manually chosen for a eye-balled
+# best fit of the weave mode for the Teensy set gain (keys).
+GAIN_MAP = {8: 3.9, 10: 5.4}
 
 if not os.path.exists(FIG_DIR):
     os.mkdir(FIG_DIR)
@@ -38,6 +41,10 @@ speeds = np.linspace(0.0, 10.0, num=2001)
 
 # control law
 def generate_gains(static_gain):
+    """
+    static_gain : float
+        Gain value for use in the controller model.
+    """
     vmin, vmin_idx = 1.5, np.argmin(np.abs(speeds - 1.5))
     vmax, vmax_idx = 4.7, np.argmin(np.abs(speeds - 4.7))
     kphidots = -static_gain*(vmax - speeds)
@@ -46,15 +53,20 @@ def generate_gains(static_gain):
     return kphidots
 
 
-static_gain = 3.8
-kphidots = generate_gains(static_gain)
-
 # FIGURE : Plot the roll rate gains versus speed.
-fig, ax = plt.subplots(layout='constrained')
-fig.set_size_inches((80/25.4, 80/25.4/golden_ratio))
-ax.plot(speeds, kphidots)
-ax.set_ylabel(r'$k_\dot{\phi}$')
-fig.savefig(os.path.join(FIG_DIR, 'gains-vs-speed.png'), dpi=300)
+def plot_gains_vs_speed():
+    fig, ax = plt.subplots(layout='constrained')
+    fig.set_size_inches((80/25.4, 80/25.4/golden_ratio))
[…]

### 2025-10-21 main.tex
Added the mean age of the particpants (does not include 4 missing ages in the data files).
Now includes two eigenvalue figures.

@@
   \centering
-  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
+  \includegraphics[width=160mm]{figures/balance-assist-gain-8-eig-vs-speeds.png}
+  \caption{Uncontrolled (upper row) and controlled with \(\kappa=8\) (lower
+    row) root locus of the eigenvalue components (real: solid, imaginary:
+    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
+    without the inertial effect of a rigid rider (left) and with a rigid rider
+    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
+    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
+    region is the linear stable speed range. Geometric and inertial parameters
+    for these plots were estimated with the methods presented in
+    \citet{Moore2012} and software packages
+    BicycleParameters~1.1.1~\citep{Moore2024} and
+    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
+    Appendix~\ref{sec:physical-parameters}.}
+  \label{fig:eigenvalues}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=160mm]{figures/balance-assist-gain-10-eig-vs-speeds.png}
   \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower
@@
 %
-We recruited 26 able-bodied young adults (20-36 years old) to participate in
-the experiments. The participants were all confident in their cycling skills
-and had cycled at least once in the last month. All participants consented to
-the experiment and could decline to continue at any time. The study was
-approved by Delft University of Technology's Human Research Ethics Council
-(record \#3897).
+We recruited 26 able-bodied young adults between 20 and 36 years old (mean 26
+years) to participate in the experiments. The participants were all confident
+in their cycling skills and had cycled at least once in the last month. All
+participants consented to the experiment and could decline to continue at any
+time. The study was approved by Delft University of Technology's Human Research
+Ethics Council (record \#3897).
 

### 2025-10-21 src/control.py

@@ -18,7 +18,7 @@ KPH2MPS = 1000.0/3600.0
 MPS2KPH = 1.0/KPH2MPS
 # NOTE : The theorectical gains (values) are manually chosen for a eye-balled
 # best fit of the weave mode for the Teensy set gain (keys).
-GAIN_MAP = {8: 3.9, 10: 5.4}
+GAIN_MAP = {8: 3.9, 10: 5.2}
 
 if not os.path.exists(FIG_DIR):
     os.mkdir(FIG_DIR)

### 2025-10-30 main.tex
Combined the eigenvalue plots for both gains into one panel plot.

@@
   \centering
-  \includegraphics[width=160mm]{figures/balance-assist-gain-8-eig-vs-speeds.png}
+  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
   \caption{Uncontrolled (upper row) and controlled with \(\kappa=8\) (lower
-    row) root locus of the eigenvalue components (real: solid, imaginary:
-    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
-    without the inertial effect of a rigid rider (left) and with a rigid rider
-    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
-    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
-    region is the linear stable speed range. Geometric and inertial parameters
-    for these plots were estimated with the methods presented in
-    \citet{Moore2012} and software packages
-    BicycleParameters~1.1.1~\citep{Moore2024} and
-    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
-    Appendix~\ref{sec:physical-parameters}.}
-  \label{fig:eigenvalues}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=160mm]{figures/balance-assist-gain-10-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled with \(\kappa=10\) (lower
     row) root locus of the eigenvalue components (real: solid, imaginary:

### 2025-10-30 src/control.py

@@ -88,11 +88,11 @@ def plot_eig(ax, model, teensy_gain, kphidots=0.0):
     print(msg.format(weave_speed, weave_speed*MPS2KPH))
     msg = 'Capsize speed: {:1.2f} [m/s], {:1.1f} [km/h]'
     print(msg.format(capsize_speed, capsize_speed*MPS2KPH))
-    if teensy_gain == 10:
+    if teensy_gain == 10 or teensy_gain == 'both':
         ymin, ymax = -6.0, 12.0
         ax.axvline(6.0*KPH2MPS, ymin=ymin, ymax=ymax, color='black',
                    linestyle=':')
-    elif teensy_gain == 8:
+    if teensy_gain == 8 or teensy_gain == 'both':
         ymin, ymax = -5, 11.0
         ax.axvline(10.0*KPH2MPS, ymin=ymin, ymax=ymax, color='black',
                    linestyle='-.')
@@ -102,58 +102,59 @@ def plot_eig(ax, model, teensy_gain, kphidots=0.0):
     return ax
 
 
-def create_four_panel(teensy_gain):
-    """
-    teensy_gain : integer
-        8 or 10
-    """
-    msg = f"For Teensy Gain of {teensy_gain}"
-    print("="*len(msg))
-    print(msg)
-    print("="*len(msg))
+def create_four_panel():
 
-    data_fname = f'weave_eigenvalues_from_experiment_gain_{teensy_gain}.csv'
-    plot_fname = f'balance-assist-gain-{teensy_gain}-eig-vs-speeds.png'
+    data8_fname = 'weave_eigenvalues_from_experiment_gain_8.csv'
+    data10_fname = 'weave_eigenvalues_from_experiment_gain_10.csv'
+    plot_fname = 'balance-assist-eig-vs-speeds.png'
 
-    try:
-        static_gain = GAIN_MAP[teensy_gain]
-    except KeyError:
-        raise ValueError(f'Gains can only be {GAIN_MAP.keys()}')
-
-    kphidots = generate_gains(static_gain)
-
-    fig_four, axes = plt.subplots(2, 2, sharex=True, sharey=True,
+    fig_four, axes = plt.subplots(3, 2, sharex=True, sharey=True,
                                   layout='constrained')
     fig.set_size_inches((160/25.4, 160/25.4/golden_ratio))
 
     msg = 'Without rigid rider and balance assist off:'
     print(msg)
[…]

### 2025-10-30 src/control.py
Add a comma.

@@ -134,7 +134,7 @@ def create_four_panel():
             linestyle='')
     ax.plot(weave_eig[:, 0], weave_eig[:, 2], color='black', marker='*',
             linestyle='')
-    ax.set_ylabel(f'Assist On $\kappa={GAIN_MAP[8]}$\nEig. Comp. [1/s]')
+    ax.set_ylabel(f'Assist On, $\kappa={GAIN_MAP[8]}$\nEig. Comp. [1/s]')
     ax.set_xlabel('')
 
     msg = '\nWithout rigid rider and balance assist on:'
@@ -148,7 +148,7 @@ def create_four_panel():
             linestyle='')
     ax.plot(weave_eig[:, 0], weave_eig[:, 2], color='black', marker='*',
             linestyle='')
-    ax.set_ylabel(f'Assist On $\kappa={GAIN_MAP[10]}$\nEig. Comp. [1/s]')
+    ax.set_ylabel(f'Assist On, $\kappa={GAIN_MAP[10]}$\nEig. Comp. [1/s]')
     ax.set_xlabel('Speed [m/s]')
 
     msg = '\nWith rigid rider and balance assist off:'

### 2025-10-30 main.tex
Started updating text in seciond 2.2 for the new Fig 2.

@@
   \item A bicycle steering motor with automatic roll rate feedback control can
-    stabilize the bicycle-rider system at speeds as low as 4.6~\si{\kph}.
+    stabilize the bicycle-rider system at speeds as low as 8~\si{\kph}.
   \item Every bicycle rider has their own perturbation resistance threshold
@@
 sensors is logged with a CanEdge2 CAN bus (CSS Electronics, Aabyhøj, Denmark)
-at least 100 Hz.
+at at least 100 Hz.
 
@@
 average stable speed predicted from the open loop bicycle rigid-rider dynamics.
-We use the gain factor \(\kappa=8\) (low) and \(\kappa=10\) (high) to tune the gain
-magnitude during the experiments. Scaling the proportional feedback gain
-linearly with respect to speed stabilizes the normally unstable weave mode of
-the bicycle down to 2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
-4.6~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
-Figure~\ref{fig:eigenvalues}.
+We use the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune
+the gain magnitude during the experiments. For the 6~\si{\kilo\meter\per\hour}
+experiment, scaling the proportional feedback gain linearly with respect to
+speed stabilizes the normally unstable weave mode of the bicycle down to
+3.4~\si{\kilo\meter\per\hour} for the riderless bicycle and
+9.7~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
+Figure~\ref{fig:eigenvalues}. Geometric and inertial parameters for these plots
+were estimated with the methods presented in \citet{Moore2012} and software
+packages BicycleParameters~1.1.1~\citep{Moore2024} and
+Yeadon~1.5.0~\citep{Dembia2015} and are shown in
+Appendix~\ref{sec:physical-parameters}. The identified weave mode uses the
+methods of Kooijman et. al~\citep{Kooijman2009a} and gives an indication of the
+performance of the system.
 %
@@
   \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
-  \caption{Uncontrolled (upper row) and controlled with \(\kappa=8\) (lower
-    row) root locus of the eigenvalue components (real: solid, imaginary:
-    dashed) plotted versus speed for the linear Carvallo-Whipple bicycle model
-    without the inertial effect of a rigid rider (left) and with a rigid rider
-    (right). Vertical dotted and dotted-dashed lines indicate the two speeds we
-    perform experiments at: 6~\si{\kph} and 10~\si{\kph}. The grey shaded
-    region is the linear stable speed range. Geometric and inertial parameters
-    for these plots were estimated with the methods presented in
-    \citet{Moore2012} and software packages
-    BicycleParameters~1.1.1~\citep{Moore2024} and
-    Yeadon~1.5.0~\citep{Dembia2015} and are shown in
-    Appendix~\ref{sec:physical-parameters}.}
+  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
+    row), and controleld with \(\kappa=5.2\) (lower row) root locus of the
+    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
+    for the linear Carvallo-Whipple bicycle model without the inertial effect
+    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
+    dotted-dashed lines indicate the two speeds we perform experiments at:
+    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
+    speed range. Black stars are experimentally identified weave mode decay and
+  frequency.}
   \label{fig:eigenvalues}

### 2025-11-12 main.tex
Adjusted text to match the new kappa values and the new Figure 2.

@@
   automatically controlled steering motor to a consumer electric bicycle that
-  stabilizes the riderless bicycle down to about 3~\si{\kph} to assist a rider
-  in balancing the vehicle. We hypothesize that a such a stabilized bicycle
-  will reduce the probability of falling. To test the system's possible
+  stabilizes the riderless bicycle down to just below 4~\si{\kph} to assist a
+  rider in balancing the vehicle. We hypothesize that a such a stabilized
+  bicycle will reduce the probability of falling. To test the system's possible
   assistance during falls, we applied varying magnitude external handlebar
@@
 Single-actor bicycle crashes are associated with a surprisingly large
-percentage of reported serious injuries~\citep{Wegman2024}. At low speeds,
-from start up to typical cruising speeds, bicycles are not self-stable and can
-be challenging for the rider to balance. Low speed crashes may be reduced if
-the bicycle was self-stable at these speeds by relieving the rider of some of
-their necessary control activity. Bicycles can be mechanically modified to
-lower the speeds at which they are self-stable~\citep{Astrom2005} and since the
-1980s it has been known that applying a motor actuated steering torque
-proportional to the vehicle's roll angular rate~\footnote{Roll angular rate is
-defined herein as per the SAE J670 vehicle standards.} can stabilize a
-single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
-control of steering can stabilize a bicycle, it may reduce the control required
-from the rider to successfully manage balancing tasks much like natural high
-speed self-stability already does. We have developed a balance-assisting
-bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
+percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
+start up to typical cruising speeds, bicycles are not self-stable and can be
+challenging for the rider to balance. Low speed crashes may be reduced if the
+bicycle was self-stable at these speeds by relieving the rider of some of their
+necessary control activity. Bicycles can be mechanically modified to lower the
+speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
+has been known that applying a motor actuated steering torque proportional to
+the vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the
+SAE J670 vehicle standards.} can stabilize a single-track vehicle down to very
+low speeds~\citep{Ruijs1986}. If automatic control of steering can stabilize a
+bicycle, it may reduce the control required from the rider to successfully
+manage balancing tasks much like natural high speed self-stability already
+does. We have developed a balance-assisting bicycle,
+Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
 hypothesize that it helps the rider in situations in which they are likely to
@@
 but \citet{Hanakam2023} recently showed rider dissatisfaction with the
-stabilization of a similar vehicle so the overall possible benefits are not
-yet definitively established.
+stabilization of a similar vehicle so the overall possible benefits are not yet
+definitively established.
 
@@
 \citet{Schwab2008} elaborate on some of the possibilities in scheduling the
-gains for such a controller for a bicycle and shows that a linear scheduling
-with respect to speed can give satisfactory stability for a low speed range. We
-use this simple feedback principle as the basis for our balance assist
-controller.
+gains for such a controller for a bicycle and show that linear scheduling with
+respect to speed can give satisfactory stability for a low speed range. We use
+this simple feedback principle as the basis for our balance assist controller.
 
@@
 where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
-average stable speed predicted from the open loop bicycle rigid-rider dynamics.
+average stable speed predicted from the open-loop bicycle rigid-rider dynamics.
 We use the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune
-the gain magnitude during the experiments. For the 6~\si{\kilo\meter\per\hour}
-experiment, scaling the proportional feedback gain linearly with respect to
-speed stabilizes the normally unstable weave mode of the bicycle down to
-3.4~\si{\kilo\meter\per\hour} for the riderless bicycle and
-9.7~\si{\kilo\meter\per\hour} for the ridden bicycle as shown in
-Figure~\ref{fig:eigenvalues}. Geometric and inertial parameters for these plots
-were estimated with the methods presented in \citet{Moore2012} and software
-packages BicycleParameters~1.1.1~\citep{Moore2024} and
-Yeadon~1.5.0~\citep{Dembia2015} and are shown in
-Appendix~\ref{sec:physical-parameters}. The identified weave mode uses the
-methods of Kooijman et. al~\citep{Kooijman2009a} and gives an indication of the
-performance of the system.
+the gain magnitude during the experiments. Figure~\ref{fig:eigenvalues} shows
+the stabilization effect of the controller via the eigenmodes of the linear
+system. Geometric and inertial parameters for these plots were estimated with
+the methods presented in \citet{Moore2012} and software packages
+BicycleParameters~1.1.1~\citep{Moore2024} and Yeadon~1.5.0~\citep{Dembia2015}
+and are shown in Appendix~\ref{sec:physical-parameters}.
+
+For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
+with respect to speed stabilizes the normally unstable weave mode of the
+bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
+the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
+unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
+bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
+Figure~\ref{fig:eigenvalues}.  Notably, the bicycle is not stable at
+6~\si{\kph} and just stable at 10~\si{\kph}. Figure~\ref{fig:eigenvalues} also
+shows the identified weave mode captured using the methods of
+Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
+performance of the system, with the model predicting the weave mode well the
+region of the tested speeds.
 %
@@
   \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
-    row), and controleld with \(\kappa=5.2\) (lower row) root locus of the
+    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
     eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
@@
 time. The study was approved by Delft University of Technology's Human Research
-Ethics Council (record \#3897).
-
-The participants were divided into two groups. The first group of fifteen
-participants performed the protocol at a belt speed of
-10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=8\) and two
-weeks later the second group of eleven participants performed the protocol at a
-belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
-\(\kappa=10\).
+Ethics Council (record \#3897). The participants were divided into two groups.
+The first group of fifteen participants performed the protocol at a belt speed
+of 10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=3.9\) and
+two weeks later the second group of eleven participants performed the protocol
+at a belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
+\(\kappa=5.2\).
 
@@
 setting on the balance assist controller indicates if the assistance is off
-\(\kappa=0\) or on at two different levels: low \(\kappa=8\) or high
-\(\kappa=10\). As mentioned earlier, a recovery from the perturbation is
+\(\kappa=0\) or on at two different levels: low \(\kappa=3.9\) or high
+\(\kappa=5.2\). As mentioned earlier, a recovery from the perturbation is
 successful if the person neither places their foot down onto the treadmill
@@
 assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(\kappa=10\) trials and one
-for the 10~\si{\kph}, \(\kappa=8\) trials and we evaluate our hypothesis for each
+separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and one
+for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis for each
 set of data. Statistics were computed with the lme4 version
@@
 The coefficient estimates for a single-level logistic regression for the data
-from the 6~\si{\kph}, gain factor \(\kappa=10\) trials are shown in
+from the 6~\si{\kph}, gain factor \(\kappa=5.2\) trials are shown in
 Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
@@
   \caption{Logistic regression intercept \(\beta\) and coefficient estimates
-    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=10\) along with the
+    \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
     standard error SE, p-value \(p\), multiplicative change in odds
@@
 The coefficient estimates for a single-level logistic regression at
-10~\si{\kph} with gain factor \(\kappa=8\) are shown in Table
+10~\si{\kph} with gain factor \(\kappa=3.9\) are shown in Table
 \ref{tab:freq-coefs-10}. The angular impulse and perturbation order are
@@
   \caption{Logistic regression intercept \(\beta\) and coefficient estimates
-    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=8\) along with the
+    \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
     standard error SE, p-value \(p\), multiplicative change in odds

### 2025-11-12 src/control.py

@@ -126,6 +126,7 @@ def create_four_panel():
     msg = '\nWithout rigid rider and balance assist on:'
     print(msg)
     print("-"*len(msg))
+    print('Model gain: {}'.format(GAIN_MAP[8]))
     kphidots = generate_gains(GAIN_MAP[8])
     ax = plot_eig(axes[1, 0], model_without, 8, kphidots=kphidots)
     weave_eig = np.loadtxt(os.path.join(DAT_DIR, data8_fname), delimiter=',',
@@ -140,6 +141,7 @@ def create_four_panel():
     msg = '\nWithout rigid rider and balance assist on:'
     print(msg)
     print("-"*len(msg))
+    print('Model gain: {}'.format(GAIN_MAP[10]))
     kphidots = generate_gains(GAIN_MAP[10])
     ax = plot_eig(axes[2, 0], model_without, 10, kphidots=kphidots)
     weave_eig = np.loadtxt(os.path.join(DAT_DIR, data10_fname), delimiter=',',
@@ -165,6 +167,7 @@ def create_four_panel():
     msg = '\nWith rigid rider and balance assist on:'
     print(msg)
     print("-"*len(msg))
+    print('Model gain: {}'.format(GAIN_MAP[8]))
     kphidots = generate_gains(GAIN_MAP[8])
     ax = plot_eig(axes[1, 1], model_with, 8, kphidots=kphidots)
     ax.set_ylabel('')
@@ -173,6 +176,7 @@ def create_four_panel():
     msg = '\nWith rigid rider and balance assist on:'
     print(msg)
     print("-"*len(msg))
+    print('Model gain: {}'.format(GAIN_MAP[10]))
     kphidots = generate_gains(GAIN_MAP[10])
     ax = plot_eig(axes[2, 1], model_with, 10, kphidots=kphidots)
     ax.set_ylabel('')

### 2025-11-12 main.tex
Added info about the fall modes and a new citation.

@@
 safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
-natural free movement pre-fall. The participants practiced riding on the
-treadmill until they indicated they were comfortable enough to have
-perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
-We then asked the rider to ride for 90~\si{\second}, while attempting to
-maintain the location of their front wheel on the center line of the treadmill
-as a baseline measure before the perturbations. We then applied perturbations
-in random directions (clockwise or counter-clockwise), starting at
-20~\si{\newton} and increasing the magnitude by 30~\si{\newton} until the
-participants fell. We defined a ``fall'' on the treadmill by two criteria: 1)
-the rider removes their foot from the pedal and places it on the ground or 2)
-the bicycle wheel exceeds the width of the treadmill belt.
+natural free movement pre-fall. The participants practiced falling in the
+harness at least once and then practiced riding on the treadmill until they
+indicated they were comfortable enough to have perturbations applied. For most,
+this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
+for 90~\si{\second}, while attempting to maintain the location of their front
+wheel on the center line of the treadmill as a baseline measure before the
+perturbations. We then applied perturbations in random directions (clockwise or
+counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
+30~\si{\newton} until the participants fell. We defined a ``fall'' on the
+treadmill by two criteria: 1) the rider removes their foot from the pedal and
+places it on the ground or 2) the bicycle wheel exceeds the width of the
+treadmill belt. These were the two actions which necessitated stopping the
+treadmill belt. We informed the participants beforehand that we would stop the
+treadmill in these situations.  Placing a foot down on the fall side of the
+bicycle is a natural reaction when steering or other body movements are not
+sufficient to keep the bicycle upright, as shown in \citep{Gildea2024a}.
+Exiting the width of the treadmill is akin to exiting the width of a curb-less
+bicycle path and arises when the rider's control actions are not sufficient to
+correct the heading of the vehicle in sufficient time.
 Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a

### 2025-11-12 main.tex
Expanded the note on the unstable model at 6 km/h.
Addressed the review comment about how much the cyclist experienced the range of forces and why we didn't try to eliminate the learning effects.

@@
 bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
-Figure~\ref{fig:eigenvalues}.  Notably, the bicycle is not stable at
-6~\si{\kph} and just stable at 10~\si{\kph}. Figure~\ref{fig:eigenvalues} also
-shows the identified weave mode captured using the methods of
+Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
+but the time constant of the unstable eigenvalue is increased by a factor of 8
+and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
+also shows the identified weave mode captured using the methods of
 Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
@@
 
-Following the initial threshold determination, we choose perturbation forces
+Following the initial threshold determination where the participant experienced
+a full range of pertubation magnitudes, we then choose perturbation forces
 according to a random and adaptive staircase procedure applying perturbations
 above and below the initial perturbation threshold, while allowing small
-progression of the perturbation threshold to accommodate learning effects. Five
-possible perturbation forces are determined based on the initial perturbation
-threshold estimate: the initial estimate itself, two perturbations lower than
-the initial estimate, and two perturbations higher than the initial estimate.
-The five possible perturbations are separated by 10~\si{\newton} steps. For
-example, if the initial estimate of the perturbation threshold of a participant
-is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
-100~\si{\newton}. Which one of these five perturbations is chosen first is
-determined at random. If the perturbation results in a fall, the estimate of
+progression of the perturbation threshold to accommodate learning effects. We
+did not try to elimate learning effects due to the time per participant that
+would be required, but factored this into the later analysis as an alternative.
+Five possible perturbation forces are determined based on the initial
+perturbation threshold estimate: the initial estimate itself, two perturbations
+lower than the initial estimate, and two perturbations higher than the initial
+estimate.  The five possible perturbations are separated by 10~\si{\newton}
+steps. For example, if the initial estimate of the perturbation threshold of a
+participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
+90 and 100~\si{\newton}. Which one of these five perturbations is chosen first
+is determined at random. If the perturbation results in a fall, the estimate of
 the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if

### 2025-11-12 main.tex
Addressed the comment about direction of perturbation mattering.

@@
 instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
-characterize the configuration of the bicycle when perturbed. The gain factor
+characterize the configuration of the bicycle when perturbed. The sense of
+these variables relative to the sense of the perturbation matters. For example,
+if there is already a large steer angle in the same direction as the
+perturbation, this will have a different effect on the probability that a fall
+will occur than if the steer angle is in the opposite direction of the
+perturbation. To account for this, the roll and steer variables at the instance
+of perturbation are normalized for perturbation direction. The gain factor
 setting on the balance assist controller indicates if the assistance is off

### 2025-11-12 main.tex
Added note about four missing ages.

@@
 We recruited 26 able-bodied young adults between 20 and 36 years old (mean 26
-years) to participate in the experiments. The participants were all confident
-in their cycling skills and had cycled at least once in the last month. All
-participants consented to the experiment and could decline to continue at any
-time. The study was approved by Delft University of Technology's Human Research
-Ethics Council (record \#3897). The participants were divided into two groups.
-The first group of fifteen participants performed the protocol at a belt speed
-of 10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=3.9\) and
-two weeks later the second group of eleven participants performed the protocol
-at a belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
-\(\kappa=5.2\).
+years~\footnote{Four participants' ages were lost but did fall within the
+stated range.}) to participate in the experiments. The participants were all
+confident in their cycling skills and had cycled at least once in the last
+month. All participants consented to the experiment and could decline to
+continue at any time. The study was approved by Delft University of
+Technology's Human Research Ethics Council (record \#3897). The participants
+were divided into two groups.  The first group of fifteen participants
+performed the protocol at a belt speed of 10~\si{\kph}(2.8~\si{\mps}) with the
+gain factor set to \(\kappa=3.9\) and two weeks later the second group of
+eleven participants performed the protocol at a belt speed of 6~\si{\kph}
+(1.7~\si{\mps}) with the gain factor set to \(\kappa=5.2\).
 

### 2025-11-13 main.tex
Started with imrpovement of introduction on types of falls and how we recreate them.

@@
 the rider to perturbations at the handlebar, which can easily cause a rider to
-fall.
+fall. From a mechanics perspective, it is not particularly important how or
+where the perturbation is applied but that it results in forcing the
+bicycle-rider system into an unrecoverable state, as all forces acting on a
+dynamical system can be resolved into an equivalent force and torque of a
+couple at any location in the system~\citep{Kane1985}. We chose to perturb the
+handlebar because of this principle and it was practical and required a low
+force to cause a fall.
 

### 2025-11-13 main.tex
Addressed the review comment about background and connection of literature to our fall perturbations.

@@
 Single-actor bicycle crashes are associated with a surprisingly large
-percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
-start up to typical cruising speeds, bicycles are not self-stable and can be
-challenging for the rider to balance. Low speed crashes may be reduced if the
-bicycle was self-stable at these speeds by relieving the rider of some of their
-necessary control activity. Bicycles can be mechanically modified to lower the
-speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
-has been known that applying a motor actuated steering torque proportional to
-the vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the
-SAE J670 vehicle standards.} can stabilize a single-track vehicle down to very
-low speeds~\citep{Ruijs1986}. If automatic control of steering can stabilize a
-bicycle, it may reduce the control required from the rider to successfully
-manage balancing tasks much like natural high speed self-stability already
-does. We have developed a balance-assisting bicycle,
-Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
+percentage (\textgreater80\%) of reported serious injuries in The Netherlands
+and values range betwen 50\% and 95\% in other surveyed
+countries~\citep{Wegman2024}. \citet{Utriainen2023} report that slippery
+surfaces, simple loss of control, and engagement with a curb, pothole, or
+bollard are most the common fall types in the seven reported countries. Most of
+these types involve perturbations that result in loss of balance. At low
+speeds, from start up to typical cruising speeds, bicycles are not self-stable
+and can be challenging for the rider to balance. Low speed crashes may be
+reduced if the bicycle was self-stable at these speeds by relieving the rider
+of some of their necessary control activity. Bicycles can be mechanically
+modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
+and since the 1980s it has been known that applying a motor actuated steering
+torque proportional to the vehicle's roll angular rate~\footnote{``Roll'' is
+defined herein as per the SAE J670 vehicle standards.} can stabilize a
+single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
+control of steering can stabilize a bicycle, it may reduce the control required
+from the rider to successfully manage balancing tasks much like natural high
+speed self-stability already does. We have developed a balance-assisting
+bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
 hypothesize that it helps the rider in situations in which they are likely to
@@
 couple at any location in the system~\citep{Kane1985}. We chose to perturb the
-handlebar because of this principle and it was practical and required a low
-force to cause a fall.
+handlebar because of this superposition principle and that it was both
+practical and required a low force to cause a fall.
 

### 2025-11-13 main.tex
Addressed the second reviewer's comments on the two fall modes.

@@
 We recruited 26 able-bodied young adults between 20 and 36 years old (mean 26
-years~\footnote{Four participants' ages were lost but did fall within the
-stated range.}) to participate in the experiments. The participants were all
-confident in their cycling skills and had cycled at least once in the last
-month. All participants consented to the experiment and could decline to
-continue at any time. The study was approved by Delft University of
+years~\footnote{Excluding four participants' ages that were misplaced but that
+did fall within the stated range.}) to participate in the experiments. The
+participants were all confident in their cycling skills and had cycled at least
+once in the last month. All participants consented to the experiment and could
+decline to continue at any time. The study was approved by Delft University of
 Technology's Human Research Ethics Council (record \#3897). The participants
@@
 places it on the ground or 2) the bicycle wheel exceeds the width of the
-treadmill belt. These were the two actions which necessitated stopping the
-treadmill belt. We informed the participants beforehand that we would stop the
-treadmill in these situations. Placing a foot down on the fall side of the
-bicycle is a natural reaction when steering or other body movements are not
-sufficient to keep the bicycle upright, as shown in \citep{Gildea2024a}.
+treadmill belt.
+These two fall criteria are not strictly the same and the latter is more likely
+to happen at higher speeds.
+We informed the participants beforehand that we would stop the treadmill in
+these situations.
+These were the two actions which necessitated stopping the treadmill belt and
+the simulate real fall modes.
+Placing a foot down on the fall side of the bicycle is a natural reaction when
+steering or other body movements are not sufficient to keep the bicycle
+upright, as shown in \citep{Gildea2024a}.
 Exiting the width of the treadmill is akin to exiting the width of a curb-less

### 2025-11-13 main.tex
Added a note on the mixed effects model and the ICC being very low for reviewer #2.

@@
 steering or other body movements are not sufficient to keep the bicycle
-upright, as shown in \citep{Gildea2024a}.
+upright, as shown in \citep{Gildea2024}.
 Exiting the width of the treadmill is akin to exiting the width of a curb-less
@@
 that they have a mean of zero and a standard deviation of one by cluster-mean
-centering, as recommended by \citet{Enders2007}, with clusters being an
-individual participant. The clusters are chosen as all data from an individual
+centering with clusters being an individual participant, as recommended by
+\citet{Enders2007}. The clusters are chosen as all data from an individual
 participant because we are only interested in the association between the state
-of the balance assist system and the outcome of the perturbation. We expected
-there to be a variation between participants in how well they are able to
-resist the perturbation; however, this was not true. Cluster-mean centering
-showed there to be no variation between participants. This fact allows us to
-utilize a simple single-level logistic regression model, instead of a
+of the balance assist system and the outcome of the perturbation for that
+participant, not how participants perform relative to each other. We expected
+there to be variation between participants in how well they are able to resist
+the perturbation; however, this was not true. If a random effect is included in
+the model for each participant, the intraclass correlation (ICC) is less that
+3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
+single-level logistic regression model without the random effect instead of a
 multilevel model. This left us with angular impulse, perturbation order,
@@
 assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and one
-for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis for each
-set of data. Statistics were computed with the lme4 version
+separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
+one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
+for each set of data. Statistics were computed with the lme4 version
 1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

### 2025-11-14 review/review-response-01.tex
Started the review response letter.

\documentclass[en]{tuletter}
\usepackage{csquotes}
\signname{Jason K. Moore, PhD\\Assistant Professor, BioMechanical
Engineering\\Delft University of Technology}
\signimage{jkm-signature}

% Reference top left
\date{17 november 2025}
\ourref{NA}
\yourref{NA}
\contact{Jason K. Moore}
\phone{+31 010-844 6853}
\email{j.k.moore@tudelft.nl}
\subject{IATSS Review Response}

% Recipient left under reference
\toname{IATSS Editor}
\toaddress{}

% Sender right under logo
\fromname{Faculty of Mechanical Engineering\\BioMechanical Engineering}
\fromaddress{Mekelweg 2\\2628 CD Delft\\The Netherlands}
\frompobox{NA}

\begin{document}
\makeheader
\opening{To Whom It May Concern,}

Thank you for the careful review of our work. We have revised the paper to
address your comments both for clarification of various methods and details and
providing background and support for some of our choices. The revision is an
improved version based on your feedback.

\begin{displayquote}
  Editor:Thank you once more for your submission to IATSS Research. The
  reviewers have now completed their assessment of your manuscript and have
  recommended major revisions before the manuscript is being considered for
  publication. I recommend you read carefully the reviewers' suggestions and
  prepare a detailed response to be submitted along your revised manuscript. To
  facilitate the reviewer's work, I also recommend to highlight the lines in
  the new version of the manuscript where changes you introduce address the
  each reviewer's specific concerns.
\end{displayquote}

\begin{displayquote}
  Reviewer \#1: This paper describe attempt to reduce probability of falls by
  providing electro-mechanical assistance to bicycles, which generally become
  unstable at very low speeds. Using an electric bicycle with an assist motor
  attached to the steering axis as the test vehicle, we first conducted
  theoretical studies and confirmed that it can reach speeds of 2.9 km/h
  without a rider and 4.6 km/h with rigid rider. Furthermore, 26 young people
  participated in an experiment in which an external force was applied to the
  handlebars while the test vehicle was running on a treadmill, and it was
  confirmed that assistance reduced the fall probability.

  Indoor experiments under controlled conditions are extremely useful for
  studying vehicle steering stability. In particular, since it is difficult to
  conduct full-scale experiments on single-track vehicles, which lack static
  upright stability, the results of indoor experiments using a treadmill are
  extremely useful. By controlling the conditions, the analysis of variables
  has been precise, and interesting findings have been obtained.

  Please add some explanation on the following points:

  One of the main themes of this research is evaluating the effectiveness of
  automatic assist systems. For this evaluation, it is extremely important to
  understand the operating characteristics of the automatic assist system. The
  paper provides detailed descriptions of control models and theories (e.g.,
  root locus), but says very little about actual performance or operating
  characteristics. While the paper states that the maximum torque is 7 Nm as a
  specific performance feature, I believe that showing transient performance
  such as instantaneous maximum torque is also important in understanding
  response characteristics. Therefore, while theoretical explanations are
  important, there should also be data on things like the responsiveness of
  actual bicycles. While the paper provides a very detailed description of
  perturbations, it says very little about the specific performance of the
  assist system, which is the subject of the paper's evaluation. I would like
  the paper to provide more information on the assist system.
\end{displayquote}

The steer motor system was developed by Bosch eBike systems and is a
proprietary, thus we are not privy to the fine details of its operation. We
have not performed our own motor testing, as that is not generally our
expertise or concern. But we have now included extra vehicle performance tests
in the form of weave mode identification of the rider-less bicycle. Figure 2
has been reworked to show the transient behavior of the steer motor controlled
bicycle compared to the theoretical model's predictions. For the two testing
speeds we set the controller gain factor to different values. We now show the
gain factors needed in the ideal bicycle dynamics model to cause matching
motion of the actual system. Note that these gain factors 3.9 and 5.2 are
different than the prior paper version of 8 and 10, respecitively. The reason
for this is that the model does not include the motor dynamics (motor constant
or winding resistance), nor does it include the scrub torque effect of the tire
on the ground. The approximate factor of 2 difference in the gain factor
accounts for these unmodeled effects. The gain factor of 8 and 10 are the
values we employ in the microcontroller of the real system. We hope that this
proves 1) that our model can predict the steer controlled bicycle's motion and
2) gives an idea of the performance of the controlled vehicle. We have adapted
the text in section 2.2 to explain this.

\begin{displayquote}
  In this study, a fall is defined as both "putting one's foot on the ground"
  and "exceeds the width of treadmill", but there is insufficient explanation
  for this. In 4.2, the relationship between "exceeds the width of treadmill"
  and actual falls is discussed, but the explanation for "putting one's foot on
  the ground" is insufficient. "Putting one's foot on the ground" seem to occur
  with a much milder degree of instability than "exceeds the width of
  treadmill". In the actual experimental situation, "putting one's foot on the
  ground" may have been extremely dangerous, equivalent to "exceeds the width
  of treadmill", but this is not clear from the manuscript. I would like this
  point to be described in the Methods or Discussion section. Furthermore, the
  occurrence of such dangerous situations is strongly related to the
  instructions given to subjects beforehand. I would like the Methods section
  to provide additional explanation about the content of the instructions, such
  as whether any explanation about risk avoidance was given or not.
\end{displayquote}

\begin{displayquote}
  According to the explanation of the experimental procedures, I understand
  that in this experiment, each force was tested only once, and the subjects
  were not allowed to fully familiarize themselves with the force. I believe
  that the interpretation of the results would be completely different if the
  number of trials with each force were small or large enough. The fact that
  the number of trials was analyzed as a significant variable in the regression
  also reflects this assumption. I would like you to add an explanation about
  this point in the experimental procedures or discussion section.
\end{displayquote}

\begin{displayquote}
  The direction of the perturbation is said to be random from left to right,
  but I cannot decide whether it is appropriate to assume that it is truly
  random when analyzing the results. I believe that when the number of trials
  is small, the factors that affect the results are not only the magnitude of
  the force, but also the direction, which has a significant impact.
  Consideration and explanation is needed as to whether the results when a
  perturbation is in the opposite direction to the previous perturbation can be
  considered the same as the results when the perturbation is in the same
  direction. Furthermore, I presume that the results will be affected if the
  bicycle's posture at the time the perturbation is applied is leaning in the
  same direction as the perturbation, or if it is leaning in the opposite
  direction. If such conditions are taken into account in the analysis, please
  explain.
\end{displayquote}

\begin{displayquote}
  The age range of the subjects is listed in 2.4 Protocol, but I would like the
  average age to be included as well.
\end{displayquote}

\begin{displayquote}
  Reviewer \#2: This paper investigates the effectiveness of an automatic
  balance-assist system designed to reduce falls at low speeds. The core
  hypothesis is that a bicycle stabilized by an automatically controlled
  steering motor will reduce the probability of a fall. To verify this
  hypothesis, a well-considered experiment was conducted, and the analysis of
  its results is clearly explained. However, there are some unclear points that
  require the authors' response. Please refer to the following comments and
  make the necessary revisions.

  Section 1: Introduction on page 3

  Regarding low-speed accidents, the authors state that "single-vehicle
  accidents are associated with a surprisingly high proportion of reported
  serious injuries," citing examples that can cause loss of control, such as
  gusts of wind, collisions between handlebars, bags hanging from handlebars,
  or simply contact with road surface irregularities. However, this qualitative
  description alone does not fully explain how the mechanical perturbations
  applied to the handlebars in the laboratory, specifically vertical forces,
  quantitatively reproduce these events. Particularly, bicycles are thought to
  weave more at low speeds, meaning the handlebars are shaking. Therefore,
  additional literature review or supplemental explanation is required to
  demonstrate that this experiment is an effective and representative method
  for studying real-world low-speed accident scenarios.
\end{displayquote}

\begin{displayquote}
  Section 2.4: Protocol on page 7

  Regarding the definition of a fall, additional explanation is needed for the
  authors' reasoning for treating the rider removing their foot from the pedal
  and placing it on the ground and the bicycle wheel exceeding the width of the
  treadmill belt as the same category in the analysis. Please explain whether
  the fall scenarios differ or are the same depending on bicycle speed or the
  method of applying disturbances. In other words, can scenarios where the
  rider places their foot on the ground and falls, and scenarios where the
  bicycle's wheel exceeds the treadmill width and falls, be considered the same
  fall process, though differing in degree?
\end{displayquote}

\begin{displayquote}
  Section 2.6: Statistics on page 11

  The authors state that "cluster-mean centering showed there to be no
  variation between participants," which allowed authors to use a simple
  single-level logistic regression model rather than a multilevel model. This
  assertion is a foundational premise for the authors' entire analysis.
  Therefore, the authors are required to present the statistical testing
  process for cluster mean centering.
\end{displayquote}

\begin{displayquote}
  Section 4.4: Rider Skill and Experience on page 14

  This relates to the above point. While the authors state that "Each
  individual participant had their own perturbation resistance threshold"
  regarding individual differences in the experiment, it is difficult for
  readers to understand how individual differences were accounted for in the
  regression analysis, beyond the evidence regarding resistance values in the
  experiment. It is necessary to present the basic statistics of the
  experimental results and provide a detailed explanation of how individual
  differences were considered in the analysis.
\end{displayquote}

\begin{displayquote}
  Figure 2 on page 6

  It is preferable to shorten the figure caption, move the legend into the
  figure itself, and move the detailed figure explanation into the main body
  text.
\end{displayquote}

\closing{Sincerely,}
\end{document}

### 2025-11-14 review/review-response-01.tex
Added start of response to modes of falling concern.

@@
 
+The two modes of falling are different. We have added more information in
+section 2.4 to explain these two modes. If the particpants were cycling on an
+infinite plane and we perturb them to fall, then we would strictly use body
+contact with the ground as the criteria to define a fall. We have cited
+research that indicates riders often attempt to remove their foot from the
+pedal and place it on the ground to lessen or prevent a complete fall to the
+ground, so using that as an indicator for defining a fall seems justified. But
+we have a limited width treadmill, so as speed increases a perturbation will
+cause greater lateral deviations. If the wheel of the bicycle exits the width
+we had to stop the treadmill because it is not possible to recover
+
 \begin{displayquote}

### 2025-11-15 review/review-response-01.tex
Review response letter now has text for each section but some TODOs.

@@
 \usepackage{csquotes}
-\signname{Jason K. Moore, PhD\\Assistant Professor, BioMechanical
-Engineering\\Delft University of Technology}
+\signname{
+  Jason K. Moore, PhD\\
+  Assistant Professor, BioMechanical Engineering\\
+  Delft University of Technology
+}
 \signimage{jkm-signature}
@@
 \makeheader
-\opening{To Whom It May Concern,}
+\opening{Dear Reviewers,}
 
@@
 
+We address each concern below and have included a copy of the updated
+manuscript with the changes highlighted for your convenience.
+
 \begin{displayquote}
-  Reviewer \#1: This paper describe attempt to reduce probability of falls by
-  providing electro-mechanical assistance to bicycles, which generally become
-  unstable at very low speeds. Using an electric bicycle with an assist motor
-  attached to the steering axis as the test vehicle, we first conducted
+  \textbf{Reviewer \#1}: This paper describe attempt to reduce probability of
+  falls by providing electro-mechanical assistance to bicycles, which generally
+  become unstable at very low speeds. Using an electric bicycle with an assist
+  motor attached to the steering axis as the test vehicle, we first conducted
   theoretical studies and confirmed that it can reach speeds of 2.9 km/h
@@
 cause greater lateral deviations. If the wheel of the bicycle exits the width
-we had to stop the treadmill because it is not possible to recover
+we had to stop the treadmill for safety reasons. Neither of the fall modes on
+the treadmill are "extremely dangerous" or "dangerous".
 
@@
 \end{displayquote}
+
+The participants experienced 40 random force magnitudes after they experienced
+all force magnitudes in the threshold determiniation stage. So some forces were
+experienced more than once. We would have liked to remove the learning effect
+by letting the participants train for a long duration, but this was not
+practical. We choose to include a measure of learning in the statistics as an
+alternative.
+
+TODO : Did I change anything in the paper about this?
 
@@
 
+We did consider this and normalized the time series measurements to the sign of
+the perturbation as well as included the configuration at the time of
+perturbation as independent variables. We added text to clarify these things in
+Section 2.5.
+
 \begin{displayquote}
@@
 
+We have included the average age, albiet that four participants' ages were
+misplaced in the two years since we did this study.
+
 \begin{displayquote}
-  Reviewer \#2: This paper investigates the effectiveness of an automatic
-  balance-assist system designed to reduce falls at low speeds. The core
-  hypothesis is that a bicycle stabilized by an automatically controlled
+  \textbf{Reviewer \#2}: This paper investigates the effectiveness of an
+  automatic balance-assist system designed to reduce falls at low speeds. The
+  core hypothesis is that a bicycle stabilized by an automatically controlled
   steering motor will reduce the probability of a fall. To verify this
@@
 
+We have included citations to papers that discuss the types and nature of
+bicycle falls and expanded the text in Section 1 to better connect our study to
+real life falls. We also explain the force superposition principle of dynamics
+to argue that it matters little how the perturbation is delivered as long as we
+force the system into an unrecoverable state.
+
 \begin{displayquote}
@@
 
+First, see the answer above to review \#1 as a partial response to your
+question. The two fall modes are not technically the same and do not simply
+differ in degree, but were necessary given the nature of the experimental
+design. We did not record the category of the fall mode, so it is not possible
+for us to investigate whether having a tenary factor instead of a binary one
+would reveal any more insight. We added text in Section 2.4 to make this clear.
+
 \begin{displayquote}
@@
 \end{displayquote}
+
+In Section 2.6, we have included the maximum intraclass correlation (ICC) of
+the models that include the participant as a random effect, which is 3\%.
+Including this random effect does lower the p-values by a small amount, since a
+bit more variation is accounted for, but we decided to use the more
+consertative approach and not include this random effect. The participants do
+not have appreciable variation in how they respond to the perturbations after
+the data is normalized per participant.
+
+TODO : Cite that this is the similar to other data in Marten's thesis.
 
@@
 
+We report the ICC of the multilevel model, as noted in the prior answer, to
+show that the variation in participants was not an important factor. We also
+were not interested in the differences among participants, we are only
+interested in whether the system reduces the probability of anyone to fall.
+The cluster mean centering removes the relative differences in the particpants
+perturbation resistance threshold.
+
 \begin{displayquote}
@@
 
+We have shortened the figure caption but prefer that the figure is entirely
+understandable from itself and the caption. We hope this difference in style is
+not a major concern.
+
 \closing{Sincerely,}

### 2025-11-15 main.tex
Spellcheck.

@@
 percentage (\textgreater80\%) of reported serious injuries in The Netherlands
-and values range betwen 50\% and 95\% in other surveyed
+and values range between 50\% and 95\% in other surveyed
 countries~\citep{Wegman2024}. \citet{Utriainen2023} report that slippery
@@
 places it on the ground or 2) the bicycle wheel exceeds the width of the
-treadmill belt.
-These two fall criteria are not strictly the same and the latter is more likely
-to happen at higher speeds.
-We informed the participants beforehand that we would stop the treadmill in
-these situations.
-These were the two actions which necessitated stopping the treadmill belt and
-the simulate real fall modes.
-Placing a foot down on the fall side of the bicycle is a natural reaction when
-steering or other body movements are not sufficient to keep the bicycle
-upright, as shown in \citep{Gildea2024}.
-Exiting the width of the treadmill is akin to exiting the width of a curb-less
-bicycle path and arises when the rider's control actions are not sufficient to
-correct the heading of the vehicle in sufficient time.
+treadmill belt. These two fall criteria are not strictly the same and the
+latter is more likely to happen at higher speeds. We informed the participants
+beforehand that we would stop the treadmill in these situations. These were
+the two actions which necessitated stopping the treadmill belt and the simulate
+real fall modes. Placing a foot down on the fall side of the bicycle is a
+natural reaction when steering or other body movements are not sufficient to
+keep the bicycle upright, as shown in \citep{Gildea2024}. Exiting the width of
+the treadmill is akin to exiting the width of a curb-less bicycle path and
+arises when the rider's control actions are not sufficient to correct the
+heading of the vehicle in sufficient time.
 Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
@@
 Following the initial threshold determination where the participant experienced
-a full range of pertubation magnitudes, we then choose perturbation forces
+a full range of perturbation magnitudes, we then choose perturbation forces
 according to a random and adaptive staircase procedure applying perturbations
@@
 progression of the perturbation threshold to accommodate learning effects. We
-did not try to elimate learning effects due to the time per participant that
+did not try to eliminate learning effects due to the time per participant that
 would be required, but factored this into the later analysis as an alternative.
@@
 single-level logistic regression model without the random effect instead of a
-multilevel model. This left us with angular impulse, perturbation order,
-balance assist state, and roll \& steer angles at the time of perturbation as
-independent variables. We also include interaction effects between the balance
-assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
-one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
-for each set of data. Statistics were computed with the lme4 version
-1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.
+multilevel model. This tracks with the independent data collected in
+\citet{Haitjema2023a} which also had negligible participant variation. This
+left us with angular impulse, perturbation order, balance assist state, and
+roll \& steer angles at the time of perturbation as independent variables. We
+also include interaction effects between the balance assist state and the other
+four variables. We divide the analysis into two separate model evaluations, one
+for the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
+\(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
+Statistics were computed with R version 4.1.3~\citep{RCoreTeam2021} the lme4
+version 1.1~\citep{Bates2015} package.
 
@@
 
-Turning the balance assist system on significantly \((p=0.02<0.05)\) reduces the
-odds that a perturbation results in a fall while cycling at a speed of
+Turning the balance assist system on significantly \((p=0.02<0.05)\) reduces
+the odds that a perturbation results in a fall while cycling at a speed of
 6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of

### 2025-11-15 src/control.py
Add legend to eigenvalue plot.

@@ -79,7 +79,7 @@ def stable_ranges(evals):
     return start_stop_idxs.reshape(-1, 2)
 
 
-def plot_eig(ax, model, teensy_gain, kphidots=0.0):
+def plot_eig(ax, model, teensy_gain, kphidots=0.0, legend=False):
     evals_, evecs_ = sort_eigenmodes(*model.calc_eigen(v=speeds,
                                                        kphidot=kphidots))
     weave_idx, capsize_idx = stable_ranges(evals_)[0]
@@ -96,21 +96,38 @@ def plot_eig(ax, model, teensy_gain, kphidots=0.0):
         ymin, ymax = -5, 11.0
         ax.axvline(10.0*KPH2MPS, ymin=ymin, ymax=ymax, color='black',
                    linestyle='-.')
+    if legend:
+        # NOTE : Fake line to make legend work.
+        ax.axvline(40.0*KPH2MPS, ymin=ymin, ymax=ymax, color='black',
+                   linestyle=':')
     model.plot_eigenvalue_parts(ax=ax, v=speeds, kphidot=kphidots,
                                 hide_zeros=True, colors=['k']*4)
     ax.set_ylim((ymin, ymax))
+    if legend:
+        # NOTE : Fake points to make legend work.
+        ax.plot([40.0, 41.0], [40.0, 41.], color='black', marker='*',
+                linestyle='')
+        ax.legend(['10 km/h', '6 km/h', 'Stable', 'Imaginary', '_none',
+                   '_none', '_none', 'Real', '_none', '_none', '_none',
+                   'Identified'],
+                  fontsize=8,
+                  bbox_to_anchor=(1.6, 0.25),
+                  bbox_transform=fig.transFigure,
+                  loc='upper center',
+                  ncol=6, labelspacing=0.0)
     return ax
 
 
-def create_four_panel():
+def create_six_panel():
 
     data8_fname = 'weave_eigenvalues_from_experiment_gain_8.csv'
     data10_fname = 'weave_eigenvalues_from_experiment_gain_10.csv'
     plot_fname = 'balance-assist-eig-vs-speeds.png'
 
-    fig_four, axes = plt.subplots(3, 2, sharex=True, sharey=True,
-                                  layout='constrained')
[…]

### 2025-11-15 src/generate_time_series_imgs.py
Make fontsize of y axis labels 8.

@@ -138,7 +138,7 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 color='black',
                 linestyle='--',
             )
-            axs[0].set_ylabel("Bump'Em\nHandlebar\nTorque\n[Nm]")
+            axs[0].set_ylabel("Bump'Em\nHandlebar\nTorque\n[Nm]", fontsize=8)
             axs[0].legend(fontsize="x-small")
 
             df['roll_angle_times_10'] = df['roll_angle']*10.0
@@ -146,10 +146,10 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 x="seconds_since_start",
                 y="roll_angle_times_10",
                 ax=axs[1],
-                ylabel="Angle\n[deg]",
                 label="Roll X 10",
                 color='black',
             )
+            axs[1].set_ylabel("Angle\n[deg]", fontsize=8)
             axs[1].plot(0.0, df['roll_angle_times_10'].iloc[0],
                         'o',
                         fillstyle='none',
@@ -174,10 +174,10 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 x="seconds_since_start",
                 y="roll_rate",
                 ax=axs[2],
-                ylabel="Roll Rate\n[deg/s]",
                 legend=False,
                 color='black',
             )
+            axs[2].set_ylabel("Roll Rate\n[deg/s]", fontsize=8)
             if "motor_current" in df.columns.values:
                 #com_torque = -30.35*np.deg2rad(df['roll_rate'])
                 #axs[3].plot(df['seconds_since_start'], com_torque,
@@ -189,11 +189,11 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 df.plot(
                     x="seconds_since_start", y="motor_torque",
                     ax=axs[3],
-                    ylabel="Commanded\nMotor Torque\n[Nm]",
                     label='__none__',
                     legend=False,
                     color='black',
                 )
+                axs[3].set_ylabel("Commanded\nMotor Torque\n[Nm]", fontsize=8)
                 axs[3].legend(fontsize="x-small")
             # 1.7 m/s is 6 km/h
             #axs[4].axhline(1.7, color='black', linestyle='--')
@@ -211,16 +211,16 @@ def generate_torque_angle_plots(perturbations_dfs, directory):
                 y="speed",
                 ax=axs[4],
                 xlabel="Time since start of perturbation [s]",
-                ylabel="Speed\n[m/s]",
                 label='__none__',
[…]

### 2025-11-15 review/review-response-01.tex
Shorted response about figure legend.

@@
 not have appreciable variation in how they respond to the perturbations after
-the data is normalized per participant.
-
-TODO : Cite that this is the similar to other data in Marten's thesis.
+the data is normalized per participant. This lack of variation was also
+observed in the independent dataset collected in Marten Haitjema's MSc thesis.
 
@@
 
-We have shortened the figure caption but prefer that the figure is entirely
-understandable from itself and the caption. We hope this difference in style is
-not a major concern.
+We have shortened the figure caption and added a legend.
 

### 2025-11-15 main.tex
Simplified units.

@@
 %
-where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
-average stable speed predicted from the open-loop bicycle rigid-rider dynamics.
-We use the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune
-the gain magnitude during the experiments. Figure~\ref{fig:eigenvalues} shows
-the stabilization effect of the controller via the eigenmodes of the linear
-system. Geometric and inertial parameters for these plots were estimated with
-the methods presented in \citet{Moore2012} and software packages
+where \(v_{\textrm{stable}} = 4.7~\si{\mps}\) is approximately the average
+stable speed predicted from the open-loop bicycle rigid-rider dynamics.  We use
+the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune the gain
+magnitude during the experiments. Figure~\ref{fig:eigenvalues} shows the
+stabilization effect of the controller via the eigenmodes of the linear system.
+Geometric and inertial parameters for these plots were estimated with the
+methods presented in \citet{Moore2012} and software packages
 BicycleParameters~1.1.1~\citep{Moore2024} and Yeadon~1.5.0~\citep{Dembia2015}
@@
 Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
-18~\si{\kilo\meter\per\hour}. The general design of this perturbation system is
-described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
-the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
-Our modifications relative to Van De Velde's design include simplifying the
-Bump'Em motor controller with an inexpensive microcontroller and the use of a
-simpler non-active safety harness.
+18~\si{\kph}. The general design of this perturbation system is described in
+detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical
+arrangement is shown in Figure~\ref{fig:participant-in-set-up}.  Our
+modifications relative to Van De Velde's design include simplifying the Bump'Em
+motor controller with an inexpensive microcontroller and the use of a simpler
+non-active safety harness.
 %

### 2025-11-16 main.tex
Change all citep and citet to cite.

@@
 and values range between 50\% and 95\% in other surveyed
-countries~\citep{Wegman2024}. \citet{Utriainen2023} report that slippery
-surfaces, simple loss of control, and engagement with a curb, pothole, or
-bollard are most the common fall types in the seven reported countries. Most of
-these types involve perturbations that result in loss of balance. At low
-speeds, from start up to typical cruising speeds, bicycles are not self-stable
-and can be challenging for the rider to balance. Low speed crashes may be
-reduced if the bicycle was self-stable at these speeds by relieving the rider
-of some of their necessary control activity. Bicycles can be mechanically
-modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
-and since the 1980s it has been known that applying a motor actuated steering
-torque proportional to the vehicle's roll angular rate~\footnote{``Roll'' is
-defined herein as per the SAE J670 vehicle standards.} can stabilize a
-single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
-control of steering can stabilize a bicycle, it may reduce the control required
-from the rider to successfully manage balancing tasks much like natural high
-speed self-stability already does. We have developed a balance-assisting
-bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
+countries~\cite{Wegman2024}. Utriainen, O’Hern, and
+Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
+control, and engagement with a curb, pothole, or bollard are most the common
+fall types in the seven reported countries. Most of these types involve
+perturbations that result in loss of balance. At low speeds, from start up to
+typical cruising speeds, bicycles are not self-stable and can be challenging
+for the rider to balance. Low speed crashes may be reduced if the bicycle was
+self-stable at these speeds by relieving the rider of some of their necessary
+control activity. Bicycles can be mechanically modified to lower the speeds at
+which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
+known that applying a motor actuated steering torque proportional to the
+vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
+J670 vehicle standards.} can stabilize a single-track vehicle down to very low
+speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
+bicycle, it may reduce the control required from the rider to successfully
+manage balancing tasks much like natural high speed self-stability already
+does. We have developed a balance-assisting bicycle,
+Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
 hypothesize that it helps the rider in situations in which they are likely to
@@
     column and data acquisition and control electronics mounted in the rear
-    rack first introduced in \citet{Alizadehsaravi2023}.}
+    rack first introduced in \cite{Alizadehsaravi2023}.}
   \label{fig:balance-assist-bicycle}
@@
 dynamical system can be resolved into an equivalent force and torque of a
-couple at any location in the system~\citep{Kane1985}. We chose to perturb the
+couple at any location in the system~\cite{Kane1985}. We chose to perturb the
 handlebar because of this superposition principle and that it was both
@@
 During the early years of developments in automatic control,
-\citet{Whipple1899} not only derived the correct equations of motion of the
-bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
-can stabilize a bicycle. Much later, attempts at automatic roll stabilization
-of single-track vehicles began in earnest after predictive motorcycle models
-were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
-influenced by Whipple's work and seems to be the first to attempt to
-robotically stabilize a small motorbike with a controlled inverted pendulum
-that mimicked rider lean, but he was not successful in demonstrating what his
-theoretic control model correctly predicted. In his model, he recognized that
-feedback of the vehicle roll angle and angular rate was essential to stabilize
-the vehicle. It was not until the early 1980s that \citet{Nagai1983}
-successfully demonstrated balancing a robotic bicycle on a treadmill with both
-steering control and a laterally moving mass. \citet{Ruijs1986} followed this
-breakthrough by demonstrating an automatically balanced motorcycle and they did
-so solely with a steering motor. Ruijs and Pacejka showed that steer torque
-driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
-feedback stabilizes the weave mode, and by steer angular rate feedback
-stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
-are defined in \citep{Sharp1971}.} They also showed how the control gains must
-change with respect to vehicle speed for favorable control across all speeds.
-Thus roll motion feedback enables the simplest controller that can stabilize a
+Whipple~\cite{Whipple1899} not only derived the correct equations of motion of
+the bicycle \cite{Meijaard2007} but realized and showed that roll motion
+feedback can stabilize a bicycle. Much later, attempts at automatic roll
+stabilization of single-track vehicles began in earnest after predictive
+motorcycle models were developed and refined throughout the 1970s. Van
+Zytveld~\cite{vanZytveld1975} was influenced by Whipple's work and seems to be
+the first to attempt to robotically stabilize a small motorbike with a
+controlled inverted pendulum that mimicked rider lean, but he was not
+successful in demonstrating what his theoretic control model correctly
+predicted. In his model, he recognized that feedback of the vehicle roll angle
+and angular rate was essential to stabilize the vehicle. It was not until the
+early 1980s that Nagai~\cite{Nagai1983} successfully demonstrated balancing a
+robotic bicycle on a treadmill with both steering control and a laterally
+moving mass. Ruijs and Pacejka~\cite{Ruijs1986} followed this breakthrough by
+demonstrating an automatically balanced motorcycle and they did so solely with
+a steering motor. Ruijs and Pacejka showed that steer torque driven by roll
+angle feedback stabilizes the capsize mode, by roll angular rate feedback
+stabilizes the weave mode, and by steer angular rate feedback stabilizes the
+wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes are defined in
+\cite{Sharp1971}.} They also showed how the control gains must change with
+respect to vehicle speed for favorable control across all speeds.  Thus roll
+motion feedback enables the simplest controller that can stabilize a
 single-track vehicle above a minimum speed when one is not concerned with
@@
 effectively change the human-controlled plant dynamics, up to some limits. In
-our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion
-during distractions and light perturbations due to the balance assist system
-but \citet{Hanakam2023} recently showed rider dissatisfaction with the
-stabilization of a similar vehicle so the overall possible benefits are not yet
-definitively established.
+our prior study~\cite{Alizadehsaravi2023}, we demonstrate reduced motion during
+distractions and light perturbations due to the balance assist system but
+Hanakam, Wehner, and Wrede~\cite{Hanakam2023} recently showed rider
+dissatisfaction with the stabilization of a similar vehicle so the overall
+possible benefits are not yet definitively established.
 
 The design of our balance assist system relies on the linear Carvallo-Whipple
-bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
+bicycle model~\cite{Carvallo1899,Whipple1899} which is the simplest bicycle
 model that exhibits both the non-minimum phase behavior ``countersteering'' and
@@
 roll motion feedback driven steer motor on the dynamics. This model is
-equivalently valid for on-road or treadmill riding~\citep{Kooijman2009a}, which
+equivalently valid for on-road or treadmill riding~\cite{Kooijman2009a}, which
 have the same fundamental dynamics. The linear version of this model can be
@@
 respect to \(v\), the speed range where the bicycle is stable can be maximized
-within any physical actuator magnitude and bandwidth limits.
-\citet{Schwab2008} elaborate on some of the possibilities in scheduling the
-gains for such a controller for a bicycle and show that linear scheduling with
-respect to speed can give satisfactory stability for a low speed range. We use
-this simple feedback principle as the basis for our balance assist controller.
+within any physical actuator magnitude and bandwidth limits.  Schwab, Kooijman,
+and Meijaard~\cite{Schwab2008} elaborate on some of the possibilities in
+scheduling the gains for such a controller for a bicycle and show that linear
+scheduling with respect to speed can give satisfactory stability for a low
+speed range. We use this simple feedback principle as the basis for our balance
+assist controller.
 
@@
 Geometric and inertial parameters for these plots were estimated with the
-methods presented in \citet{Moore2012} and software packages
-BicycleParameters~1.1.1~\citep{Moore2024} and Yeadon~1.5.0~\citep{Dembia2015}
-and are shown in Appendix~\ref{sec:physical-parameters}.
+methods presented in \cite{Moore2012} and software packages
+BicycleParameters~1.1.1~\cite{Moore2024} and Yeadon~1.5.0~\cite{Dembia2015} and
+are shown in Appendix~\ref{sec:physical-parameters}.
 
@@
 also shows the identified weave mode captured using the methods of
-Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
-performance of the system, with the model predicting the weave mode well the
-region of the tested speeds.
+Kooijman~et~al.~\cite{Kooijman2009a} and gives an indication of the performance
+of the system, with the model predicting the weave mode well the region of the
+tested speeds.
 %
@@
 We apply longitudinal forces to the ends of each handlebar from earth-fixed
-posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
-four EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
+posts using an adapted Bump'Em system~\cite{Tan2020} which we arrange with four
+EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
 Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
@@
 18~\si{\kph}. The general design of this perturbation system is described in
-detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical
+detail in Van De Velde's MSc thesis~\cite{vandeVelde2022a} and the physical
 arrangement is shown in Figure~\ref{fig:participant-in-set-up}.  Our
@@
 latter is more likely to happen at higher speeds. We informed the participants
-beforehand that we would stop the treadmill in these situations. These were
-the two actions which necessitated stopping the treadmill belt and the simulate
+beforehand that we would stop the treadmill in these situations. These were the
+two actions which necessitated stopping the treadmill belt and the simulate
 real fall modes. Placing a foot down on the fall side of the bicycle is a
 natural reaction when steering or other body movements are not sufficient to
-keep the bicycle upright, as shown in \citep{Gildea2024}. Exiting the width of
+keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
 the treadmill is akin to exiting the width of a curb-less bicycle path and
@@
 Based on findings from measuring riders without balance assist in
-\citep{Haitjema2023a}, we calculate several variables that we hypothesize may
+\cite{Haitjema2023a}, we calculate several variables that we hypothesize may
 influence fall probability. We use the angular impulse \(L\) of the
@@
 centering with clusters being an individual participant, as recommended by
-\citet{Enders2007}. The clusters are chosen as all data from an individual
+\cite{Enders2007}. The clusters are chosen as all data from an individual
 participant because we are only interested in the association between the state
@@
 multilevel model. This tracks with the independent data collected in
-\citet{Haitjema2023a} which also had negligible participant variation. This
-left us with angular impulse, perturbation order, balance assist state, and
-roll \& steer angles at the time of perturbation as independent variables. We
-also include interaction effects between the balance assist state and the other
-four variables. We divide the analysis into two separate model evaluations, one
-for the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
+\cite{Haitjema2023a} which also had negligible participant variation. This left
+us with angular impulse, perturbation order, balance assist state, and roll \&
+steer angles at the time of perturbation as independent variables. We also
+include interaction effects between the balance assist state and the other four
+variables. We divide the analysis into two separate model evaluations, one for
+the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
 \(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
-Statistics were computed with R version 4.1.3~\citep{RCoreTeam2021} the lme4
-version 1.1~\citep{Bates2015} package.
+Statistics were computed with R version 4.1.3~\cite{RCoreTeam2021} the lme4
+version 1.1~\cite{Bates2015} package.
 
@@
 stable system, i.e. cannot fall, but having plant eigenvalues with very small
-unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}
+unstable eigenvalue real parts correlates to ease of control~\cite{Hess2012}
 which can be beneficial for avoiding falls.
@@
 bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
-human's neuromuscular system~\citep{Magdaleno1971}, but may feel unusual as it
+human's neuromuscular system~\cite{Magdaleno1971}, but may feel unusual as it
 is more akin to what the steering would feel like at in the
-\SIrange{30}{40}{\kph} speed range for a normal bicycle. \Citet{Hanakam2023}
-reported dissatisfaction in subjective rider feeling on their similar bicycle
-to ours and this high frequency effect to the human-controlled plant dynamics
-could be an explanation to their findings.
+\SIrange{30}{40}{\kph} speed range for a normal bicycle. Hanakam, Wehner, and
+Wrede~\cite{Hanakam2023} reported dissatisfaction in subjective rider feeling
+on their similar bicycle to ours and this high frequency effect to the
+human-controlled plant dynamics could be an explanation to their findings.
 
@@
 These are found using the measurement approach for geometric and inertial
-parameters for a bicycle explained in \citet{Moore2012}. The rider inertial
+parameters for a bicycle explained in \cite{Moore2012}. The rider inertial
 parameters are derived from the 83.5~\si{\kilogram} rider ``Jason'' available
-in BicycleParameters~1.1.1~\citep{Moore2024}. The variable names match the
+in BicycleParameters~1.1.1~\cite{Moore2024}. The variable names match the
 notation and definitions from the benchmark bicycle model presented in
-\citet{Meijaard2007}.
+\cite{Meijaard2007}.
 %

### 2025-11-18 main.tex
Switch to bibtex instead of biblatex + biber.
bibtex instead of biblatex now working.
Added sentence explaining the exclusion of two participants' data.

@@
 \usepackage[margin=25mm]{geometry}
-\usepackage[natbib=true,style=numeric,sorting=none,giveninits=true]{biblatex}
-% NOTE : this file is automatically generated from Zotero, do not edit
-% manually!
-\addbibresource{references.bib}
 \usepackage{siunitx}  % use for all units
@@
 \usepackage[]{todonotes}  % use [disable] to hide notes
+\usepackage[hidelinks]{hyperref}
 
@@
 Following the initial threshold determination where the participant experienced
-a full range of perturbation magnitudes, we then choose perturbation forces
+a range of perturbation magnitudes, we then choose perturbation forces
 according to a random and adaptive staircase procedure applying perturbations
@@
 did not try to eliminate learning effects due to the time per participant that
-would be required, but factored this into the later analysis as an alternative.
-Five possible perturbation forces are determined based on the initial
-perturbation threshold estimate: the initial estimate itself, two perturbations
-lower than the initial estimate, and two perturbations higher than the initial
-estimate.  The five possible perturbations are separated by 10~\si{\newton}
-steps. For example, if the initial estimate of the perturbation threshold of a
-participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
-90 and 100~\si{\newton}. Which one of these five perturbations is chosen first
-is determined at random. If the perturbation results in a fall, the estimate of
+would be required, but factored this into the later statistical analysis.  Five
+possible perturbation forces are determined based on the initial perturbation
+threshold estimate: the initial estimate itself, two perturbations lower than
+the initial estimate, and two perturbations higher than the initial estimate.
+The five possible perturbations are separated by 10~\si{\newton} steps. For
+example, if the initial estimate of the perturbation threshold of a participant
+is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
+100~\si{\newton}. Which one of these five perturbations is chosen first is
+determined at random. If the perturbation results in a fall, the estimate of
 the perturbation threshold is decreased by 10~\si{\newton}, and vice versa if
@@
 
-Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
-that they have a mean of zero and a standard deviation of one by cluster-mean
-centering with clusters being an individual participant, as recommended by
-\cite{Enders2007}. The clusters are chosen as all data from an individual
-participant because we are only interested in the association between the state
-of the balance assist system and the outcome of the perturbation for that
-participant, not how participants perform relative to each other. We expected
-there to be variation between participants in how well they are able to resist
-the perturbation; however, this was not true. If a random effect is included in
-the model for each participant, the intraclass correlation (ICC) is less that
-3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
-single-level logistic regression model without the random effect instead of a
-multilevel model. This tracks with the independent data collected in
-\cite{Haitjema2023a} which also had negligible participant variation. This left
-us with angular impulse, perturbation order, balance assist state, and roll \&
-steer angles at the time of perturbation as independent variables. We also
-include interaction effects between the balance assist state and the other four
-variables. We divide the analysis into two separate model evaluations, one for
-the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
-\(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
-Statistics were computed with R version 4.1.3~\cite{RCoreTeam2021} the lme4
-version 1.1~\cite{Bates2015} package.
+Due to data quality issues that prevented time synchronization of the
+measurements, we excluded one participant from the 6~\si{\kps} dataset and one
+from the 10~\si{\kps} dataset, leaving 10 and 14 included participants with 40
+perturbations each. Before fitting the model, we scale each independent
+variable \(x_{ij}^k\) such that they have a mean of zero and a standard
+deviation of one by cluster-mean centering with clusters being an individual
+participant, as recommended by \cite{Enders2007}. The clusters are chosen as
+all data from an individual participant because we are only interested in the
+association between the state of the balance assist system and the outcome of
+the perturbation for that participant, not how participants perform relative to
+each other. We expected there to be variation between participants in how well
+they are able to resist the perturbation; however, this was not true. If a
+random effect is included in the model for each participant, the intraclass
+correlation (ICC) is less that 3\% for both the 6~\si{\kph} and 10~\si{\kph}
+datasets, so we utilize a simple single-level logistic regression model without
+the random effect instead of a multilevel model. This tracks with the
+independent data collected in \cite{Haitjema2023a} which also had negligible
+participant variation. This left us with angular impulse, perturbation order,
+balance assist state, and roll \& steer angles at the time of perturbation as
+independent variables. We also include interaction effects between the balance
+assist state and the other four variables. We divide the analysis into two
+separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
+one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
+for each set of data.  Statistics were computed with R version
+4.1.3~\cite{RCoreTeam2021} the lme4 version 1.1~\cite{Bates2015} package.
 
@@
 
-\printbibliography
+\bibliographystyle{plain}
+% NOTE : this file is automatically generated from Zotero, do not edit
+% manually!
+\bibliography{references}
 

### 2025-11-21 main.tex
Completed a pass of review and editing, manually adjusted some references.

@@
 and values range between 50\% and 95\% in other surveyed
-countries~\cite{Wegman2024}. Utriainen, O’Hern, and
+countries~\cite{Wegman2024}. Utriainen, O'Hern, and
 Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
-control, and engagement with a curb, pothole, or bollard are most the common
+control, and engagement with a curb, pothole, or bollard are the most common
 fall types in the seven reported countries. Most of these types involve
@@
     column and data acquisition and control electronics mounted in the rear
-    rack first introduced in \cite{Alizadehsaravi2023}.}
+    rack; first introduced in \cite{Alizadehsaravi2023}.}
   \label{fig:balance-assist-bicycle}
@@
 
-Riders fall when their control authority is not able to maintain a stable
+Riders fall when their control authority is not sufficient to maintain a stable
 vehicle state. There are many real world scenarios that can put the vehicle
@@
 couple at any location in the system~\cite{Kane1985}. We chose to perturb the
-handlebar because of this superposition principle and that it was both
-practical and required a low force to cause a fall.
+handlebar because of this superposition principle, it was practical, and it
+required a low force to cause a fall.
 
@@
 Whipple~\cite{Whipple1899} not only derived the correct equations of motion of
-the bicycle \cite{Meijaard2007} but realized and showed that roll motion
+the bicycle~\cite{Meijaard2007} but realized and showed that roll motion
 feedback can stabilize a bicycle. Much later, attempts at automatic roll
@@
 bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
-Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
-but the time constant of the unstable eigenvalue is increased by a factor of 8
-and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
-also shows the identified weave mode captured using the methods of
-Kooijman~et~al.~\cite{Kooijman2009a} and gives an indication of the performance
-of the system, with the model predicting the weave mode well the region of the
-tested speeds.
+Figure~\ref{fig:eigenvalues}. Notably, the controlled bicycle is not stable at
+6~\si{\kph} but the time constant of the unstable eigenvalue is increased by a
+factor of eight and at 10~\si{\kph} the system is just stable.
+Figure~\ref{fig:eigenvalues} also shows the identified weave mode captured
+using the methods of Kooijman~et~al.~\cite{Kooijman2009a} and gives an
+indication of the performance of the system, with the model predicting the
+weave mode well the region of the tested speeds.
 %
@@
 beforehand that we would stop the treadmill in these situations. These were the
-two actions which necessitated stopping the treadmill belt and the simulate
+two actions which necessitated stopping the treadmill belt and they simulate
 real fall modes. Placing a foot down on the fall side of the bicycle is a
@@
 successful if the person neither places their foot down onto the treadmill
-surface nor their wheel of the bicycle exits the width of the treadmill belt.
-We record this as a binary variable \(f\) for ``fall''.  All measured variables
+surface nor the wheel of the bicycle exits the width of the treadmill belt.  We
+record this as a binary variable \(f\) for ``fall''.  All measured variables
 are reported in Table~\ref{tab:raw}.
@@
 Due to data quality issues that prevented time synchronization of the
-measurements, we excluded one participant from the 6~\si{\kps} dataset and one
-from the 10~\si{\kps} dataset, leaving 10 and 14 included participants with 40
+measurements, we excluded one participant from the 6~\si{\kph} dataset and one
+from the 10~\si{\kph} dataset, leaving 10 and 14 included participants with 40
 perturbations each. Before fitting the model, we scale each independent
@@
     \toprule
-    Variable & \(\beta\) / \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) / \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
+    Variable & \(\beta\) or \(\alpha_k\) & SE & \(p\) & \(e^{\beta}\) or \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
     \midrule
@@
 trials, the balance assist state is not a significant predictor (\(p=0.07\))
-but the effect would be a reduction in the probability to fall if it were. At
-both speeds, the angular impulse has about twice the magnitude in effect as the
-perturbation order.
+for our \(p\)-value threshold but the effect would be a reduction in the
+probability to fall if it were. At both speeds, the angular impulse has about
+twice the magnitude in effect as the perturbation order.
 %
@@
 fall with the balance assist system turned on, but the effect is not
-significant.
+significant \((p=0.07 > 0.05)\).
 %
@@
 The controller design we utilize, Equation~\ref{eq:implemented-controller},
-also increases the weave mode frequency by a factor of about three in a
-bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
-human's neuromuscular system~\cite{Magdaleno1971}, but may feel unusual as it
-is more akin to what the steering would feel like at in the
-\SIrange{30}{40}{\kph} speed range for a normal bicycle. Hanakam, Wehner, and
-Wrede~\cite{Hanakam2023} reported dissatisfaction in subjective rider feeling
-on their similar bicycle to ours and this high frequency effect to the
-human-controlled plant dynamics could be an explanation to their findings.
+also increases the weave mode frequency by a factor of about three to almost
+1~\si{\hertz}. This bandwidth is still controllable by the human's
+neuromuscular system~\cite{Magdaleno1971}, but may feel unusual as it is more
+akin to what the steering would feel like at 25~\si{\kph} for a normal bicycle.
+Hanakam, Wehner, and Wrede~\cite{Hanakam2023} reported dissatisfaction in
+subjective rider feeling on their similar bicycle to ours and this high
+frequency effect to the human-controlled plant dynamics could be an explanation
+to their findings.
 
@@
 in BicycleParameters~1.1.1~\cite{Moore2024}. The variable names match the
-notation and definitions from the benchmark bicycle model presented in
-\cite{Meijaard2007}.
+notation and definitions from the benchmark bicycle model presented
+in~\cite{Meijaard2007}.
 %

### 2025-11-21 review/review-response-01.tex
Finalized the review question responses.

@@
 Thank you for the careful review of our work. We have revised the paper to
-address your comments both for clarification of various methods and details and
-providing background and support for some of our choices. The revision is an
-improved version based on your feedback.
+address your comments both for clarification of various methods and details as
+well as providing background and support for some of our choices. The revision
+is an improved version based on your feedback.
 
@@
 have not performed our own motor testing, as that is not generally our
-expertise or concern. But we have now included extra vehicle performance tests
+expertise or interest. But we have now included extra vehicle performance tests
 in the form of weave mode identification of the rider-less bicycle. Figure 2
@@
 motion of the actual system. Note that these gain factors 3.9 and 5.2 are
-different than the prior paper version of 8 and 10, respecitively. The reason
+different than the prior paper version of 8 and 10, respectively. The reason
 for this is that the model does not include the motor dynamics (motor constant
@@
 on the ground. The approximate factor of 2 difference in the gain factor
-accounts for these unmodeled effects. The gain factor of 8 and 10 are the
+accounts for these un-modeled effects. The gain factor of 8 and 10 are the
 values we employ in the microcontroller of the real system. We hope that this
@@
 The two modes of falling are different. We have added more information in
-section 2.4 to explain these two modes. If the particpants were cycling on an
+section 2.4 to explain these two modes. If the participants were cycling on an
 infinite plane and we perturb them to fall, then we would strictly use body
-contact with the ground as the criteria to define a fall. We have cited
+contact with the ground as the criteria to define a fall. We have now cited
 research that indicates riders often attempt to remove their foot from the
@@
 we had to stop the treadmill for safety reasons. Neither of the fall modes on
-the treadmill are "extremely dangerous" or "dangerous".
+the treadmill are "extremely dangerous" or "dangerous", based on our
+experience, so we did not consider this aspect.
 
@@
 The participants experienced 40 random force magnitudes after they experienced
-all force magnitudes in the threshold determiniation stage. So some forces were
+all force magnitudes in the threshold determination stage. So some forces were
 experienced more than once. We would have liked to remove the learning effect
@@
 practical. We choose to include a measure of learning in the statistics as an
-alternative.
-
-TODO : Did I change anything in the paper about this?
+alternative. We have now noted this in Section 2.4.
 
@@
 
-We have included the average age, albiet that four participants' ages were
-misplaced in the two years since we did this study.
+We have included the average age, albeit that four participants' ages were
+misplaced in the two years since we did this study, which is declared in a
+footnote.
 
@@
 
-We have included citations to papers that discuss the types and nature of
+We have now included citations to papers that discuss the types and nature of
 bicycle falls and expanded the text in Section 1 to better connect our study to
@@
 
-First, see the answer above to review \#1 as a partial response to your
-question. The two fall modes are not technically the same and do not simply
-differ in degree, but were necessary given the nature of the experimental
-design. We did not record the category of the fall mode, so it is not possible
-for us to investigate whether having a tenary factor instead of a binary one
-would reveal any more insight. We added text in Section 2.4 to make this clear.
+First, see the answer above to reviewer \#1's similar question as a partial
+response to this question. The two fall modes are not technically the same and
+do not simply differ in degree, but were necessary given the nature of the
+experimental design. We did not record the category of the fall mode, so it is
+not possible for us to investigate whether having a ternary factor instead of a
+binary one would reveal any more insight. We added text in Section 2.4 to make
+this clear.
 
@@
 bit more variation is accounted for, but we decided to use the more
-consertative approach and not include this random effect. The participants do
+conservative approach and not include this random effect. The participants do
 not have appreciable variation in how they respond to the perturbations after
@@
 interested in whether the system reduces the probability of anyone to fall.
-The cluster mean centering removes the relative differences in the particpants
-perturbation resistance threshold.
+The cluster mean centering removes the relative differences in the
+participants' perturbation resistance threshold.
 
@@
 
+We hope that these changes satisfy your concerns and look forward to your
+further feedback.
+
 \closing{Sincerely,}

### 2025-11-23 review/review-response-01.tex
Editing pass on the review response letter.

@@
 % Reference top left
-\date{17 november 2025}
+\date{23 november 2025}
 \ourref{NA}
@@
 % Recipient left under reference
-\toname{IATSS Editor}
-\toaddress{}
+\toname{Editor}
+\toaddress{International Association of\\Traffic and Safety Sciences}
 
@@
 \fromaddress{Mekelweg 2\\2628 CD Delft\\The Netherlands}
-\frompobox{NA}
+%\frompobox{}
 
@@
 \begin{displayquote}
-  Editor:Thank you once more for your submission to IATSS Research. The
-  reviewers have now completed their assessment of your manuscript and have
+  \textbf{Editor}: Thank you once more for your submission to IATSS Research.
+  The reviewers have now completed their assessment of your manuscript and have
   recommended major revisions before the manuscript is being considered for
@@
 
-The steer motor system was developed by Bosch eBike systems and is a
-proprietary, thus we are not privy to the fine details of its operation. We
-have not performed our own motor testing, as that is not generally our
-expertise or interest. But we have now included extra vehicle performance tests
-in the form of weave mode identification of the rider-less bicycle. Figure 2
-has been reworked to show the transient behavior of the steer motor controlled
-bicycle compared to the theoretical model's predictions. For the two testing
-speeds we set the controller gain factor to different values. We now show the
-gain factors needed in the ideal bicycle dynamics model to cause matching
-motion of the actual system. Note that these gain factors 3.9 and 5.2 are
-different than the prior paper version of 8 and 10, respectively. The reason
-for this is that the model does not include the motor dynamics (motor constant
-or winding resistance), nor does it include the scrub torque effect of the tire
-on the ground. The approximate factor of 2 difference in the gain factor
-accounts for these un-modeled effects. The gain factor of 8 and 10 are the
-values we employ in the microcontroller of the real system. We hope that this
-proves 1) that our model can predict the steer controlled bicycle's motion and
-2) gives an idea of the performance of the controlled vehicle. We have adapted
-the text in section 2.2 to explain this.
+The steer motor was developed by Bosch eBike Systems and is a proprietary, thus
+we are not privy to the fine details of its operation. We have not performed
+our own motor testing, as that is not generally our expertise or interest. But
+we have now included extra vehicle performance tests in the form of weave mode
+identification of the rider-less bicycle. Figure 2 has been reworked to show
+the transient behavior of the steer motor controlled bicycle compared to the
+theoretical model's predictions. For the two testing speeds we set the
+controller gain factor to different values. We now show the gain factors needed
+in the ideal bicycle dynamics model to cause matching motion of the actual
+system. Note that these gain factors 3.9 and 5.2 are different than the prior
+paper version of 8 and 10, respectively. The reason for this is that the model
+does not include the motor dynamics (motor constant or winding resistance), nor
+does it include the scrub torque effect of the tire on the ground. The
+approximate factor of 2 difference in the gain factor accounts for these
+un-modeled effects. The gain factor of 8 and 10 are the values we employ in the
+microcontroller of the real system. We hope that this proves 1) that our model
+can predict the steer controlled bicycle's motion and 2) gives an idea of the
+performance of the controlled vehicle. We have adapted the text in section 2.2
+to explain this.
 
 \begin{displayquote}
-  In this study, a fall is defined as both "putting one's foot on the ground"
-  and "exceeds the width of treadmill", but there is insufficient explanation
-  for this. In 4.2, the relationship between "exceeds the width of treadmill"
-  and actual falls is discussed, but the explanation for "putting one's foot on
-  the ground" is insufficient. "Putting one's foot on the ground" seem to occur
-  with a much milder degree of instability than "exceeds the width of
-  treadmill". In the actual experimental situation, "putting one's foot on the
-  ground" may have been extremely dangerous, equivalent to "exceeds the width
-  of treadmill", but this is not clear from the manuscript. I would like this
+  In this study, a fall is defined as both ``putting one's foot on the ground''
+  and ``exceeds the width of treadmill'', but there is insufficient explanation
+  for this. In 4.2, the relationship between ``exceeds the width of treadmill''
+  and actual falls is discussed, but the explanation for ``putting one's foot
+  on the ground'' is insufficient. ``Putting one's foot on the ground'' seem to
+  occur with a much milder degree of instability than ``exceeds the width of
+  treadmill''. In the actual experimental situation,``putting one's foot on the
+  ground'' may have been extremely dangerous, equivalent to ``exceeds the width
+  of treadmill'', but this is not clear from the manuscript. I would like this
   point to be described in the Methods or Discussion section. Furthermore, the
@@
 pedal and place it on the ground to lessen or prevent a complete fall to the
-ground, so using that as an indicator for defining a fall seems justified. But
-we have a limited width treadmill, so as speed increases a perturbation will
-cause greater lateral deviations. If the wheel of the bicycle exits the width
-we had to stop the treadmill for safety reasons. Neither of the fall modes on
-the treadmill are "extremely dangerous" or "dangerous", based on our
+ground, so using that as an indicator for defining a fall is justified. But we
+have a limited width treadmill, so as speed increases a perturbation will cause
+greater lateral deviations. If the wheel of the bicycle exits the width we had
+to stop the treadmill for safety reasons. Neither of the fall modes on the
+treadmill are ``extremely dangerous'' or ``dangerous'', based on our
 experience, so we did not consider this aspect.
@@
 the perturbation as well as included the configuration at the time of
-perturbation as independent variables. We added text to clarify these things in
-Section 2.5.
+perturbation as independent variables. A variable could be included to indicate
+if a perturbation was preceded by the same sign perturbation or not, but we
+have chosen not to include this, as it was not in our original hypothesis. We
+added text to clarify these things in Section 2.5.
 
@@
 
-  Regarding low-speed accidents, the authors state that "single-vehicle
+  Regarding low-speed accidents, the authors state that ``single-vehicle
   accidents are associated with a surprisingly high proportion of reported
-  serious injuries," citing examples that can cause loss of control, such as
+  serious injuries,'' citing examples that can cause loss of control, such as
   gusts of wind, collisions between handlebars, bags hanging from handlebars,
@@
 to argue that it matters little how the perturbation is delivered as long as we
-force the system into an unrecoverable state.
+force the system into an unrecoverable state. We did not apply vertical forces
+to the handlebars, only forces in the longitudinal direction, maybe this was a
+misunderstanding as it is stated in the paper and shown in Figure 3a.
 
@@
 binary one would reveal any more insight. We added text in Section 2.4 to make
-this clear.
+this clear. We recommend attention be paid to this in future similar studies.
 
@@
 
-  The authors state that "cluster-mean centering showed there to be no
-  variation between participants," which allowed authors to use a simple
+  The authors state that ``cluster-mean centering showed there to be no
+  variation between participants,'' which allowed authors to use a simple
   single-level logistic regression model rather than a multilevel model. This
@@
 
-  This relates to the above point. While the authors state that "Each
-  individual participant had their own perturbation resistance threshold"
+  This relates to the above point. While the authors state that ``Each
+  individual participant had their own perturbation resistance threshold''
   regarding individual differences in the experiment, it is difficult for
@@
 The cluster mean centering removes the relative differences in the
-participants' perturbation resistance threshold.
+participants' perturbation resistance threshold. We have clarified this in
+Section 2.6.
 

### 2025-11-23 main.tex
Edited new text.

@@
 Single-actor bicycle crashes are associated with a surprisingly large
-percentage (\textgreater80\%) of reported serious injuries in The Netherlands
+percentage of reported serious injuries in The Netherlands (\textgreater80\%)
 and values range between 50\% and 95\% in other surveyed
@@
 couple at any location in the system~\cite{Kane1985}. We chose to perturb the
-handlebar because of this superposition principle, it was practical, and it
-required a low force to cause a fall.
+handlebar because of this superposition principle, because it was practical,
+and because it required a low force to cause a fall.
 
@@
 beforehand that we would stop the treadmill in these situations. These were the
-two actions which necessitated stopping the treadmill belt and they simulate
-real fall modes. Placing a foot down on the fall side of the bicycle is a
+two actions which necessitated stopping the treadmill belt and they are proxies
+for real fall modes. Placing a foot down on the fall side of the bicycle is a
 natural reaction when steering or other body movements are not sufficient to
@@
 random effect is included in the model for each participant, the intraclass
-correlation (ICC) is less that 3\% for both the 6~\si{\kph} and 10~\si{\kph}
+correlation (ICC) is less than 3\% for both the 6~\si{\kph} and 10~\si{\kph}
 datasets, so we utilize a simple single-level logistic regression model without
@@
 for each set of data. Statistics were computed with R version
-4.1.3~\cite{RCoreTeam2021} the lme4 version 1.1~\cite{Bates2015} package.
+4.1.3~\cite{RCoreTeam2021} and the lme4 version 1.1~\cite{Bates2015} package.
 
@@
     \(\alpha_k\) at 6~\si{\kph} and gain factor \(\kappa=5.2\) along with the
-    standard error SE, p-value \(p\), multiplicative change in odds
+    standard error SE, \(p\)-value \(p\), multiplicative change in odds
     \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}
@@
     \(\alpha_k\) at 10~\si{\kph} and gain factor \(\kappa=3.9\) along with the
-    standard error SE, p-value \(p\), multiplicative change in odds
+    standard error SE, \(p\)-value \(p\), multiplicative change in odds
     \(e^{\beta}\) or \(e^{\alpha_k}\), and the 5\% confidence interval bounds.}

### 2025-11-24 main.tex
Jason's final edit pass.

@@
     riding at 6~\si{\kph} when mechanically perturbed.
-  \item Balance assist technology can reduce natural single-actor falls around
-    the limits of the rider's control authority.
+  \item Balance assist technology could reduce natural single-actor falls
+    around the limits of the rider's control authority.
 \end{itemize}
@@
 rider if the motor controlled steer torque and the rider applied steer torque
-act in parallel. The effect of this automatic control gives the ability to
-effectively change the human-controlled plant dynamics, up to some limits. In
-our prior study~\cite{Alizadehsaravi2023}, we demonstrate reduced motion during
+act in parallel. This automatic control provides the ability to effectively
+change the human-controlled plant dynamics, up to some limits. In our prior
+study~\cite{Alizadehsaravi2023}, we demonstrated reduced motion during
 distractions and light perturbations due to the balance assist system but
 Hanakam, Wehner, and Wrede~\cite{Hanakam2023} recently showed rider
-dissatisfaction with the stabilization of a similar vehicle so the overall
+dissatisfaction with the stabilization of a similar vehicle. So, the overall
 possible benefits are not yet definitively established.
@@
 along with their time derivatives and the inputs \(\vec{u}\) are roll torque
-\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
-function of the equilibrium forward speed \(v\). It and the input matrix
-\(\mathbf{B}\) are otherwise populated with expressions that are functions of
-the geometric and inertial parameters of the nonholonomic multibody system made
-up of four rigid bodies: two wheels, front frame, and rear frame.
+\(T_\phi\) and steer torque \(T_\delta\) as defined in ~\cite{Meijaard2007}.
+The state matrix \(\mathbf{A}\) is a function of the equilibrium forward speed
+\(v\). It and the input matrix \(\mathbf{B}\) are otherwise populated with
+expressions that are functions of the geometric and inertial parameters of the
+nonholonomic multibody system made up of four rigid bodies: two wheels, front
+frame, and rear frame.
 
@@
 
-Based on findings from measuring riders without balance assist in
-\cite{Haitjema2023a}, we calculate several variables that we hypothesize may
-influence fall probability. We use the angular impulse \(L\) of the
+Based on our prior findings from measuring riders without balance
+assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
+may influence fall probability. We use the angular impulse \(L\) of the
 perturbation forces over a 0.3~\si{\second} duration to characterize the
@@
 perturbation. To account for this, the roll and steer variables at the instance
-of perturbation are normalized for perturbation direction. The gain factor
-setting on the balance assist controller indicates if the assistance is off
-\(\kappa=0\) or on at two different levels: low \(\kappa=3.9\) or high
-\(\kappa=5.2\). As mentioned earlier, a recovery from the perturbation is
-successful if the person neither places their foot down onto the treadmill
-surface nor the wheel of the bicycle exits the width of the treadmill belt.  We
-record this as a binary variable \(f\) for ``fall''.  All measured variables
-are reported in Table~\ref{tab:raw}.
+of perturbation are normalized for perturbation direction as was done in
+\cite{Haitjema2023a}. The gain factor setting on the balance assist controller
+indicates if the assistance is off \(\kappa=0\) or on at two different levels:
+low \(\kappa=3.9\) or high \(\kappa=5.2\). As mentioned earlier, a recovery
+from the perturbation is successful if the person neither places their foot
+down onto the treadmill surface nor the wheel of the bicycle exits the width of
+the treadmill belt.  We record this as a binary variable \(f\) for ``fall''.
+All measured variables are reported in Table~\ref{tab:raw}.
 %
@@
 %
-We evaluate our hypothesis using a multivariate logistic regression model that
-represents the probability of this binary outcome. The log-odds of the
-probability is then a linear function of our independent variables with
-\(\beta\) being the intercept and \(\alpha_k\) the linear coefficients to the
-\(K\) independent variables \(x_{ij}^k\), i.e. all variables in
-Table~\ref{tab:stat-model-variables} except \(f\).
+We evaluate our hypothesis using a multivariate logistic regression model,
+Equation~\ref{eq:log-regress}, that represents the probability of this binary
+outcome. The log-odds of the probability is then a linear function of our
+independent variables with \(\beta\) being the intercept and \(\alpha_k\) the
+linear coefficients to the \(K\) independent variables \(x_{ij}^k\), i.e. all
+variables in Table~\ref{tab:stat-model-variables} except \(f\).
 %
@@
 
-\subsection{Utility of the Logistic Model and Extrapolation to Natural Falls}
+\subsection{Utility of the Logistic Model to Natural Falls}
 %
@@
 from the limited data collected from natural cycling would be needed to
-populate the model.
-
-The positive effect of the balance assist system is coupled to the assumptions
-and experimental scenarios we implemented and there is unfortunately no simple
-way to extrapolate our results to reductions of single-actor crashes we may see
-if such a system were deployed widely to bicyclists. Although, our results do
-indicate that we would see such a reduction, even if only in a class of
-single-actor crashes that most resemble our experimental design. If there were
-more comprehensive and detailed natural data of how people fall we could make
-estimates on the number of falls reduced if everyone rode a balance assist
-bicycle.
+populate the model. The positive effect of the balance assist system is coupled
+to the assumptions and experimental scenarios we implemented and there is
+unfortunately no simple way to extrapolate our results to reductions of
+single-actor crashes we may see if such a system were deployed widely to
+bicyclists. Although, our results do indicate that we would see such a
+reduction, even if only in a class of single-actor crashes that most resemble
+our experimental design. If there were more comprehensive and detailed natural
+data of how people fall we could make estimates on the number of falls reduced
+if everyone rode a balance assist bicycle.
 
@@
 higher speeds, the same perturbation magnitude causes larger lateral
-deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
+deviations. At 10~\si{\kph}, almost all falls were due to the bicycle exiting
 the maximum width of the treadmill. If the same experiment was performed on an
 infinitely wide plane, the riders may have recovered from more perturbations.
-At 6~\si{\kph} the riders could often recover in the allotted treadmill width
+At 6~\si{\kph}, the riders could often recover in the allotted treadmill width
 due to the smaller lateral deviations. We believe our results are very much
 dependent on the two modes of falling, i.e. exiting the treadmill width or
-placing a foot on the belt. On the other hand, cycle paths are a similar width
-as the treadmill, so riders are often limited in width when recovering from a
-fall thus exiting the treadmill width may be an appropriate measure for
-indicating a fall. While the treadmill simulates narrow cycle paths, real-world
-paths may offer more lateral recovery space, providing riders with additional
-opportunities to regain balance after a perturbation. However, testing in such
-narrow conditions is still highly relevant, as system design and validation
-should focus on extreme conditions like narrow paths, because even in wider
-paths, obstacles such as parked cars or barriers can limit lateral space.
+placing a foot on the belt. Future studies should consider coding fall mode by
+type. On the other hand, cycle paths are a similar width as the treadmill, so
+riders are often limited in width when recovering from a fall thus exiting the
+treadmill width may be an appropriate measure for indicating a fall. While the
+treadmill simulates narrow cycle paths, real-world paths may offer more lateral
+recovery space, providing riders with additional opportunities to regain
+balance after a perturbation. However, testing in such narrow conditions is
+still highly relevant, as system design and validation should focus on extreme
+conditions like narrow paths, because even in wider paths, obstacles such as
+parked cars or barriers can limit lateral space.
 
@@
 an individual's perturbation resistance threshold, otherwise we would observe
-bicyclists falling much more but low skilled riders may experience more
+bicyclists falling much more, but low skilled riders may experience more
 perturbations that exceed their threshold.
@@
 learning effect that occurs during the experiment is not strongly dependent on
-the speed but that similar experiments should consider the learning effect.
+the speed but that similar experiments should consider the learning effect. An
+experiment that tests professional cyclists or one that trains each participant
+to their maximum performance could be insightful.
 
@@
 limits of the rider's control authority. Given that similar effects cause falls
-during bicycling, use of the balance assist system in real world use cases will
-reduce the number of falls at low bicycling speeds.
+during bicycling, use of the balance assist system in real world use cases
+should reduce the number of falls at low bicycling speeds.
 
@@
 Thanks to Wouter Gregoor, Hans van der Does, and Cor Meijneke for reviving the
-bicycles at a critical moment. Thanks to Christoph Schmidt for motivation to
-do yet another round of trials. We thank Peter Stahlecker, Benjamin Gonzalez,
-and Riender Happee for valuable feedback in early revisions. Lastly, we thank
-Frans van der Helm for the consistent encouragement to execute this experiment
-and his early vision for obtaining the funding.
+bicycles at a critical moment. Thanks to Christoph Schmidt for motivation to do
+yet another round of trials. We thank Peter Stahlecker, Benjamin Gonzalez, and
+Riender Happee for valuable feedback in early revisions. Lastly, we thank Frans
+van der Helm for the consistent encouragement to execute this experiment and
+his early vision for obtaining the funding.
 

### 2025-11-25 main.tex
Added the exact mean age of all particpants and removed footnote about missing ages.

@@
 %
-We recruited 26 able-bodied young adults between 20 and 36 years old (mean 26
-years~\footnote{Excluding four participants' ages that were misplaced but that
-did fall within the stated range.}) to participate in the experiments. The
-participants were all confident in their cycling skills and had cycled at least
-once in the last month. All participants consented to the experiment and could
-decline to continue at any time. The study was approved by Delft University of
-Technology's Human Research Ethics Council (record \#3897). The participants
-were divided into two groups. The first group of fifteen participants
-performed the protocol at a belt speed of 10~\si{\kph}(2.8~\si{\mps}) with the
-gain factor set to \(\kappa=3.9\) and two weeks later the second group of
-eleven participants performed the protocol at a belt speed of 6~\si{\kph}
-(1.7~\si{\mps}) with the gain factor set to \(\kappa=5.2\).
+We recruited 26 able-bodied young adults between 20 and 36 years old (mean 25
+years) to participate in the experiments. The participants were all confident
+in their cycling skills and had cycled at least once in the last month. All
+participants consented to the experiment and could decline to continue at any
+time. The study was approved by Delft University of Technology's Human Research
+Ethics Council (record \#3897). The participants were divided into two groups.
+The first group of fifteen participants performed the protocol at a belt speed
+of 10~\si{\kph}(2.8~\si{\mps}) with the gain factor set to \(\kappa=3.9\) and
+two weeks later the second group of eleven participants performed the protocol
+at a belt speed of 6~\si{\kph} (1.7~\si{\mps}) with the gain factor set to
+\(\kappa=5.2\).
 

### 2025-11-27 review/review-response-01.tex
Adjusted review response from Marten's comments.

@@
 
-The participants experienced 40 random force magnitudes after they experienced
-all force magnitudes in the threshold determination stage. So some forces were
-experienced more than once. We would have liked to remove the learning effect
-by letting the participants train for a long duration, but this was not
-practical. We choose to include a measure of learning in the statistics as an
-alternative. We have now noted this in Section 2.4.
+The participants experienced the same magnitude perturbations more than once
+throughout the warm-up, the perturbation threshold determination, and the
+adaptive staircase procedure. But they did not experience the same force
+repeatedly many times in order to gain familiarity. We would have liked to
+remove the learning effect by letting the participants train for a long
+duration, but this was not practical. We choose to include a measure of
+learning in the statistics as an alternative. We have now noted this in Section
+2.4.
 
@@
 
-We did consider this and normalized the time series measurements to the sign of
-the perturbation as well as included the configuration at the time of
-perturbation as independent variables. A variable could be included to indicate
-if a perturbation was preceded by the same sign perturbation or not, but we
-have chosen not to include this, as it was not in our original hypothesis. We
-added text to clarify these things in Section 2.5.
+Each participant experienced an equal number of clockwise and counterclockwise
+perturbations. So 10 clockwise, 10 counterclockwise for both balance-assist on
+and off, making 40 perturbations in total. Within the 20 perturbations of each
+of the balance-assist states, the order of the directions was randomized.
+
+In the analysis we normalized the time series measurements to the sign of the
+perturbation as well as included the configuration at the time of perturbation
+as independent variables. A variable could be included to indicate if a
+perturbation was preceded by the same sign perturbation or not, but we have
+chosen not to include this, as it was not in our original hypothesis. We added
+text to clarify these things in Section 2.5.
 

### 2025-11-27 main.tex
Adjusted review response from Marten's comments.

--- 2025-11-24 main.tex
@@
 %
-We recruited 26 able-bodied young adults between 20 and 36 years old (mean 26
-years~\footnote{Excluding four participants' ages that were misplaced but that
-did fall within the stated range.}) to participate in the experiments. The
-participants were all confident in their cycling skills and had cycled at least
-once in the last month. All participants consented to the experiment and could
-decline to continue at any time. The study was approved by Delft University of
+We recruited 26 able-bodied young adults between 20 and 36 years old (mean of
+25 years) to participate in the experiments. The participants were all
+confident in their cycling skills and had cycled at least once in the last
+month. All participants consented to the experiment and could decline to
+continue at any time. The study was approved by Delft University of
 Technology's Human Research Ethics Council (record \#3897). The participants
@@
 perturbation is selected at random. This process iterates until twenty
-perturbations are applied. The goal of this adaptive staircase procedure is to
-have participants fall for approximately 50\% of the time. After the first set
-of 20 perturbations, we let the cyclist rest and then perform another 20
+perturbations are applied. The sense of the perturbation was randomized within
+the 20 perturbations. The goal of this adaptive staircase procedure is to have
+participants fall for approximately 50\% of the time. After the first set of
+20 perturbations, we let the cyclist rest and then perform another 20
 perturbations. We randomize whether the balance assist system is on or off

### 2025-11-30 main.tex
State that 24 participants of 26 were availabe to analyze.

@@
 measurements, we excluded one participant from the 6~\si{\kph} dataset and one
-from the 10~\si{\kph} dataset, leaving 10 and 14 included participants with 40
-perturbations each. Before fitting the model, we scale each independent
-variable \(x_{ij}^k\) such that they have a mean of zero and a standard
-deviation of one by cluster-mean centering with clusters being an individual
-participant, as recommended by \cite{Enders2007}. The clusters are chosen as
-all data from an individual participant because we are only interested in the
-association between the state of the balance assist system and the outcome of
-the perturbation for that participant, not how participants perform relative to
-each other. We expected there to be variation between participants in how well
-they are able to resist the perturbation; however, this was not true. If a
-random effect is included in the model for each participant, the intraclass
-correlation (ICC) is less than 3\% for both the 6~\si{\kph} and 10~\si{\kph}
-datasets, so we utilize a simple single-level logistic regression model without
-the random effect instead of a multilevel model. This tracks with the
-independent data collected in \cite{Haitjema2023a} which also had negligible
-participant variation. This left us with angular impulse, perturbation order,
-balance assist state, and roll \& steer angles at the time of perturbation as
-independent variables. We also include interaction effects between the balance
-assist state and the other four variables. We divide the analysis into two
-separate model evaluations, one for the 6~\si{\kph}, \(\kappa=5.2\) trials and
-one for the 10~\si{\kph}, \(\kappa=3.9\) trials and we evaluate our hypothesis
-for each set of data. Statistics were computed with R version
-4.1.3~\cite{RCoreTeam2021} and the lme4 version 1.1~\cite{Bates2015} package.
+from the 10~\si{\kph} dataset, leaving 24 participants for the analysis. Before
+fitting the model, we scale each independent variable \(x_{ij}^k\) such that
+they have a mean of zero and a standard deviation of one by cluster-mean
+centering with clusters being an individual participant, as recommended by
+\cite{Enders2007}. The clusters are chosen as all data from an individual
+participant because we are only interested in the association between the state
+of the balance assist system and the outcome of the perturbation for that
+participant, not how participants perform relative to each other. We expected
+there to be variation between participants in how well they are able to resist
+the perturbation; however, this was not true. If a random effect is included in
+the model for each participant, the intraclass correlation (ICC) is less than
+3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
+single-level logistic regression model without the random effect instead of a
+multilevel model. This tracks with the independent data collected in
+\cite{Haitjema2023a} which also had negligible participant variation. This left
+us with angular impulse, perturbation order, balance assist state, and roll \&
+steer angles at the time of perturbation as independent variables. We also
+include interaction effects between the balance assist state and the other four
+variables. We divide the analysis into two separate model evaluations, one for
+the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
+\(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
+Statistics were computed with R version 4.1.3~\cite{RCoreTeam2021} and the lme4
+version 1.1~\cite{Bates2015} package.
 

### 2025-11-30 main.tex
Adjust the line arrangement of the subcaption figures so that pandoc parses them in the html conversion.

@@
 %
+% NOTE : The html conversion with pandoc fails to include these subcaption
+% figures if the includegraphics line is on the same line as the subcaptionbox
+% line (or maybe simply if the curly braces containing the includegraphics call
+% is on the same line as the subcaptionbox.
 \begin{figure}
   \centering
-  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
-  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
-  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
-  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
+  \subcaptionbox{Before perturbation}
+    {\includegraphics[width=60mm]{figures/start.png}}
+  \subcaptionbox{During perturbation}
+    {\includegraphics[width=60mm]{figures/during.png}}
+  \subcaptionbox{After perturbation}
+    {\includegraphics[width=60mm]{figures/after.png}}
+  \subcaptionbox{Recovery from perturbation}
+    {\includegraphics[width=60mm]{figures/recovery.png}}
   \caption{Video frames depicting the application of a perturbation and the
@@
   \centering
-  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
-    \label{fig:probability-6kph}}{
-    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
-  }
-  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
-    \label{fig:probability-10kph}}{
-    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
-  }
+  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.\label{fig:probability-6kph}}
+    {\includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}}
+  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.\label{fig:probability-10kph}}
+    {\includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}}
   \caption{Comparison of predicted fall probability for balance assist on

### 2025-11-30 main.tex
Added descriptions to the parameters table in the appendix.

@@
 notation and definitions from the benchmark bicycle model presented
-in~\cite{Meijaard2007}.
+in~\cite{Meijaard2007} with the \(xyz\) axes following the SAE J670 standard
+with positive \(z\) directed down into the ground.
 %
@@
     the inertia effects of a rigid rider.}
-  \begin{tabular}{llrr}
+  \begin{tabular}{lllrr}
     \toprule
-    Variable & Unit & Without Rider & With Rider \\
+    Description & Variable & Unit & Without Rider & With Rider \\
     \midrule
-    \(c\) & \si{\meter} & 0.04 & 0.04 \\
-    \(g\) & \si{\meter\per\second\squared} & 9.80665 & 9.80665 \\
-    \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 19.7 \\
-    \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & -3.5 \\
-    \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 22.7 \\
-    \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 5.0 \\
-    \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0995 \\
-    \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.1902 \\
-    \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.298 \\
-    \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & -0.038 \\
-    \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.257 \\
-    \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.057 \\
-    \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.1023 \\
-    \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.1887 \\
-    \(\lambda\) & \si{\radian} & 0.25 & 0.25 \\
-    \(m_B\) & \si{\kilogram} & 22.50 & 106.00 \\
-    \(m_F\) & \si{\kilogram} & 2.23 & 2.23 \\
-    \(m_H\) & \si{\kilogram} & 4.30 & 4.30 \\
-    \(m_R\) & \si{\kilogram} & 4.08 & 4.08 \\
-    \(r_F\) & \si{\meter} & 0.3523 & 0.3523 \\
-    \(r_R\) & \si{\meter} & 0.3489 & 0.3489 \\
-    \(w\) & \si{\meter} & 1.113 & 1.113 \\
-    \(x_B\) & \si{\meter} & 0.52 & 0.374 \\
-    \(x_H\) & \si{\meter} & 0.92 & 0.92 \\
-    \(z_B\) & \si{\meter} & -0.52 & -1.009 \\
-    \(z_H\) & \si{\meter} & -0.860 & -0.860 \\
+    trail & \(c\) & \si{\meter} & 0.04 & 0.04 \\
+    acceleration due to gravity & \(g\) & \si{\meter\per\second\squared} & 9.80665 & 9.80665 \\
+    rear frame moment of inertia & \(I_{Bxx}\) & \si{\kilogram\meter\squared} & 1.12 & 19.7 \\
+    rear frame product of inertia & \(I_{Bxz}\) & \si{\kilogram\meter\squared} & 0.05 & -3.5 \\
+    rear frame moment of inertia & \(I_{Byy}\) & \si{\kilogram\meter\squared} & 3.16 & 22.7 \\
+    rear frame product of inertia & \(I_{Bzz}\) & \si{\kilogram\meter\squared} & 2.12 & 5.0 \\
+    front wheel moment of inertia & \(I_{Fxx}\) & \si{\kilogram\meter\squared} & 0.0995 & 0.0995 \\
+    front wheel moment of inertia & \(I_{Fyy}\) & \si{\kilogram\meter\squared} & 0.1902 & 0.1902 \\
+    front frame moment of inertia & \(I_{Hxx}\) & \si{\kilogram\meter\squared} & 0.298 & 0.298 \\
+    front frame product of inertia & \(I_{Hxz}\) & \si{\kilogram\meter\squared} & -0.038 & -0.038 \\
+    front frame moment of inertia & \(I_{Hyy}\) & \si{\kilogram\meter\squared} & 0.257 & 0.257 \\
+    front frame moment of inertia & \(I_{Hzz}\) & \si{\kilogram\meter\squared} & 0.057 & 0.057 \\
+    rear wheel moment of inertia & \(I_{Rxx}\) & \si{\kilogram\meter\squared} & 0.1023 & 0.1023 \\
+    rear wheel moment of inertia & \(I_{Ryy}\) & \si{\kilogram\meter\squared}& 0.1887 & 0.1887 \\
+    steer axis tilt & \(\lambda\) & \si{\radian} & 0.25 & 0.25 \\
+    rear frame mass & \(m_B\) & \si{\kilogram} & 22.50 & 106.00 \\
+    front wheel mass & \(m_F\) & \si{\kilogram} & 2.23 & 2.23 \\
+    front frame mass & \(m_H\) & \si{\kilogram} & 4.30 & 4.30 \\
+    rear wheel mass & \(m_R\) & \si{\kilogram} & 4.08 & 4.08 \\
+    front wheel radius & \(r_F\) & \si{\meter} & 0.3523 & 0.3523 \\
+    rear wheel radius & \(r_R\) & \si{\meter} & 0.3489 & 0.3489 \\
+    wheelbase & \(w\) & \si{\meter} & 1.113 & 1.113 \\
+    rear frame mass center coordinate & \(x_B\) & \si{\meter} & 0.52 & 0.374 \\
+    front frame mass center coordinate & \(x_H\) & \si{\meter} & 0.92 & 0.92 \\
+    rear frame mass center coordinate & \(z_B\) & \si{\meter} & -0.52 & -1.009 \\
+    front frame mass center coordinate & \(z_H\) & \si{\meter} & -0.860 & -0.860 \\
     \bottomrule
